---
title: "Causal Effects of Religious Service Attendance: Evidence Using Novel Measures From A National Longitudinal Panel"
abstract: |
   Causal investigations for the effects of religion on prosociality must be precise. One must articulate a specific causal contrast for a feature of religion, select appropriate prosociality measures, define the target population, gather time-series data, and, only after meeting identification assumptions, conduct statistical and sensitivity analyses. Here, we examine three distinct interventions on religious service attendance (increase, decrease, maintain) across a longitudinal dataset of 33,198 New Zealanders from 2018 to 2021. Study 1 quantifies the effect of religious service attendance on charitable contributions and volunteerism. Studies 2 and 3 investigate the effects of religious service on the relative risk of receiving aid and financial support from others during the past week, employing measures designed to minimise self-reporting bias. Across all studies, the observed effects are substantially less pronounced than cross-sectional regressions might imply. Nonetheless, regular attendance across the population would enhance charitable donations by 4% of the New Zealand Government's annual spending. This research underscores the essential role of formulating precise causal questions and establishes a methodological framework for answering them in the scientific study of cultural practices.
author: 
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University,  Matheny Center for the Study of Stress, Trauma, and Resilience 
    orcid_id: 0000-0003-3169-6576 
  - name: Kenneth G. Rice
    affiliation: Georgia State University,  Matheny Center for the Study of Stress, Trauma, and Resilience  
    orcid_id: 0000-0002-0558-2818
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Geoffrey Troughton
    affiliation: Victoria University of Wellington
    orcid_id: 0000-0001-7423-0640
keywords:
  - Causal Inference
  - Charity
  - Church
  - Cooperation
  - Religion
  - Shift Intervention
  - Volunteering
editor_options: 
  chunk_output_type: console
format:
  pdf:
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

## WARNING SET THIS PATH TO YOUR DATA ON YOUR SECURE MACHINE. 
# pull_path <-
#   fs::path_expand(
#     #"/Users/joseph/v-project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs_refactor/nzavs_data_23"
#     "/Users/joseph/Library/CloudStorage/Dropbox-v-project/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs-current/r-data/nzavs_data_qs"
#   )
# 


push_mods <-  fs::path_expand(
  "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/nzvs_mods/24/church-prosocial-v7"
)


#tinytext::tlmgr_update()

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
#source("/Users/joseph/GIT/templates/functions/libs2.R")

library(margot)
library(ggplot2)
library(tidyverse)
library(gtsummary)
# # WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
# source("/Users/joseph/GIT/templates/functions/funs.R")

#ALERT: UNCOMMENT THIS AND DOWNLOAD THE FUNCTIONS FROM JB's GITHUB

# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )
# 
# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/funs.R"
# )

# check path:is this correct?  check so you know you are not overwriting other directors
#push_mods

# for latex graphs
# for making graphs
library("tinytex")
library("extrafont")
library("tidyverse")
library("kableExtra")

loadfonts(device = "all")

## Read in data from analysis 
# imports
n_participants <- readRDS(here::here(push_mods, "n_participants"))
n_total <-  margot::here_read("n_total")

# make pretty number


n_participants<- prettyNum(n_participants,big.mark=",")
n_total<- prettyNum(n_total,big.mark=",")



#conflicts_prefer(stats::filter)

# propensity score graph 
love_plot_one <- margot::here_read("love_plot_one")
# love_plot

# # summary of matching
# summary_match_ebal_base <- margot::here_read("summary_match_ebal")
# # summary_match_ebal$summary_match_ebal_base
# 
# summary_match_ebal <- margot::here_read("summary_match_ebal")
# summary_match_ebal$effective.sample.size

# verify positivity 
#church
transition_table <- margot::here_read("transition_table")
#transition_table
# binary
transition_table_binary_gain <- margot::here_read("transition_table_binary_gain")
transition_table_binary_zero <- margot::here_read("transition_table_binary_zero")

#transition_table_binary
# 
# # ordinary regressions
fit_church_on_hours_charity<- margot::here_read('fit_church_on_hours_charity')
fit_church_on_charity_donate<-margot::here_read('fit_church_on_charity_donate')


# fit_church_on_community_time_binary<- margot::here_read('fit_church_on_community_time_binary')
# fit_church_on_community_money_binary<- margot::here_read('fit_church_on_community_money_binary')
library(gtsummary)
library(glue)

# lm_fit_church_on_hours_charity <- tbl_regression(fit_church_on_hours_charity)
# 
# # 
#b_church_on_hours_charity <-inline_text(lm_fit_church_on_hours_charity,
                                        # variable = religion_church_round,
                                        # pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
#b_church_on_hours_charity
# here_save(b_church_on_hours_charity, "b_church_on_hours_charity")

b_church_on_hours_charity<-margot::here_read("b_church_on_hours_charity")

# 
# 
#lm_coef_fit_church_on_charity_donate <- tbl_regression(fit_church_on_charity_donate)
#b_church_on_charity_donate <-inline_text(lm_coef_fit_church_on_charity_donate, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
#b_church_on_charity_donate
#here_save(b_church_on_charity_donate, "b_church_on_charity_donate")
b_church_on_charity_donate<-margot::here_read("b_church_on_charity_donate")
#b_church_on_charity_donate

#push_mods
# # lm_coef_fit_church_on_community_time_binary <- tbl-regression(fit_church_on_community_time_binary)
# # b_church_on_community_time_binary <-inline_text(lm_coef_fit_church_on_community_time_binary, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
# # here_save(b_church_on_community_time_binary, "b_church_on_community_time_binary")
# b_church_on_community_time_binary<-margot::here_read("b_church_on_community_time_binary")
# #b_church_on_community_time_binary
# 
# # lm_coef_fit_church_on_community_money_binary <- tbl-regression(fit_church_on_community_money_binary)
# # b_church_on_community_money_binary <-inline_text(lm_coef_fit_church_on_community_money_binary, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
# # here_save(b_church_on_community_money_binary, "b_church_on_community_money_binary")
# b_church_on_community_money_binary<-margot::here_read("b_church_on_community_money_binary")
# 

library(margot)
library(glue)
# works but not used
#testlm_coef_church_on_donate <- glue("b = {fit_church_on_donate_x[2, 1]}, SE = {fit_church_on_donate_x[2, 3]}")
# not this table is too long in most instance 
# table_baseline<-margot::here_read("table_baseline")

# baseline
table_baseline <-margot::here_read("table_baseline")

table_exposures <-margot::here_read("table_exposures")

table_outcomes <-margot::here_read("table_outcomes")
table_baseline
# density of the exposure 
graph_density_of_exposure_up<-margot::here_read("graph_density_of_exposure_up")
graph_density_of_exposure_down<-margot::here_read("graph_density_of_exposure_down")
graph_density_of_exposure_both<-margot::here_read("graph_density_of_exposure_both")



# examples
mean_volunteer <- margot::here_read("mean_volunteer")

# sample mean at end of study (won't be population mean)
mean_donations <- margot::here_read("mean_donations")

#n_participants
# read standard deviations units of outcome (volunteering/donations) in 2020
sd_donations <- margot::here_read("sd_donations")
sd_volunteer <- margot::here_read("sd_volunteer")

# import models

# population mean
t2_charity_donate_z_null <- margot::here_read("t2_charity_donate_z_null")

# individual models
t2_charity_donate_z_zero <- margot::here_read("t2_charity_donate_z_zero")
t2_charity_donate_z_gain <- margot::here_read("t2_charity_donate_z_gain")
t2_hours_charity_z_gain <-  margot::here_read("t2_hours_charity_z_gain")
t2_hours_charity_z_zero <- margot::here_read("t2_hours_charity_z_zero")
t2_hours_charity_z_null<- margot::here_read("t2_hours_charity_z_null")


table_baseline

# results
# import to manuscript -------------------------------------------------------

tab_all_prosocial <- here_read("tab_all_prosocial")
tab_all_prosocial_null <- here_read("tab_all_prosocial_null")
tab_all_prosocial_null_zero <- here_read("tab_all_prosocial_null_zero")
tab_all_prosocial
tab_all_prosocial_null
tab_all_prosocial_null_zero


group_tab_all_prosocial<- here_read("group_tab_all_prosocial")
group_tab_all_prosocial_null<- here_read("group_tab_all_prosocial_null")
group_tab_all_prosocial_null_zero<- here_read("group_tab_all_prosocial_null_zero")

group_tab_all_prosocial
group_tab_all_prosocial_null
group_tab_all_prosocial_null_zero


tab_received_time <- here_read( "tab_received_time")
tab_received_time_null <- here_read( "tab_received_time_null")
tab_received_time_null_zero <- here_read( "tab_received_time_null_zero")
push_mods

group_tab_received_time <- here_read( "group_tab_received_time")
group_tab_received_time_null <- here_read( "group_tab_received_time_null")
group_tab_received_time_null_zero <- here_read( "group_tab_received_time_null_zero")

tab_received_time
group_tab_received_time
group_tab_received_time

tab_all_received_money <- here_read( "tab_all_received_money")
tab_all_received_money_null <- here_read( "tab_all_received_money_null")
tab_all_received_money_null_zero <- here_read( "tab_all_received_money_null_zero")


group_tab_all_received_money <- here_read( "group_tab_all_received_money")
group_tab_all_received_money_null <- here_read( "group_tab_all_received_money_null")
group_tab_all_received_money_null_zero <- here_read( "group_tab_all_received_money_null_zero")

# Compute for manuscript
diff_charity <- round( ( sd_donations *  tab_all_prosocial[1,1]), 2)
diff_volunteer <- round(  60 * (sd_volunteer *  tab_all_prosocial[2,1]), 2)

loss_charity <- round( ( sd_donations *  tab_all_prosocial_null_zero[1,1]), 2)
loss_volunteer <- round(  60 * (sd_volunteer *  tab_all_prosocial_null_zero[2,1]), 2)


# make pretty numbers for prosocial outcomes
diff_charity<- prettyNum(diff_charity,big.mark=",")
diff_volunteer <- prettyNum(diff_volunteer,big.mark=",")
loss_charity <- prettyNum(loss_charity,big.mark=",")
loss_volunteer <- prettyNum(loss_volunteer,big.mark=",")

charity_null <- round( ( sd_donations * tab_all_prosocial_null[1,1]), 2)
charity_null<- prettyNum(charity_null,big.mark=",")
volunteer_null <- round( sd_volunteer * tab_all_prosocial_null[2,1] * 60, 2)
volunteer_null<- prettyNum(volunteer_null,big.mark=",")

charity_null_zero <- round( ( sd_donations * tab_all_prosocial_null_zero[1,1]), 2)
charity_null_zero <- prettyNum(charity_null_zero,big.mark=",")
volunteer_null_zero <- round( sd_volunteer * tab_all_prosocial_null_zero[2,1] * 60, 2)
volunteer_null_zero<- prettyNum(volunteer_null_zero,big.mark=",")

# graphs ------------------------------------------------------------------
title = "Weekly Religious Service vs. No Religious Service"
title_null = "Weekly Religious Service vs. Status Quo"
title_null_zero = "No Religious Service vs. Status Quo"


plot_group_tab_all_prosocial <- margot_plot(
  group_tab_all_prosocial,
  type = "RD",
  title = title,
  subtitle = NULL,
  #"Self-Reported Prosociality",
  estimate_scale = 3,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 14,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.07,
  x_lim_lo = -.2,
  x_lim_hi =  .2
)+ xlab("")

plot_group_tab_all_prosocial

plot_group_tab_all_prosocial_null <- margot_plot(
  group_tab_all_prosocial_null,
  type = "RD",
  title = title_null,
  subtitle = NULL,
  #"Self-Reported Prosociality",
  estimate_scale = 3,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 14,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.07,
  x_lim_lo = -.2,
  x_lim_hi =  .2
)+ xlab("")

plot_group_tab_all_prosocial_null


plot_group_tab_all_prosocial_null_zero <- margot_plot(
  group_tab_all_prosocial_null_zero,
  type = "RD",
  title = title_null_zero,
  subtitle = NULL,#"Self-Reported Prosociality",
  #"Self-Reported Prosociality",
  estimate_scale = 3,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 14,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.07,
  x_lim_lo = -.2,
  x_lim_hi =  .2
) 

plot_group_tab_all_prosocial_null_zero

# ggsave(
#   plot_group_tab_all_prosocial_null_zero,
#   path = here::here(here::here(push_mods)),
#   width = 16,
#   height = 9,
#   units = "in",
#   filename = "plot_group_tab_all_prosocial_null_zero.png",
#   device = 'png',
#   limitsize = FALSE,
#   dpi = 300
# )



# pathwork
library(patchwork)
study_1 <- plot_group_tab_all_prosocial /
plot_group_tab_all_prosocial_null/
  plot_group_tab_all_prosocial_null_zero + 
  plot_annotation(title = "Study 1: Self-Reported Prosociality", tag_levels = 'A',
             theme = theme(plot.title = element_text(size = 16, face = "bold")))



plot_group_tab_all_received_time <- margot_plot(
  group_tab_received_time,
  type = "RR",
  title = title,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)+ xlab("")
plot_group_tab_all_received_time


plot_group_tab_all_received_time_null <- margot_plot(
  group_tab_received_time_null,
  type = "RR",
  title = title_null,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)+ xlab("")

plot_group_tab_all_received_time_null


plot_group_tab_all_received_time_null_zero <- margot_plot(
  group_tab_received_time_null_zero,
  type = "RR",
  title = title_null_zero,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)

plot_group_tab_all_received_time_null_zero


study_2 <- plot_group_tab_all_received_time /
plot_group_tab_all_received_time_null/
  plot_group_tab_all_received_time_null_zero + 
  plot_annotation(title = "Study 2: Relative Risk of Help Received Last Week: Time", tag_levels = 'A',
                  theme = theme(plot.title = element_text(size = 16, face = "bold")))
study_2




plot_group_tab_all_received_money<- margot_plot(
  group_tab_all_received_money,
  type = "RR",
  title = title,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)+ xlab("")
plot_group_tab_all_received_money


plot_group_tab_all_received_money_null <- margot_plot(
  group_tab_all_received_money_null,
  type = "RR",
  title = title_null,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)+ xlab("")

plot_group_tab_all_received_money_null


plot_group_tab_all_received_money_null_zero <- margot_plot(
  group_tab_all_received_money_null_zero,
  type = "RR",
  title = title_null_zero,
  subtitle = NULL,#"Support Received From Others: Time",
  estimate_scale = 6,
  base_size = 14,
  text_size = 4,
  point_size = 2,
  title_size = 12,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset =  0,
  x_lim_lo = -.05,
  x_lim_hi =  1.75
)

plot_group_tab_all_received_money_null_zero


study_3<- plot_group_tab_all_received_money /
plot_group_tab_all_received_money_null/
  plot_group_tab_all_received_money_null_zero + 
  plot_annotation(title = "Study 3: Relative Risk of Help Received Last Week: Money", tag_levels = 'A',
                  theme = theme(plot.title = element_text(size = 16, face = "bold")))
study_3


```


## Introduction

A central question in the scientific study of religion is whether religion fosters cooperation [@johnson2005; @norenzayan2016; @bulbuliaj.2013; @watts2016; @watts2015; @sosis2003cooperation; @sibley2012a]. However, quantifying causal effects for religion, and many other social behaviours, presents significant challenges. Much of religion eludes randomised experiments. Additionally, valid causal conclusions from observational data require both high-resolution time-series data and robust methods for causal inference. Few studies combine these characteristics (refer to @kelly2024religiosity for a recent research summary and @chen2020religious for a well-designed observational study on religious service and mental health).

Moreover, the question "Does religion cause prosociality?" lacks specificity. To refine this question, we must articulate clear causal contrasts and their scale, select specific measures of "prosociality," define our target population, gather appropriate time-series data, and, assuming that causal assumptions and identification criteria are met, calculate statistical estimates. Having obtained these estimates, which rely on assumptions, we must evaluate robustness using sensitivity analyses [@hernan2024WHATIF; @ogburn2021; @bulbulia2022; @linden2020EVALUE].

Here, we use comprehensive panel data from `r n_participants` participants in the New Zealand Attitudes and Values Study from 2018-2021 to quantify the effects of clearly defined interventions in religious attendance across the population of New Zealanders on two features of prosociality: charitable financial donations and volunteering, measured both by self-reported giving as well as help received.

We define causal effects as quantitative contrasts between potential outcomes under intervention on religious service attendance, the "treatment," within a specified population (see @rubin2005; @neyman1923, @robins1986; our approach is consistent with the counterfactual approaches @pearl2009, @vanderlaan2018). We express these interventions as "modified treatment policies" and obtain causal inferences by contrasting inferred population treatment averages [@haneuse2013estimation; @diaz2023lmtp; @díaz2021; @hoffman2023].

Our initial causal contrast investigates: "What would be the average difference across the New Zealand population if everyone attended religious services regularly (at least four times per month) versus if no one attended?" This theoretical question simulates a hypothetical experiment with random assignment to regular or non-attendance.

The second contrast investigates: "What would be the average difference across the New Zealand population if everyone attended religious services regularly compared with maintaining the status quo?" This scenario does not require shifting regular attendees to non-attendance; here, results inform practical policies targeting non-regular attendees who might start.

The third contrast examines: "What would be the average difference across the New Zealand population if no one attended religious services compared with the status quo?"  This scenario does not require shifting those who never attend; here, results inform practical policies targeting regular attendees who might stop.

Although the set of causal contrasts analysts might consider is unbounded, those we have selected here specifically address scientific and policy interests.

Consider that our approach does not focus on testing predefined hypotheses; instead, we aim to compute causal effects with high accuracy by combining appropriate time-series data and robust methods for causal inference [@hernan2024stating].


## Method

### Sample

Data were collected by the New Zealand Attitudes and Values Study (NZAVS), an annual longitudinal national probability panel study of social attitudes, personality, ideology, and health outcomes in New Zealand. Chris G. Sibley started the New Zealand Attitudes and Values Study in 2009, which has grown to include a community of over fifty researchers. Since its inception, The New Zealand Attitudes and Values Study has accumulated questionnaire responses from `r n_total` New Zealand residents. The study operates independently of political or corporate funding and is based in a university setting. Data summaries for our study sample are found in **Appendices B-D**.  For more details about the New Zealand Attitudes and Values Study see: [OSF.IO/75SNB](https://doi.org/10.17605/OSF.IO/75SNB).

### Treatment Indicator

Religious service attendance is assessed in the New Zealand Attitudes and Values Study with the following questions:

- *Do you identify with a religion and/or spiritual group? If yes...How many times did you attend a church or place of worship during the last month?*

We rounded responses to the nearest whole number. Because there were few responses greater than eight, we coded all above eight as eight (see *Appendix B*: we label this variable `religion_church_round`). Note that contrasts with responses greater than four were not intervened upon in the regular church service condition; our decision was to ease the computational burden during estimation.

### Measures of Prosociality


**Study 1: Self-reported charity** The New Zealand Attitudes and Values Study includes two self-reported measures of pro-sociality:

- Volunteering:
  *"Please estimate how many hours you spent doing each of the following things last week...Volunteer/charitable work"*
  
- Annual charitable financial donations:
  *"How much money have you donated to charity in the last year?"*


**Study 2: Help received from others in the last week: *time* ** 

Participants were asked:

*"Please estimate how much help you have received from the following sources in the last week."*

- *Family...TIME (hours)*
- *Friends...TIME (hours)*
- *Community...TIME (hours)*

Owing to the high variability of responses, we transformed responses into binary indicators: *0 = none/ 1 = any*.

**Study 3: Help received from others in the last week: *money* ** 

Similarly, participants were asked:

*"Please estimate how much help you have received from the following sources in the last week."*

- *Family...MONEY (dollars)*
- *Friends...MONEY (dollars)*
- *Community...MONEY (dollars)*

These measures were also highly variable. Hence, we converted responses to binary indicators: *0 = none/ 1 = any*.

Studies 2 and 3 aim to minimise self-presentation bias by using external measures of prosocial outcomes. We assume that if religious institutions foster prosociality, the initiation of regular attendance --controlling for past religious service, past measures of the prosocial outcomes, and rich array of demographic, personality, and health measures recorded at baseline --  will increase exposure to prosocial behaviours. These self-reported measures of dependency on others are robust to self-presentation biases that might associate religious service attendance with indicators of prosociality in the absence of causation.

We provide comprehensive details of all measures in **Appendix A**.


### Causal Interventions

We define three targeted causal contrasts (*causal estimands*) as interventions on prespecified modified treatment policies [@haneuse2013estimation; @diaz2021nonparametric; @diaz2023lmtp]. Let $A_t$ denote the treatment, the monthly frequency of religious service. There are three time points, $t\in{0,1,2}$, where $t=0$ denotes the baseline wave, $t=1$, the treatment wave, and $t=2$ the end of study. $\mathbf{d}(\cdot)$ denotes a modified treatment policy $f_\mathbf{d}$. When a treatment is fixed to a level defined by the treatment, perhaps contrary to the observed treatment, we use the lower case symbol $a_1$. In this study the functions defined by $f_\mathbf{d}$ are inteventions that fix $A_1$ to $a_1$.

We define three targeted causal contrasts (*causal estimands*) using modified treatment policies as outlined in [@haneuse2013estimation; @diaz2021nonparametric; @diaz2023lmtp]. Let $A_t$ represent the treatment variable, the monthly frequency of religious service. This study has three distinct measurement intervals, $t \in \{0, 1, 2\}$, where $t=0$ is the baseline wave, $t=1$ is the treatment wave, and $t=2$ marks the end of the study. The notation $\mathbf{d}(\cdot)$ denotes to a modified treatment policy $f_\mathbf{d}$. When the policy prescribes a treatment level, perhaps contrary to the observed treatment level, we denote this fixed level by the lowercase symbol $a_1$. In this study, the functions defined by $f_\mathbf{d}$ are interventions that set $A_1$ to $a_1$.

1. **Regular Religious Service Treatment**: Administer regular religious service attendance to everyone in the adult population. If an individual's religious service attendance is below four times per month, shift to four; otherwise, maintain their current attendance:

$$
\mathbf{d}^\lambda (a_1) = \begin{cases} 4 & \text{if } a_1 < 4 \\ 
a_1 & \text{otherwise} \end{cases}
$$

2. **Zero Religious Service Treatment**: Ensure no religious service attendance for everyone in the adult population of New Zealand. If an individual's religious service attendance is greater than zero, shift to zero; otherwise, make no change:

$$
\mathbf{d}^\phi (a_1) = \begin{cases} 0 & \text{if } a_1 > 0 \\ 
a_1 & \text{otherwise} \end{cases}
$$

3. **Status Quo -- No Treatment**: Apply no treatment. Each expected mean outcome is calculated using the natural (observed) value of religious service attendance for each individual.


$$
\mathbf{d}(a_1) = a_1
$$


### Causal Contrasts

From these policies, we computed the following causal contrasts.

**Target Contrast A: 'Regular vs. Zero'**: How do the prosocial effects of a society with regular religious service attendance differ from those of a society with zero religious service attendance?

$$ \text{Regular Religious Service vs. Zero Religious Service} = E[Y(\mathbf{d}^\lambda) - Y(\mathbf{d}^\phi)] $$

This contrast simulates a scientifically interesting hypothetical experiment where we could randomise individuals to either regular religious service or none, assessing the differences in prosociality outcomes measured one year after the intervention.

**Target Contrast B: 'Regular vs. Status Quo'**: How does a society with regular religious service attendance compare to its status quo?

$$ \text{Regular Religious Service vs. No Treatment} = E[Y(\mathbf{d}^\lambda) - Y(\mathbf{d})] $$

This contrast reflects a policy-relevant hypothetical experiment examining the effect of transitioning people to regular religious service, assessing whether this intervention reliably alters societal prosociality compared to the current state.

**Target Contrast C: 'Zero vs. Status Quo'**: What are the social consequences for society of zero religious service attendance compared to its status quo?

$$ \text{Zero Religious Service vs. No Treatment} = E[Y(\mathbf{d}^\phi) - Y(\mathbf{d})] $$

This scenario investigates the policy implications of eliminating religious services entirely, questioning whether such a shift would meaningfully affect the prosocial outcomes we measure.

### Identification Assumptions

To consistently estimate causal effects, we must satisfy three assumptions:

1. **Causal consistency:** potential outcomes must correspond with observed outcomes under the treatments in the data. Essentially, we assume potential outcomes do not depend on how the treatment was administered, conditional on measured covariates [@vanderweele2009; @vanderweele2013]. 

2. **Exchangeability** given observed covariates, we assume treatment assignment is independent of the potential outcomes to be contrasted. In other words, there is "no unmeasured confounding" [@hernan2024WHATIF; @chatton2020].

3. **Positivity:** for unbiased estimation, every subject must have a non-zero chance of receiving the treatment, regardless of their covariate values @westreich2010. We evaluate this assumption in each study by examining changes in religious service attendance from baseline (NZAVS time 10) to the treatment wave (NZAVS time 11). For further discussion of these assumptions in the context of NZAVS studies, see @bulbulia2023a.


### Target Population

The target population for this study comprises New Zealand residents as represented in the baseline wave of the New Zealand Attitudes and Values Study (NZAVS) during the years 2018-2019. The NZAVS is a national probability study designed to reflect the broader New Zealand population accurately. Despite its comprehensive scope, the NZAVS does have some limitations in its demographic representation. Notably, it tends to under-sample males and individuals of Asian descent while over-sampling females and Māori (the indigenous peoples of New Zealand). To address these disparities and enhance the accuracy of our findings, we apply New Zealand Census survey weights to the sample data. These weights adjust for variations in age, gender, and ethnicity to better approximate the national demographic composition. Detailed methodologies related to the application of these weights are discussed in @sibley2021.


### Eligibility Criteria

To be included in the analysis of this study, participants needed to meet the following eligibility criteria:

- Enrolled in the 2018 wave of the New Zealand Attitudes and Values Study (NZAVS time 10).
- Complete data on religious service attendance at baseline (NZAVS time 10), with no missing information in this area.
- Missing covariate data at baseline were permitted and subjected to imputation methods to ensure comprehensive analysis.
- Participants who were lost to follow-up or did not respond at the study's conclusion were still included in the initial sample set.

In total, `r n_participants` individuals fulfilled these criteria and were included in the study.



###  Causal Identification 

::: {#tbl-02}
```{=latex}
\lmtptablethree
```
This table presents three Single World Intervention Graphs (SWIGs), one for each of the treatment conditions we compare. Note that we obtain strong confounding control by including baseline measures for both the treatments and outcomes, see @vanderweele2020, @bulbulia2024PRACTICAL.
:::



@tbl-02 presents three Single World Intervention Graphs (SWIGs) that outline our identification strategy [@robins2010alternative; @richardson2013swigsprimer; @shpitser2022multivariate; @richardson2023nested; @richardson2023potential; @shpitser2016causal]. Our approach consistently applies the same identification strategy across all functions estimated in this study. Note that the natural value of the treatment $A$ is obtained from its observed instances and baseline historical data. This method ensures that our analysis accurately captures the causal effects of flexible treatment regimes [@diaz2012population; @young2014identification; @diaz2021nonparametric].



### Confounding Control

To manage confounding in our analysis, we implemented @vanderweele2019's *modified disjunctive cause criterion* by following these steps:

  1. **Identified all common causes** of both the treatment and outcomes to ensure a comprehensive approach to confounding control.
  2. **Excluded instrumental variables** that affect the exposure but not the outcome. Instrumental variables do not contribute to controlling confounding and can reduce the efficiency of the estimates.
  3. **Included proxies for unmeasured confounders** affecting both exposure and outcome. According to the principles of d-separation, using proxies allows us to indirectly control for their associated unmeasured confounders.
  4. **Controlled for baseline exposure** and **baseline outcome**. Both are used as proxies for unmeasured common causes, enhancing the robustness of our causal estimates.

[Appendix B](#appendix-demographics) details the covariates we included for confounding control. These methods adhere to the guidelines provided in [@bulbulia2024PRACTICAL] and were pre-specified in our study protocol [https://osf.io/ce4t9/](https://osf.io/ce4t9/).



### Missing Data


To mitigate the influence of missing data on our study results and enhance the robustness of our analyses, we implemented the following strategies:

**Baseline missingness**: we employed the `ppm` algorithm from the `mice` package in R [@vanbuuren2018] to impute missing baseline data. This method allowed us to reconstruct incomplete datasets by estimating a plausible value for missing observation. Because we could only pass one data set to the `lmtp`, we employed a single imputation. About 2% of covariate values were missing at baseline. Eligibility for the study required fully observed baseline treatment measures as well as treatment wave treatment measures. 

**Outcome missingness**: to address confounding and selection bias arising from missing responses and panel attrition, we applied non-parametrically estimated censoring weights using the `lmtp` package in R [@williams2021]. 

### Statistical Estimator

<!-- We use a Sequentially Doubly-Robust (SDR) estimator designed for the analysis of longitudinal data within causal inference frameworks [@díaz2021]. The estimator is grounded in the principles of robust semi-parametric statistics and is characterised first by its sequential approach to handling time-varying treatments and outcomes and, secondly, its double robustness property. -->

<!-- Fundamentally, the SDR estimator operates by iteratively regressing pseudo-outcomes on observed treatments and covariates at each time point, starting from the end of the observation period and proceeding backwards. This iterative process employs the concept of pseudo-outcomes, which are updated at each step based on the outcome of the regression, incorporating information from both the treatment mechanism and the outcome model. -->

<!-- A key feature of the SDR estimator is its 'multiply robust' property, implying that the estimator remains consistent if, for each time point, either the model for the treatment assignment or the model for the outcome, conditional on past treatments and covariates, is correctly specified. This property significantly enhances the estimator's applicability to real-world scenarios such as ours, where model misspecification is a common challenge. Furthermore, the marriage of SDR and machine learning reduces the dependence of inference on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021]. Estimation was performed using the `lmtp` package [@williams2021].  -->


We perform statistical estimation using non-parametric Targeted Learning, specifically a Targeted Minimum Loss-based Estimation (TMLE) estimator. TMLE is a robust method that combines machine learning techniques with traditional statistical models to estimate causal effects while providing valid statistical uncertainty measures for these estimates [@van2014targeted; @van2012targeted].

TMLE operates through a two-step process that involves modelling both the outcome and treatment (exposure). Initially, TMLE employs machine learning algorithms to flexibly model the relationship between treatments, covariates, and outcomes. This flexibility allows TMLE to account for complex, high-dimensional covariate spaces *efficiently* without imposing restrictive model assumptions [@van2014discussion; @vanderlaan2011; @vanderlaan2018]. The outcome of this step is a set of initial estimates for these relationships.

The second step of TMLE involves "targeting" these initial estimates by incorporating information about the observed data distribution to improve the accuracy of the causal effect estimate. TMLE achieves this precision through an iterative updating process, which adjusts the initial estimates towards the true causal effect. This updating process is guided by the efficient influence function, ensuring that the final TMLE estimate is as close as possible, given the measures and data, to the true causal effect while still being robust to model-misspecification in either the outcome or the treatment model.

Again, a central feature of TMLE is its double-robustness property. If either the treatment model or the outcome model is correctly specified, the TMLE estimator will consistently estimate the causal effect. Additionally, TMLE uses cross-validation to avoid over-fitting and ensure that the estimator performs well in finite samples. Each step contributes to a robust methodology for examining an intervention's *causal* effects on outcomes. The marriage of TMLE and machine learning technologies reduces the dependence on restrictive modelling assumptions and introduces an additional layer of robustness. For further details of the specific targeted learning strategy we favour, see [@hoffman2022; @hoffman2023; @díaz2021]. We performed estimation using the `lmtp` package [@williams2021]. We used the `superlearner` library for non-parametric estimation with the predefined libraries `SL.ranger`, `SL.glmnet`, and `SL.xgboost` [@polley2023; @xgboost2023; @Ranger2017]. We created graphs, tables and output reports using the `margot` package [@margot2024]. All analysis methods follow pre-stated protocols in @bulbulia2024PRACTICAL. 



### Sensitivity Analysis Using the E-value

To assess the sensitivity of results to unmeasured confounding, we report VanderWeele and Ding's "E-value" in all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain away the observed exposure-outcome association [@vanderweele2020; @linden2020EVALUE]. To evaluate the strength of evidence, we use the bound of the E-value 95% confidence interval closest to 1.


### Scope of Interventions

To illustrate the magnitude of the shift interventions we contrast, we provide histograms in @fig-hist, which display the distribution of treatment levels during the treatment wave. @fig-hist *A*: The intervention for regular religious service, represented in these histograms, affects a larger portion of the sample than the zero religious service intervention. @fig-hist *B*: presents the intervention for zero religious service. It involves a smaller portion of the sample compared to the regular religious service intervention. The comparative analysis of the 'Regular' versus 'Zero' interventions addresses the scientifically intriguing question: what is the effect difference in a scenario where religious service is universal versus completely absent?  The intervention that increases attendance to regular service levels from the status quo allows us to consider the potential costs and benefits of widespread religious practice across society. The intervention that eliminates religious service allows us to consider the potential costs and benefits of the widespread loss of religious practice across society.


```{r}
#| label: fig-hist
#| fig-cap: "This figure shows a histogram of responses to religious service frequency in the baseline + 1 wave. Responses above eight were assigned to eight, and values were rounded to the nearest whole number. The red dashed line shows the population average. (A) Responses in the gold bars are shifted to four on the Regular Religious Service intervention. All those responses in grey (four and above) remain unchanged. (B) On the zero intervention, responses in the blue bars are shifted."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 12

graph_density_of_exposure_both <- margot::here_read("graph_density_of_exposure_both")
graph_density_of_exposure_both + theme_classic()
```
{{< pagebreak >}}
### Evidence for Change in the Treatment Variable

@tbl-transition clarifies the change in the treatment variable from the baseline wave to the baseline + 1 wave across the sample.  Assessing change in a variable is essential for evaluating the positivity assumption and recovering evidence for the incident exposure effect of the treatment variable [@vanderweele2020; @danaei2012; @hernan2024WHATIF]. We find that state 4 (weekly attendance) and state 0 present the highest overall. However, we also find that movement between these states reveals they are not deterministic. States 1, 2, 3, and 5 present relatively frequent jumps in and out of these states, suggesting lower stability or measurement error.  

```{r}
#| label: tbl-transition
#| tbl-cap: "This transition matrix captures stability and change in religious service between the baseline and treatment wave. Each cell in the matrix represents the count of individuals transitioning from one state to another. The rows correspond to the state at baseline (From), and the columns correspond to the state at the treatment wave (To). **Diagonal entries** (in **bold**) signify the number of individuals who remained in their initial state across both waves. **Off-diagonal entries** signify the transitions of individuals from their baseline state to a different state in the treatment wave. A higher number on the diagonal relative to the off-diagonal entries in the same row indicates greater stability in a state. Conversely, higher off-diagonal numbers suggest more frequent shifts in the sample from the baseline state to other states."
#| eval: true
#| echo: false

transition_table <- margot::here_read("transition_table")

transition_table$table
```




{{< pagebreak >}}


## Results

### Study 1: Causal Effects of Regular Church Attendance on Self-Reported Volunteering and Self-Reported Volunteering and Donations

#### Regular Religious Service vs. Zero Treatment Contrast for Donations and Volunteering

Results for the treatment contrasts between Regular Religious Service and Zero Religious Service, focusing on self-reported volunteering and charitable donations, are displayed in @fig-1_1 *A* and @tbl-1_1. These results are measured on the difference scale.



```{r}
#| label: tbl-1_1
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal gain of weekly religious service vs. a universal loss of weekly religious service on reported charitable behaviours at the end of the study. Contrasts are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false

tab_all_prosocial |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")


#(group_tab_all_prosocial, "causal_difference", "LMTP")

# t2_charity_donate_z_gain<- here_read("t2_charity_donate_z_gain")
# 
# round( ( sd_donations * tab_all_prosocial[1,1]), 2)
```

For 'donations', the effect estimate is 0.132 [0.102, 0.161]. The E-value for this estimate is 1.507, with a lower bound of 1.426. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.426 to negate the observed effect. Weaker associations would not overturn it. We infer **evidence for causality**.  **On the data scale, this intervention represents a difference of NZD `r diff_charity` per adult per year in charitable giving.**


The effect estimate for 'hours volunteer' is 0.123 [0.09, 0.156]. The E-value for this estimate is 1.482, with a lower bound of 1.389. We infer **evidence for causality**. **On the data scale, this intervention represents a difference of NZD `r diff_volunteer` minutes per adult per week in charitable giving.**


#### Regular Religious Service vs. Status Quo Treatment Contrast for Donations and Volunteering

@fig-1_1 *B* and @tbl-1_2 present results for the treatment contrasts between Regular Religious Service and Status Quo, focusing on self-reported volunteering and charitable donations. These results are measured on the difference scale.


```{r}
#| label: tbl-1_2
#| tbl-cap: "This table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs. the status quo on reported charitable behaviours at the end of the study. Contrasts are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false

library(tidyverse)
library(margot)

tab_all_prosocial_null |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_prosocial_null, "causal_difference", "lmtp")
#round( ( sd_donations * tab_all_prosocial_null[1,1]), 2)
```

For 'donations', the effect estimate is 0.121 [0.102, 0.14]. The E-value for this estimate is 1.477, with a lower bound of 1.422. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.422 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**. **On the data scale, this intervention represents a difference of NZD `r charity_null` per adult per year in charitable giving.**


For 'hours volunteer', the effect estimate is 0.095 [0.066, 0.123]. The E-value for this estimate is 1.404, with a lower bound of 1.317. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.317 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.  **On the data scale, this intervention represents a difference of `r volunteer_null` minutes per adult per week in charitable giving.**


#### Zero Religious Service vs. The Status Quo Treatment Contrast for Donations and Volunteering

@fig-1_1 *C* and @tbl-1_3 present results for the treatment contrasts between Zero Religious Service and Status Quo, focusing on self-reported volunteering and charitable donations. These results are measured on the difference scale.


```{r}
#| label: tbl-1_3
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal loss of weekly religious service vs. the status quo on reported charitable behaviours at the end of the study. Contrasts are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false

#margot::margot_interpret_table(group_tab_all_prosocial_null_zero, "causal_difference", "lmtp")
tab_all_prosocial_null_zero |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")
```


For 'donations', the effect estimate is -0.011 [-0.029, 0.008]. The E-value for this estimate is 1.111, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer that **the evidence for causality is not reliable**. **On the data scale, this intervention represents a difference of NZD `r charity_null_zero` per adult per year in charitable giving, but again, the effect is not reliable.**


The effect estimate for 'hours volunteer' is -0.028 [-0.042, -0.014]. The E-value for this estimate is 1.189, with a lower bound of 1.128. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.128 to negate the observed effect. Again, weaker confounding would not overturn it. We infer **evidence for causality**. **On the data scale, this intervention represents a difference of `r volunteer_null_zero` in volunteering minutes.**


```{r}
#| label: fig-1_1
#| fig-cap: "This figure graphs the results of model estimates for the causal effects of the three causal contrasts of interest on reported charitable behaviours at the study's end. The causal contrasts are: (A) Regular vs. Zero Religious Service (B) Regular Religious Service vs. Status Quo; (C) Zero  Religious Service vs. Status Quo. Contrasts are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 12

# plot_group_tab_all_prosocial <- margot::margot_plot(
#   group_tab_all_prosocial,
#   type = "RD",
#   title = title,
#   subtitle = "Self-Reported Prosociality",
#   estimate_scale = 3,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 2,
#   title_size = 16,
#   subtitle_size = 14,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.1,
#   x_lim_lo = -.35,
#   x_lim_hi =  .25
# )
# 
# plot_group_tab_all_prosocial
# study_1 <- here_read(study_1, "study_1")
study_1
```

{{< pagebreak >}}

### Study 2: Causal Effects of Regular Church Attendance on Support Received From Others -- Time

#### Regular vs. Zero Causal Treatment Contrast for Time Received From Others


@fig-study2 *A* and @tbl-study2A present results for the treatment contrasts between Regular Religious Service and Zero, focusing on voluntary help received from others during the past week (yes/no). These results are measured on the risk ratio scale.


```{r}
#| label: tbl-study2A
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal gain of weekly religious service vs. a universal loss of weekly religious service on voluntary help received from others during the past week (yes/no) at the end of the study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false

tab_received_time |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_received_time, causal_scale = "risk_ratio" , estimand = "lmtp")

```

For 'community gives time', the effect estimate is 1.378 [1.231, 1.541]. The E-value for this estimate is 2.1, with a lower bound of 1.764. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.764 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'friends give time', the effect estimate is 1.187 [1.108, 1.271]. The E-value for this estimate is 1.658, with a lower bound of 1.454. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.454 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'family gives time', the effect estimate is 0.95 [0.901, 1.003]. The E-value for this estimate is 1.288, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

#### Regular Religious Service vs. Status Quo Treatment Contrast for Time Received From Others

@fig-study2*B* and @tbl-2-2 present results for the treatment contrasts between Regular Religious Service and Status Quo, focusing on voluntary help received from others during the past week (yes/no). These results are measured on the risk ratio scale.

```{r}
#| label: tbl-2-2
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal gain of weekly religious service vs. the status quo on voluntary help received from others during the past week (yes/no) at the end of the study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false


tab_received_time_null |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_received_time_null, "risk_ratio", "lmtp")

```



For 'community gives time', the effect estimate is 1.289 [1.174, 1.415]. The E-value for this estimate is 1.899, with a lower bound of 1.626. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.626 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'friends give time', the effect estimate is 1.128 [1.061, 1.199]. The E-value for this estimate is 1.508, with a lower bound of 1.315. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.315 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'family gives time', the effect estimate is 0.958 [0.913, 1.006]. The E-value for this estimate is 1.258, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.


#### Zero Religious Service vs. Status Quo Treatment Contrast for Time Received From Others

@fig-study2 *C* and @tbl-2_3 present results for the treatment contrasts between Zero Religious Service and Status Quo, focusing on voluntary help received from others during the past week (yes/no). These results are measured on the risk ratio scale.


```{r}
#| label: tbl-2_3
#| tbl-cap: "This table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs. the status quo on voluntary help received from others during the past week (yes/no) at the end of the study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false

tab_received_time_null_zero |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_received_time_null_zero, "risk_ratio", "lmtp")

```

For 'family gives time', the effect estimate is 1.008 [0.991, 1.026]. The E-value for this estimate is 1.098, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

For 'friends give time', the effect estimate is 0.95 [0.928, 0.973]. The E-value for this estimate is 1.288, with a lower bound of 1.197. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.197 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'community gives time', the effect estimate is 0.936 [0.889, 0.985]. The E-value for this estimate is 1.339, with a lower bound of 1.14. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.14 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

```{r}
#| label: fig-study2
#| fig-cap: "This figure reports the results of model estimates for the three causal contrasts of interest on help received from others during the past week (yes/no). The causal contrasts are: (A) Regular vs. Zero Religious Service (B) Regular Religious Service vs. Status Quo; (C) Zero  Religious Service vs. Status Quo. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 12

study_2
```


{{< pagebreak >}}

### Study 3: Causal Effects of Regular Church Attendance on Support Received From Others -- Money

#### Regular vs. Zero Causal Contrast on Money Received From Others

@fig-study_3 *A* and @tbl-3_1 present results for the treatment contrasts between Regular Religious Service and Zero, focusing on money received from others during the past week (yes/no). These results are measured on the risk ratio scale.


```{r}
#| label: tbl-3_1
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal gain of weekly religious service vs. universal loss of weekly religious service on financial help received from others during the past week (yes/no) at the end of the study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false


tab_all_received_money |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money, "risk_ratio", "lmtp")

```


For 'community gives money', the effect estimate is 1.376 [1.112, 1.703]. The E-value for this estimate is 2.095, with a lower bound of 1.465. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.465 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'family gives money', the effect estimate is 1.137 [1.028, 1.258]. The E-value for this estimate is 1.532, with a lower bound of 1.198. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.198 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'friends give money', the effect estimate is 1.137 [0.964, 1.342]. The E-value for this estimate is 1.532, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.


#### Regular vs. Status Quo Causal Contrast on Money Received From Others

@fig-study_3 *B* and @tbl-3_2 present results for the treatment contrasts between Regular Religious Service and Status Quo, focusing on money received from others during the past week (yes/no). These results are measured on the risk ratio scale.


```{r}
#| label: tbl-3_2
#| tbl-cap: "This table reports the results of model estimates for the causal effects of a universal gain of weekly religious service vs. the status quo on financial help received from others during the past week (yes/no) at the end of the study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false



tab_all_received_money_null|>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money_null, "risk_ratio", "lmtp")

```

For 'community gives money', the effect estimate is 1.254 [1.098, 1.432]. The E-value for this estimate is 1.818, with a lower bound of 1.426. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.426 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'family gives money', the effect estimate is 1.13 [1.037, 1.232]. The E-value for this estimate is 1.513, with a lower bound of 1.233. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.233 to negate the observed effect. Weaker confounding would not overturn it. We infer **evidence for causality**.

For 'friends give money', the effect estimate is 1.041 [0.951, 1.139]. The E-value for this estimate is 1.248, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

#### Zero vs. Status Quo Causal Contrast on Money Received From Others

@fig-study_3 *C* and @tbl-3_3 present results for the treatment contrasts between Zero Religious Service and Status Quo, focusing on money received from others during the past week (yes/no). These results are measured on the risk ratio scale.



```{r}
#| label: tbl-3_3
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs. the status quo on financial help received from others during the past week (yes/no) at the end of study. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false


tab_all_received_money_null_zero|>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money_null_zero, "risk_ratio", "lmtp")
```


For 'family gives money', the effect estimate on the risk ratio scale is 0.993 [0.953, 1.035]. The E-value for this estimate is 1.091, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

For 'friends give money', the effect estimate on the risk ratio scale is 0.915 [0.809, 1.036]. The E-value for this estimate is 1.412, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

For 'community gives money', the effect estimate on the risk ratio scale is 0.911 [0.796, 1.042]. The E-value for this estimate is 1.425, with a lower bound of 1. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1 to negate the observed effect. Weaker confounding would not overturn it. We infer **that evidence for causality is not reliable**.

```{r}
#| label: fig-study_3
#| fig-cap: "This figure reports the results of model estimates for the three causal contrasts of interest on help received from others during the past week (yes/no). The causal contrasts are: (A) Regular vs. Zero Religious Service (B) Regular Religious Service vs. Status Quo; (C) Zero  Religious Service vs. Status Quo. Contrasts are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 12

study_3
```
{{< pagebreak >}}
### Additional Study: Comparison of Causal Inference Results with Cross-Sectional Regressions


```{r}
#| label: regressions
#| eval: true
#| include: false
#| echo: false
# regression 
# b_church_on_hours_charity<-margot::here_read("b_church_on_hours_charity")
# b_church_on_hours_charity
# lm_coef_fit_church_on_charity_donate
# verify

# contrast for charity

contrast_charity <-
  lmtp::lmtp_contrast(t2_charity_donate_z_gain , ref =  t2_charity_donate_z_zero, type = "additive")
contrast_volunteer <-
  lmtp::lmtp_contrast(t2_hours_charity_z_gain , ref =  t2_hours_charity_z_zero, type = "additive")


b_church_on_hours_charity<-margot::here_read("b_church_on_hours_charity")
b_church_on_charity_donate<-margot::here_read("b_church_on_charity_donate")

#b_church_on_hours_charity

fit_church_on_hours_charity<-margot::here_read("fit_church_on_hours_charity")
fit_church_on_hours_charity
fit_church_on_charity_donate<-margot::here_read("fit_church_on_charity_donate")
fit_church_on_charity_donate

# community time
b_church_on_community_time_binary_exp_false <- here_read("b_church_on_community_time_binary_exp_false")


regression_regular_church_charity <- round( (fit_church_on_charity_donate$coefficients[2]  *  4.2), 2) # monthly
regression_regular_church_charity

regression_regular_church_volunteer_minutes <- round( (fit_church_on_hours_charity$coefficients[2]  *  4.2  * 60), 2) # monthly
regression_regular_church_volunteer_minutes

# donate data scale 
donate_lmtp <- round( contrast_charity$vals$theta* sd_donations, 2)
donate_lmtp
# voluneer_data scale
volunteer_lmtp_mins = round( (contrast_volunteer$vals$theta * sd_volunteer * 60), 2)
volunteer_lmtp_mins
# perc greater

percent_greater_regression_donate  <- round( (regression_regular_church_charity/donate_lmtp), 2) 
percent_greater_regression_volunteer <- round( (regression_regular_church_volunteer_minutes/volunteer_lmtp_mins), 2) 
percent_greater_regression_donate
percent_greater_regression_volunteer


## Ratio
t2_community_time_binary_gain <- here_read("t2_community_time_binary_gain")
t2_community_time_binary_zero <- here_read("t2_community_time_binary_zero")
t2_community_money_binary_gain<- here_read("t2_community_money_binary_gain")
t2_community_money_binary_zero <- here_read("t2_community_money_binary_zero")

contrast_community_time <-
  lmtp::lmtp_contrast(t2_community_time_binary_gain,
                ref =  t2_community_time_binary_zero,
                type = "rr")

community_time_rr <- contrast_community_time$vals$theta

contrast_community_money <-
  lmtp::lmtp_contrast(t2_community_money_binary_gain,
                ref = t2_community_money_binary_zero ,
                type = "rr")

community_money_rr <- contrast_community_money$vals$theta


fit_church_on_community_time_binary<- margot::here_read('fit_church_on_community_time_binary')
fit_church_on_community_money_binary<- margot::here_read('fit_church_on_community_money_binary')

community_time_risk_ratio <- round( exp( fit_church_on_community_time_binary$coefficients[2]  *  4.2 ), 3)
community_time_risk_ratio

community_money_risk_ratio <- round( exp( fit_church_on_community_money_binary$coefficients[2]  *  4.2 ), 3)
community_money_risk_ratio



percent_greater_rr_community_money  <- round( (community_money_risk_ratio/community_money_rr), 2) 
percent_greater_rr_community_time <- round( (community_time_risk_ratio/community_time_rr), 2) 

percent_greater_rr_community_money
percent_greater_rr_community_time

percent_greater_regression_volunteer <- round( (regression_regular_church_volunteer_minutes/volunteer_lmtp_mins), 2) 
percent_greater_regression_donate
percent_greater_regression_volunteer

b_church_on_community_time_binary_exp_false <- here_read("b_church_on_community_time_binary_exp_false")
b_church_on_community_money_binary_exp_false <- here_read("b_church_on_community_money_binary_exp_false")

b_church_on_community_time_binary_exp_true <- here_read("b_church_on_community_time_binary_exp_true")
b_church_on_community_money_binary_exp_true <- here_read("b_church_on_community_money_binary_exp_true")



# key values

b_church_on_community_time_binary_exp_false
b_church_on_community_money_binary_exp_false
percent_greater_rr_community_money
percent_greater_rr_community_time


# make pretty numbers
regression_regular_church_volunteer_minutes <- prettyNum(regression_regular_church_volunteer_minutes,big.mark=",")


```


To better evaluate the contributions of our methodology to current practice, we conducted a series of cross-sectional analyses using the baseline wave data. We quantified the statistical associations between religious service attendance and our focal prosocial outcomes. We included all regression covariates from the causal models (including sample weights) for each analysis, obviously omitting the outcome measured at baseline, i.e. the response variable, as a predictor.

**Cross-sectional volunteering result**: the change in expected hours of volunteer work for a one-unit increase in religious service attendance is `r b_church_on_hours_charity`. Multiplying this by 4.2 gives a monthly estimate of `r regression_regular_church_volunteer_minutes` minutes. This result is `r percent_greater_regression_volunteer` per cent greater than the effect estimated from the 'regular vs. zero' causal contrast, indicating an overstatement in the regression model.

**Cross-sectional charitable donations result**: The coefficient for religious service on annual charitable donations suggests a change in expected donation amount per unit increase in attendance is: `r b_church_on_charity_donate`. When adjusted to a monthly rate by multiplying by 4.2, this value equals NZ Dollars `r regression_regular_church_charity`. It is `r percent_greater_regression_donate` per cent greater than our causal contrast estimate, again indicating an overstatement by the regression.

For Studies 2 and 3, which focus on community help received, we adjusted our analysis for the non-collapsibility of odds ratios by assuming a Poisson distribution for the outcome variables, obtaining a rate ratio that approximates a risk ratio [@huitfeldt2019collapsibility;@vanderweele2020]:

**Cross-sectional community assistance received result: Time**: the exponentiated change in expectation for a one-unit change in religious service attendance is `r b_church_on_community_time_binary_exp_true` approximate risk rate ratio. The monthly rate ratio derived by multiplying this coefficient by 4.2 is `r community_time_risk_ratio`. This estimate is `r percent_greater_rr_community_time` per cent greater than the 'regular vs. zero' causal estimate, pointing to an overestimation in the regression.

**Cross-sectional community assistance received result: Money**: similarly, the exponentiated change for money received yields an approximate risk ratio of `r b_church_on_community_money_binary_exp_true`. The monthly risk ratio, after adjustment, is `r community_money_risk_ratio`. This rate ratio is `r percent_greater_rr_community_money` per cent greater than the causal estimate, again revealing the bias of cross-sectional regressions.

*These findings underscore that the results of cross-sectional regressions, although suggestive, can considerably diverge from those obtained from the causal analysis of panel data.*
{{< pagebreak >}}
### What is the "Cash Value" of Religious Service Attendance for Charitable Donations in New Zealand? 

```{r}
#| label: cashvalue
#| eval: true
#| include: false
#| echo: false

# avoid scientific notation
options(scipen = 999)

# # model estimates for charity under each intervention
# average annual donations 

# sample mean at end of study (won't be population mean)
mean_donations <- margot::here_read("mean_donations")
mean_volunteer <- margot::here_read("mean_volunteer")

# import models
# population mean
t2_charity_donate_z_null <- here_read("t2_charity_donate_z_null")

# all loss
t2_charity_donate_z_zero <- here_read("t2_charity_donate_z_zero")

# all gain
t2_charity_donate_z_gain <- here_read("t2_charity_donate_z_gain")

# calculate  population estimates for means of annual charitable donations using the models
# note these correctly account for sample weights
mean_donations_gain <-round( mean_donations  + ( t2_charity_donate_z_gain$theta * sd_donations ), 2)
mean_donations_null <-round( mean_donations  + ( t2_charity_donate_z_null$theta * sd_donations ), 2)

mean_donations_zero<- round( mean_donations  + ( t2_charity_donate_z_zero$theta * sd_donations ), 2)


mean_volunteer_gain<- round( mean_volunteer  + ( t2_hours_charity_z_gain$theta * sd_volunteer ), 2)
mean_volunteer_null<- round( mean_volunteer  + ( t2_hours_charity_z_null$theta * sd_volunteer ), 2)
mean_volunteer_zero <- round( mean_volunteer  + ( t2_hours_charity_z_zero$theta * sd_volunteer ), 2)

# sanity checks for differences 
donations_lmtp_up = round( abs(mean_donations_null - mean_donations_gain), 2)
donations_lmtp_down = round(mean_donations_null - mean_donations_zero, 2)

volunteer_lmtp_up = round(abs(mean_volunteer_null - mean_volunteer_gain), 2)
volunteer_lmtp_down = round((mean_volunteer_null - mean_volunteer_zero), 2)


# verify
contrast_char <- lmtp::lmtp_contrast(t2_charity_donate_z_gain , ref =  t2_charity_donate_z_zero, type = "additive")

contrast_vol <- lmtp::lmtp_contrast(t2_hours_charity_z_gain , ref =  t2_hours_charity_z_zero, type = "additive")



# verified: this is exactly correct
#contrast_char$vals$theta * sd_donations

# #number of people in NZ > 18 year olds in 2021/22
nz_adult_population <- 3989000
nz_20_over <- 3.8391
#source; https://www.stats.govt.nz/information-releases/national-population-estimates-at-30-june-2021


# annual charity estimate for the population = Status Quo
factual_charity_estimate <- round( nz_adult_population * mean_donations_null, 2)

mean_donations_null
# counterfactual charity estimate if all were to attend religious service regularly
counterfactual_factual_charity_estimate_up = round(nz_adult_population * mean_donations_gain, 2)

factual_charity_estimate_pop = round(nz_adult_population * mean_donations_null, 2)


# counterfactual charity estimate if all were to lose religious service regularly
counterfactual_factual_charity_estimate_down = round(nz_adult_population * mean_donations_zero, 2)
net_loss = round(counterfactual_factual_charity_estimate_down - factual_charity_estimate, 2)
net_gain = round(counterfactual_factual_charity_estimate_up - factual_charity_estimate, 2)
# net_loss
# net_gain
 
# compute differences at the national level: gain
counterfactual_difference_charity_donate_up = round( counterfactual_factual_charity_estimate_up - factual_charity_estimate, 2)
#counterfactual_difference_charity_donate_up

# compute differences at the national level: loss
counterfactual_difference_charity_donate_down =  round(factual_charity_estimate - counterfactual_factual_charity_estimate_down, 2)
#counterfactual_difference_charity_donate_down

# differences  - absolute scale
#print(counterfactual_difference_charity_donate_up)
#print(counterfactual_difference_charity_donate_down)
# 
# # ~ 2.5 billion NZD 
# # 2503913110
# 
# nz annual budget in 2021

nz_annual_budget = 14494000000 * 4 # treasury (https://www.treasury.govt.nz/publications/budgets/budget-2021)
# 
# # Intervention as a percentage of gov't annual budget 

# percentage of gain/loss as a function of nz government annual budget
per_gain <- round( counterfactual_difference_charity_donate_up/ nz_annual_budget, 3)
per_lost <- round( counterfactual_difference_charity_donate_down/ nz_annual_budget, 3)

# per_gain
# per_lost
# prop of values
prop_loss_of_gain <- round( abs(net_loss/net_gain), 2)

# make pretty numbers
nz_adult_population <- prettyNum(nz_adult_population,big.mark=",")
net_gain <- prettyNum(net_gain,big.mark=",")
net_loss <- prettyNum(net_loss,big.mark=",")
nz_annual_budget <-  prettyNum(nz_annual_budget,big.mark=",")
factual_charity_estimate_pop <- prettyNum(factual_charity_estimate_pop,big.mark=",")
factual_charity_estimate <- prettyNum(factual_charity_estimate,big.mark=",")


```

We leveraged our findings to estimate the economic value of religious service attendance by comparing expected donation amounts under different scenarios— 'Regular Religious Service', 'Zero Religious Service', and the 'Status Quo'.

- **Regular Religious Service**: an increase in religious service attendance yields **an individual average donation rate of NZD `r mean_donations_gain`.**
- **Zero Religious Service**: Reducing religious service attendance to zero yields **an average donation rate of NZD `r mean_donations_zero`.**
- **Status Quo**: the expected individual **average donation rate is currently NZD `r mean_donations_null`.**

With `r nz_adult_population` adult residents in New Zealand in 2021.[^pop] 

- Multiplying the adult population by the average donation sum gives a status quo national estimate for charitable giving of NZD `r factual_charity_estimate_pop`. 
- The net gain to charity from country-wide regular attendance at religious services, compared to the status quo, is NZD `r net_gain`.
- Conversely, the net cost to charity from a complete cessation of regular religious service attendance is NZD `r net_loss`. However, recall the confidence interval crosses zero, and this effect is not reliable.

To provide context, consider these economic consequences against the New Zealand government’s annual budget in year outcomes were measured (2021-2022) is NZD `r nz_annual_budget`


- **The expected gain from a nationwide adoption of regular religious service represents `r per_gain` of New Zealand's annual government budget 2021.**
- We do not obtain a reliable effect from the loss intervention.

<!-- - A nationwide discontinuation of all religious services constitutes `r per_lost` of the annual government budget. Although, again, we observed that the estimate in under this observation is not statistically reliable. We cannot exclude a difference between the zero intervention and the status quo. -->

[^pop]: [National Population Estimates at 30 June 2021](https://www.stats.govt.nz/information-releases/national-population-estimates-at-30-june-2021)
[^source]: [Budget 2021](https://www.treasury.govt.nz/publications/budgets/budget-2021)


Clearly, some amount of charity is always better than none. We do not devalue any form of charitable giving; *every act matters.* Furthermore, our population-wide estimates for the aggregated effects of *religious behaviour* do not clarify the effects of gaining or losing *religious institutions*; that is, our analysis reflects the contributions made by those who participate in religious institutions. 

Focusing on the individual level and aggregating effects across the adult population, the counterfactual scenario in which all New Zealand adults regularly attend religious services shows a substantial increase in society-wide charitable support compared to the status quo, one year after the intervention. However, suppose New Zealand were to experience a complete cessation of religious service attendance. We do not find evidence that the landscape of charitable giving would change from its current state one year later.

## Discussion 

### Considerations

First, consider **unmeasured confounding**: although we use robust methods for causal inference, our results are contingent upon the adequacy of our identification strategy. We employ non-parametric machine learning ensembles for causal estimation, enhancing modelling efficiency and reducing threats of model misspecification; moreover, our sensitivity analyses describe the potential effects of unmeasured confounding. However, the presence and influence of such confounders remains uncertain because we have not, and clearly cannot, randomise individuals into treatment conditions.

Second, **nonparametric estimation/machine learning is new**: although the threat of model misspecification is greatly diminished in non-parametric machine learning, whether our estimators obtain valid standard errors is a function of sample size. Against this concern, our relatively large sample size of `r n_participants` participants offers confidence. However, investigators should be mindful that the application of non-parametric machine learning ensembles to psychological data is new and untested, and those who rely on small samples should use these estimators with caution because standard errors are uncertain (so to speak) in small samples [@van2014discussion].

Third, our study confronts the spectre of **measurement error**: both direct and correlated measurement errors can introduce biases, either by implying effects where none exist or by attenuating true effects [@vanderweele2012MEASUREMENT]. Although evaluating prosociality using multiple measures helps to mitigate concerns about whether religious service affects charitable donations and volunteering, unknown combinations of measurement error might nevertheless bias our results. The outcomes and estimates we report here are best-considered approximations.

Fourth, we do not examine **treatment effect heterogeneity**: identifying which subgroups experience the strongest responses remains a task for future research. Such investigations are crucial for making informed policy decisions and tailoring advice relevant to those subgroups of the population who might benefit most. Perhaps the most obvious stratum of interest are those who already affiliate with a religion at baseline.

Sixth, the **transportability of our findings remains unclear**: New Zealand is our target population. Our findings generalise to this population. The transportability of our findings to other settings -- that is, whether our results generalise beyond this target New Zealand population -- remains an open question, a matter for future investigations. 

### Observations and Recommendations

First, we emphasise that **different causal questions will lead to different causal answers.** Our study obtains causal effect estimates by (1) stating contrasts for specific interventions that increase and decrease religious service attendance across the population; (2) balancing these counterfactual interventions on confounders through doubly robust estimators that flexibly employ machine learning emsembles and cross-validation, and (3) evaluating contrasts of counterfactual projections one year later. In the scientifically interesting contrast ‘Regular vs Zero’, we observe that the social benefits of regular religious attendance on charitable donations and volunteering are considerable. Moreover, we also find the expected effect of regular religious service enacted in contemporary New Zealand would strongly enhance charity and volunteering above the status quo, a point that should shape discussions among critics of religion. Conversely, the overall social effects of altogether ceasing religious services are minimal compared with the status quo. Should New Zealand's society entirely forego regular religious attendance, we would not anticipate significant differences in the aggregate of charitable donations and volunteering one year later. Although we did not investigate longer-term effects, this finding should alleviate at least some concerns among those wary of New Zealand's longstanding secular trend. **We recommend that future research distinguish between models of cultural gains and losses** (as exemplified by @vantongeren2020). More generally, **we recommend that investigators clearly define, compute, and discuss their prespecified causal contrasts in light of well-defined scientific theories and policy interests.**   


Second, we caution that traditional **measures of effect such as Cohen's *D* or $R^2$ may present misleading pictures of practical significance.**  Our study computed standardised differences in the continuous charity and volunteering outcomes. The contrast between regular religious service and no service yeilds and effects size of 0.132 [0.102, 0.161]. This effect would be categorised as 'small' effect by contemporary experimental conventions. Nonetheless, we project that the difference in annual charitable donations between the regular service condition and the status quo amounts to NZD `r mean_donations_gain` versus NZD `r mean_donations_null`. Not only is this effect practically significant at the individual level, the difference amounts to over 4% of New Zealand's Annual Government Budget (2021). **We recommend using causal inference to prioritise, evaluate and communicate practical effect sizes. Conventional statistical measures should not be mistaken for metrics of practical significance.**   


Third, our novel measures of prosociality—**assistance and financial support received from one's community**--align with the expectations of self-reported data on charitable donations and volunteering. We detect marginal causal effects of religious service attendance on **self-reported prosocial dependence and benefit** compared to non-attendance. Notice that just as our measures of prosocial actions (charity/volunteering) do not capture the targets of prosocial help, likewise our measures of community help received (time/money) do not capture the specific sources of help beyond the categories of "community," "freinds," and "family". Nevertheless, the evolutionary theories of religious prosociality that motivate this study are grounded on within group-cooperative benefits [@sosis2003cooperation; @johnson2015; @bulbulia2009charismatic;  @whitehouse2023; @schloss2011evolutionary]. This consistency between self-reported giving and help received aligns with public data in New Zealand;[^charities] where religious institutional charity accounts for 40% of the charitable sector (refer to @McLeod2020, p.17, and also @brooks2004faith; @woodyard2014doing; @monsma2007religion). Additionally, it has been suggested that religious institutions are particularly efficient due to low administrative costs and high volunteer engagement [@khanna1995charity; @bekkers2011literature; @McLeod2020, p.26]. We note that the community signals of religious giving are consistent with the theory that religion promotes altruism outside families, friendships, and even among strangers [@mccullough2020kindness]. **In contexts where investigators are concerned about self-presentation biases when surveying charitable donations and volunteering, we recommend piloting and deploying measures that capture community-received assistance.**

[^charities]: [https://www.charities.govt.nz/view-data/](https://www.charities.govt.nz/view-data/)

Fourth, our secondary analysis reveals that standard cross-sectional regressions overstate the causal effects estimates obtained from a conscientious application of causal methods to time-series data. Importantly, associational models might also *understate* true causal relationships, especially if adjustments involve mediators [@westreich2013; @mcelreath2020]. Despite their effectiveness and relevance, causal inference methods remain uncommon in many human sciences, including the psychological sciences, where traditional associational methods still dominate. Transitioning to causal data science might seem daunting. Yet, causation inherently unfolds in time, with causes preceding effects. Estimating average causal effects involves multi-stepped workflows and robust time-series data [@neal2020introduction; @bulbulia2023]. The need for researchers to develop new skills and the requirement for robust time-series data have profound implications for research design, funding, and the accepted pace of scientific progress. However, causal inference methods have become standard in epidemiology [@lash2020] and are rapidly gaining traction in economics [@angrist2009mostly; @athey2019; @athey2021] and sociology [@montgomery2018]. Moreover, new global panel studies suited for researching religion @johnson2022global offer exciting opportunities for psychological scientists to apply causal methods to those psychological questions at the heart of our discipline. The intellectual benefits of retooling, then, are considerable (refer also to @major2023exploring). 

Suppose these intellectual benefits were not sufficient motivation. It is vital to remember that investigators should be able to coherently evaluate the practical interest of their findings. Absent causal inferential workflows, associational methods cannot deliver. This *practical significance gap* remains even when associational methods are applied to time series data and even when the associational methods are sophisticated—as in the statistical structural equation tradition and multilevel modelling traditions. A *practical significance gap* arises because we cannot interpret associations as causation unless we have obtained balance on the treatments to be compared, unless we know that causes precede effects in time, and unless we ensure that all disabling assumptions have been addressed. 

Because we assume investigators seek both scientific knowledge and practical understanding, **we strongly recommend the broader adoption of causal inferential methods in the psychological sciences.** (refer to @vanderweele2015)


Finally, it is essential to frame our contribution. This study is one of the first to combine robust causal methods with national panel data to quantitatively investigate the social consequences of religious behaviour. It provides evidence that religious service attendance causally affects charitable donations and volunteering. These findings establish *how much* religious service affects charity and volunteering in New Zealand and possibly in similar cultural settings. We hope this research will serve as a guide for studies investigating the social consequences of religious behaviour in other countries. 

However, much remains to be discovered about the mechanisms by which religious service attendance affects social behaviours, even within New Zealand.  Here, we have raised a series of *how much?* questions about the influence of religious service on charity and volunteering. We do not attempt to elucidate the pathways by which such effects operate; nor do we describe their heterogeneity. A substantial body of research has examined the psychological and cultural features through which religious behaviours affect individuals [@norenzayan2016; @shaver2020church; @watts2016; @mccullough2016christian; @konvalinka2011synchronized; @lang2015effects; @cristofori2016neural; @shaver2015evolution; @sosis2005does]. Yet obtaining a data-driven causal understanding about the pathways through which these features operate presents both conception and data challenges [@robins1992; @vanderweele2015; @Diaz2023; @diaz2021nonparametric]. Such challenges are underappreciated (refer to @bulbulia2023). However, by carefully navigating the uncertainties inherent in observational data and leveraging the insights from comprehensive time-series data, we should eventually seek inferences about these mechanisms of operation and their variations. In the years ahead, **we recommend a broader application of causal inferential methods over multiple repeated waves to address questions of causal mediation and interaction**. 


{{< pagebreak >}}

### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study (NZAVS). Full copies of the NZAVS data files are held by members of the NZAVS management team and research group. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author, or any member of the NZAVS advisory board for the purposes of replication or checking of any published study using NZAVS data. The code for the analysis can be found at: [https://github.com/go-bayes/models/blob/main/scripts/24-bulbulia-church-prosocial.R](https://github.com/go-bayes/models/blob/main/scripts/24-bulbulia-church-prosocial.R). 


### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or the decision to publish.


### Author Statement

JB conceived of the study and approach. CS led NZAVS data collection. All authors contributed to the manuscript.


{{< pagebreak >}}

## Appendix A: Measures {#appendix-measures}

#### Age (waves: 1-15)

We asked participants' ages in an open-ended question ("What is your age?" or "What is your date of birth?").


#### Born in New Zealand


#### Charitable Donations  (Study 1 outcome)

Using one item from @hoverd_religious_2010, we asked participants, "How much money have you donated to charity in the last year?". 

<!-- To stabilise this indicator, we took the natural log of the response + 1. -->


#### Charitable Volunteering  (Study 1 outcome)


We measured hours of volunteering using one item from @sibley2011: "Hours spent … voluntary/charitable work."

<!-- To stabilise this indicator, we took the natural log of the response + 1. -->

#### Children Number (waves: 1-3, 4-15)

We measured the number of children using one item from @Bulbulia_2015. We asked participants, "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, or adopted?" or "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, and/or parented?" (waves: 12-15).

<!-- Here, we used responses to create a binary indicator for parenthood: Parent Yes = 1/ No = 0.  -->



<!-- #### Community: Sense of Community (Study 2 outcome) -->

<!-- We measured a sense of community with a single item from @sengupta2013: "I feel a sense of community with others in my local neighbourhood." Participants answered on a scale of 1 (strongly disagree) to 7 (strongly agree). -->


<!-- #### Community: Felt Belongingness  (Study 2 outcome) -->

<!-- We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument [@hagerty1995]: (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded. -->

<!-- #### Community: Social Support (Study 2 outcome) -->

<!-- Participants perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated how much they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item. -->

#### Disability

We assessed disability with a one-item indicator adapted from @verbrugge1997. It asks, "Do you have a health condition or disability that limits you and that has lasted for 6+ months?" (1 = Yes, 0 = No).


#### Education Attainment (waves: 1, 4-15)

We asked participants, "What is your highest level of qualification?". We coded participants' highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing) See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf

#### Employment (waves: 1-3, 4-11)

We asked participants, "Are you currently employed? (This includes self-employed or casual work)".

<!-- #### European (waves: 1-15) -->

<!-- Participants were asked, "Which ethnic group do you belong to (NZ census question)?" or "Which ethnic group(s) do you belong to? (Open-ended)" (wave: 3). Europeans were coded as 1, whereas other ethnicities were coded as 0. -->

#### Ethnicity

Based on the New Zealand Census, we asked participants, "Which ethnic group(s) do you belong to?". The responses were: (1) New Zealand European; (2) Māori; (3) Samoan; (4) Cook Island Māori; (5) Tongan; (6) Niuean; (7) Chinese; (8) Indian; (9) Other such as DUTCH, JAPANESE, TOKELAUAN. Please state:. We coded their answers into four groups: Maori, Pacific, Asian, and Euro (except for Time 3, which used an open-ended measure).


#### Fatigue

We assessed subjective fatigue by asking participants, "During the last 30 days, how often did ... you feel exhausted?" Responses were collected on an ordinal scale (0 = None of The Time, 1 = A little of The Time, 2 = Some of The Time, 3 = Most of The Time, 4 = All of The Time).


#### Honesty-Humility-Modesty Facet (waves: 10-14)

Participants indicated the extent to which they agree with the following four statements from @campbell2004 , and @sibley2011 (1 = Strongly Disagree to 7 = Strongly Agree)

```
i.  I want people to know that I am an important person of high status, (Waves: 1, 10-14)
ii. I am an ordinary person who is no better than others.
iii. I wouldn't want people to treat me as though I were superior to them.
iv. I think that I am entitled to more respect than the average person is.
```

<!-- #### Has Siblings -->

<!-- "Do you have siblings?" [@stronge2019onlychild] -->

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stabilise this indicator, we took the natural log of the response + 1.

<!-- #### Hours Volunteering -->

<!-- We measured hours of volunteering using one item from @sibley2011: "Hours spent … voluntary/charitable work." -->

<!-- To stabilise this indicator, we took the natural log of the response + 1. -->

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Sleep

Participants were asked, "During the past month, on average, how many hours of *actual sleep* did you get per night?".


#### Hours of Work

We measured work hours using one item from @sibley2011: "Hours spent … working in paid employment."

To stabilise this indicator, we took the natural log of the response + 1.


#### Income (waves: 1-3, 4-15)

Participants were asked, "Please estimate your total household income (before tax) for the year XXXX". To stabilise this indicator, we first took the natural log of the response + 1, and then centred and standardised the log-transformed indicator.

<!-- #### Job Security (waves: 1-3,4-7,9-15) -->

<!-- Participants indicated their feeling of job security by answering "How secure do you feel in your current job?" on a scale from 1 (not secure) to 7 (very secure). -->


#### Kessler-6: Psychological Distress  (waves: 2-3,4-15)

We measured psychological distress using the Kessler-6 scale (kessler2002?), which exhibits
strong diagnostic concordance for moderate and severe psychological distress in large, crosscultural
samples (kessler2010?; prochaska2012?). Participants rated during the past 30
days, how often did… (1) “… you feel hopeless”; (2) “… you feel so depressed that nothing
could cheer you up”; (3) “… you feel restless or fidgety”; (4)“… you feel that everything was
an effort”; (5) “… you feel worthless”; (6) ” you feel nervous?” Ordinal response alternatives
for the Kessler-6 are: “None of the time”; “A little of the time”; “Some of the time”; “Most of
the time”; “All of the time.”


#### Male Gender (waves: 1-15)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male as 1, and gender diverse coded as 3 [@fraser_coding_2020]. (or 0.5 = neither female nor male)

Here, we coded all those who responded as Male as 1, and those who did not as 0.


#### Mini-IPIP 6 (waves: 1-3,4-15)

We measured participants' personalities with the Mini International Personality Item Pool 6 (Mini-IPIP6) [@sibley2011], which consists of six dimensions and each dimension is measured with four items:

1.  agreeableness,

    i.  I sympathize with others' feelings.
    ii. I am not interested in other people's problems. (r)
    iii. I feel others' emotions.
    iv. I am not really interested in others. (r)

2.  conscientiousness,

    i.  I get chores done right away.
    ii. I like order.
    iii. I make a mess of things. (r)
    iv. I often forget to put things back in their proper place. (r)

3.  extraversion,

    i.  I am the life of the party.
    ii. I don't talk a lot. (r)
    iii. I keep in the background. (r)
    iv. I talk to a lot of different people at parties.

4.  honesty-humility,

    i.  I feel entitled to more of everything. (r)
    ii. I deserve more things in life. (r)
    iii. I would like to be seen driving around in a very expensive car. (r)
    iv. I would get a lot of pleasure from owning expensive luxury goods. (r)

5.  neuroticism, and

    i.  I have frequent mood swings.
    ii. I am relaxed most of the time. (r)
    iii. I get upset easily.
    iv. I seldom feel blue. (r)

6.  openness to experience

    i.  I have a vivid imagination.
    ii. I have difficulty understanding abstract ideas. (r)
    iii. I do not have a good imagination. (r)
    iv. I am not interested in abstract ideas. (r)

Each dimension was assessed with four items and participants rated the accuracy of each item as it applies to them from 1 (Very Inaccurate) to 7 (Very Accurate). Items marked with (r) are reverse coded.


#### NZ-Born (waves: 1-2,4-15)

We asked participants, "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15).


#### NZ Deprivation Index (waves: 1-15)

We used the NZ Deprivation Index to assign each participant a score based on where they live [@atkinson2019]. This score combines data such as income, home ownership, employment, qualifications, family structure, housing, and access to transport and communication for an area into one deprivation score.


#### NZSEI Occupational Prestige and Status (waves: 8-15)

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017a]. This index uses the income, age, and education of a reference group, in this case the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant an NZSEI-13 score based on their occupation.

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017]. This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant an NZSEI-13 score based on their occupation.

<!-- We asked participants, "If you are a parent, what is the birth date of your eldest child?". -->
<!-- #### Parent (waves: 5-15) -->

<!-- We asked participants, "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-15). Parents were coded as 1, while the others were coded as 0. --> 




#### Opt-in

The New Zealand Attitudes and Values Study allows opt-ins to the study. Because the opt-in population may differ from those sampled randomly from the New Zealand electoral roll; although the opt-in rate is low, we include an indicator (yes/no) for this variable.



#### Partner (No/Yes)

"What is your relationship status?" (e.g., single, married, de-facto, civil union, widowed, living together, etc.)


#### Politically Conservative 

We measured participants' political conservative orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically liberal versus conservative you see yourself as being."

(1 = Extremely Liberal to 7 = Extremely Conservative)


##### Religious Service Attendance

If participants answered *yes* to "Do you identify with a religion and/or spiritual group?" we measured their frequency of church attendence using one item from @sibley2012: "how many times did you attend a church or place of worship in the last month?". Those participants who were not religious were imputed a score of "0".

#### Rural/Urban Codes

Participants residence locations were coded according to a five-level ordinal categorisation ranging from "Urban" to Rural, see @sibley2021.

<!-- #### Politically Right Wing  -->

<!-- We measured participants' political right-wing orientation using a single item adapted from @jost_end_2006-1. -->

<!-- "Please rate how politically left-wing versus right-wing you see yourself as being." -->

<!-- (1 = Extremely left-wing to 7 = Extremely right-wing) -->


<!-- #### Religion Affiliation -->

<!-- Participants were asked to indicate their religion identification ("Do you identify with a religion and/or spiritual group?") on a binary response (1 = Yes, 0 = No). We then asked, "What religion or spiritual group?" These questions are used in the New Zealand Census -->


<!-- #### Support -->

<!-- Participants' perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated how much they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item. -->

#### Short-Form Health

Participants' subjective health was measured using one item ("Do you have a health condition or disability that limits you, and that has lasted for 6+ months?"; 1 = Yes, 0 = No) adapted from @verbrugge1997.


#### Sample Origin

Wave enrolled in NZAVS, see @sibley2021.

<!-- #### Socialising (time 10, time 11) -->

<!-- As part of the time usage measures, participants were asked to state how many hour the spent in activities related to socialising with the following groups: -->

<!--   - Hours spent … socialising with family -->
<!--   - Hours spent … socialising with friends -->
<!--   - Hours spent … socialising with community groups -->

<!--   <!-- - Hours spent … socialising with religious groups -->


#### Support received: money (waves 10-12) (Study 4 outcomes)

The NZAVS has a 'revealed' measure of received help and support measured in hours of support in the previous week. The items are: 


*Please estimate how much help you have received from the following sources in the last week?*

- *family...MONEY (hours)*
- *friends...MONEY (hours)*
- *members of my community...MONEY (hours)*


Because this measure is highly variable, we convert responses to binary indicators: *0 = none/1 any*


#### Support received: time (waves 10-13) (Study 3 outcomes)

*Please estimate how much help you have received from the following sources in the last week.*

- *family...TIME (hours)*
- *friends...TIME (hours)*
- *members of my community...TIME (hours)*

Because this measure is highly variable, we convert responses to binary indicators: *0 = none/1 any*
  

#### Total Siblings

Participants were asked the following questions related to sibling counts: 
 
 - Were you the 1st born, 2nd born, or 3rd born, etc, child of your mother?
 - Do you have siblings?
 - How many older sisters do you have?
 - How many younger sisters do you have?
 - How many older brothers do you have?
 - How many younger brothers do you have?

A single score was obtained from sibling counts by summing responses to the "How many..." items. From these scores, an ordered factor was created ranging from 0 to 7, where participants with more than 7 siblings were grouped into the highest category. 


<!-- #### Volunteers -->

<!-- We asked participants,"Please estimate how many hours you spent doing each of the following things last week" and responded to an item ("voluntary/charitable work") from [@sibley2011]. -->


{{< pagebreak >}}

## Appendix B. Baseline Demographic Statistics {#appendix-demographics}


```{r}
#| label: tbl-table-demography
#| tbl-cap: "Baseline demographic statistics"
#| eval: true
#| include: true
#| echo: false

table_baseline <- margot::here_read("table_baseline")
table_baseline |>  kbl(format = "markdown")
```


@tbl-table-demography baseline demographic statistics for couples who met inclusion criteria.

{{< pagebreak >}}


##  Appendix C: Treatment Statistics {#appendix-exposures}


```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and baseline + 1 (treatment) wave"
#| eval: true
#| include: true
#| echo: false

table_exposures <-margot::here_read("table_exposures")
table_exposures |>  kbl(format = "markdown")
```



tbl-table-exposures-code presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: religious service attendance (range 0-8). Responses coded as eight or above were coded as "8". This decision to avoid spare treatments was based on theoretical grounds, namely, that daily exposure would be similar in its effects to more than daily exposure. We note that causal contrasts were obtained for projects with either no attendance or four or more visits per month. Hence this simplification of the measure is unlikely to affect theoretical and practical inferences. All models adjusted for the pandemic alert level because the treatment wave (NZAVS time 11) occurred during New Zealand's COVID-19 pandemic. The pandemic is not a "confounder" because a confounder must be related to the treatment and the outcome. At the end of the study, all participants had been exposed to the pandemic.  However, to satisfy the causal consistency assumption, all treatments must be conditionally equivalent within levels of all covariates [@vanderweele2013]. Because COVID affected the ability or willingness of individuals to attend religious service, we included the lockdown condition as a covariate [@sibley2021]. To better enable conditional independence within levels of the treatment variable, we conditioned on the lead value of COVID-alert level at baseline. To mitigate systematic biases arising from attrition and missingness, the `lmtp` package uses inverse probability of censoring weights, which were used when estimating the causal effects of the exposure on the outcome. 



### Binary Transition Table for The Treatment

```{r}
#| label: tbl-transition-tablegain
#| tbl-cap: "Transition table for stability and change in regular religious service (4x per month) between baseline and treatment wave."
#| eval: true
#| include: true
#| echo: false

transition_table_binary_gain <-margot::here_read("transition_table_binary_gain")

transition_table_binary_gain$table
```

@tbl-transition-tablegain presents a transition matrix to evaluate treatment shifts between baseline and treatment wave. Here, we focus on the shift from/to monthly attendance at four or more visits per month. Entries along the diagonal (in bold) indicate the number of individuals who **stayed** in their initial state. By contrast, the off-diagonal shows the transitions from the initial state (bold) to another state in the following wave (off diagonal). Thus the cell located at the intersection of row $i$ and column $j$, where $i \neq j$, gives us the counts of individuals moving from state $i$ to state $j$.





```{r}
#| label: tbl-transition-tableloss
#| tbl-cap:  "Transition table for stability and change in zero religious service (0 x per month) between baseline and treatment wave."
#| eval: true
#| include: true
#| echo: false

transition_table_binary_zero <-margot::here_read("transition_table_binary_zero")
transition_table_binary_zero$table
```

@tbl-transition-tableloss presents a transition matrix to evaluate treatment shifts between baseline and treatment wave. Here, we focus on the shift from/to zero religious service attendance. Again, entries along the diagonal (in bold) indicate the number of individuals who **stayed** in their initial state. By contrast, the off-diagonal shows the transitions from the initial state (bold) to another state in the following wave (off diagonal). Thus the cell located at the intersection of row $i$ and column $j$, where $i \neq j$, gives us the counts of individuals moving from state $i$ to state $j$.


### Imbalance of Confounding Covariates Treatments 

@fig-match_1 shows imbalance of covariates on the treatment at the treatment wave. The variable on which there is strongest imbalance is the baseline measure of religious service attendance. It is important to adjust for this measure both for confounding control and to better estimate an incident exposure effect for the religious service at the treatment wave (in contrast to merely estimating a prevalence effect). See @vanderweele2020.

```{r}
#| label: fig-match_1
#| fig-cap: "This figure shows the imbalance in covariates on the treatment"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 10

love_plot_one <- margot::here_read("love_plot_one")
love_plot_one
```


##  Appendix D: Baseline and End of Study Outcome Statistics {#appendix-outcomes}


```{r}
#| label: tbl-table-outcomes
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: true
#| include: true
#| echo: false
table_outcomes <-margot::here_read("table_outcomes")
table_outcomes |>  kbl(format = "markdown")
```



@tbl-table-outcomes presents baseline and end-of-study descriptive statistics for the outcome variables.  


{{< pagebreak >}}



## References {.appendix}








