---
title: "Causal effects of religious service attendance on prosociality evidence from a national longitudinal study"
abstract: |
   The question of whether religion causes prosociality is poorly defined. To make it precise we must state a causal contrasts of interest, specific our measures of "prosociality," and clarify the target population for whom results are meant to generalise. Here we investigate the causal effects of three hypothetical interventions on religious service attendance across the New Zealand population where the levels of intervention address different theoretical and policy questions. Study 1 investigates the effects of these pre-specified interventions on self-reported social behaviours; Study 2 investigates effects on perceived social connection; Study 3 investigates effects voluntary help received from others; Study 4: investigates effects on money received from others . Notably the measures in studies 3 and 4 avoid self presentation bias. Overall we find evidence that religion supports pro-social behaviours across all measurement domains of prosociality; however, the quality and magnitude of benefit varies depending on the estimand of interest -- whether religious service is gained or lost, and whether the contrast condition is the status quo. Moreover magnitudes of benefit are lower than suggested by unmodelled associations in the data. Nevertheless, the gain of regular religious service has economic considerable economic benefits, it's loss, deficits -- which we compute. These findings illustrate a pathway for quantitative investigations of the causal effects of religious behaviours using longitudinal modified treatment policies. 
author: 
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University
    orcid_id: 0000-0003-3169-6576 
  - name: Ken Rice
    affiliation: Georgia State University 
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Geoffrey Troughton
    affiliation: Victoria University of Wellington
    orcid_id: 0000-0001-7423-0640
keywords:
  - Causal Inference
  - Charity
  - Church
  - Cooperation
  - Religion
  - Shift Intervention
  - Volunteering
editor_options: 
  chunk_output_type: console
format:
  pdf:
    sanitize: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [singlecolumn]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true



## WARNING SET THIS PATH TO YOUR DATA ON YOUR SECURE MACHINE. 
pull_path <-
  fs::path_expand(
    #"/Users/joseph/v-project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs_refactor/nzavs_data_23"
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs-current/r-data/nzavs_data_qs"
  )



push_mods <-  fs::path_expand(
  "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/nzvs_mods/24/ow-coop-church-v5"
)


#tinytex::tlmgr_update()

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
#source("/Users/joseph/GIT/templates/functions/libs2.R")

library(margot)
library(ggplot2)

# # WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
# source("/Users/joseph/GIT/templates/functions/funs.R")

# ALERT: UNCOMMENT THIS AND DOWNLOAD THE FUNCTIONS FROM JB's GITHUB
# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )

# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )

# check path:is this correct?  check so you know you are not overwriting other directors
#push_mods

# for latex graphs
# for making graphs
library("tinytex")
library("extrafont")
library("tidyverse")
library("kableExtra")

loadfonts(device = "all")

## Read in data from analysis 
# imports
n_participants <-  margot::here_read("n_participants")
n_total <-  margot::here_read("n_total")

#n_participants
# read standard deviations units of outcome (volunteering/donations) in 2020
sd_donations <- margot::here_read("sd_donations")
sd_volunteer <- margot::here_read("sd_volunteer")

#conflicts_prefer(stats::filter)

# propensity score graph 

# love_plot_base <- margot::here_read("love_plot_base")
# love_plot <- margot::here_read("love_plot")
# # love_plot
# 
# # summary of matching
# summary_match_ebal_base <- margot::here_read("summary_match_ebal")
# # summary_match_ebal$summary_match_ebal_base
# 
# summary_match_ebal <- margot::here_read("summary_match_ebal")
# summary_match_ebal$effective.sample.size

# verify positivity 
#church
transition_table <- margot::here_read("transition_table")
#transition_table
# binary
transition_table_binary <- margot::here_read("transition_table_binary")
#transition_table_binary
# 
# # ordinary regressions
# fit_church_on_hours_charity<- margot::here_read('fit_church_on_hours_charity')
# fit_church_on_charity_donate<-margot::margot::here_read('fit_church_on_charity_donate')
# fit_church_on_community_time_binary<- margot::here_read('fit_church_on_community_time_binary')
# fit_church_on_community_money_binary<- margot::here_read('fit_church_on_community_money_binary')
# 
# 
# lm_fit_church_on_hours_charity <- tbl_regression(fit_church_on_hours_charity)
# 
# 
# # b_church_on_hours_charity <-inline_text(lm_fit_church_on_hours_charity, 
# #                                         variable = religion_church_round, 
# #                                         pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # b_church_on_hours_charity
# #here_save(b_church_on_hours_charity, "b_church_on_hours_charity")
# b_church_on_hours_charity<-margot::here_read("b_church_on_hours_charity")
# 
# 
# # lm_coef_fit_church_on_charity_donate <- tbl_regression(fit_church_on_charity_donate)
# # b_church_on_charity_donate <-inline_text(lm_coef_fit_church_on_charity_donate, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
# # b_church_on_charity_donate
# # here_save(b_church_on_charity_donate, "b_church_on_charity_donate")
# b_church_on_charity_donate<-margot::here_read("b_church_on_charity_donate")
# 
# 
# # lm_coef_fit_church_on_community_time_binary <- tbl_regression(fit_church_on_community_time_binary)
# # b_church_on_community_time_binary <-inline_text(lm_coef_fit_church_on_community_time_binary, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
# # here_save(b_church_on_community_time_binary, "b_church_on_community_time_binary")
# b_church_on_community_time_binary<-margot::here_read("b_church_on_community_time_binary")
# #b_church_on_community_time_binary
# 
# # lm_coef_fit_church_on_community_money_binary <- tbl_regression(fit_church_on_community_money_binary)
# # b_church_on_community_money_binary <-inline_text(lm_coef_fit_church_on_community_money_binary, variable = religion_church_round, pattern = "b = {estimate}; (95% CI {conf.low}, {conf.high})")
# # 
# # here_save(b_church_on_community_money_binary, "b_church_on_community_money_binary")
# b_church_on_community_money_binary<-margot::here_read("b_church_on_community_money_binary")
# 


library(glue)
# works but not used
# testlm_coef_church_on_donate <- glue("b = {fit_church_on_donate_x[2, 1]}, SE = {fit_church_on_donate_x[2, 3]}")



# not this table is too long in most instance 
# table_baseline<-margot::here_read("table_baseline")

# baseline
table_baseline <-margot::here_read("table_baseline")

table_exposures <-margot::here_read("table_exposures")

table_outcomes <-margot::here_read("table_outcomes")

# density of the exposure 
graph_density_of_exposure_up<-margot::here_read("graph_density_of_exposure_up")
graph_density_of_exposure_down<-margot::here_read("graph_density_of_exposure_down")

# results
# import to manuscript -------------------------------------------------------

tab_all_prosocial <-margot::here_read("tab_all_prosocial")
tab_all_prosocial_null <-margot::here_read("tab_all_prosocial_null")
tab_all_prosocial_null_zero <-margot::here_read("tab_all_prosocial_null_zero")


group_tab_all_prosocial<-margot::here_read("group_tab_all_prosocial")
group_tab_all_prosocial_null<-margot::here_read("group_tab_all_prosocial_null")
group_tab_all_prosocial_null_zero<-margot::here_read("group_tab_all_prosocial_null_zero")
## 
# tab_all_prosocial_rr <-margot::here_read("tab_all_prosocial_rr")
# group_tab_all_prosocial_rr <-margot::here_read("group_tab_all_prosocial_rr")

tab_all_perceived_support <-margot::here_read("tab_all_perceived_support")
tab_all_perceived_support_null <-margot::here_read("tab_all_perceived_support_null")
tab_all_perceived_support_null_zero <-margot::here_read("tab_all_perceived_support_null_zero")


group_tab_all_perceived_support <-margot::here_read("group_tab_all_perceived_support")
group_tab_all_perceived_support_null <-margot::here_read("group_tab_all_perceived_support_null")
group_tab_all_perceived_support_null_zero <-margot::here_read("group_tab_all_perceived_support_null_zero")


tab_all_received_money <-margot::here_read( "tab_all_received_money")
tab_all_received_money_null <-margot::here_read( "tab_all_received_money_null")
tab_all_received_money_null_zero <-margot::here_read( "tab_all_received_money_null_zero")


group_tab_all_received_money <-margot::here_read( "group_tab_all_received_money")
group_tab_all_received_money_null <-margot::here_read( "group_tab_all_received_money_null")
group_tab_all_received_money_null_zero <-margot::here_read( "group_tab_all_received_money_null_zero")


tab_contrast_received_time <-margot::here_read( "tab_contrast_received_time")
tab_contrast_received_time_null <-margot::here_read( "tab_contrast_received_time_null")
tab_contrast_received_time_null_zero <-margot::here_read( "tab_contrast_received_time_null_zero")


group_tab_contrast_received_time <-margot::here_read( "group_tab_contrast_received_time")
group_tab_contrast_received_time_null <-margot::here_read( "group_tab_contrast_received_time_null")
group_tab_contrast_received_time_null_zero <-margot::here_read( "group_tab_contrast_received_time_null_zero")



# graphs ------------------------------------------------------------------
title = "Weekly Religious Service vs No Religious Service"
title_null = "Weekly Religious Service vs Status Quo"
title_null_zero = "No Religious Service vs Status Que"
# 
# 
# plot_group_tab_all_prosocial <- margot_plot(
#   group_tab_all_prosocial,
#   type = "RD",
#   title = title,
#   subtitle = "Self-Reported Prosociality",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# 
# plot_group_tab_all_prosocial

# regression scale
# (sd_donations * .173)
# (sd_volunteer * .091)
# .37 * 60
# 
# ggsave(
#   plot_group_tab_all_prosocial,
#   path = here::here(here::here(push_mods)),
#   width = 16,
#   height = 9,
#   units = "in",
#   filename = "plot_group_tab_all_prosocial.png",
#   device = 'png',
#   limitsize = FALSE,
#   dpi = 300
# )

# # # null 
# plot_group_tab_all_prosocial_null <- margot_plot(
#   group_tab_all_prosocial_null,
#   type = "RD",
#   title = title_null,
#   subtitle = "Self-Reported Prosociality",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# 
# plot_group_tab_all_prosocial_null
# 
# #( sd_donations * .096) / 4

# 
# plot_group_tab_all_prosocial_null_zero <- margot_plot(
#   group_tab_all_prosocial_null_zero,
#   type = "RD",
#   title = title_null_zero,
#   subtitle = "Self-Reported Prosociality",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# 
# plot_group_tab_all_prosocial_null_zero
# 
# ( sd_donations * -.022) 
# 
# ( sd_volunteer * -.077) 
# 
# (.31 * 60) 
# 
# plot_group_tab_all_perceived_support <- margot_plot(
#   group_tab_all_perceived_support,
#   type = "RD",
#   title = title,  
#   subtitle = "Perceived Social Support",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# plot_group_tab_all_perceived_support
# 
# plot_group_tab_all_perceived_support_null <- margot_plot(
#   group_tab_all_perceived_support_null,
#   type = "RD",
#   title = title_null,
#   subtitle = "Perceived Social Support",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# plot_group_tab_all_perceived_support_null
# ggsave(
#   plot_group_tab_all_perceived_support_null,
#   path = here::here(here::here(push_mods)),
#   width = 16,
#   height = 9,
#   units = "in",
#   filename = "plot_group_tab_all_perceived_support_null.png",
#   device = 'png',
#   limitsize = FALSE,
#   dpi = 300
# )
# 
# 
# 
# plot_group_tab_all_perceived_support_null_zero <- margot_plot(
#   group_tab_all_perceived_support_null_zero,
#   type = "RD",
#   title = title_null_zero,
#   subtitle = "Perceived Social Support",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = -.5,
#   x_lim_lo = -.5,
#   x_lim_hi =  .5
# )
# plot_group_tab_all_perceived_support_null_zero


# plot_group_tab_all_received_time <- margot_plot(
#   group_tab_contrast_received_time,
#   type = "RR",
#   title = title,
#   subtitle = "Support Received From Others: Time",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# 
# plot_group_tab_all_received_time
# # 
# 
# ggsave(
#   plot_group_tab_all_received_time,
#   path = here::here(here::here(push_mods)),
#   width = 16,
#   height = 9,
#   units = "in",
#   filename = "plot_group_tab_all_received_time.png",
#   device = 'png',
#   limitsize = FALSE,
#   dpi = 300
# )
# 
# 
# 
# plot_group_tab_all_received_time_null <- margot_plot(
#   group_tab_contrast_received_time_null,
#   type = "RR",
#   title = title_null,
#   subtitle = "Support Received From Others: Time",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# 
# plot_group_tab_all_received_time_null
# 
# 
# plot_group_tab_all_received_time_null_zero <- margot_plot(
#   group_tab_contrast_received_time_null_zero,
#   type = "RR",
#   title = title_null_zero,
#   subtitle = "Support Received From Others: Time",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# 
# plot_group_tab_all_received_time_null_zero
# 
# # 
# # 
# # ggsave(
# #   plot_group_tab_all_received_time_null,
# #   path = here::here(here::here(push_mods)),
# #   width = 16,
# #   height = 9,
# #   units = "in",
# #   filename = "plot_group_tab_all_received_time_null.png",
# #   device = 'png',
# #   limitsize = FALSE,
# #   dpi = 300
# # )
# 
# 
# 
# plot_group_tab_all_received_money <- margot_plot(
#   group_tab_all_received_money,
#   type = "RR",
#   title = title,  
#   subtitle = "Support Received From Others: Money",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# plot_group_tab_all_received_money
# #plot_group_tab_all_received_money
# # ggsave(
# #   plot_group_tab_all_received_money,
# #   path = here::here(here::here(push_mods)),
# #   width = 16,
# #   height = 9,
# #   units = "in",
# #   filename = "plot_group_tab_all_received_money.png",
# #   device = 'png',
# #   limitsize = FALSE,
# #   dpi = 300
# # )
# plot_group_tab_all_received_money_null <- margot_plot(
#   group_tab_all_received_money_null,
#   type = "RR",
#   title = title_null,
#   subtitle = "Support Received From Others: Money",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# plot_group_tab_all_received_money_null
# 
# plot_group_tab_all_received_money_null_zero <- margot_plot(
#   group_tab_all_received_money_null_zero,
#   type = "RR",
#   title = title_null_zero,
#   subtitle = "Support Received From Others: Money",
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 18,
#   text_size = 4.5,
#   point_size = 3.5,
#   title_size = 20,
#   subtitle_size = 16,
#   legend_text_size = 10,
#   legend_title_size = 10,
#   x_offset = 0,
#   x_lim_lo = 0,
#   x_lim_hi =  4
# )
# plot_group_tab_all_received_money_null_zero

# 
# ggsave(
#   plot_group_tab_all_received_money_null,
#   path = here::here(here::here(push_mods)),
#   width = 16,
#   height = 9,
#   units = "in",
#   filename = "plot_group_tab_all_received_money_null.png",
#   device = 'png',
#   limitsize = FALSE,
#   dpi = 300
# )
# 



```


## Introduction

A central question in the scientific study of religion is whether religion causes cooperation. However, to address this question is challenging. Most features of religion are inaccessible to experiments. On the other hand, to quantify causality from non-experimental data requires the appropriate modelling of time-series data. Obtaining valid causal inferences from observational data requires a combination of rich time-series data and robust causal inferential methods. Presently such combinations are rare.

Perhaps more fundamental, the question of whether religion causes prosociality is poorly defined. To make this question precise we must to state the causal contrasts of interest, specific measures of "prosociality," and the target population for whom results are meant to generalise. 

Here, we leverage comprehensive panel data from `n_participants` who participated in the New Zealand Attitudes and Values Study from years 2018-2021 and measure the one-year effect of religious attendance across four domains of prosociality. For each domain we compute three estimands: (1) weekly (or more) religious service attendance vs zero attendance. This estimand addresses the question: "what would be the difference if everyone attended weekly religious service compared with if no-one attended?"; (2) weekly (or more) religious service attendance vs the observed attenendance rate, which addresses the question: "what would be the difference if everyone in the country that was not attending weekly religious service switched to regular attendance?" (3) No religious service attendance vs the observed attenendance rate, which addresses the question: "what would be the difference if religious service attendance disappeared altogether?"  The first estimand holds theoretical interest.  The second estimand allows us to evaluate the policy effect of supporting greater religious service attendance. The second estimand allows us to evaluate the policy effect of supporting lower religious service attendance. It is interesting to separately investigate the social consequences of a gain and loss of religious service attendance because these quantities and advice might differ. 


## Method

### Sample

Data were collected as part of The New Zealand Attitudes and Values Study (NZAVS), an annual longitudinal national probability panel study of social attitudes, personality, ideology and health outcomes. The NZAVS began in 2009. It includes questionnaire responses `r n_total` New Zealand residents. Because the NZAVS follows the same people over time, with strong retention, it can track subtle change in attitudes and values over time. The NZAVS is university-based, not- for-profit and independent of political or corporate funding: see: https://doi.org/10.17605/OSF.IO/75SNB

### Outcome-wide measures of prosociality

Rather than cheery picking one or several measures of prosociality, we adopted an unrestrictive "outcome-wide" approach that used a rich array of measures to assess multiple-dimensions of a potentially multi-dimensional construct. The measures we use to investigate this construct: 

**Study 1. Self-reported prosocial behaviours** as measured by:

(a) self-reported hours of weekly volunteering; 
(b) self-reported annual charitable financial donations (dollars);

**Study 2. Perceived social connection** as measured by:

(a) social belonging (ordinal 1-7), 
(b) social support (ordinal 1-7), 
(c) neighbourhood belonging (ordinal 1-7)

**Study 3: Reported personal help received** in the previous week from:

(a) family (yes/no) 
(b) friends (yes/no)
(c) community (yes/no)

**Study 4: Reported financial help received** in the previous week from:

(a) family (yes/no) 
(b) friends (yes/no)
(c) community (yes/no)


We note that Studies 2-4 avoid self-presentation bias by using individuals as measures of prosociality. That is, if religious service attendance were to induce greater within-group prosociality, then regular religious service attendance would increase prosocial exposures. Studies 3 and 4 are especially robust to self-presentation bias because they rely on self-reported dependency.


### Causal Contrasts


To answer a causal question requires that we first state it. Here, we ask three distinct causal questions.

Let $A$  denote religiosu service, the treatment of inteest. We calibrate $f(A)$ according to following rules:

$$
f(A) = 
\begin{cases} 
a, & \text{where } a \text{ is the observed value.}\\
a^* = 4, & \text{if } a < 4 \times \text{monthly religious service attendance; otherwise, } a^* = a.\\
a' = 0, & \text{if } a > 0 \text{ and monthly religious service attendance is positive; otherwise, } a' = a. 
\end{cases}
$$

This function operates as follows:

- **for $a$**: $a$ remains unchanged; estimation is performed for observed value.
  
- **for $a^*$**: this conditionally modifies $a$ to $a^* = 4$ if $a$ is less than four times the monthly religious service attendance. If $a$ does not meet this condition, it remains unaltered as $a^*$ equals $a$. This function enforces a minimum threshold for $a^*$ of religious attendance but allows those above this threshold to remain unchanged.
  
- **for $a^\prime$**: this condition modifies $a$ to $a' = 0$ when the observed value of $a$ is greater than 0, thus setting all values of religious attendance to 0.


We may now formulate our causal contrasts, $Y(\bar{a})$ denote a potential outcome under the treatment rule $f(A)$; $L$ denotes the set of measured covariates we assume to be sufficient for confounding control.

1. Our first causal question asks: "what would be the social consequences of everyone attending religious service compared with no-one attending religious service" which computes the following causal contrast (on the difference scale):

$$ \text{Longitudinal Modified Treatment Policy}^1 = E[Y(a*,a*)|\textcolor{red}{f(A)},L] - E[Y(a^\prime,a^\prime)|\textcolor{red}{f(A)},L] $$

This causal question reflects the theoretical interest. Everyone attending regularly vs. non-one attending regularly. 


2. Our second causal question asks: "what would be the social consequences if everyone who was not attending weekly religious service in New Zealand were to do so?"  The contrast conditions for this estimand are given by (a) regular religious service attendance and (b) the status quo, which can be written:

$$ \text{Longitudinal Modified Treatment Policy}^2 = E[Y(a*,a*)|\textcolor{red}{f(A)},L] - E[Y(a,a)|\textcolor{red}{f(A)},L] $$

This contrasts helps us to evaluate the social effects of a policy that promoted the society-wide adoption of religious service attendance across the contemporary target population. 


3. What would be the social consequences if everyone who was attending religious service in New Zealand were stop.  The contrast conditions are thus no religious service attendance and the status quo, which can be written: 


$$ \text{Longitudinal Modified Treatment Policy}^3 = E[Y(a*,a*)|\textcolor{red}{f(A)},L] - E[Y(a,a)|\textcolor{red}{f(A)},L] $$

This contrasts helps us to evaluate the social effects of a policy that promoted the society-wide loss of religious service attendance in the contemporary population. 


### Identification assumptions

To obtain valid causal effect estimates we must satisfy three identification assumptions.

1.  **Causal consistency**: it must be credible in our study that each outcomes observed in the data corresponds to at least one of the potential outcomes to be compared. Relatedly, this assumption demands that potential outcomes we consider are independent of the versions of treatment administered, conditional on measured covariates. 

This assumption cannot be verified by data.

2.  **Conditional exchangeability**: it must be credible in our study that given the observed covariates, treatment assignment is independent of each of the potential outcomes that we consider. Put differently, it must be credible that our statistical models obtain balance in confounding co-variates in the treatments to be compared.

To improve crediblity for this assumption we conduct analysis using semi-parametric doubly robust estimators. Such estimators minimise restrictive assumptions about a statisitcal model's functional form; they are "doubly-robust" in the sense two models are deployed, one that seeks to obtain balance for all confounders on the treatment, and the other that seeks obtain balance for all confounders in the distribution of potential outcome -- and only one of the two models needs to be correctly estimated to consistently estimate a causal effect. 

Nevertheless, as with the causal consistency assumption, the conditional exchangeability assumption cannot be verified by data. Hence we perform sensitivity analyses, as described below. 

3.  **Positivity**: it must be credible in our study that every subject has a non-zero probability of receiving the treatment, irrespective of their covariates. Below, we evaluate the positivity assumption by reporting the number of instances in which individuals transitioned from one state of religious service in the baseline wave to another state of religious service in the treatment wave.  

<!-- However, although we observe the relevant state changes it is arguably not plausible that everyone would does not attend religious service could be made to do so, weekly, and likewise, that everyone who is regularly attending could be made to desist. Note that wWereas typically a causal estimate requires projecting the entire population into one of two states to be contrasted, a longitudinal modified treatment policy is less restrictive. We do not require of those attending religious service more than weekly that they reduce attendance under the intervention:  $f(A = a^* = a)$.[^note_attend].  -->

<!-- [^note_attend]: Similarly, for those who never attend religious service  $f(A = a^\prime = a)$, however because this intervention is bounded by zero, the zero intervention corresponds to the familiar average treatment effect projection.  -->


<!-- However, to the extent that religious behaviours are deterministic, results lack a clear causal interpretation.  -->


### Target Population 

The target population is the population of New Zealand at the baseline wave. Although the New Zealand Attitudes and Values Study is a national probability study with good representation of the New Zealand population as a whole, its coverage is not perfect.  For example the study under samples males and asians and over samples women and Māori (New Zealand's indigenous peoples).  The `lmtp` package allows us to pass survey weights when computing marginal treatment effects; in this study we weighted the sample to age, gender and ethnicity, following the pre-stated protocols in @bulbulia2024PRACTICAL. 



### Eligibility criteria 

-   Enrolled in the New Zealand Attitudes and Values Study in 2018 (NZAVS time 10).
-   Missing data for all variables at baseline were allowed, and imputed using the `ppm` algorith from the `mice` package [@vanbuuren2018].
-   Attrition/loss to follow up was permented. Inverse probability of censoring weights were calculated as part of estimation in `lmtp` to adjust for missing outcomes at NZAVS Time 12 (years 2020-2021, the outcome wave). See details of the `lmtp` package [@williams2021])

There were `r n_participants` who met these criteria.


### Confounding control

We use @vanderweele2020 *modified disjunctive cause criterion* in which we (1) identified all common causes of the treatment and outcomes; (2) removed any variables identified that might influence the exposure but not the outcome -- instrumental variables -- because instrumental variables are known to reduce efficiency; (3) included proxies for any unmeasured variables affecting both exposure and outcome -- including instrumental variables, because by the rules of d-separation, conditioning on a proxy is akin to conditioning on its parent (in this case, a confounder) (4) **control for baseline exposure** and (5) **control for baseline outcome** both of which serve as proxies for unmeasured common causes [@vanderweele2020]. Note that our shift interventions obtain controlled effects for baseline-treatment and exposure wave-treatment, with the naturally observed treatment levels being explicitly modelled from the data, see [] for details. 



#### Sensitivity Analysis Using the E-value

To assess sensitivity of results to unmeasured confounding, we report VanderWeele and Ding's "E-value" in all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain away the observed exposure-outcome association [@vanderweele2020; @linden2020EVALUE]. To evaluate strength of evidence, we use the bound of the E-value's 95% confidence interval that is closest to 1.

### Analysis

<!-- We use a Sequentially Doubly-Robust (SDR) estimator designed for the analysis of longitudinal data within causal inference frameworks [@díaz2021]. The estimator is grounded in the principles of robust semi-parametric statistics, and is characterised first by its sequential approach to handling time-varying treatments and outcomes, and secondly by its double robustness property. -->

<!-- Fundamentally, the SDR estimator operates by iteratively regressing pseudo-outcomes on observed treatments and covariates at each time point, starting from the end of the observation period and proceeding backwards. This iterative process employs the concept of pseudo-outcomes, which are updated at each step based on the outcome of the regression, incorporating information from both the treatment mechanism and the outcome model. -->

<!-- A key feature of the SDR estimator is its 'multiply robust' property, implying that the estimator remains consistent if, for each time point, either the model for the treatment assignment or the model for the outcome, conditional on past treatments and covariates, is correctly specified. This property significantly enhances the estimator's applicability to real-world scenarios such as ours, where model misspecification is a common challenge. Furthermore, the marriage of SDR and machine learning  reduces the dependence of inference on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021]. Estimation was performed using `lmtp` package [@williams2021].  -->


We employ a semi-parametric estimator known as Targeted Minimum Loss-based Estimation (TMLE). TMLE is a robust method that combines machine learning techniques with traditional statistical models to estimate causal effects while providing valid statistical uncertainty measures for these estimates.

TMLE operates through a two-step process involving both outcome and treatment (exposure) models. Initially, it employes machine learning algorithms to flexibly model the relationship between treatments, covariates, and outcomes. This flexibility allows TMLE to account for complex, high-dimensional covariate spaces without imposing restrictive model assumptions. The outcome of this step is a set of initial estimates for these relationships.

The second step of TMLE involves "targeting" these initial estimates by incorporating information about the observed data distribution to improve the accuracy of the causal effect estimate. This is achieved through an iterative updating process, which adjusts the initial estimates towards the true causal effect. This updating process is guided by the efficient influence function, ensuring that the final TMLE estimate is as close as possible given the measures and data to the true causal effect, while still being robust to model misspecification in either the outcome model or the treatment model.

Again, a central feature of TMLE is its double-robustness property, meaning that if either the model for the treatment or the outcome is correctly specified, the TMLE estimator will still consistently estimate the causal effect. Additionally, TMLE uses cross-validation to avoid over-fitting and ensure that the estimator performs well in finite samples. Each of these steps contributes to a robust methodology for examining the *causal* effects on of interventions on outcomes. The marriage of TMLE and machine learning technologies reduces the dependence on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021]. We performed estimation using `lmtp` package [@williams2021]. We used the `superlearner` library for non-parametric estimation with the predefined libraries `SL.ranger`, `SL.glmnet`, `SL.xboost`. Graphs, tables and output reorts were created using the `margot` package [@margot2024]. 


{{< pagebreak >}}

<!-- ```{tikz} -->
<!-- #| label: fig-outcomewide-dag -->
<!-- #| fig-cap: "Causal graph: three-wave panel design with selection bias. By controling for confounding at baseline we block back-door paths that might lead to an association between the treatment and the outcome. Adding previous measurments of the exposure and outcome at baseline further reduces confounding. Nevertheless, because we cannot ensure all confounding paths have been blocked, we perform sensitivity analysis. These allow us to ask, how strong would the unmeasured confounder need to be in its association with the treatment and the outcome to explain away the observed association." -->
<!-- #| out-width: 80% -->
<!-- #| echo: false -->
<!-- #| include: true -->
<!-- #| eval: true -->
<!-- #| fig-height: 8 -->

<!-- \usetikzlibrary{positioning} -->
<!-- \usetikzlibrary{shapes.geometric} -->
<!-- \usetikzlibrary{arrows} -->
<!-- \usetikzlibrary{decorations} -->
<!-- \tikzstyle{Arrow} = [->, thin, preaction = {decorate}] -->
<!-- \tikzset{>=latex} -->
<!-- % Define a simple decoration -->
<!-- \tikzstyle{cor} = [-, dotted, preaction = {decorate}] -->

<!-- \begin{tikzpicture}[{every node/.append style}=draw] -->

<!-- \node [rectangle, draw=white] (U) at (0, 0) {U}; -->
<!-- \node [rectangle, draw=black](C1) at (0, -1) {C}; -->
<!-- \node [rectangle, draw=black, align=left] (L) at (2, 0) {L$_{t0}$ \\A$_{t0}$ \\Y$_{t0}$}; -->
<!-- \node [rectangle, draw=white] (A) at (4, 0) {A$_{t1}$}; -->
<!-- \node [ellipse, draw=white] (US) at (0, -2) {U}; -->
<!-- \node [rectangle, draw=black](C2) at (6, 0) {C}; -->
<!-- \node [ellipse, draw=white] (Y) at (8, 0) {Y$_{t2}$}; -->

<!-- \draw [-latex, draw=black] (U) to (L); -->
<!-- \draw [-latex, draw=black] (L) to (A); -->
<!-- \draw [-latex, bend left=50, draw=black] (L) to (Y); -->
<!-- \draw [-latex, bend right=50, draw=black, dotted] (U) to (Y); -->
<!-- \draw [-latex, bend left=50, draw=black, dotted] (U) to (A); -->
<!-- \draw [-latex, draw=black] (A) to (C2); -->
<!-- draw [-latex, draw=black] (US) to (C1); -->
<!-- \draw [-latex, draw=black] (US) to (C2); -->
<!-- \draw [-latex, draw=black] (US) to (Y); -->
<!-- \draw [-latex, bend left = 40, draw=red, dashed] (A) to (Y); -->
<!-- \draw [cor, draw=red, bend right=20, dashed] (A) to (US); -->

<!-- \end{tikzpicture} -->


<!-- ``` -->



### Scope of Interventions



### Evidence for Change in the Treatment Variable



## Results



{{< pagebreak >}}

## Study 1: Causal Effects of Regular Church Attendance on Self-Reported Volunteering and Self Reported Charitable Donations

```{r}
#| label: fig_1_1
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_prosocial <- margot_plot(
  group_tab_all_prosocial,
  type = "RD",
  title = title,
  subtitle = "Self-Reported Prosociality",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_prosocial
```


::: {#tbl_1_1}
```{r}
#| label: tbl_11
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false


tab_all_prosocial |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_prosocial, "causal_difference", "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and constrast groups for the target population.

For the outcome 'relig service: donations', the lmtp is 0.173 [0.097,0.25]. The E-value for this effect estimate is 1.617 with a lower bound of 1.409. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.409, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.409) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: hours volunteer' is positive, however, given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

{{< pagebreak >}}

```{r}
#| label: fig_1_2
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_prosocial_null <- margot_plot(
  group_tab_all_prosocial_null,
  type = "RD",
  title = title_null,
  subtitle = "Self-Reported Prosociality",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_prosocial_null
```


::: {#tbl_1_2}
```{r}
#| label: tbl_1_2
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false


tab_all_prosocial_null |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

# margot::margot_interpret_table(group_tab_all_prosocial_null, "causal_difference", "lmtp")

```
:::

For the outcome 'relig service: donations', the lmtp is 0.096 [0.02,0.171]. The E-value for this effect estimate is 1.407 with a lower bound of 1.154. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.154, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.154) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: hours volunteer', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

{{< pagebreak >}}

```{r}
#| label: fig_1_3
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_prosocial_null_zero <- margot_plot(
  group_tab_all_prosocial_null_zero,
  type = "RD",
  title = title_null_zero,
  subtitle = "Self-Reported Prosociality",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_prosocial_null_zero
```



::: {#tbl_1_3}
```{r}
#| label: tbl_1_3
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on reported charitable behaviours at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false


tab_all_prosocial_null_zero |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_prosocial_null_zero, "causal_difference", "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: hours volunteer', the lmtp is -0.022 [-0.033,-0.01]. The E-value for this effect estimate is 1.164 with a lower bound of 1.107. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.107, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.107) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: donations', the lmtp is -0.077 [-0.099,-0.056]. The E-value for this effect estimate is 1.352 with a lower bound of 1.285. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.285, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.285) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

{{< pagebreak >}}

## Study 2: Causal Effects of Regular Church Attendance on Self-Reported Social Connection

```{r}
#| label: fig_2_1
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on perceived social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_perceived_support <- margot_plot(
  group_tab_all_perceived_support,
  type = "RD",
  title = title,
  subtitle = "Perceived social connection",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_perceived_support
```

::: {#tbl_2_1}
```{r}
#| label: tbl_2_1
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on perceived social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false


tab_all_perceived_support |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_perceived_support, "causal_difference", "lmtp")

```
:::


This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: social belonging', the lmtp is 0.121 [0.031,0.211]. The E-value for this effect estimate is 1.477 with a lower bound of 1.2. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.2, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.2) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: social suport', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

For the outcome 'relig service: neighbourhood community', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

{{< pagebreak >}}

```{r}
#| label: fig_2_2
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on perceived social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_perceived_support_null<- margot_plot(
  group_tab_all_perceived_support_null,
  type = "RD",
  title = title_null,
  subtitle = "Self-Reported Prosociality",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_perceived_support_null
```


::: {#tbl_2_2}
```{r}
#| label: tbl_2_2
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on perceved social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false


tab_all_perceived_support_null |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")


#margot::margot_interpret_table(group_tab_all_perceived_support_null, "causal_difference", "lmtp")

```
:::


This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: social belonging', the lmtp is 0.108 [0.019,0.197]. The E-value for this effect estimate is 1.441 with a lower bound of 1.155. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.155, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.155) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: social suport', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

For the outcome 'relig service: neighbourhood community', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.
> 


{{< pagebreak >}}


```{r}
#| label: fig_2_3
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on perceived social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_perceived_support_null_zero<- margot_plot(
  group_tab_all_perceived_support_null_zero,
  type = "RD",
  title = title_null_zero,
  subtitle = "Self-Reported Prosociality",
  xlab = "",
  ylab = "",
  estimate_scale = 3,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -1.5,
  x_lim_hi =  .5
)

plot_group_tab_all_perceived_support_null_zero
```


::: {#tbl_2_3}
```{r}
#| label: tbl_2_3
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on perceved social connection at the end of study. Outcomes are expressed in standard deviation units."
#| eval: true
#| include: true
#| echo: false

tab_all_perceived_support_null_zero |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_perceived_support_null_zero, causal_scale = "causal_difference", estimand = "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: social suport', the lmtp is -0.011 [-0.018,-0.004]. The E-value for this effect estimate is 1.111 with a lower bound of 1.073. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.073, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.073) would not be sufficient to fully account for the observed effect.Here, we find the evidence for causality is weak.

For the outcome 'relig service: social belonging', the lmtp is -0.013 [-0.019,-0.006]. The E-value for this effect estimate is 1.122 with a lower bound of 1.087. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.087, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.087) would not be sufficient to fully account for the observed effect.Here, we find the evidence for causality is weak.

For the outcome 'relig service: neighbourhood community', the lmtp is -0.016 [-0.023,-0.009]. The E-value for this effect estimate is 1.137 with a lower bound of 1.094. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.094, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.094) would not be sufficient to fully account for the observed effect.Here, we find the evidence for causality is weak.


{{< pagebreak >}}

## Study 3: Causal Effects of Regular Church Attendance on Support Received From Others -- Time


```{r}
#| label: fig_3_1
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

margot_plot <- function(.data,
                        type = c("RD", "RR"),
                        title,
                        subtitle,
                        ylab = "",
                        estimate_scale = 1,
                        base_size = 11,
                        text_size = 2.75,
                        point_size = .5,
                        title_size = 10,
                        subtitle_size = 9,
                        legend_text_size = 6,
                        legend_title_size = 6,
                        x_offset = ifelse(type == "RR", 0, -1.75),
                        x_lim_lo = ifelse(type == "RR", .1, -1.75),
                        x_lim_hi = ifelse(type == "RR", 2.5, 1),
                        simplify_axis = FALSE) {
  type <- match.arg(type)
  xintercept <- if (type == "RR") 1 else 0
  x_axis_label <- if (type == "RR") "Causal risk ratio scale (vertical line marks threshold: risk ratio = 1)" else "Causal difference scale: (vertical line marks threshold: difference = 0)."
  
# Define Reliability based on type
  if (type == "RR") {
    .data$Reliability <-
      ifelse(
        .data$`2.5 %` > 1 & .data$`97.5 %` > 1,
        "positive",
        ifelse(
          .data$`2.5 %` < 1 &
            .data$`97.5 %` < 1,
          "negative",
          "zero_crossing"
        )
      )
  } else {
    .data$Reliability <-
      ifelse(
        .data$`2.5 %` > 0 & .data$`97.5 %` > 0,
        "positive",
        ifelse(
          .data$`2.5 %` < 0 &
            .data$`97.5 %` < 0,
          "negative",
          "zero_crossing"
        )
      )
  }
  # Add a condition to apply axis simplifications
  if (simplify_axis) {
    simplified_theme <- theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "top",
      legend.direction = "horizontal",
      axis.ticks.y = element_blank(),
      plot.title = element_text(face = "bold", size = title_size, hjust = 0),
      plot.subtitle = element_text(face = "bold", size = subtitle_size, hjust = 0),
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt")
    )
  } else {
    simplified_theme <- theme(
      legend.position = "top",
      legend.direction = "horizontal",
      axis.ticks.y = element_blank(),
      plot.title = element_text(face = "bold", size = title_size, hjust = 0),
      plot.subtitle = element_text(face = "bold", size = subtitle_size, hjust = 0),
      legend.text = element_text(size = legend_text_size),
      legend.title = element_text(size = legend_title_size),
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt")
    )
  }
  
  out <- ggplot(
    data = .data,
    aes(
      y = reorder(outcome, .data[[paste0("E[Y(1)]", ifelse(type == "RR", "/", "-"), "E[Y(0)]")]]),
      x = .data[[paste0("E[Y(1)]", ifelse(type == "RR", "/", "-"), "E[Y(0)]")]],
      xmin = .data$`2.5 %`,
      xmax = .data$`97.5 %`,
      group = Estimate,
      color = Reliability
    )
  ) +
    geom_errorbarh(aes(color = Reliability), height = .3, position = position_dodge(width = .3)) +
    geom_point(size = point_size, position = position_dodge(width = 0.3)) +
    geom_vline(xintercept = xintercept, linetype = "solid") +
    theme_classic(base_size = base_size) +
    scale_color_manual(values = c(
      "positive" = "dodgerblue",
      "zero_crossing" = "black",
      "negative" = "orange"
    )) +
    labs(
      x = x_axis_label,
      y = NULL,
      title = title,
      subtitle = subtitle
    ) +
      geom_text(
      aes(x = x_offset * estimate_scale, label = estimate_lab),
      size = text_size,
      hjust = 0,
      fontface = ifelse(.data$Estimate == "unreliable", "plain", "bold")
    ) +
    coord_cartesian(xlim = c(x_lim_lo, x_lim_hi)) +
    simplified_theme 
  # apply the simplified theme conditionally
  return(out)
}
plot_group_tab_all_received_time <- margot_plot(
  group_tab_contrast_received_time,
  type = "RR",
  title = title,
  subtitle = "Support Received From Others: Time",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)
plot_group_tab_all_received_time
```

::: {#tbl_3_1}
```{r}
#| label: tbl_3_1
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false



tab_contrast_received_time |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")




#margot::margot_interpret_table(group_tab_contrast_received_time, causal_scale = "risk_ratio" , estimand = "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: community gives time', the lmtp is 1.439 [1.046,1.979]. The E-value for this effect estimate is 2.234 with a lower bound of 1.265. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.265, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.265) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: friends gives time', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

For the outcome 'relig service: family gives time', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.


{{< pagebreak >}}

```{r}
#| label: fig_3_2
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_received_time_null <- margot_plot(
  group_tab_contrast_received_time_null,
  type = "RR",
  title = title_null,
  subtitle = "Support Received From Others: Time",
  xlab = "",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)

plot_group_tab_all_received_time_null

```


::: {#tbl_3_2}
```{r}
#| label: tbl_3_2
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false


tab_contrast_received_time_null |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_contrast_received_time_null, "risk_ratio", "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: community gives time', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

For the outcome 'relig service: friends gives time', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.

For the outcome 'relig service: family gives time', given the lower bound of the E-value equals 1, we infer there is no reliable evidence for causality.


{{< pagebreak >}}


```{r}
#| label: fig_3_3
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_received_time_null_zero <- margot_plot(
  group_tab_contrast_received_time_null_zero,
  type = "RR",
  title = title_null_zero,
  subtitle = "Support Received From Others: Time",
  xlab = "",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)

plot_group_tab_all_received_time_null_zero
```


::: {#tbl_3_3}
```{r}
#| label: tbl_3_3
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on volunatary help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false



tab_contrast_received_time_null_zero |>
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_contrast_received_time_null_zero, "risk_ratio", "lmtp")


```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: family gives time', the lmtp is 0.976 [0.964,0.987]. The E-value for this effect estimate is 1.183 with a lower bound of 1.129. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.129, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.129) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: friends gives time', the lmtp is 0.966 [0.95,0.982]. The E-value for this effect estimate is 1.226 with a lower bound of 1.155. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.155, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.155) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: community gives time', the lmtp is 0.897 [0.866,0.93]. The E-value for this effect estimate is 1.473 with a lower bound of 1.36. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.36, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.36) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

{{< pagebreak >}}

## Study 4: Causal Effects of Regular Church Attendance on Support Received From Others -- Money


```{r}
#| label: fig_4_1
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6
plot_group_tab_all_received_money <- margot_plot(
  group_tab_all_received_money,
  type = "RR",
  title = title,
  subtitle = "Support Received From Others: Money",
  xlab = "",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)
plot_group_tab_all_received_money
```

::: {#tbl_4_1}
```{r}
#| label: tbl_4_1
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs universal loss of weekly religious service on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false


tab_all_received_money |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money, "risk_ratio", "lmtp")

```
:::


This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: community gives money', the lmtp is 2.663 [1.932,3.671]. The E-value for this effect estimate is 4.767 with a lower bound of 3.274. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 3.274, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 3.274) would not be sufficient to fully account for the observed effect.Here, we find strong evidence for causality.

For the outcome 'relig service: friends gives money', the lmtp is 2.052 [1.323,3.184]. The E-value for this effect estimate is 3.521 with a lower bound of 1.977. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.977, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.977) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: family gives money', the lmtp is 1.656 [1.219,2.249]. The E-value for this effect estimate is 2.698 with a lower bound of 1.736. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.736, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.736) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.


{{< pagebreak >}}

```{r}
#| label: fig_4_2
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_received_money_null <- margot_plot(
  group_tab_all_received_money_null,
  type = "RR",
  title = title_null,
  subtitle = "Support Received From Others: Money",
  xlab = "",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)
plot_group_tab_all_received_money_null
```


::: {#tbl_4_2}
```{r}
#| label: tbl_4_2
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal gain of weekly religious service vs status quo on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false



tab_all_received_money_null|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money_null, "risk_ratio", "lmtp")

```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: community gives money', the lmtp is 2.042 [1.59,2.623]. The E-value for this effect estimate is 3.501 with a lower bound of 2.559. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 2.559, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 2.559) would not be sufficient to fully account for the observed effect.Here, we find strong evidence for causality.

For the outcome 'relig service: friends gives money', the lmtp is 1.718 [1.129,2.616]. The E-value for this effect estimate is 2.829 with a lower bound of 1.511. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.511, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.511) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: family gives money', the lmtp is 1.467 [1.086,1.981]. The E-value for this effect estimate is 2.295 with a lower bound of 1.392. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.392, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.392) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.



{{< pagebreak >}}


```{r}
#| label: fig_4_3
#| fig-cap: "Figure reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_all_received_money_null_zero <- margot_plot(
  group_tab_all_received_money_null_zero,
  type = "RR",
  title = title_null_zero,
  subtitle = "Support Received From Others: Money",
  xlab = "",
  ylab = "",
  estimate_scale = 6,
  base_size = 18,
  text_size = 4.5,
  point_size = 3.5,
  title_size = 20,
  subtitle_size = 16,
  legend_text_size = 10,
  legend_title_size = 10,
  x_offset = -.5,
  x_lim_lo = -3,
  x_lim_hi =  4
)
plot_group_tab_all_received_money_null_zero
```


::: {#tbl_4_3}
```{r}
#| label: tbl_4_3
#| tbl-cap: "Table reports results of model estimates for the causal effects of a universal loss of weekly religious service vs status quo on financial help received from others during the past week (yes/no) at the end of study. Outcomes are expressed on the risk ratio scale."
#| eval: true
#| include: true
#| echo: false



tab_all_received_money_null_zero|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

#margot::margot_interpret_table(group_tab_all_received_money_null_zero, "risk_ratio", "lmtp")
```
:::

This longitudinal modified treatment policy computes the expected difference in outcomes between treatment and contrast groups for the target population.

For the outcome 'relig service: family gives money', the lmtp is 0.886 [0.863,0.91]. The E-value for this effect estimate is 1.51 with a lower bound of 1.429. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.429, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.429) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: friends gives money', the lmtp is 0.837 [0.784,0.894]. The E-value for this effect estimate is 1.677 with a lower bound of 1.483. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.483, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.483) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.

For the outcome 'relig service: community gives money', the lmtp is 0.767 [0.672,0.875]. The E-value for this effect estimate is 1.933 with a lower bound of 1.547. In this context, if there exists an unmeasured confounder that is associated with both the treatment and the outcome, and this association has a risk ratio of 1.547, then it is possible for such a confounder to negate the observed effect. Conversely, any confounder with a weaker association (i.e., a risk ratio of less than 1.547) would not be sufficient to fully account for the observed effect.Here, we find evidence for causality.



## Summary of findings


### Assumptions and limitations

1.  **Causality and confounding**: we employs rigorous causal inference techniques, but these are contingent on the assumption of no unmeasured confounding. (Say more about E-values...)

2.  **Measurement error**: the variables under consideration---exercise and sleep---are self-reported, which might introduce both systematic and random measurement errors. It should be specifically noted that the apparent modesty of the practical effects could arise, in part, to measurement inaccuracy. Given the limitations of self-report measures, the true effect sizes may differ from those estimated. Therefore, while the evidence does suggest a modest impact, the actual real-world effects may be either smaller or larger than the estimates suggest. Given the modest effect sizes and the limitations of self-report measures, future research should explore the use of more objective measures for variables like exercise and sleep. (... etc.)

3.  **Generalisability and transportability**: our findings should be interpreted within the context of the New Zealand population from which the data were sourced. Although the results may have broader relevance, direct extrapolation to different populations or sociocultural settings should be undertaken cautiously.

### Theoretical and practical relevance


{{< pagebreak >}}

### Ethics

The NZAVS is reviewed every three years by the University of Auckland Human Participants Ethics Committee. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.

### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the TempletoReligion Trust (TRT0196; TRT0418). JB received support from the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or the decision to publish.

### Author Statement

TBA

{{< pagebreak >}}

## Appendix A. Measures

## Appendix A: Measurues {#appendix-measures}

#### Age (waves: 1-15)

We asked participants' ages in an open-ended question ("What is your age?" or "What is your date of birth").


#### Charitable Donations  (Study 1 outcome)

Using one item from @hoverd_religious_2010, we asked participants "How much money have you donated to charity in the last year?". 

<!-- To stablise this indicator, we took the natural log of the response + 1. -->


#### Charitable Volunteering  (Study 1 outcome)


We measured hours of volunterering using one item from @sibley2011: "Hours spent … voluntary/charitable work."

<!-- To stablise this indicator, we took the natural log of the response + 1. -->


#### Community: Sense of Community (Study 2 outcome)

We measured sense of community with a single item from @sengupta2013: "I feel a sense of community with others in my local neighbourhood." Participants answered on a scale of 1 (strongly disagree) to 7 (strongly agree).


#### Community: Felt Belongingness  (Study 2 outcome)

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument [@hagerty1995]: (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded.

#### Community: Social Support (Study 2 outcome)

Participants' perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated the extent to which they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item.

#### Disability

We assessed disability with a one item indicator adapted from @verbrugge1997, that asks "Do you have a health condition or disability that limits you, and that has lasted for 6+ months?" (1 = Yes, 0 = No).


#### Education Attainment (waves: 1, 4-15)

We asked participants, "What is your highest level of qualification?". We coded participans highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing) See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf

#### Employment (waves: 1-3, 4-11)

We asked participants, "Are you currently employed? (This includes self-employed or casual work)". \* note: This question disappeared in the updated NZAVS Technical documents (Data Dictionary).

<!-- #### European (waves: 1-15) -->

<!-- Participants were asked "Which ethnic group do you belong to (NZ census question)?" or "Which ethnic group(s) do you belong to? (Open-ended)" (wave: 3). Europeans were coded as 1, whereas other ethnicities were coded as 0. -->

#### Ethnicity

Based on the New Zealand Census, we asked participants, "Which ethnic group(s) do you belong to?". The responses were: (1) New Zealand European; (2) Māori; (3) Samoan; (4) Cook Island Māori; (5) Tongan; (6) Niuean; (7) Chinese; (8) Indian; (9) Other such as DUTCH, JAPANESE, TOKELAUAN. Please state:. We coded their answers into four groups: Maori, Pacific, Asian, and Euro (except for Time 3, which used an open-ended measure).


#### Fatigue

We assessed subjective fatigue by asking participants, "During the last 30 days, how often did ... you feel exhausted?" Responses were collected on an ordinal scale (0 = None of The Time, 1 = A little of The Time, 2 = Some of The Time, 3 = Most of The Time, 4 = All of The Time).

#### Gender (waves: 1-15)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male was coded as 1, and gender diverse coded as 3 [@fraser_coding_2020]. (or 0.5 = neither female nor male)

Here, we coded all those who responded as Male as 1, and those who did not as 0.


#### Honesty-Humility-Modesty Facet (waves: 10-14)

Participants indicated the extent to which they agree with the following four statements from @campbell2004 , and @sibley2011 (1 = Strongly Disagree to 7 = Strongly Agree)

```
i.  I want people to know that I am an important person of high status, (Waves: 1, 10-14)
ii. I am an ordinary person who is no better than others.
iii. I wouldn't want people to treat me as though I were superior to them.
iv. I think that I am entitled to more respect than the average person is.
```
#### Has Siblings

"Do you have siblings?" [@stronge2019onlychild]

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stabilise this indicator, we took the natural log of the response + 1.

<!-- #### Hours Volunteering -->

<!-- We measured hours of volunteering using one item from @sibley2011: "Hours spent … voluntary/charitable work." -->

<!-- To stabilise this indicator, we took the natural log of the response + 1. -->

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stablise this indicator, we took the natural log of the response + 1.


#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stablise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stablise this indicator, we took the natural log of the response + 1.


#### Hours of Sleep

Participants were asked "During the past month, on average, how many hours of *actual sleep* did you get per night".


#### Socialising (time 10, time 11)

As part of the time usage measures, participants were asked to state how many hour the spent in activities related to socialising with the following groups:

  - Hours spent … socialising with family
  - Hours spent … socialising with friends
  - Hours spent … socialising with community groups
  - Hours spent … socialising with religious groups

#### Hours of Work

We measured hours of work using one item from @sibley2011:"Hours spent … working in paid employment."

To stablise this indicator, we took the natural log of the response + 1.


#### Income (waves: 1-3, 4-15)

Participants were asked "Please estimate your total household income (before tax) for the year XXXX". To stabilise this indicator, we first took the natural log of the response + 1, and then centred and standardised the log-transformed indicator.

<!-- #### Job Security (waves: 1-3,4-7,9-15) -->

<!-- Participants indicated their feeling of job security by answering "How secure do you feel in your current job?" on a scale from 1 (not secure) to 7 (very secure). -->


#### Living in an Urban Area (waves: 1-15)

We coded whether they are living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

We coded whether they were living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

#### Mini-IPIP 6 (waves: 1-3,4-15)

We measured participants' personalities with the Mini International Personality Item Pool 6 (Mini-IPIP6) [@sibley2011], which consists of six dimensions and each dimension is measured with four items:

1.  agreeableness,

    i.  I sympathize with others' feelings.
    ii. I am not interested in other people's problems. (r)
    iii. I feel others' emotions.
    iv. I am not really interested in others. (r)

2.  conscientiousness,

    i.  I get chores done right away.
    ii. I like order.
    iii. I make a mess of things. (r)
    iv. I often forget to put things back in their proper place. (r)

3.  extraversion,

    i.  I am the life of the party.
    ii. I don't talk a lot. (r)
    iii. I keep in the background. (r)
    iv. I talk to a lot of different people at parties.

4.  honesty-humility,

    i.  I feel entitled to more of everything. (r)
    ii. I deserve more things in life. (r)
    iii. I would like to be seen driving around in a very expensive car. (r)
    iv. I would get a lot of pleasure from owning expensive luxury goods. (r)

5.  neuroticism, and

    i.  I have frequent mood swings.
    ii. I am relaxed most of the time. (r)
    iii. I get upset easily.
    iv. I seldom feel blue. (r)

6.  openness to experience

    i.  I have a vivid imagination.
    ii. I have difficulty understanding abstract ideas. (r)
    iii. I do not have a good imagination. (r)
    iv. I am not interested in abstract ideas. (r)

Each dimension was assessed with four items and participants rated the accuracy of each item as it applies to them from 1 (Very Inaccurate) to 7 (Very Accurate). Items marked with (r) are reverse coded.


#### NZ-Born (waves: 1-2,4-15)

We asked participants, "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15).


#### NZ Deprivation Index (waves: 1-15)

We used the NZ Deprivation Index to assign each participant a score based on where they live [@atkinson2019]. This score combines data such as income, home ownership, employment, qualifications, family structure, housing, and access to transport and communication for an area into one deprivation score.

#### Opt-in

The New Zealand Attitudes and Values Study allows opt-ins to the study. Because the opt-in population may differ from those sampled randomly from the New Zealand electoral roll; although the opt-in rate is low, we include an indicator (yes/no) for this variable.


#### NZSEI Occupational Prestige and Status (waves: 8-15)

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017a]. This index uses the income, age, and education of a reference group, in this case the 2013 New Zealand census, to calculate an score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation.

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017]. This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation.

<!-- We asked participants, "If you are a parent, what is the birth date of your eldest child?". -->

<!-- #### Parent (waves: 5-15) --> -->

<!-- We asked participants, "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-15). Parents were coded as 1, while the others were coded as 0. --> -->

#### Number of Children (waves: 1-3, 4-15)

We measured the number of children using one item from @Bulbulia_2015. We asked participants, "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, or adopted?" or ""How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, and/or parented?" (waves: 12-15).

<!-- Here, we used responses to create a binary indicator for parenthood: Parent Yes = 1/ No = 0.  -->

#### Partner: Has

"What is your relationship status?" (e.g., single, married, de-facto, civil union, widowed, living together, etc.)



#### Politically Conservative 

We measured participants' political conservative orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically liberal versus conservative you see yourself as being."

(1 = Extremely Liberal to 7 = Extremely Conservative)

#### Politically Right Wing 

We measured participants' political right-wing orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically left-wing versus right-wing you see yourself as being.."

(1 = Extremely left-wing to 7 = Extremely right-wing)



<!-- #### Religion Affiliation -->

<!-- Participants were asked to indicate their religion identification ("Do you identify with a religion and/or spiritual group?") on a binary response (1 = Yes, 0 = No). We then asked, "What religion or spiritual group?" These questions are used in the New Zealand Census -->


<!-- #### Support -->

<!-- Participants' perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated the extent to which they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item. -->

#### Short-Form Health

Participants' subjective health was measured using one item ("Do you have a health condition or disability that limits you, and that has lasted for 6+ months?"; 1 = Yes, 0 = No) adapted from @verbrugge1997.


#### Support received: money (waves 10-13) (Study 4 outcomes)

The NZAVS has a 'revealed' measure of received help and support measured in hours of support in the previous week. The items are: 

Received help and support - money
  - family
  - friends
  - others in my community
  
  
#### Support received: time (waves 10-13) (Study 3 outcomes)

The NZAVS has a 'revealed' measure of received help and support measured in hours of support in the previous week. The items are: 

Received help and support - hours
  - family
  - friends
  - others in my community


#### Total Siblings

Participants were asked the following questions related to sibling counts: 
 
 - Were you the 1st born, 2nd born, or 3rd born, etc, child of your mother?
 - Do you have siblings?
 - How many older sisters do you have?
 - How many younger sisters do you have?
 - How many older brothers do you have?
 - How many younger brothers do you have?

A single score was obtained from sibling counts by summing responses to the "How many..." items. From these scores an ordered factor was created ranging from 0 to 7, where participants with more than 7 siblings were grouped into the highest category. 


<!-- #### Volunteers -->

<!-- We asked participants,"Please estimate how many hours you spent doing each of the following things last week" and responded to an item ("voluntary/charitable work") from [@sibley2011]. -->


## Appendix B. Baseline Demographic Statistics {#appendix-demographics}

::: {#tbl-B}

```{r}
#| label: tbl-table-demography-code
#| tbl-cap: "Baseline demographic statistics"
#| eval: false
#| echo: false
#| include: false

table_baseline <- read_here("table_baseline")
#table_baseline |>  kbl(format = "markdown")
```
:::

|**Exposure + Demographic Variables** |**N = 47,948** |
|:------------------------------------|:--------------|
|__Age__                              |NA             |
|Mean (SD)                            |49 (14)        |
|Range                                |18, 99         |
|IQR                                  |39, 60         |
|__Agreeableness__                    |NA             |
|Mean (SD)                            |5.35 (0.99)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.00     |
|Unknown                              |452            |
|__Alert Level Combined Lead__        |NA             |
|no_alert                             |24,789 (71%)   |
|early_covid                          |3,902 (11%)    |
|alert_level_1                        |3,010 (8.7%)   |
|alert_level_2                        |865 (2.5%)     |
|alert_level_2_5_3                    |569 (1.6%)     |
|alert_level_4                        |1,648 (4.7%)   |
|Unknown                              |13,165         |
|__Born Nz__                          |37,082 (78%)   |
|Unknown                              |585            |
|__Children Num__                     |NA             |
|Mean (SD)                            |1.74 (1.46)    |
|Range                                |0.00, 14.00    |
|IQR                                  |0.00, 3.00     |
|Unknown                              |393            |
|__Conscientiousness__                |NA             |
|Mean (SD)                            |5.11 (1.06)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.50, 6.00     |
|Unknown                              |443            |
|__Education Level Coarsen__          |NA             |
|no_qualification                     |1,239 (2.7%)   |
|cert_1_to_4                          |16,700 (36%)   |
|cert_5_to_6                          |5,969 (13%)    |
|university                           |12,595 (27%)   |
|post_grad                            |5,118 (11%)    |
|masters                              |3,924 (8.4%)   |
|doctorate                            |1,125 (2.4%)   |
|Unknown                              |1,278          |
|__Employed__                         |38,024 (79%)   |
|Unknown                              |107            |
|__Eth Cat__                          |NA             |
|euro                                 |38,158 (81%)   |
|maori                                |5,457 (12%)    |
|pacific                              |1,137 (2.4%)   |
|asian                                |2,543 (5.4%)   |
|Unknown                              |653            |
|__Extraversion__                     |NA             |
|Mean (SD)                            |3.91 (1.20)    |
|Range                                |1.00, 7.00     |
|IQR                                  |3.00, 4.75     |
|Unknown                              |443            |
|__Hlth Disability__                  |10,540 (22%)   |
|Unknown                              |926            |
|__Hlth Fatigue__                     |NA             |
|0                                    |7,326 (15%)    |
|1                                    |15,242 (32%)   |
|2                                    |14,803 (31%)   |
|3                                    |7,448 (16%)    |
|4                                    |2,572 (5.4%)   |
|Unknown                              |557            |
|__Hlth Sleep Hours__                 |NA             |
|Mean (SD)                            |6.94 (1.13)    |
|Range                                |2.50, 16.00    |
|IQR                                  |6.00, 8.00     |
|Unknown                              |2,482          |
|__Honesty Humility__                 |NA             |
|Mean (SD)                            |5.41 (1.19)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.25     |
|Unknown                              |448            |
|__Hours Children log__               |NA             |
|Mean (SD)                            |1.17 (1.61)    |
|Range                                |0.00, 5.13     |
|IQR                                  |0.00, 2.40     |
|Unknown                              |1,538          |
|__Hours Community log__              |NA             |
|Mean (SD)                            |0.34 (0.65)    |
|Range                                |0.00, 4.80     |
|IQR                                  |0.00, 0.69     |
|Unknown                              |1,538          |
|__Hours Exercise log__               |NA             |
|Mean (SD)                            |1.54 (0.85)    |
|Range                                |0.00, 4.39     |
|IQR                                  |1.10, 2.08     |
|Unknown                              |1,538          |
|__Hours Family log__                 |NA             |
|Mean (SD)                            |1.61 (1.04)    |
|Range                                |0.00, 5.13     |
|IQR                                  |1.10, 2.40     |
|Unknown                              |1,538          |
|__Hours Friends log__                |NA             |
|Mean (SD)                            |1.45 (0.87)    |
|Range                                |0.00, 5.02     |
|IQR                                  |1.10, 1.95     |
|Unknown                              |1,538          |
|__Hours Housework log__              |NA             |
|Mean (SD)                            |2.14 (0.78)    |
|Range                                |0.00, 5.13     |
|IQR                                  |1.61, 2.71     |
|Unknown                              |1,538          |
|__Hours Religious Community log__    |NA             |
|Mean (SD)                            |0.17 (0.49)    |
|Range                                |0.00, 5.04     |
|IQR                                  |0.00, 0.00     |
|Unknown                              |1,538          |
|__Hours Work log__                   |NA             |
|Mean (SD)                            |2.66 (1.58)    |
|Range                                |0.00, 4.62     |
|IQR                                  |1.39, 3.71     |
|Unknown                              |1,538          |
|__Household Inc log__                |NA             |
|Mean (SD)                            |11.39 (0.77)   |
|Range                                |0.69, 14.92    |
|IQR                                  |11.00, 11.92   |
|Unknown                              |3,726          |
|__Kessler6 Sum__                     |NA             |
|Mean (SD)                            |5 (4)          |
|Range                                |0, 24          |
|IQR                                  |2, 8           |
|Unknown                              |489            |
|__Male__                             |17,779 (37%)   |
|__Modesty__                          |NA             |
|Mean (SD)                            |5.98 (0.95)    |
|Range                                |1.00, 7.00     |
|IQR                                  |5.50, 6.75     |
|Unknown                              |34             |
|__Neuroticism__                      |NA             |
|Mean (SD)                            |3.49 (1.15)    |
|Range                                |1.00, 7.00     |
|IQR                                  |2.75, 4.25     |
|Unknown                              |454            |
|__Nz Dep2018__                       |NA             |
|Mean (SD)                            |4.77 (2.73)    |
|Range                                |1.00, 10.00    |
|IQR                                  |2.00, 7.00     |
|Unknown                              |310            |
|__Nzsei 13 l__                       |NA             |
|Mean (SD)                            |54 (17)        |
|Range                                |10, 90         |
|IQR                                  |41, 69         |
|Unknown                              |676            |
|__Openness__                         |NA             |
|Mean (SD)                            |4.96 (1.12)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.25, 5.75     |
|Unknown                              |445            |
|__Partner__                          |34,818 (75%)   |
|Unknown                              |1,539          |
|__Political Conservative__           |NA             |
|1                                    |2,515 (5.6%)   |
|2                                    |8,676 (19%)    |
|3                                    |8,820 (20%)    |
|4                                    |13,913 (31%)   |
|5                                    |6,694 (15%)    |
|6                                    |3,299 (7.4%)   |
|7                                    |741 (1.7%)     |
|Unknown                              |3,290          |
|__Religion Church Round__            |NA             |
|0                                    |38,479 (83%)   |
|1                                    |1,517 (3.3%)   |
|2                                    |1,125 (2.4%)   |
|3                                    |907 (2.0%)     |
|4                                    |2,478 (5.3%)   |
|5                                    |475 (1.0%)     |
|6                                    |326 (0.7%)     |
|7                                    |106 (0.2%)     |
|8                                    |964 (2.1%)     |
|Unknown                              |1,571          |
|__Rural Gch 2018 l__                 |NA             |
|1                                    |29,479 (62%)   |
|2                                    |9,014 (19%)    |
|3                                    |5,865 (12%)    |
|4                                    |2,694 (5.7%)   |
|5                                    |588 (1.2%)     |
|Unknown                              |308            |
|__Sample Frame Opt in__              |1,384 (2.9%)   |
|__Sample Origin__                    |NA             |
|1-2                                  |2,970 (6.2%)   |
|3-3.5                                |2,052 (4.3%)   |
|4                                    |2,700 (5.6%)   |
|5-6-7                                |4,506 (9.4%)   |
|8-9                                  |5,802 (12%)    |
|10                                   |29,918 (62%)   |
|__Short Form Health__                |NA             |
|Mean (SD)                            |5.04 (1.17)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.33, 6.00     |
|Unknown                              |9              |
|__Total Siblings__                   |NA             |
|Mean (SD)                            |2.55 (1.87)    |
|Range                                |0.00, 23.00    |
|IQR                                  |1.00, 3.00     |
|Unknown                              |1,205          |
|__Urban__                            |38,493 (81%)   |
|Unknown                              |308            |

: Baseline demography statistics {#tbl-table-demography}


@tbl-table-demography presents baseline demographic statistics for couples who met inclusion criteria.




##  Appendix C: Treatment Statistics {#appendix-exposures}


```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and baseline + 1 (treatment) wave"
#| eval: false
#| echo: false
#| include: false
table_exposures <-margot::here_read("table_exposures")

#table_exposures |>  kbl(format = "markdown")
```



|**Exposure Variables by Wave** |**2018**, N = 47,948 |**2019**, N = 47,948 |
|:------------------------------|:--------------------|:--------------------|
|__Religion Church Round__      |NA                   |NA                   |
|0                              |38,479 (83%)         |28,671 (84%)         |
|1                              |1,517 (3.3%)         |916 (2.7%)           |
|2                              |1,125 (2.4%)         |751 (2.2%)           |
|3                              |907 (2.0%)           |651 (1.9%)           |
|4                              |2,478 (5.3%)         |1,700 (5.0%)         |
|5                              |475 (1.0%)           |312 (0.9%)           |
|6                              |326 (0.7%)           |209 (0.6%)           |
|7                              |106 (0.2%)           |71 (0.2%)            |
|8                              |964 (2.1%)           |655 (1.9%)           |
|Unknown                        |1,571                |14,012               |
|__Alert Level Combined__       |NA                   |NA                   |
|no_alert                       |47,948 (100%)        |24,789 (71%)         |
|early_covid                    |0 (0%)               |3,902 (11%)          |
|alert_level_1                  |0 (0%)               |3,010 (8.7%)         |
|alert_level_2                  |0 (0%)               |865 (2.5%)           |
|alert_level_2_5_3              |0 (0%)               |569 (1.6%)           |
|alert_level_4                  |0 (0%)               |1,648 (4.7%)         |
|Unknown                        |0                    |13,165               |

: Baseline and treatment wave descriptive statistics {#tbl-table-exposures}


@tbl-table-exposures presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: religious service attendance (range 0-8). Responses coded as eight or above were coded as "8". This decision to avoid spare treatments was based on theoretical grounds, namel, that daily exposure would be similar in its effects to more than daily exposure. We note that causal contrasts were obtained for projects at either no attendance or four-or-more visits per month. Hence this simplification of the measure is unlikely to affect theoretical and practical inferences. Because the treatment wave (NZAVS time 11) occurred to New Zealand's COVID-19 pandemic, all models adjusted for the pandemic alert-level. The pandemic is a not a "confounder" because a confounder must be related to the treatment and the outcome. At the end of the study, all participants had been exposed to the pandemic.  However, to satisfy the causal consistency assumption, all treatments must be conditionally equivalent within levels of all covariates [@vanderweele2013]. Because COVID affected the ability or willingness of individuals to attend religious service, we included the lockdown condition as a covariate [@sibley2012a]. To better enable conditional independence within levels of the treatment variable, we conditioned on the lead value of COVID-alert level at baseline. To mitigate systematic biases arising from attrition, and missingness, the `lmtp` package uses inverse probability of censoring weights, which were use when estimating the causal effects of the exposure on the outcome. 





### Transition Table for The Treatment

```{r}
#| label: transition-table
#| eval: false
#| echo: false

transition_table <-margot::here_read("transition_table")
#transition_table
```

@tbl-transition shows a transition matrix that captures stability and movement between disinhibition responses from the baseline (NZAVS time 10) wave and exposure wave (NZAVS time 11) among all participants who responded to the religion question at baseline and in the following year (the censored sample). Entries on the diagonal (in bold) indicate the number of individuals who stayed in their initial state. In contrast, the off-diagonal shows the transitions from the initial state (bold) to another state the following wave (off-diagonal). A cell located at the intersection of row $i$ and column $j$, where $i \neq j$, shows the count of individuals moving from state $i$ to state $j$.


|     From      | Not Religious | Religious |
|:-------------:|:-------------:|:---------:|
| Not Religious |   **20959**   |   1183    |
|   Religious   |     1977      | **10624** |

: Transition matrix for change in treatment from baseline to the treatment wave. The diagonal shows stability in response. The off-diagonal shows the number of transitions from the initial state (bold) to another state in the following wave (off-diagonal). {#tbl-transition}





##  Appendix D: Baseline and End of Study Outcome Statistics {#appendix-outcomes}


```{r}
#| label: tbl-table-demographic_vars-code
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: false
#| echo: false
#| include: false
table_outcomes <-margot::here_read("table_outcomes")
table_outcomes |>  kbl(format = "markdown")

```

|**Outcome Variables by Wave**     |**2018**, N = 47,948 |**2020**, N = 47,948 |
|:---------------------------------|:--------------------|:--------------------|
|__Annual Charity__                |120 (25, 500)        |200 (20, 520)        |
|Unknown                           |3,181                |16,514               |
|__Community Gives Money Binary__  |205 (0.4%)           |144 (0.5%)           |
|Unknown                           |2,074                |16,836               |
|__Community Gives Time Binary__   |2,389 (5.2%)         |1,934 (6.2%)         |
|Unknown                           |2,074                |16,836               |
|__Family Gives Money Binary__     |2,802 (6.1%)         |1,477 (4.7%)         |
|Unknown                           |2,074                |16,836               |
|__Family Gives Time Binary__      |13,590 (30%)         |8,867 (29%)          |
|Unknown                           |2,074                |16,836               |
|__Friends Give Money Binary__     |589 (1.3%)           |321 (1.0%)           |
|Unknown                           |2,074                |16,836               |
|__Friends Give Time Binary__      |8,201 (18%)          |5,664 (18%)          |
|Unknown                           |2,074                |16,836               |
|__Sense Neighbourhood Community__ |NA                   |NA                   |
|1                                 |3,025 (6.3%)         |1,383 (4.3%)         |
|2                                 |5,953 (12%)          |3,252 (10%)          |
|3                                 |7,033 (15%)          |4,283 (13%)          |
|4                                 |9,911 (21%)          |6,961 (22%)          |
|5                                 |9,924 (21%)          |7,625 (24%)          |
|6                                 |8,055 (17%)          |5,921 (19%)          |
|7                                 |3,781 (7.9%)         |2,503 (7.8%)         |
|Unknown                           |266                  |16,020               |
|__Social Belonging__              |5.33 (4.33, 6.00)    |5.00 (4.33, 6.00)    |
|Unknown                           |450                  |16,114               |
|__Social Support__                |6.33 (5.33, 7.00)    |6.33 (5.33, 7.00)    |
|Unknown                           |38                   |15,919               |
|__Volunteering Hours__            |0.00 (0.00, 1.00)    |0.00 (0.00, 1.00)    |
|Unknown                           |1,538                |16,738               |
|__Volunteers Binary__             |12,905 (28%)         |8,004 (26%)          |
|Unknown                           |1,538                |16,738               |

: Outcomes measured at baseline and end-of-study  {#tbl-table-outcomes}

@tbl-table-outcomes presents baseline and end-of-study descriptive statistics for the outcome variables.  









## References {.appendix}
