---
title: "Declining Trust in Science: Longitudinal Evidence From A National Panel Study in New Zealand Years 2019-2024"
abstract: |
  Public perceptions of science shape a range of behaviours, from following health advice to supporting climate action. Using nationally representative panel data from New Zealand (2019–2024, N = 42,681; New Zealand Attitudes and Values Study), we examine changes in how highly people value science and trust scientists. **Study 1** estimates longitudinal trajectories of these attitudes, using multiple imputation to reduce bias from systematic attrition. **Study 2** analyses trust as a categorical variable, distinguishing low (1–3), medium (4–5), and high (6–7) trust. Results show that both the value placed on science and trust in scientists rose following New Zealand’s response to COVID-19, then later declined. Categorical results reveal that the proportion reporting low value placed on science remained steady, while high value decreased. In contrast, low trust in scientists increased and high trust declined. Comparisons of multiply imputed with observed data indicate that even minimal bias can shift population estimates by about 2–3% for low valuations and trust, and by 6–8% for high valuations and trust. These findings suggest caution when interpreting population-level estimates of trust in science.
  
   **KEYWORDS**: *Institutional Trust*; *Longitudinal*; *Measurement*; *Missing Data*; *Political*; *Values*.
author: 
  - name: John Kerr
    affiliation:  University of Otago, New Zealand
    orcid: 0000-0000-0000-0000
    email: john.kerr@otago.ac.nz
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-4064-8800
    email: c.sibley@auckland.ac.nz
  - name: Carla Houkamau 
    affiliation: University of Auckland, New Zealand
    orcid: 0000-0002-3449-3726
    email: c.houkamau@auckland.ac.nz
  - name: Danny Osborne
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-8513-4125
    email: d.osborne@auckland.ac.nz
  - name: Marc Wilson
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: marc.wilson@vuw.ac.nz
  - name: Kumar Yogeswaaran
    affiliation: University of Canterbury New Zealand
    orcid: 0000-0002-1978-5077
    email: 
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
editor_options: 
    chunk_output_type: console
format:
  docx:
    default: false
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % adjust this value as needed
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true


# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
library(margot)
packageVersion(pkg = "margot")
library(boilerplate)

# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# combo_marginal_graph_ord
# combo_pols_graph_ord
# combo_eth_graph_ord
# combo_male_graph_ord


# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("tinytable")
# set seed
options(tinytable_tt_digits = 3)

# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)


# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-john-kerr-trust-science-growth-simple")

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)

measures_path_rstudio <- here::here("manuscripts", "24-john-kerr-growth-trust-science")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))


baseline_vars <- margot::here_read("baseline_vars")
exposure_vars <- margot::here_read("exposure_vars")
outcome_vars <- margot::here_read("outcome_vars")
n_participants <-here_read("n_participants")
# read waves
all_waves <- c(2019:2023)

# get table/graphs
summary_vars <- margot::here_read_qs("summary_vars", push_mods)

# read timeline
timeline_histogram <- here_read_qs("timeline_histogram", push_mods)

# hist outcomes
histogram_outcomes <- here_read_qs("histogram_trust_science_trust_scientists_by_wave", push_mods)

# get range of pol conservative
max_pol <- here_read('max_pol')
min_pol<- here_read('min_pol')

max_pol <- round(max_pol,2)
min_pol <- round(min_pol,2)
#outcome table means

# # read cat
# tbl_combined_cat <- here_read_qs("tbl_combined_cat", push_mods)
# 
# # first model
# tbl_combined_cat
# # first imputed dataset
# tbl_combined_imp <- here_read_qs("tbl_combined_imp", push_mods)
# tbl_combined_imp
# 
# # mean
# summary_table_mean <- here_read_qs("summary_table_mean", push_mods)
# summary_table_mean_imp <- here_read_qs("summary_table_mean_imp", push_mods)
# 
# summary_table_mean
# summary_table_mean_imp

tt_summary_mean_response_sample <- here_read("tt_summary_mean_response_sample")
tt_summary_mean_response_sample
tt_summary_mean_response_sample_imp <- here_read("tt_summary_mean_response_sample_imp")
tt_summary_mean_response_sample_imp


# alluvial plots
# tba
sankey_plot_science_sample<-  here_read_qs("sankey_plot_science_sample", push_mods)
sankey_plot_st_sample <-  here_read_qs("sankey_plot_st_sample", push_mods)
sankey_plot_science_imp<-  here_read_qs("sankey_plot_science_imp", push_mods)
sankey_plot_st_imp<-  here_read_qs("sankey_plot_st_imp", push_mods)
tt_summary_table_cat<- here_read("tt_summary_table_cat")
tt_summary_table_cat_imp <- here_read("tt_summary_table_cat_imp")
tt_summary_table_cat


# tt_summary_table_cat_imp |> print("markdown")
sankey_plot_science_sample/  sankey_plot_st_sample
# n_total
n_total <- here_read("n_total")

scale_outcome <- glue::glue("(1-7)")
outcome_lower_bound <- glue::glue("1")
outcome_upper_bound <- glue::glue("7")
# shift =   glue::glue('two points')
nice_baseline_wave <- glue::glue("New Zealand Attitudes and Values Study Time 11, years 2019-2024")
nice_all_waves <- glue::glue("New Zealand Attitudes and Values Study Time Time 11-14, years 2019-2024")
nice_end_of_study_wave <- glue::glue("Zealand Attitudes and Values Study Time 14, years 2022-2024")

# not used
prop_missing_baseline <- here_read("prop_missing_baseline")
# percentage_missing_baseline <- glue::glue(round(percent_missing_baseline,2))
# percentage_missing_baseline
# labels
#

# read models

# read models -------------------------------------------------------------
# nnet. 
# trust_science_marginal <- here_read("trust_science_marginal")
# trust_scientists_marginal <- here_read("trust_scientists_marginal")
# trust_science_pols <- here_read("trust_science_pols")
# trust_scientists_pols <- here_read("trust_scientists_pols")
# trust_science_eth <- here_read("trust_science_eth")
# trust_scientists_eth <- here_read("trust_scientists_eth")
# trust_science_male <- here_read("trust_science_male")
# trust_scientists_male <- here_read("trust_scientists_male")
# trust_science_marginal

# gee
# trust_science_marginal_gee <- here_read("trust_science_marginal_gee")
# trust_scientists_marginal_gee <- here_read("trust_scientists_marginal_gee")
# trust_science_pols_gee <- here_read("trust_science_pols_gee")
# trust_scientists_pols_gee <- here_read("trust_scientists_pols_gee")
# trust_science_eth_gee <- here_read("trust_science_eth_gee")
# trust_scientists_eth_gee <- here_read("trust_scientists_eth_gee")
# trust_science_male_gee <- here_read("trust_science_male_gee")
# trust_scientists_male_gee <- here_read("trust_scientists_male_gee")

predictions_gee_science_observed <- here_read("predictions_gee_science_observed")
predictions_gee_scientists_observed <- here_read("predictions_gee_scientists_observed")


predictions_gee_science_imputed <- here_read("predictions_gee_science_imputed")
predictions_gee_scientists_imputed <- here_read("predictions_gee_scientists_imputed")


# plot ordinal marginal ---------------------------------------------------

plot_trust_science_observed_continuous <- plot (
  predictions_gee_science_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  line_size = 1.5,
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Social Value of Science (1-7)", title = "Average Social Value of Science: Retained Sample") + theme_bw() 

plot_trust_scientists_observed_continuous <- plot (
  predictions_gee_scientists_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  line_size = 1.5,
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Average Trust in Scientists: Retained Sample") + theme_bw() 


plot_trust_science_imputed_continuous <- plot (
  predictions_gee_science_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  line_size = 1.5,
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Social Value of Science (1-7)", title = "Average Social Value of Science: Imputed Missing Responses") + theme_bw() 



plot_trust_scientists_imputed_continuous <- plot (
  predictions_gee_scientists_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  line_size = 1.5,
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Average Trust in Scientists: Imputed Missing Responses") + theme_bw()



# patchwork graph
combo_science_continuous<-   plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study): ", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Missing Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  

combo_scientists_continuous <- plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous   + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom') 

combo_all_continuous <- (plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous)/ (plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous) + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2023), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science and Scientists over Time: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  

# margot::margot_save_png(combo_all,base_filename = "trust_science_observed_vs_imputed", save_path = here::here(push_mods), height=20,width=20, dpi = 500)


# Categorical-------------------------------------------------------------------
# 
observed_trust_science_factor <- here_read("observed_trust_science_factor")
observed_trust_scientists_factor <- here_read("observed_trust_scientists_factor")

predictions_ord_science_imp <- here_read("predictions_ord_science_imp")
predictions_ord_scientists_imp <- here_read("predictions_ord_scientists_imp")

observed_trust_science_factor |> print_md()

# plot
observed_trust_science_factor_plot =  plot(
  observed_trust_science_factor,
  jitter = .4,
  dot_alpha  = .025,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Social Value of Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Social Value of Science: Retained Sample", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()

observed_trust_scientists_factor_plot =  plot(
  observed_trust_scientists_factor,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Retained Sample",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()


imputed_trust_science_factor_plot <- plot(
  predictions_ord_science_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
  )+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Social Value of Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Social Value of Science: Imputed Missing Responses", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()

imputed_trust_scientists_factor_plot <- plot(
  predictions_ord_scientists_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Imputed Missing Responses",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()  +   theme(
    plot.title = element_text(face = "bold", size = 14))


combo_science_factor <- observed_trust_science_factor_plot  + imputed_trust_science_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Missing Responses: Multinomial Response Model\n'Our Society Places Too Much Emphasis on Science'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

combo_scientists_factor <- observed_trust_scientists_factor_plot  + imputed_trust_scientists_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Scientists: Observed vs Imputed Missing Responses: Multinomial Response Model\n'I have a high degree of confidence in the scientific community'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

# 
# combo_science_continuous
# combo_scientists_continuous
# 
#combo_science_factor
# combo_scientists_factor


# number of people in NZ > 18 year olds 
# method: use 2023 - 2024. Compare the raw frequencies at the point predictions for the # unadjusted and adjusted modes. 
# https://www.stats.govt.nz/information-releases/national-population-estimates-at-30-june-2024-2018-base/
# https://infoshare.stats.govt.nz/SelectVariables.aspx?pxID=591c2174-6602-4a34-b09c-e6632b63bac1

#  data frame with ages and their corresponding populations
age_data <- data.frame(
  age = c(18:94),  # individual from 18-94+ (i.e. adults)
  population = c(64630, 64520, 64890, 64670, 65530, 67670, 67520, 66940, 69260, 
                71030, 73710, 75910, 78480, 82610, 84700, 85540, 84050, 81890, 
                80670, 77110, 74690, 73350, 72050, 71290, 69150, 67670, 66310, 
                64340, 62700, 62540, 62280, 63930, 65310, 67480, 68680, 67990, 
                66290, 65770, 63690, 62470, 62250, 63160, 64470, 65050, 63600, 
                61750, 59310, 57740, 55850, 53770, 51900, 50090, 47950, 46570, 
                44650, 43350, 42140, 40880, 40180, 37450, 31820, 28810, 25850, 
                23730, 24110, 21970, 18950, 16480, 14250, 12150, 10270, 9020, 
                7560, 6210, 5120, 4260, 3360)
)

# compupted the weighted sum (age * population)
# total_age_years <- sum(age_data$age * age_data$population)

# calculate total population for these age groups
total_population <- sum(age_data$population)
total_population
# calc average age
# average_age <- total_age_years / total_population

# results
# cat("Total person-years:", format(total_age_years, scientific = FALSE), "\n")
cat("Total population:", format(total_population, scientific = FALSE), "\n")

#total
total_population
# 
# # results for unadjusted
# observed_trust_science_factor
# 
# # results for adjusted (imputed)
# predictions_ord_science_imp
# 
# # observed_trust_scientists_factor
# # predictions_ord_scientists_imp
# 
# # unadjusted population estimate in low trust in science
# trust_science_naive = round(total_population * .07)
# 
# # imputation adjusted population estimate in low trust in science
# trust_science_imputed = round(total_population * .09)
# 
# # numerical difference 
# low_trust_science_difference = trust_science_imputed - trust_science_naive
# 
# # make a pretty number with custom function
# low_trust_science_difference =  margot::pretty_number(low_trust_science_difference)
# 
# # underestimating low trust
# low_trust_science_difference 
# 
# # high trust in science calculations
# high_trust_science_naive = round(total_population * .69)  # unadjusted mean prob
# high_trust_science_imputed = round(total_population * .61)  # adjusted mean prob
# high_trust_science_difference =  high_trust_science_naive - high_trust_science_imputed
# high_trust_science_difference = margot::pretty_number(high_trust_science_difference)
# 
# # over-estimating high trust
# high_trust_science_difference 
# 
# # reminder results for unadjusted
# observed_trust_scientists_factor
# 
# # results for adjusted (imputed)
# predictions_ord_scientists_imp
# 
# 
# # unadjusted population estimate in low trust in scientists
# trust_scientists_naive = round(total_population * .09)
# 
# # imputation adjusted population estimate in low trust in scientists
# trust_scientists_imputed = round(total_population * .13)
# 
# # numerical difference 
# low_trust_scientists_difference = trust_scientists_imputed - trust_scientists_naive
# 
# # make a pretty number with custom function
# low_trust_scientists_difference =  margot::pretty_number(low_trust_scientists_difference)
# 
# 
# # underestimating low trust in scientists by
# low_trust_scientists_difference
# 
# 
# # high trust in scientists calculations
# high_trust_scientists_naive = round(total_population * 0.58)  # unadjusted
# high_trust_scientists_imputed = round(total_population * 0.52)  # adjusted
# high_trust_scientists_difference =  high_trust_scientists_naive - high_trust_scientists_imputed
# 
# high_trust_scientists_difference = margot::pretty_number(high_trust_scientists_difference)
# 
# # over-estimating high trust
# high_trust_scientists_difference 



# new ------------------------------------------------------------------------

# define confidence intervals for trust in science
observed_trust_science_low <- list(mean = 0.07, lower = 0.07, upper = 0.08)
imputed_trust_science_low <- list(mean = 0.09, lower = 0.09, upper = 0.09)

observed_trust_science_high <- list(mean = 0.69, lower = 0.68, upper = 0.70)
imputed_trust_science_high <- list(mean = 0.61, lower = 0.61, upper = 0.62)

# define confidence intervals for trust in scientists
observed_trust_scientists_low <- list(mean = 0.12, lower = 0.11, upper = 0.12)
imputed_trust_scientists_low <- list(mean = 0.13, lower = 0.13, upper = 0.13)

observed_trust_scientists_high <- list(mean = 0.58, lower = 0.58, upper = 0.59)
imputed_trust_scientists_high <- list(mean = 0.52, lower = 0.51, upper = 0.52)


# total adjult population 
# total_population  # about 4.13 million adults
# 
# # minimal and worst bias low trust in science
minimal_bias_low_trust_science <- (imputed_trust_science_low$lower - observed_trust_science_low$upper) * total_population
worst_case_bias_low_trust_science <- (imputed_trust_science_low$upper - observed_trust_science_low$lower) * total_population

# minimal and worst bias high trust in science
minimal_bias_high_trust_science <- (observed_trust_science_high$lower - imputed_trust_science_high$upper) * total_population
worst_case_bias_high_trust_science <- (observed_trust_science_high$upper - imputed_trust_science_high$lower) * total_population

# minimal and worst bias low trust in scientists
minimal_bias_low_trust_scientists <- (imputed_trust_scientists_low$lower - observed_trust_scientists_low$upper) * total_population
worst_case_bias_low_trust_scientists <- (imputed_trust_scientists_low$upper - observed_trust_scientists_low$lower) * total_population

# minimal and worst bias high trust in scientists
minimal_bias_high_trust_scientists <- (observed_trust_scientists_high$lower - imputed_trust_scientists_high$upper) * total_population
worst_case_bias_high_trust_scientists <- (observed_trust_scientists_high$upper - imputed_trust_scientists_high$lower) * total_population


# format bias estimates
minimal_bias_low_trust_science_fmt <- pretty_number(round(minimal_bias_low_trust_science))
worst_case_bias_low_trust_science_fmt <- pretty_number(round(worst_case_bias_low_trust_science))

minimal_bias_high_trust_science_fmt <- pretty_number(round(minimal_bias_high_trust_science))
worst_case_bias_high_trust_science_fmt <- pretty_number(round(worst_case_bias_high_trust_science))

minimal_bias_low_trust_scientists_fmt <- pretty_number(round(minimal_bias_low_trust_scientists))
worst_case_bias_low_trust_scientists_fmt <- pretty_number(round(worst_case_bias_low_trust_scientists))

minimal_bias_high_trust_scientists_fmt <- pretty_number(round(minimal_bias_high_trust_scientists))
worst_case_bias_high_trust_scientists_fmt <- pretty_number(round(worst_case_bias_high_trust_scientists))


```



## Introduction

Whether people are becoming more sceptical of science is a matter of substantial practical interest and concern [@sturgis2021trust; @kennedy2022americans; @pagliaro2021trust]. Public trust in science shapes behaviours from the uptake of vaccination and other public health measures [@rueger2021perception] to support for climate action [@bogert2024effect]. Yet, accurately documenting population trends is challenging because individuals with lower trust in science may be less likely to participate in scientific studies, biasing results [@kreps2020model; @reif2021representative].

This study draws on five waves of national panel data collected between 2019 and 2024 from the New Zealand Attitudes and Values Study (NZAVS). We focus on change over time in the 2019 baseline cohort of `r n_participants` participants, estimating annual response trajectories for two domains of institutional trust: (1) attitudes about the social value of science, and (2) trust in scientists. To address attrition and systematic nonresponse, we apply multiple imputation models that are appropriate for panel data. By comparing the pooled results of these models to unadjusted models in a restricted sample, we clarify how observed trends among the retained sample may be biased. Although this bias estimate only represents a minimum bound—given that those who never participated might be even less trusting—at least some individuals with low trust were apparently willing to stick around. Because our objective is to include the perspectives of the mistrusting, this finding is reassuring. 

In **Study 1a**, we analyse annual trends in the average value placed on science, comparing observed responses with results that adjust for missing data. **Study 1b** applies the same framework to trust in scientists, again treating responses as continuous variables. In **Study 2a**, we classify respondents into low (1–3), medium (4–5), or high (6–7) categories for the value placed on science. **Study 2b** extends this categorical approach to trust in scientists, focussing on the probabilities of falling into each category. By considering these categorical models, we identify patterns that remain hidden when examining means alone, potentially uncovering meaningful social divisions in attitudes toward science and scientists.

This work is descriptive and exploratory. Our analysis does not identify causal drivers of these trends that we report, but rather provides descriptive insights into how trust distributions evolve over time among those already willing to participate in a scientific study, in a setting where there is likely bias from selective attrition among science's discontents. 


## Method


```{r}
#| eval: false
#| include: false
outcome_measures <- boilerplate::boilerplate_report_measures(
  baseline_vars = baseline_vars,
  outcome_vars = outcome_vars,
  measure_data = measure_data
)
cat(outcome_measures)

```

### Target Population

The target population for this study comprises the cohort of New Zealand residents in wave 2019 whose mistrust of science was not so great at that time as to prevent participation in the New Zealand Attitudes and Values Study (NZAVS). Our objective is to infer population dynamics for this cohort by using responses from individuals who took part in the NZAVS during the 2019 baseline wave, applying weights derived from the 2018 New Zealand Census to account for age, gender, and ethnicity distributions [@sibley2021].


### Sample

The NZAVS is a national probability study designed to represent the New Zealand population accurately. Although it employs prize draws to incentivise participation, response rates and demographic balances vary. For example, the NZAVS tends to under-sample males and individuals of Asian descent and to over-sample females and Māori (the Indigenous people of New Zealand). To enhance representativeness of the target population, we apply Census-based survey weights that adjust for age, gender, and ethnicity (New Zealand European, Asian, Māori, Pacific), thus improving the alignment of our sample with the national demographic composition [@sibley2021].

### Eligibility Criteria

To be eligible for inclusion in this study, participants needed to have responded to the `r nice_baseline_wave`. Individuals meeting this criterion were then tracked over the baseline and the subsequent four waves, permitting missing responses in any wave through `r nice_end_of_study_wave`. A total of `r n_participants` individuals satisfied these conditions and were included in the analyses.

[Appendix A](#appendix-a) (@tbl-demography) presents demographic characteristics of this cohort and details patterns of missing data. [Appendix B](#appendix-b) (@fig-timeline) illustrates the daily data collection periods for NZAVS waves 11–15 (2019–2024), providing additional context for the timing of responses and potential nonresponse.

## Measures

```{r}
#| eval: false
#| include: false
outcome_measures <- boilerplate::boilerplate_report_measures(
  outcome_vars = outcome_vars,
  measure_data = measure_data
)
```

Our analysis distinguishes between attitudes about the social value of science and and trusting attitudes towards scientists. 

It is plausible that at least some who disagree with society's emphasis on science are nevertheless trusting of scientific institutions. 

On the other hand, it is the solomn duty of every scientist to doubt scientists. That is, mistrust of scientists is an institutional cornerstone of science. (Refer to @schoor2021science and @gundersen2022science on the importance of distinguishing between types of institutional mistrust of science).  

Item wording is as follows.

#### Social Value of Science

*Our society places too much emphasis on science (reversed).*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@hartman2017].

#### Trust in Scientists

*I have a high degree of confidence in the scientific community.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@nisbet2015].


### Missing Data

We face two primary challenges when estimating population-level trends for institutional trust in science.

First, although the New Zealand Attitudes and Values Study uses a national probability sampling strategy, only about 10% of those invited to the study participate. It is credible that the proportion of those who do not value science or who mistrust scientists are under-represented in the sample because the New Zealand Attitudes and Values Study is run by scientists. Our sample population inferences are unlikely to directly transport to the population of New Zealand [@reif2021representative].

Second, although the NZAVS retains between 70–80% of its sample each year, panel attrition is inevitable. It is credible that a proportion of those who devalue science or mistrust scientists is increasingly diminished over time. Obtaining valid inferences for baseline cohort trajectories of mistrust is challenging because people do not report their reasons for withdrawing, and some loss will be undirected (for example mortality).

We cannot overcome the first challenge. We lack direct information about mistrust in the population of non-participants. 

However, for those who participated in 2019 but dropped out later, we can adjust estimates under the assumption that the probability of attrition or missing response is predictable given observed covariates, including previous attitudes to science, together with an appropriately specified statistical model for imputing missing responses. Although these estimates carry modelling assumptions, the development of robust multiple imputation strategies suitable for panel data means that undertainty about missing data can be incorporated into models of longitudinal trends [@blackwell_2017_unified; @bulbulia2023a].

Here, we used the `Amelia` package in R [@amelia_2011] to create ten multiply imputed datasets for the NZAVS 2019 cohort. These cover NZAVS time 11–15 (waves 2019–2023, years 2019-2024). We treated attitudes to science and scientists as ordinal responses bounded by 1 and 7 on the response scale. The Amelia algorithm is designed for within-unit imputation in repeated-measures time-series data. All covariates in @tbl-demography were included. The time variable (year) was modelled as a cubic spline to capture non-linear trends over the five waves. The code for the analysis is deposited here: [OSF link](https://osf.io/wgtz4/).


### Statistical Estimator

In Studies 1a and 1b, we estimate mean responses for (1a) values placed on science and (1b) trust in scientists over time using generalised estimating equations (GEE). We implement these models with the `geepack` package [@geepack_2006]. By specifying each participant's id as the clustering variable, we adjust for repeated observations and obtained robust standard errors [@mcneish_2017unnecessary]. We weight the data using post-stratification sample weights derived from the 2018 New Zealand Census for age, gender, and ethnicity. We fit separate models for each of the ten imputed datasets and pool results using Rubin's rule, thereby incorporating uncertainty from the missing data process into our population estimates. We generate predicted means and their associated confidence intervals using `ggeffects` [@ggeffects_2018].

In Studies 2a and 2b, we examine predicted probabilities for attitudes to science and scientists in three discrete categories: low (1–3), medium (4–5), and high (6–7). For these analyses, we employ neural network models using the `nnet` package [@nnet_2002]. These models estimate category-specific probabilities at each time point. We applied sample weights constructed from the 2018 Census data to better ensure population representativeness, and again used Rubin's rule to combine inferences across imputed datasets. We incorporate robust standard errors appropriate for clustered data and graph predictions and confidence intervals using `ggeffects` [@ggeffects_2018]. The tables and figures reported below were created using `tinytable` [@tinytable_2024], `table1` [@table1_2023], `parameters` [@parameters_2020], `tidyverse` [@tidyverse_2019], `ggplot2` [@ggplot2_2016], and `margot` [@margot2024].



#### Attitudes to Science and Scientists: Sample Means in Retained Sample

@tbl-sample-continuous present sample average responses and categorical distributions of value placed on science and trust in scientists across consecutive waves from 2019 to 2023, along with sample sizes and missing data counts for the wave 2019 cohort that dropped out or did not respond. 


Although average trust levels generally increased following New Zealand’s initial response to COVID-19 [@sibley2020a], we also note increasing missingness over time. This suggests that straightforward interpretations of sample averages may be misleading, because, individuals who become more sceptical of science may be less likely to remain in the study run by scientists.

@tbl-sample-continuous shows that mean value placed on science rose from 5.56 in 2019 to 5.86 in wave 2023, while mean trust in scientists increased from 5.30 to 5.55 between 2019 and 2020, stabilised through wave 2022, and then declined to 5.45 in wave 2023, still higher than the baseline average in wave 2019. These observed increases, particularly for value placed on science following the New Zealand COVID-19 response, appear positive. However, missing responses grew markedly, from 563 in wave 2019 to 20,600 in wave 2023 for values of science, and from 1,257 in 2019 to 21,177 in wave 2023 for trust in scientists. Yet the trend will be biased if attrition reflects systematic non-response related to mistrust.


::: {#tbl-sample-continuous}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_mean_response_sample |> print("markdown")
```
Retained sample average response by wave.
:::

{{< pagebreak >}}


### Sample Categorical Distributions Over Time


::: {#tbl-sample-cat}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_table_cat |> print("markdown")
```
Retained sample responses by wave by response category classified as low (1-3), medium (4-5), or high (6-7).
:::

@tbl-sample-cat categorises responses into low (1–3), medium (4–5), and high (6–7) ranges. For value placed on science, the proportion reporting low trust declined from 8.2% in 2019 to 6.7% in wave 2023, while the proportion reporting high trust rose from 60.5% to 70.1% over the same period.  [Appendix C](#appendix-c) @fig-alluv-science-sample graphically displays these transitions.

Similar patterns emerged for trust in scientists, with initial declines in the low category and increases in the high category, followed by some reversal in later waves. [Appendix C](#appendix-c) @fig-alluv-scientists-sample graphically displays these transitions.

These shifts might convey a narrative of strengthening value of science during early pandemic periods that continued to grow, with no evidence of increasing mistrust of scientists. Yet the steady increase in missing responses over time underscores the importance of careful interpretation. The rising average trust estimates may in part be an artefact of differential attrition.

{{< pagebreak >}}

## Results 


### Multiply Imputed Full Cohort Means in Full 2019 Cohort 

@tbl-sample-means-imp shows the estimated mean responses for value placed on science and trust in scientists, pooled over multiple imputations to adjust for systematic missingness. Compared with the observed sample means, the imputed estimates suggest moderate trends. For valuing science, the average response is higher in wave 2020 than in wave 2019, but then steadily declines from 5.7 in 2021 to 5.66 by wave 2023. For trust in scientists, the average increases from 5.29 to 5.47 between wave 2019 and wave 2020 but then declines to 5.28 by wave 2023. These imputed estimates reveal a less pronounced increase in trust than the unadjusted data might suggest, with evidence of  erosion in trust in both domains over the later waves.


::: {#tbl-sample-means-imp}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_mean_response_sample_imp |> print("markdown")
```
Imputed average responses for the full wave 2019 cohort
:::




::: {#tbl-sample-cat-imp}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_table_cat_imp|> print("markdown")
```
Cohort proportions classified as low, medium, or high (imputed data)
:::


@tbl-sample-cat-imp classifies imputed responses into low, medium, and high categories. For valuing science, the proportion of respondents reporting low trust remains relatively stable, hovering around 8%-9% over time. In contrast, the initially large high-trust category (around 60%) rises to 65.8% in 2020, then recedes to 62.8% by 2023, suggesting that some of the early gains in trust may have been short-lived. For trust in scientists a trajector of increasing mistrust. Initially, 11.9% reported low trust in wave 2019, decreasing to 10.1% in wave 2020, but this figure jumps to 12.8% low mistrust by wave 2023. The medium and high categories show corresponding shifts, reflecting a more complex pattern than the non-imputed data would suggest, with diminishing trust at the high end of the scale.

[Appendix D](#appendix-d) @fig-alluv-science-imp and @fig-alluv-scientists-imp visually present these transitions. Although early gains in high trust appeared promising -- for the perspective of science and scientists anyway -- inferred shifts over time are not uniformly in one direction. Instead, the imputed data indicate gradual redistributions of respondents across low, medium, and high categories, with low mistrust increasing and high trust declining. Such patterns underscore the importance of correcting for missingness before drawing conclusions about population-level changes in how people value science and trust scientists.

{{< pagebreak >}}


### Study 1a: Longitudinal Trajectories for Value Placed on Science: Continuous Analysis


```{r}
#| label: fig-science-continuous
#| fig-cap: "(A) Predicted value of science without adjusting for missingness (B) Predicted value of science adjusting for missingness, with points averaged over the ten imputed datasets (point alpha = 0.025). Note there is considerable variation of response above and below the mean, and there is slightly higher point density in panel (B) because we have adjusted for missingness."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_science_continuous
```


::: {#tbl-marginal-gee-science-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# predictions_gee_science_observed |> print_md()
predictions_gee_science_observed_x <- format(predictions_gee_science_observed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)
predictions_gee_science_observed_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
   latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted population average value placed on science over time, without adjusting for missing data.
:::


::: {#tbl-marginal-gee-science-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false

# predictions_gee_science_imputed |> print_md()

predictions_gee_science_imputed_x <- format(predictions_gee_science_imputed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)
predictions_gee_science_imputed_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted population average value placed on science over time, adjusting for missing data.
:::


@fig-science-continuous presents the predicted population average value place on science over time. Panel **A** presents estimates without adjusting for missing responses. Panel **B** presents estimates that incorporate multiple imputation. These predictions, derived from generalised estimating equations, reflect the estimated population trajectories, either without or with adjustments for missing responses. As shown in @tbl-marginal-gee-science-observed, when missingness is not taken into account, valuing science appears to increase steadily from the baseline period to the final wave. However, after modelling missingness as reported in @tbl-marginal-gee-science-imputed, the predicted trajectory implies a more modest upward trend that does not continue to climb. This result shows that ignoring missing responses can inflate the perceived value of science in a national longitudinal cohort.


{{< pagebreak >}}

### Study 1b: Longitudinal Trajectories for Trust in Scientists: Continuous Analysis


```{r}
#| label: fig-scientists-continuous
#| fig-cap: "(A) Predicted trust in scientists without adjusting for missingness (B) Predicted trust in scientists modelling missingness, with points averaged over the ten imputed datasets. Points are dimmed to alpha = .025. Note there is again considerable variation of response above and below the mean, and again there is slightly higher point density in panel (B) because we have adjusted for missingness."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_scientists_continuous
```

::: {#tbl-marginal-gee-scientists-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

#predictions_gee_scientists_observed |> print_md()

predictions_gee_scientists_observed_x <- format(predictions_gee_scientists_observed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)
predictions_gee_scientists_observed_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )

```
Predicted population average trust in scientists over time, without adjusting for missing data.
:::


::: {#tbl-marginal-gee-scietists-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false

# predictions_gee_scientists_imputed |> print_md()

predictions_gee_scientists_imputed_x <- format(predictions_gee_scientists_imputed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)

predictions_gee_scientists_imputed_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )

```
Predicted population average value placed on science over time adjusting for missing data.
:::


A similar pattern emerges for trust in scientists, presented in @fig-scientists-continuous. As shown in @tbl-marginal-gee-scientists-observed, the predicted mean trust increases when missingness is not accounted for. Yet when we adjust for missing data using multiple imputation, as reported in @tbl-marginal-gee-scietists-imputed, the results suggest that the initial gains in trust may be smaller and less sustained than the unadjusted models suggest.


{{< pagebreak >}}


### Study 2a: Longitudinal Trajectories for Value Placed on Science: Categorical Analysis


```{r}
#| label: fig-science-categorical
#| fig-cap: "(A) Predicted probability of classification for value placed on science without adjusting for missingness (B) Predicted probability of classification for value placed on science adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_science_factor
```

::: {#tbl-categorical-science-observed}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false

# observed_trust_science_factor |> print_md()

observed_trust_science_factor_x <- format(observed_trust_science_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)

observed_trust_science_factor_x|>
     as.data.frame() %>%
    dplyr::rename(Years = years) %>%
    dplyr::rename(Response = groups) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for value placed on science without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::



::: {#tbl-categorical-science-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false

# predictions_ord_science_imp |> print_md()

predictions_ord_science_imp_x <- format(predictions_ord_science_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)

predictions_ord_science_imp_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
    dplyr::rename(Response = groups) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
# 
# # cbind(observed_trust_science_factor_x, predictions_ord_science_imp_x) |>  as.data.frame() %>% 
# #   kbl(
# #     format = "latex",
# #     booktabs = TRUE,
# #     digits = 2,
# #     row.names = FALSE                      # remove row names to eliminate any unintended columns
# #   ) |>
# #   kable_styling(
# #     font_size = 14,
# #     latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
# #   )
```
Predicted probability of classification for value placed on science, here adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


Note that in Studies 1a and 1b, the response distributions are scattered across the response scale 1-7 (refer to @fig-science-continuous). We are encouraged that sampling in the New Zealand Attitudes and Values Study reaches science's discontents, even in the retained sample. Yet is not suprising that discontent leads to differential attrition. How bad is the bias in attrition? 

@fig-science-categorical displays predicted probabilities for classification as low (1–3), medium (4–5), or high (6–7) in the value placed in science over time. @tbl-categorical-science-observed presents these probabilities without adjusting for missing responses. Under these conditions, the proportion of respondents in the high category appears to increase from the baseline and remain relatively elevated. However, after adjusting for missingness (see @tbl-categorical-science-imputed), the predicted probabilities show a less stable pattern, with gains in the high category tempered and the medium category becoming more prominent. These differences underscore that ignoring missing responses can create an artificial appearance of increasing value placed on science.


{{< pagebreak >}}


### Study 2b: Longitudinal Trajectories for Trust in Scientists: Categorical Analysis


```{r}
#| label: fig-scientists-categorical
#| fig-cap: "(A) Predicted probability of classification for trust in scientists without adjusting for missingness (B) Predicted probability of classification for trust in scientists adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_scientists_factor
```

::: {#tbl-categorical-scientists-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false


# observed_trust_scientists_factor |> print_md()
observed_trust_scientists_factor_x <- format(observed_trust_scientists_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)

observed_trust_scientists_factor_x|>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
    dplyr::rename(Response = groups) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


::: {#tbl-categorical-scientists-imputed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# predictions_ord_scientists_imp |> print_md()


predictions_ord_scientists_imp_x <- format(predictions_ord_scientists_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2)

predictions_ord_scientists_imp_x |>
 as.data.frame() %>%
    dplyr::rename(Years = years) %>%
    dplyr::rename(Response = groups) %>%
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists, now adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


A similar pattern emerges for trust in scientists (see @fig-scientists-categorical). Without adjusting for missingness, the results (shown in @tbl-categorical-scientists-observed) suggest that the proportion classified as high in trust remains consistently large. Once we incorporate multiple imputation to address missing responses, as presented in @tbl-categorical-scientists-imputed, the predicted probabilities shift. The high category becomes less dominant over time, and more respondents fall into the medium category than would have been inferred without proper adjustment. In other words, accounting for nonresponse reveals a more nuanced pattern of trust that does not simply trend upward, and this correction prevents inflated estimates of stable or increasing trust in scientists.

## Discussion 

These findings illustrate that failing to adjust for missing responses can systematically distort estimates of public attitudes toward science and scientists. How bad is this distortion?

Compared to models that ignore missingness, our imputation-adjusted analyses indicate substantial population-level discrepancies. For example:

### Trust in Science

Without adjusting for missing attitudes, we **underestimate** low trust by approximately `r minimal_bias_low_trust_science_fmt` (minimal bias) to `r worst_case_bias_low_trust_science_fmt` (worst case) adults.

Simultaneously, we **overestimate** high trust by about `r minimal_bias_high_trust_science_fmt` (minimal bias) to `r worst_case_bias_high_trust_science_fmt` (worst case) adults.

### Trust in Scientists

For trust in scientists, the discrepancy in the low-trust population ranges from `r minimal_bias_low_trust_scientists_fmt` (minimal bias) to `r worst_case_bias_low_trust_scientists_fmt` (worst case) adults.

The high-trust category may be inflated by about `r minimal_bias_high_trust_scientists_fmt` (minimal bias) to `r worst_case_bias_high_trust_scientists_fmt` (worst case) adults.


These differences are not trivial, given a total adult population of about 4.1 million. Moreover, these bounds are likely conservative because we did not measure trust among individuals who declined participation at the outset. 

This is not a trivial matter. Trust in science can influence actions with far-reaching consequences, such as efforts to mitigate climate change [@bogert2024effect], vaccine uptake [@pagliaro2021trust], and health-related decisions [@rueger2021perception]. By accounting for missing data, we show that failing to adjust can substantially skew perceived levels of public trust in science and scientists. In practice, this means that decision-makers relying on unadjusted surveys risk misinterpreting the extent of public trust—and thus may not fully appreciate the scale of challenges or opportunities when shaping policy and communication strategies.


{{< pagebreak >}}

### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study. Members of the NZAVS management team and research group hold full copies of the NZAVS data. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author or any member of the NZAVS advisory board for replication or checking of any published study using NZAVS data. The code for the analysis can be found at: [OSF link](https://osf.io/wgtz4/). 


### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Plank Institute for the Science of Human History. [**To be filled out for all authors**]. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement

**TBA:** 

JK led the study. JK, CGS, DO, MW, and KY developed the measures for the New Zealand Attitudes and Values Study. JB developed the analytic approach and did the analysis. CGS led data collection. JB and CGS obtained the funding. **All authors had input into the manuscript.**


{{< pagebreak >}}




## Appendix A: Sample Descriptive Statistics {#appendix-a}

@tbl-demography presents descriptive statistics for the sample.


::: {#tbl-demography}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

library(gtsummary)
library(kableExtra)

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


summary_vars |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Demographic statistics for New Zealand Attitudes and Values Cohort waves 2019-2023 (years 2019-2024).
:::

{{< pagebreak >}}

## Appendix B: New Zealand Attitudes and Values Study Data Collection waves 2019-2023 (years 2019-2024) {#appendix-b}


```{r}
#| label: fig-timeline
#| fig-cap: "Historgram of New Zealand Attitudes and Values Study Daily Data Collection Wave 2019 (time 11) - Wave 2023 (time 15)."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12
#| 


timeline_histogram
```

{{< pagebreak >}}


## Appendix C: Alluvial Graph of Showing Categorical Responses to Value of Science and Trust in Scientists: No Adjustment For Missing Data. {#appendix-c}

```{r}
#| label: fig-alluv-scientists-sample
#| fig-cap: "Alluvial plot suggests increasing flow of value of science for the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_science_sample
```


```{r}
#| label: fig-alluv-science-sample
#| fig-cap: "Alluvial plot suggests increasing flow of trust in scientists for the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_st_sample

```



{{< pagebreak >}}


## Appendix D: Alluvial Graphs of Showing Categorical Responses to Social Value of Science and Trust in Scientists: Adjusting For Missing Data. {#appendix-d}

@fig-alluv-science-imp and @fig-alluv-scientists-imp depict the dynamics of categorical responses for attitudes to science and scientists. Following 2020, there is a decline in the value placed on scientific institutions and a growth of mistrust in scientists. Our analysis presents the a minimal bound for such mistrust because we cannot account for the mistrust of the population who declined to praticipate in New Zealand Attitudes and Values Study wave 2019 (time 11).

```{r}
#| label: fig-alluv-science-imp
#| fig-cap: "Alluvial plot suggests declines in value of science for the imputed cohort."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_science_imp
```



```{r}
#| label: fig-alluv-scientists-imp
#| fig-cap: "Alluvial plot suggests declines in trust in scientists for the imputed cohort"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_st_imp
```


{{< pagebreak >}}

## Appendix E: Histogram of Responses By Wave Shows Declining Sample. {#appendix-E}

@fig-hist-outcomes presents histograms of attitudes to science and scientists over time in the observed sample. As evident in the graphs, there is increasing average value placed on science over time in the retained sample, with a large boost evident in wave 2020 (years 2020-2021). This boost coincides with the initially popular New Zealand COVID-19 pandemic response [@sibley2020a]. Similarly, in the retained sample, trust in scientists grew, coinciding with New Zealand's COVID-19 pandemic response, and then stabalised, before falling in the most recent wave (years 2023-2024).


{{< pagebreak >}}

```{r}
#| label: fig-hist-outcomes
#| fig-cap: "Historgram of sample attitudes to science over time in the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 20
#| fig-height: 9
#| 
histogram_outcomes
```


## References {.appendix}











