---
title: "Declining Trust in Science: Longitudinal Evidence From A National Panel Study in New Zealand Years 2019-2024"
abstract: |
  The public’s perception of science influences a wide range of behaviours, from adherence to public health guidance to support for climate action. This study uses nationally representative panel data from New Zealand to examine changes in the value placed on science and trust in scientists between 2019 and 2024 (*N* = 42,681, New Zealand Attitudes and Values Study). **Study 1** estimates longitudinal trajectories for attitudes to science and scientists, applying multiple imputation to address biases arising from systematic attrition. **Study 2** models trust responses categorically, distinguishing low (1–3), medium (4–5), and high (6–7) trust. Results indicate that the value placed on science and trust in scientists rose after New Zealand’s response to COVID-19, then subsequently declined. Categorical analysis suggests that the proportion reporting low value placed on science remained stable, while the proportion reporting high value decreased. For trust in scientists, low trust increased, and high trust declined. Comparing multiply imputed outcomes with observed data reveals that even a minimum bound of bias leads to misestimations of approximately 2–3% of the population for low valuations and low trust, and 6–8% for high valuations and high trust. These findings underscore the importance of exercising caution when interpreting population-based estimates of trust in science. 
   **KEYWORDS**: *Institutional Trust*; *Longitudinal*; *Measurement*; *Missing Data*; *Political*; *Values*.
author: 
  - name: John Kerr
    affiliation:  University of Otago, New Zealand
    orcid: 0000-0000-0000-0000
    email: john.kerr@otago.ac.nz
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-4064-8800
    email: c.sibley@auckland.ac.nz
  - name: Carla Houkamau 
    affiliation: University of Auckland, New Zealand
    orcid: 0000-0002-3449-3726
    email: c.houkamau@auckland.ac.nz
  - name: Danny Osborne
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-8513-4125
    email: d.osborne@auckland.ac.nz
  - name: Marc Wilson
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: marc.wilson@vuw.ac.nz
  - name: Kumar Yogeswaaran
    affiliation: University of Canterbury New Zealand
    orcid: 0000-0002-1978-5077
    email: 
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
editor_options: 
    chunk_output_type: console
format:
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % adjust this value as needed
  docx:
    default: false
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true


# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
library(margot)
packageVersion(pkg = "margot")
library(boilerplate)

# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# combo_marginal_graph_ord
# combo_pols_graph_ord
# combo_eth_graph_ord
# combo_male_graph_ord


# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("tinytable")
# set seed
options(tinytable_tt_digits = 3)

# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)


# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-john-kerr-trust-science-growth-simple")

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)

measures_path_rstudio <- here::here("manuscripts", "24-john-kerr-growth-trust-science")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))


baseline_vars <- margot::here_read("baseline_vars")
exposure_vars <- margot::here_read("exposure_vars")
outcome_vars <- margot::here_read("outcome_vars")
n_participants <-here_read("n_participants")
# read waves
all_waves <- c(2019:2023)

# get table/graphs
summary_vars <- margot::here_read_qs("summary_vars", push_mods)

# read timeline
timeline_histogram <- here_read_qs("timeline_histogram", push_mods)

# hist outcomes
histogram_outcomes <- here_read_qs("histogram_trust_science_trust_scientists_by_wave", push_mods)

# get range of pol conservative
max_pol <- here_read('max_pol')
min_pol<- here_read('min_pol')

max_pol <- round(max_pol,2)
min_pol <- round(min_pol,2)
#outcome table means

# # read cat
# tbl_combined_cat <- here_read_qs("tbl_combined_cat", push_mods)
# 
# # first model
# tbl_combined_cat
# # first imputed dataset
# tbl_combined_imp <- here_read_qs("tbl_combined_imp", push_mods)
# tbl_combined_imp
# 
# # mean
# summary_table_mean <- here_read_qs("summary_table_mean", push_mods)
# summary_table_mean_imp <- here_read_qs("summary_table_mean_imp", push_mods)
# 
# summary_table_mean
# summary_table_mean_imp

tt_summary_mean_response_sample <- here_read("tt_summary_mean_response_sample")
tt_summary_mean_response_sample
tt_summary_mean_response_sample_imp <- here_read("tt_summary_mean_response_sample_imp")
tt_summary_mean_response_sample_imp


# alluvial plots
# tba
sankey_plot_science_sample<-  here_read_qs("sankey_plot_science_sample", push_mods)
sankey_plot_st_sample <-  here_read_qs("sankey_plot_st_sample", push_mods)
sankey_plot_science_imp<-  here_read_qs("sankey_plot_science_imp", push_mods)
sankey_plot_st_imp<-  here_read_qs("sankey_plot_st_imp", push_mods)
tt_summary_table_cat<- here_read("tt_summary_table_cat")
tt_summary_table_cat_imp <- here_read("tt_summary_table_cat_imp")
tt_summary_table_cat


# tt_summary_table_cat_imp |> print("markdown")
sankey_plot_science_sample/  sankey_plot_st_sample
# n_total
n_total <- here_read("n_total")

scale_outcome <- glue::glue("(1-7)")
outcome_lower_bound <- glue::glue("1")
outcome_upper_bound <- glue::glue("7")
# shift =   glue::glue('two points')
nice_baseline_wave <- glue::glue("New Zealand Attitudes and Values Study Time 11, years 2019-2024")
nice_all_waves <- glue::glue("New Zealand Attitudes and Values Study Time Time 11-14, years 2019-2024")
nice_end_of_study_wave <- glue::glue("Zealand Attitudes and Values Study Time 14, years 2022-2024")

# not used
prop_missing_baseline <- here_read("prop_missing_baseline")
# percentage_missing_baseline <- glue::glue(round(percent_missing_baseline,2))
# percentage_missing_baseline
# labels
#

# read models

# read models -------------------------------------------------------------
# nnet. 
# trust_science_marginal <- here_read("trust_science_marginal")
# trust_scientists_marginal <- here_read("trust_scientists_marginal")
# trust_science_pols <- here_read("trust_science_pols")
# trust_scientists_pols <- here_read("trust_scientists_pols")
# trust_science_eth <- here_read("trust_science_eth")
# trust_scientists_eth <- here_read("trust_scientists_eth")
# trust_science_male <- here_read("trust_science_male")
# trust_scientists_male <- here_read("trust_scientists_male")
# trust_science_marginal

# gee
# trust_science_marginal_gee <- here_read("trust_science_marginal_gee")
# trust_scientists_marginal_gee <- here_read("trust_scientists_marginal_gee")
# trust_science_pols_gee <- here_read("trust_science_pols_gee")
# trust_scientists_pols_gee <- here_read("trust_scientists_pols_gee")
# trust_science_eth_gee <- here_read("trust_science_eth_gee")
# trust_scientists_eth_gee <- here_read("trust_scientists_eth_gee")
# trust_science_male_gee <- here_read("trust_science_male_gee")
# trust_scientists_male_gee <- here_read("trust_scientists_male_gee")

predictions_gee_science_observed <- here_read("predictions_gee_science_observed")
predictions_gee_scientists_observed <- here_read("predictions_gee_scientists_observed")


predictions_gee_science_imputed <- here_read("predictions_gee_science_imputed")
predictions_gee_scientists_imputed <- here_read("predictions_gee_scientists_imputed")


# plot ordinal marginal ---------------------------------------------------

plot_trust_science_observed_continuous <- plot (
  predictions_gee_science_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Social Value of Science (1-7)", title = "Average Social Value of Science: Retained Sample") + theme_bw() 

plot_trust_scientists_observed_continuous <- plot (
  predictions_gee_scientists_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Average Trust in Scientists: Retained Sample") + theme_bw() 


plot_trust_science_imputed_continuous <- plot (
  predictions_gee_science_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Social Value of Science (1-7)", title = "Average Social Value of Science: Imputed Missing Responses") + theme_bw() 



plot_trust_scientists_imputed_continuous <- plot (
  predictions_gee_scientists_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Average Trust in Scientists: Imputed Missing Responses") + theme_bw()



# patchwork graph
combo_science_continuous<-   plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study): ", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Missing Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  


combo_scientists_continuous <- plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous   + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom') 


combo_all_continuous <- (plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous)/ (plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous) + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2023), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science and Scientists over Time: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  

# margot::margot_save_png(combo_all,base_filename = "trust_science_observed_vs_imputed", save_path = here::here(push_mods), height=20,width=20, dpi = 500)


# Categorical-------------------------------------------------------------------
# 
observed_trust_science_factor <- here_read("observed_trust_science_factor")
observed_trust_scientists_factor <- here_read("observed_trust_scientists_factor")

predictions_ord_science_imp <- here_read("predictions_ord_science_imp")
predictions_ord_scientists_imp <- here_read("predictions_ord_scientists_imp")

observed_trust_science_factor |> print_md()

# plot
observed_trust_science_factor_plot =  plot(
  observed_trust_science_factor,
  jitter = .4,
  dot_alpha  = .025,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Social Value of Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Social Value of Science: Retained Sample", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()

observed_trust_scientists_factor_plot =  plot(
  observed_trust_scientists_factor,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Retained Sample",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()


imputed_trust_science_factor_plot <- plot(
  predictions_ord_science_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
  )+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Social Value of Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Social Value of Science: Imputed Missing Responses", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()

imputed_trust_scientists_factor_plot <- plot(
  predictions_ord_scientists_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Imputed Missing Responses",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()  +   theme(
    plot.title = element_text(face = "bold", size = 14))


combo_science_factor <- observed_trust_science_factor_plot  + imputed_trust_science_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Social Value of Science: Observed vs Imputed Missing Responses: Multinomial Response Model\n'Our Society Places Too Much Emphasis on Science'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

combo_scientists_factor <- observed_trust_scientists_factor_plot  + imputed_trust_scientists_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Scientists: Observed vs Imputed Missing Responses: Multinomial Response Model\n'I have a high degree of confidence in the scientific community'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

# 
# combo_science_continuous
# combo_scientists_continuous
# 
#combo_science_factor
# combo_scientists_factor


# number of people in NZ > 18 year olds 
# method: use 2023 - 2024. Compare the raw frequencies at the point predictions for the # unadjusted and adjusted modes. 
# https://www.stats.govt.nz/information-releases/national-population-estimates-at-30-june-2024-2018-base/
# https://infoshare.stats.govt.nz/SelectVariables.aspx?pxID=591c2174-6602-4a34-b09c-e6632b63bac1

#  data frame with ages and their corresponding populations
age_data <- data.frame(
  age = c(18:94),  # individual from 18-94+ (i.e. adults)
  population = c(64630, 64520, 64890, 64670, 65530, 67670, 67520, 66940, 69260, 
                71030, 73710, 75910, 78480, 82610, 84700, 85540, 84050, 81890, 
                80670, 77110, 74690, 73350, 72050, 71290, 69150, 67670, 66310, 
                64340, 62700, 62540, 62280, 63930, 65310, 67480, 68680, 67990, 
                66290, 65770, 63690, 62470, 62250, 63160, 64470, 65050, 63600, 
                61750, 59310, 57740, 55850, 53770, 51900, 50090, 47950, 46570, 
                44650, 43350, 42140, 40880, 40180, 37450, 31820, 28810, 25850, 
                23730, 24110, 21970, 18950, 16480, 14250, 12150, 10270, 9020, 
                7560, 6210, 5120, 4260, 3360)
)

# compupted the weighted sum (age * population)
# total_age_years <- sum(age_data$age * age_data$population)

# calculate total population for these age groups
total_population <- sum(age_data$population)
total_population
# calc average age
# average_age <- total_age_years / total_population

# results
# cat("Total person-years:", format(total_age_years, scientific = FALSE), "\n")
cat("Total population:", format(total_population, scientific = FALSE), "\n")

#total
total_population

# results for unadjusted
observed_trust_science_factor

# results for adjusted (imputed)
predictions_ord_science_imp

# observed_trust_scientists_factor
# predictions_ord_scientists_imp

# unadjusted population estimate in low trust in science
trust_science_naive = round(total_population * .07)

# imputation adjusted population estimate in low trust in science
trust_science_imputed = round(total_population * .09)

# numerical difference 
low_trust_science_difference = trust_science_imputed - trust_science_naive

# make a pretty number with custom function
low_trust_science_difference =  margot::pretty_number(low_trust_science_difference)

# underestimating low trust
low_trust_science_difference 

# high trust in science calculations
high_trust_science_naive = round(total_population * .69)  # unadjusted
high_trust_science_imputed = round(total_population * .61)  # adjusted
high_trust_science_difference =  high_trust_science_naive - high_trust_science_imputed
high_trust_science_difference = margot::pretty_number(high_trust_science_difference)

# over-estimating high trust
high_trust_science_difference 

# reminder results for unadjusted
observed_trust_scientists_factor

# results for adjusted (imputed)
predictions_ord_scientists_imp


# unadjusted population estimate in low trust in scientists
trust_scientists_naive = round(total_population * .09)

# imputation adjusted population estimate in low trust in scientists
trust_scientists_imputed = round(total_population * .13)

# numerical difference 
low_trust_scientists_difference = trust_scientists_imputed - trust_scientists_naive

# make a pretty number with custom function
low_trust_scientists_difference =  margot::pretty_number(low_trust_scientists_difference)


# underestimating low trust in scientists by
low_trust_scientists_difference


# high trust in scientists calculations
high_trust_scientists_naive = round(total_population * 0.58)  # unadjusted
high_trust_scientists_imputed = round(total_population * 0.52)  # adjusted
high_trust_scientists_difference =  high_trust_scientists_naive - high_trust_scientists_imputed

high_trust_scientists_difference = margot::pretty_number(high_trust_scientists_difference)

# over-estimating high trust
high_trust_scientists_difference 



```



## Introduction

Whether people are becoming more sceptical of science is a matter of substantial practical interest and concern [@sturgis2021trust; @kennedy2022americans; @pagliaro2021trust]. Public trust in science helps shapes behaviour from the uptake of vaccination and other public health measures [@rueger2021perception] to support for climate action [@bogert2024effect]. Yet, accurately documenting trends in scientific mistrust is challenging because individuals with lower trust in science are less likely to participate in scientific studies, biasing results [@kreps2020model; @reif2021representative].

This study draws on five waves of national panel data collected between 2019 and 2024 from `r n_participants` participants in the New Zealand Attitudes and Values Study (NZAVS). Focusing on a 2019 baseline cohort, we estimate annual response trajectories for attitudes towards science and trust in scientists. To address attrition and systematic nonresponse, we apply multiple imputation models. By comparing the results of these adjusted models with unadjusted models in the restricted sample, we obtain a minimum bound for bias from differential attrition. This estimate is useful for gauging bias from cross-sectional data.

In **Study 1a**, we examine changes in the average value placed on science each year, comparing observed data with results that adjust for missingness. **Study 1b** mirrors this approach for trust in scientists, treating response as continuous variables. In **Study 2a**, we classify responses into low (1–3), medium (4–5), or high (6–7) categories for the value placed on science, and **Study 2b** applies the same categorical approach to trust in scientists, again forcussing on probabilities for classification. These categorical models enable us to capture patterns not evident from mean scores alone, revealing potential social divisions in attitudes toward science and scientists.

This work is descriptive and exploratory. Our primary objective is to characterise changes in institutional trust over time, focussing on how the value placed on science and trust in scientists evolve. By comparing raw estimates with those adjusted for missing data, we clarify how nonresponse can bias conclusions about trends in public trust. 

## Method


```{r}
#| eval: false
#| include: false
outcome_measures <- boilerplate::boilerplate_report_measures(
  baseline_vars = baseline_vars,
  outcome_vars = outcome_vars,
  measure_data = measure_data
)
cat(outcome_measures)

```

### Target Population

The target population for this study comprises the cohort of New Zealand residents in wave 2019 whose mistrust of science was not so great as to prevent participation in the New Zealand Attitudes and Values Study (NZAVS) that year. Our objective is to infer population dynamics for this cohort by using responses from individuals who took part in the NZAVS during the 2019 baseline wave, applying weights derived from the 2018 New Zealand Census to account for age, gender, and ethnicity distributions [@sibley2021].

### Sample

The NZAVS is a national probability study designed to represent the New Zealand population accurately. Although it employs prize draws to incentivise participation, response rates and demographic balances vary. For example, the NZAVS tends to under-sample males and individuals of Asian descent and to over-sample females and Māori (the Indigenous people of New Zealand). To enhance representativeness of the target population, we apply Census-based survey weights that adjust for age, gender, and ethnicity (New Zealand European, Asian, Māori, Pacific), thus improving the alignment of our sample with the national demographic composition [@sibley2021].

### Eligibility Criteria

To be eligible for inclusion in this study, participants needed to have responded to the `r nice_baseline_wave`. Individuals meeting this criterion were then tracked over the baseline and the subsequent four waves, permitting missing responses in any wave through `r nice_end_of_study_wave`. A total of `r n_participants` individuals satisfied these conditions and were included in the analyses.

[Appendix A](#appendix-a) (@tbl-demography) presents demographic characteristics of this cohort and details patterns of missing data. [Appendix B](#appendix-b) (@fig-timeline) illustrates the daily data collection periods for NZAVS waves 11–15 (2019–2024), providing additional context for the timing of responses and potential nonresponse.

## Measures

```{r}
#| eval: false
#| include: false
outcome_measures <- boilerplate::boilerplate_report_measures(
  outcome_vars = outcome_vars,
  measure_data = measure_data
)
```

We estimated target population average responses for two indicators attitudes to science, one that relates to the social value of science and another that measures trust in scientists (refer to @schoor2021science and @gundersen2022science on the importance of these distinctions). Item wording is as follows.

#### Social Value of Science

*Our society places too much emphasis on science (reversed).*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@hartman2017].

#### Trust in Scientists

*I have a high degree of confidence in the scientific community.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@nisbet2015].


Note again that each item addresses a different attitude to science. It is plausible that at least some who disagree with society's emphasis on science are nevertheless trusting of scientific institutions. On the other hand, and notably, the mistrust of scientists is an institutional cornerstone of science. It is the solomn duty of every scientists to doubt.

### Missing Data

We face two primary challenges when estimating population-level trends.

First, although the New Zealand Attitudes and Values Study uses a national probability sampling strategy, only about 10% of those invited participate. It is credible that the proportion of those who do not value science or who mistrust scientists are under-represented in the sample. After all, the  New Zealand Attitudes and Values Study is run by scientists aiming to foster scientific understanding. For this reason, it is challenging to estimate mistrust of science using surveys of the general population [@reif2021representative].

Second, although the NZAVS retains between 70–80% of its sample each year, attrition is inevitable. It is credible that the proportion of those who do not value science or who mistrust scientists becames increasingly under-represented in the longitudinal sample. For this reason, it is challenging to estimate mistrust of science in a longitudinal cohort. 

We cannot overcome the first challenge. We lack direct information about mistrust in the population of non-participants. 

However, for those who participated in 2019 but dropped out later, we can adjust estimates under the assumption that the probability of attrition or missing response is predictable given observed covariates, including previous attitudes to science. Methods for multiple imputation adjust for bias by incorporating missing data uncertainty into estimates [@blackwell_2017_unified; @bulbulia2023a].

Here, we used the `Amelia` package in R [@amelia_2011] to create ten multiply imputed datasets for the NZAVS 2019 cohort. These cover NZAVS time 11–15 (waves 2019–2023, years 2019-2024). We treated attitudes to science and scientists as ordinal responses bounded by 1 and 7 on the response scale. The Amelia algorithm is designed for within-unit imputation in repeated-measures time-series data. All covariates in @tbl-demography were included. The time variable (year) was modelled as a cubic spline to capture non-linear trends over four years. The code for the analysis is deposited here: (**OSF Link HERE**).


### Statistical Estimator

In Study 1a and Study 1b, we estimated mean responses for value placed on science and trust in scientists over time using generalised estimating equations (GEE). We implemented these models with the geepack package [@geepack_2006]. By specifying participant id as the clustering variable, we adjusted for repeated observations and obtained robust standard errors [@mcneish_2017unnecessary]. We weighted the data using sample weights derived from the 2018 New Zealand Census, ensuring that our estimates reflect population-level trends. We fitted separate models for each of the ten imputed datasets and pooled the results using Rubin's rule, thereby incorporating uncertainty from the missing data process into all population estimates. We generated predicted means and their associated confidence intervals using `ggeffects` [@ggeffects_2018].

In Study 2a and Study 2b, we examined predicted probabilities of attitudes to science and scientists in three discrete categories: low (1–3), medium (4–5), and high (6–7). For these analyses, we employed neural network models using the `nnet` package [@nnet_2002]. These models provided category-specific probability estimates at each time point. We applied sample weights constructed from the 2018 Census data to ensure population representativeness, and again used Rubin's rule to combine inferences across imputed datasets. We incorporated robust standard errors appropriate for clustered data and produced predictions and confidence intervals with `ggeffects` [@ggeffects_2018]. The final tables and figures were created using `tinytable` [@tinytable_2024], `ggplot2` [@ggplot2_2016], and `margot` [@margot2024].



#### Attitudes to Science and Scientists: Sample Means in Retained Sample


@fig-hist-outcomes presents histograms of attitudes to science and scientists over time in the observed sample. As evident in the graphs, there is increasing average value placed on science over time in the retained sample, with a large boost evident in wave 2020 (years 2020-2021). This boost coincides with the initially popular New Zealand COVID-19 pandemic response [@sibley2020a]. Similarly, in the retained sample, trust in scientists grew, coinciding with New Zealand's COVID-19 pandemic response, and then stabalised, before falling in the most recent wave (years 2023-2024).


{{< pagebreak >}}

```{r}
#| label: fig-hist-outcomes
#| fig-cap: "Historgram of sample attitudes to science over time in the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 20
#| fig-height: 9
#| 
histogram_outcomes
```


@tbl-sample-continuous present sample average responses and categorical distributions of value placed on science and trust in scientists across consecutive waves from 2019 to 2023, along with sample sizes and missing data counts for the wave 2019 cohort that dropped out or did not respond. 

Although average trust levels generally increased following New Zealand’s initial response to COVID-19, we also note rising missingness over time. This suggests that straightforward interpretations of sample averages may be misleading, because, as noted earlier, individuals who become more sceptical of science may be less likely to remain in the study.

@tbl-sample-continuous shows that mean value placed on science rose from 5.56 in 2019 to 5.86 in 2023, while mean trust in scientists increased from 5.30 to 5.55 between 2019 and 2020, stabilised through 2022, and then declined slightly to 5.45 in 2023. These observed increases, particularly for value placed on science following the New Zealand COVID-19 response, appear positive. However, missing responses grew markedly, from 563 in 2019 to 20,600 in 2023 for values of science, and from 1,257 in 2019 to 21,177 in 2023 for trust in scientists. Yet, as mentioned, this attrition may reflect systematic non-response related to mistrust.


::: {#tbl-sample-continuous}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_mean_response_sample |> print("markdown")
```
Retained sample average response by wave.
:::

{{< pagebreak >}}


### Sample Categorical Distributions Over Time


::: {#tbl-sample-cat}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_table_cat |> print("markdown")
```
Retained sample responses by wave by response category classified as low (1-3), medium (4-5), or high (6-7).
:::

@tbl-sample-cat categorises responses into low (1–3), medium (4–5), and high (6–7) ranges. For value placed on science, the proportion reporting low trust declined from 8.2% in 2019 to 6.7% in 2023, while the proportion reporting high trust rose from 60.5% to 70.1% over the same period.  [Appendix C](#appendix-c) @fig-alluv-science-sample graphically displays these transitions.

Similar patterns emerged for trust in scientists, with initial declines in the low category and increases in the high category, followed by some reversal in later waves. [Appendix C](#appendix-c) @fig-alluv-scientists-sample graphically displays these transitions.

These shifts might convey a narrative of strengthening value of sceince during early pandemic periods that continued to grow, with no evidence of increasing mistrust of scientists. Yet the steady increase in missing responses over time underscores the importance of careful interpretation. The observed improvements in attidutes towards science and scientists could be overstated if sceptical individuals disproportionately discontinued their participation. 

In other words, the rising average trust estimates may in part be an artefact of differential attrition.

{{< pagebreak >}}

## Results 


### Multiply Imputed Full Cohort Means in Full 2019 Cohort 

@tbl-sample-means-imp shows the estimated mean responses for value placed on science and trust in scientists, pooled over multiple imputations to adjust for systematic missingness. Compared with the observed sample means, the imputed estimates suggest moderate trends. For valuing science, the average response is higher in 2020 than in 2019, but then steadily declines from 5.7 in 2021 to 5.66 by 2023. For trust in scientists, the average increases from 5.29 to 5.47 between 2019 and 2020 but then declines to 5.28 by 2023. These imputed estimates reveal a less linear increase in trust than the unadjusted data might suggest, with evidence of gradual erosion in trust in both domains over the later waves.


::: {#tbl-sample-means-imp}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_mean_response_sample_imp |> print("markdown")
```
Imputed average responses for the full 2019 cohort
:::




::: {#tbl-sample-cat-imp}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
tt_summary_table_cat_imp|> print("markdown")
```
Cohort proportions classified as low, medium, or high (imputed)
:::


@tbl-sample-cat-imp classifies imputed responses into low, medium, and high categories. For valuing science, the proportion of respondents reporting low trust remains relatively stable, hovering around 8%-9% over time. In contrast, the initially large high-trust category (around 60%) rises to 65.8% in 2020, then recedes to 62.8% by 2023, suggesting that some of the early gains in trust may have been short-lived once nonresponse patterns are accounted for. For trust in scientists, imputation indicates a somewhat different trajectory. Initially, 11.9% reported low trust in 2019, decreasing to 10.1% in 2020, but this figure rises back to 12.8% by 2023. The medium and high categories show corresponding shifts, reflecting a more complex pattern than the non-imputed data suggested.

[Appendix d](#appendix-d) @fig-alluv-science-imp and @fig-alluv-scientists-imp visually present these transitions. Although early gains in high trust appeared promising, shifts over time are not uniformly in one direction. Instead, the imputed data indicate gradual redistributions of respondents across low, medium, and high categories. Such patterns underscore the importance of correcting for missingness before drawing conclusions about population-level changes in trust.

{{< pagebreak >}}


### Study 1a: Longitudinal Trajectories for Value Placed on Science: Continuous Analysis


```{r}
#| label: fig-science-continuous
#| fig-cap: "(A) Predicted value of science without adjusting for missingness (B) Predicted value of science adjusting for missingness, with points averaged over the ten imputed datasets (point alpha = 0.025). Note there is considerable variation of response above and below the mean, and there is slightly higher point density in panel (B) because we have adjusted for missingness."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_science_continuous
```


::: {#tbl-marginal-gee-science-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
predictions_gee_science_observed_x <- format(predictions_gee_science_observed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 
predictions_gee_science_observed_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted population average value placed on science over time, without adjusting for missing data.
:::


::: {#tbl-marginal-gee-science-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false
predictions_gee_science_imputed_x <- format(predictions_gee_science_imputed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 
predictions_gee_science_imputed_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted population average value placed on science over time, adjusting for missing data.
:::


@fig-science-continuous presents the predicted population average value place on science over time. Panel **A** shows estimates without adjusting for missing responses. Panel **B** shows estimates that incorporate multiple imputation. These predictions, derived from generalised estimating equations, reflect the estimated population trajectories, either without or with adjustments for missing responses. As shown in @tbl-marginal-gee-science-observed, when missingness is not taken into account, valuing science appears to increase steadily from the baseline period to the final wave. However, after modelling missingness as reported in @tbl-marginal-gee-science-imputed, the predicted trajectory suggests a more modest upward trend that does not continue to climb. This result shows that ignoring missing responses can inflate the perceived value of science in a longitudinal cohort.


{{< pagebreak >}}

### Study 1b: Longitudinal Trajectories for Trust in Scientists: Continuous Analysis


```{r}
#| label: fig-scientists-continuous
#| fig-cap: "(A) Predicted trust in scientists without adjusting for missingness (B) Predicted trust in scientists modelling missingness, with points averaged over the ten imputed datasets. Points are dimmed to alpha = .025. Note there is again considerable variation of response above and below the mean, and again there is slightly higher point density in panel (B) because we have adjusted for missingness."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_scientists_continuous
```

::: {#tbl-marginal-gee-scientists-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

predictions_gee_scientists_observed_x <- format(predictions_gee_scientists_observed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 
predictions_gee_scientists_observed_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )

```
Predicted population average trust in scientists over time, without adjusting for missing data.
:::


::: {#tbl-marginal-gee-scietists-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false

predictions_gee_scientists_imputed_x <- format(predictions_gee_scientists_imputed,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 
predictions_gee_scientists_imputed_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )

```
Predicted population average value placed on science over time adjusting for missing data.
:::


A similar pattern emerges for trust in scientists, presented in @fig-scientists-continuous. As shown in @tbl-marginal-gee-scientists-observed, the predicted mean trust increases when missingness is not accounted for. Yet when we adjust for missing data using multiple imputation, as reported in @tbl-marginal-gee-scietists-imputed, the results suggest that the initial gains in trust may be smaller and less sustained than the unadjusted models imply. By properly adjusting for missingness, these analyses provide a clearer picture of true population-wide changes in trust, rather than relying on observed data that may disproportionately represent individuals who remain engaged in the study.


{{< pagebreak >}}


### Study 2a: Longitudinal Trajectories for Value Placed on Science: Categorical Analysis


```{r}
#| label: fig-science-categorical
#| fig-cap: "(A) Predicted probability of classification for value placed on science without adjusting for missingness (B) Predicted probability of classification for value placed on science adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_science_factor
```

::: {#tbl-categorical-science-observed}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
observed_trust_science_factor_x <- format(observed_trust_science_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

observed_trust_science_factor_x|> 
     as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for value placed on science without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::



::: {#tbl-categorical-science-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false
predictions_ord_science_imp_x <- format(predictions_ord_science_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

predictions_ord_science_imp_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )


```
Predicted probability of classification for value placed on science, here adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


Note that in Studies 1a and 1b, the response distributions are scattered across the response scale 1-7 (refer to @fig-science-continuous).

@fig-science-categorical displays predicted probabilities for classification as low (1–3), medium (4–5), or high (6–7) in the value placed in science over time. @tbl-categorical-science-observed presents these probabilities without adjusting for missing responses. Under these conditions, the proportion of respondents in the high category appears to increase from the baseline and remain relatively elevated. However, after adjusting for missingness (see @tbl-categorical-science-imputed), the predicted probabilities show a less stable pattern, with gains in the high category tempered and the medium category becoming more prominent. These differences underscore that ignoring missing responses can create an artificial appearance of stronger sustained trust.


{{< pagebreak >}}


### Study 2b: Longitudinal Trajectories for Trust in Scientists: Categorical Analysis


```{r}
#| label: fig-scientists-categorical
#| fig-cap: "(A) Predicted probability of classification for trust in scientists without adjusting for missingness (B) Predicted probability of classification for trust in scientists adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 14
#| fig-height: 12

combo_scientists_factor
```

::: {#tbl-categorical-scientists-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
observed_trust_scientists_factor_x <- format(observed_trust_scientists_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

observed_trust_scientists_factor_x|> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


::: {#tbl-categorical-scientists-imputed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
predictions_ord_scientists_imp_x <- format(predictions_ord_scientists_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

predictions_ord_scientists_imp_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 14,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists, now adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


A similar pattern emerges for trust in scientists (see @fig-scientists-categorical). Without adjusting for missingness, the results (shown in @tbl-categorical-scientists-observed) suggest that the proportion classified as high in trust remains consistently large. Once we incorporate multiple imputation to address missing responses, as presented in @tbl-categorical-scientists-imputed, the predicted probabilities shift. The high category becomes less dominant over time, and more respondents fall into the medium category than would have been inferred without proper adjustment. In other words, accounting for nonresponse reveals a more nuanced pattern of trust that does not simply trend upward, and this correction prevents inflated estimates of stable or increasing trust in scientists.

## Discussion 

These findings illustrate that failing to adjust for missing responses can systematically distort estimates of public attitudes to science and scientists. Compared to models that ignore missingness, our imputation-adjusted analyses indicate that the unadjusted models overestimate the proportion of the population with high trust by several hundred thousand New Zealand adults and underestimate the proportion with low trust by tens to hundreds of thousands. For example, when the missing attitudes to science are not adjusted, we underestimate the share of the population with low trust by about 82,000 people, and simultaneously overstate the size of the high-trust group by about 330,000. Similarly, for trust in scientists, the discrepancy in the low-trust population is about 165,000 individuals, and the high-trust category may be inflated by about 248,000. These differences represent substantial distortions of attitudes to science and scientists in the general population of 4.1 million adults.

It is important to note, once again, that these are minimum bounds of potential distortion, as we cannot measure trust among the population who declined participation at the study’s outset. Moreover, we have not established any causal relationships. Our objective is not to identify determinants of changing trust, but rather to describe and characterise differences in trust distributions over time in that part of the general population whose attitudes to science did not prevent them from enlisting in a scientific study prior to COVID-19. Still, given that trust in science can influence critical actions related to climate change [@bogert2024effect], vaccine uptake [@pagliaro2021trust], and following health advice [@rueger2021perception], accurately documenting trust and its temporal shifts is a pressing concern. By accounting for missing data, we gain a clearer and more credible picture.


{{< pagebreak >}}

### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study. Members of the NZAVS management team and research group hold full copies of the NZAVS data. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author or any member of the NZAVS advisory board for replication or checking of any published study using NZAVS data. The code for the analysis can be found at: OSF-lINK TBA. 


### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Plank Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement

**TBA**: JK, CGS, DO, MW, and KY developed the measures for the New Zealand Attitudes and Values Study. JB developed the analytic approach and did the analysis. CGS led data collection. JB and CGS obtained the funding. **All authors had input into the manuscript.**


{{< pagebreak >}}




## Appendix A: Sample Descriptive Statistics {#appendix-a}

@tbl-demography presents descriptive statistics for the sample.


::: {#tbl-demography}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

library(gtsummary)
library(kableExtra)

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 
summary_vars |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Demographic statistics for New Zealand Attitudes and Values Cohort waves 2019-2023 (years 2019-2024).
:::

{{< pagebreak >}}

## Appendix B: New Zealand Attitudes and Values Study Data Collection waves 2019-2023 (years 2019-2024) {#appendix-b}


```{r}
#| label: fig-timeline
#| fig-cap: "Historgram of New Zealand Attitudes and Values Study Daily Data Collection Wave 2019 (time 11) - Wave 2023 (time 15)."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12
#| 


timeline_histogram
```

{{< pagebreak >}}


## Appendix C: Alluvial Graph of Showing Categorical Responses to Value of Science and Trust in Scientists: No Adjustment For Missing Data. {#appendix-c}

```{r}
#| label: fig-alluv-scientists-sample
#| fig-cap: "Alluvial plot suggests increasing flow of value of science for the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_science_sample
```


```{r}
#| label: fig-alluv-science-sample
#| fig-cap: "Alluvial plot suggests increasing flow of trust in scientists for the retained sample."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_st_sample

```



{{< pagebreak >}}


## Appendix D: Alluvial Graphs of Showing Categorical Responses to Social Value of Science and Trust in Scientists: Adjusting For Missing Data. {#appendix-d}

@fig-alluv-science-imp and @g-alluv-scientists-imp depict the dynamics of categorical responses for attitudes to science and scientists. Following 2020, there is a decline in the value placed on scientific institutions and a growth of mistrust in scientists. Our analysis presents the a minimal bound for such mistrust because we cannot account for the mistrust of the population who declined to praticipate in New Zealand Attitudes and Values Study wave 2019 (time 11).

```{r}
#| label: fig-alluv-science-imp
#| fig-cap: "Alluvial plot suggests declines in value of science for the imputed cohort."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_science_imp
```



```{r}
#| label: fig-alluv-scientists-imp
#| fig-cap: "Alluvial plot suggests declines in trust in scientists for the imputed cohort"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

sankey_plot_st_imp
```

## References {.appendix}











