---
title: "One-year Causal Effect of Religious Deconversion on Political Orientation in New Zealand"
abstract: |
  We leverage longitudinal (panel) data from a large national probability sample of New Zealanders to estimate three one-year causal effects of religious disaffiliation on political conservativism and political right-wing orientation (N = 47,202). In Study 1, we investigate the average treatment effect of a shift from religious affiliation to disaffiliation, contrasted with a condition in which the population was disaffiliated at baseline and remained so. We observe that religious disaffiliation reduces conservative/right political orientations. This effect is considerably stronger for political conservativism than for right-wing orientation. Study 2 replicates Study 1 in the religious-at-baseline sub-sample (N = 17,141) to obtain conditional treatment-effect estimates that require fewer assumptions. The conditional effect-estimates in Study 2 closely approximate the marginal effect estimates in Study 1, revealing robustness. In Study 3, we return to a population-wide marginal estimand, contrasting the transition to religious disaffiliation from affiliation with constant disaffiliation. This contrast clarifies whether recent religious disaffiliates are subject to "religious residue" effects in which attitudes the effects religious affiliation linger after disaffiliation. We detect a religious residue effect for political conservativism but not for right-wing orientation. Overall, these findings suggest that in New Zealand, religious disaffiliation affects political orientations by reducing conservative and (to a lesser extent) right-wing political orientations, with some evidence for religious residue effects for political conservativism.
  
author: 
  - name: Daryl Van Tongeren
    affiliation: Hope College
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University
    orcid_id: 0000-0003-3169-6576
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
keywords:
  - Causal Inference
  - Religion
  - Political Orientation 
editor_options: 
  chunk_output_type: console
format:
  pdf:
    sanitize: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [singlecolumn]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

#tinytex::tlmgr_update()

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
source("/Users/joseph/GIT/templates/functions/libs2.R")

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
source("/Users/joseph/GIT/templates/functions/funs.R")

# ALERT: UNCOMMENT THIS AND DOWNLOAD THE FUNCTIONS FROM JB's GITHUB
# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )

# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )


## WARNING SET THIS PATH TO YOUR DATA ON YOUR SECURE MACHINE. 
pull_path <-
  fs::path_expand(
    #"/Users/joseph/v-project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs_refactor/nzavs_data_23"
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs-current/r-data/nzavs_data"
  )

# read data: note that you need use the arrow package in R
### WARNING: THIS PATH WILL NOT WORK FOR YOU. PLEASE SET A PATH TO YOUR OWN COMPUTER!! ###
### WARNING: FOR EACH NEW STUDY SET UP A DIFFERENT PATH OTHERWISE YOU WILL WRITE OVER YOUR MODELS
push_mods <-
  fs::path_expand(
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/nzvs_mods/24/daryl-dones"
  )



# tables
library(gtsummary)
transition_table <-here_read("transition_table")
table_baseline <- here_read("table_baseline")

# exposures
table_exposures <- here_read("table_exposures")

# outcomes
table_outcomes <- here_read("table_outcomes")


# n participants
n_participants <-  here_read("n_participants")
n_participants_att <- here_read( "n_participants_att")

# models
# full sample conservative
tab_contrast_t2_political_conservative_z <- here_read("tab_contrast_t2_political_conservative_z")
output_tab_contrast_t2_political_conservative_z <- here_read("output_tab_contrast_t2_political_conservative_z")


# full sample pol wing
tab_contrast_t2_pol_wing_z <- here_read("tab_contrast_t2_pol_wing_z")
output_tab_contrast_t2_pol_wing_z <- here_read("output_tab_contrast_t2_pol_wing_z")



# # att conservative
# tab_contrast_t2_political_conservative_z_rels <- here_read("tab_contrast_t2_political_conservative_z_rels")
# output_tab_contrast_t2_political_conservative_z_rels <- here_read("output_tab_contrast_t2_political_conservative_z_rels")
# tab_contrast_t2_political_conservative_z_rels
# output_tab_contrast_t2_political_conservative_z_rels
# 
# # att pol wing
# tab_contrast_t2_pol_wing_z_rels <- here_read("tab_contrast_t2_pol_wing_z_rels")
# output_tab_contrast_t2_pol_wing_z_rels <- here_read("output_tab_contrast_t2_pol_wing_z_rels")
# tab_contrast_t2_pol_wing_z_rels
# output_tab_contrast_t2_pol_wing_z_rels
# # 
# 
# 
# #### GRAPHS
# tab_full <- here_read("tab_full")
# group_tab_full <- here_read("group_tab_full")
# 
# tab_att <- here_read("tab_att")
# tab_att
# group_tab_att <- here_read("group_tab_att")
# group_tab_att

# Combined
# N = n_participants
# N
# sub_title_ate = "ATE: N = 47,202"
# plot_group_tab_full <- margot_plot(
#   group_tab_full,
#   type = "RD",
#   title = "+1 year Effect of Religious Disaffiliation on
#   Political Orientation ",
#   subtitle = sub_title_ate,
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 11,
#   text_size = 3.0,
#   point_size = .5,
#   title_size = 15,
#   subtitle_size = 12,
#   legend_text_size = 12,
#   legend_title_size = 12,
#   x_offset = -1,
#   x_lim_lo = -1,
#   x_lim_hi =  .5
# )
# plot_group_tab_full
# here_save(plot_group_tab_full, "plot_group_tab_full")
# 
# sub_title_att = "CATE: N = 17,141"
# plot_group_tab_att <- margot_plot(
#   group_tab_att,
#   type = "RD",
#   title = "+1 year Effect of Religious Disaffiliation on
#   Political Orientation",
#   subtitle = sub_title_att,
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 11,
#   text_size = 3.0,
#   point_size = .5,
#   title_size = 15,
#   subtitle_size = 12,
#   legend_text_size = 12,
#   legend_title_size = 12,
#   x_offset = -1,
#   x_lim_lo = -1,
#   x_lim_hi =  .5
# )
# plot_group_tab_att
# here_save(plot_group_tab_att, "plot_group_tab_att")
# plot_group_tab_att

# tab_full <- here_read("tab_full")
# group_tab_full <- here_read("group_tab_full")
# 
# tab_att <- here_read("tab_att")
# tab_att
# group_tab_att <- here_read("group_tab_att")
# group_tab_att
# 
# tab_contrast <- here_read("tab_contrast")
# tab_contrast
# group_tab_contrast <- here_read("group_tab_contrast")
# group_tab_contrast
# 
# # Combined
# # N = n_participants
# # N
# sub_title_ate = "ATE: N = 47,202"
# plot_group_tab_contrast <- margot_plot(
#   group_tab_contrast,
#   type = "RD",
#   title = "+1 year Effect of Religious Disaffiliation on
#   Political Orientation Contrasted with Constant Disaffiliation",
#   subtitle = sub_title_ate,
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 11,
#   text_size = 3.0,
#   point_size = .5,
#   title_size = 15,
#   subtitle_size = 12,
#   legend_text_size = 12,
#   legend_title_size = 12,
#   x_offset = -1,
#   x_lim_lo = -1,
#   x_lim_hi =  .5
# )
# plot_group_tab_contrast
# here_save(plot_group_tab_contrast, "plot_group_tab_contrast")

# tab_full <- here_read("tab_full")
# group_tab_full <- here_read("group_tab_full")
# plot_group_tab_full <- here_read("plot_group_tab_full")
# 
# 
# tab_att <- here_read("tab_att")
# group_tab_att <- here_read("group_tab_att")
# plot_group_tab_att <- here_read("plot_group_tab_att")

#####
#library(glue)

# not this table is too long in most instance 
# table_baseline<- here_read("table_baseline")
# 
# # personality
# table_baseline_personality <- here_read("table_baseline_personality")
# 
# # demographic vars
# table_demographic_vars <- here_read("table_demographic_vars")
# 
# # self reported behaviours 
# table_virtue_vars <- here_read("table_virtue_vars")
# 
# # prejudice 
# table_acceptance_vars<- here_read("table_acceptance_vars")
# 
# # help received 
# table_selected_sorted_names_help_received_vars <- here_read("table_selected_sorted_names_help_received_vars")
# 1] "The table presents a transition matrix describing stability and movement between the treatment from the baseline to the treatment wave. Entries on the diagonal (in bold) indicate the number of individuals who stayed in their initial state. In contrast, the off-diagonal shows the transitions from the initial state (bold) to another state in the following wave (off-diagonal). A cell located at the intersection of row $i$ and column $j$, where $i \neq j$, presents the count of individuals moving from state $i$ to state $j$."

```

## Introduction

Throughout the west, religious disaffiliation is commonplace [@wilson2014differences]. How does religious disaffiliation affect political orientation? Because correlation is not causation, this question cannot be straightforwardly addressed with cross-sectional data. 

Longitudinal (panel) data may help to clarify causation, however, only if certain data and modelling assumptions are satisfied.

Here, we systematically investigate three distinct one-year causal effects of religious disaffiliation on two dimensions of political orientation. Our results draw from a national longitudinal probability study of New Zealanders (New Zealand Attitudes and Values Study, years 2018-2021, N = `r n_participants`). To obtain valid inference, we performing doubly-robust non-parametric estimation controlling for a rich array of baseline confounders, including the baseline treatment and outcomes. Finally, because confounding cannot be controlled with certainty, we provide sensitivity analyses for unmeasured confounding. 


## Method

### Sample

Data were collected as part of The New Zealand Attitudes and Values Study (NZAVS) is an annual longitudinal national probability panel study of social attitudes, personality, ideology and health outcomes. The NZAVS began in 2009. It includes questionnaire responses from a total of 72,290 New Zealand residents. The NZAVS is university-based, not-for-profit and independent of political or corporate funding. see [https://doi.org/10.17605/OSF.IO/75SNB](https://doi.org/10.17605/OSF.IO/75SNB)

[Appendix A](#appendix-measures) provides information about the measures used in this study.

[Appendix B](#appendix-demographics) reports sample statistics for all covariates at the baseline wave (NZAVS time 10, years 2018-2019).

[Appendix C](#appendix-exposures) reports sample statistics for the treatment at the baseline wave (NZAVS time 10, years 2018-2019) and the treatment wave (NZAVS time 11, years 2019-2020).

[Appendix D](#appendix-outcomes) reports reports sample statistics for the outcomes responses at baseline (NZAVS time 10, years 2018-2019) and the outcome wave (NZAVS time 12, years 2020-2021).

### The Treatment/Exposure: Religious Affiliation

**Study 1 and Study 2**: we emulate a "target trial" [@hernán2016a] that contrasts two treatments: (1) all start as religiously affiliated in the baseline wave and then lose their religious affiliation in the following wave; (2) all start as religiously affiliated in the baseline wave and remain religiously affiliated. These contrasts answer the question: "Does religious disaffiliation affect change in political orientations?" 

**Study 3**: we emulate a "target trial" [@hernán2016a] that contrasts religious affiliation to disaffiliation: (1) all start as religiously affiliated in the baseline wave and then lose their religious affiliation in the following wave; (2) all start as religiously disaffiliated in the baseline wave and then remain religiously disaffiliated: "Does religious disaffiliation affect change in political orientations such that the those who disaffiliate are different from those who were already religiously disaffiliated at baseline?"  Put differently, is there a religious residue effect?; see @vantongeren2020.

### Outcomes

The New Zealand Attitudes and Values Study has two measures of political orientation, both adapted from @jost_end_2006-1:

**Politically Conservative**: "Please rate how politically liberal versus conservative you see yourself as being." (1 = Extremely Liberal to 7 = Extremely Conservative)

**Politically Right Wing**: "Please rate how politically left-wing versus right-wing you see yourself as being" (1 = Extremely left-wing to 7 = Extremely right-wing)

Outcomes under the different treatments were assessed one year after the treatment wave. 

### Causal Contrast of Interest

Our causal estimands assume the form of "shift functions" or "modified treatment policies":

In **Study 1**, the causal estimand is: 

$$ \text{Average Treatment Effect} = E[Y(a^*,a)|\textcolor{red}{f(A)},L] -  E[Y(a^*, a^*)|\textcolor{red}{f(A)},L] $$

where the shift functions for this causal contrast are as specified as follows:

* **TREATMENT: start affiliated and then disaffiliate:**
   $$
   f(A) = \begin{cases} a^* & \text{if religious affiliation is absent, shift to religious affiliation} \\ 
   a & \text{if religious affiliation is present, shift to disaffiliation} \end{cases}
   $$

* **CONTRAST: start affiliated and remain affiliated:**
   $$
   f(A) = a^* ~ \text{if religious affiliation is absent, shift to religious affiliation} \\ 
  $$

The target population for Study 1 is the New Zealand population.

In **Study 2**, the causal estimand is the Conditional Average Treatment Effect in the Religious Population, $R = r$

$$ 
\text{Conditional Average Treatment Effects} = E[Y(a^*,a)|\textcolor{red}{f(A)},L, R = r] -  E[Y(a^*, a^*)|\textcolor{red}{f(A)},L, R = r] 
$$

where the shift functions for the causal contrast are the same as in Study 1. 

The target population for Study 2 is those who identify as religious in the New Zealand population.

In **Study 3** we investigate the average treatment effect in the full population, as is in Study 1, however, here we contrast the shift from religious affiliation to religious disaffiliation with a consistent pattern of religious disaffiliation:

$$ 
\text{Average Treatment Effect} = E[Y(a^*,a)|\textcolor{red}{f(A)},L] -  E[Y(a, a)|\textcolor{red}{f(A)},L] 
$$

where the shift functions for this causal contrast are as specified as follows:

* **TREATMENT: start affiliated and then disaffiliate (as in Study 1):**

   $$
   f(A) = \begin{cases} a^* & \text{if religious affiliation is absent, shift to religious affiliation} \\ 
   a & \text{if religious affiliation is present, shift to disaffiliation} \end{cases}
   $$

* **CONTRAST: always disaffiliate:**

   $$
   f(A) = a  ~ \text{if religious affiliation is present, shift to disaffiliation} \\ 
   $$
   
The target population for Study 1 is the New Zealand population.


### Eligibility criteria 

The sample consisted of respondents to NZAVS times 10 (baseline, years 2018-2019), time 11 (treatment wave, years 2019-2020), and time 12 (outcome wave, years 2020-2021). For sampling information please see: @sibley2021.

Criteria for inclusion were: enrolled in the New Zealand Attitudes and Values Study at Time 10 (baseline) and provided complete responses to the religious affiliation question ("Do you identify with a religious or spiritual group: y/n"). We allow that participants may have been lost to follow up, and adjusted for missingness using inverse probability of censoring model -- implemented during estimation in `lmtp` [@williams2021].

In Study 1, there were `r n_participants` NZAVS participants who met these criteria.

Study 2 had the same eligibility criteria, with the further requirement that, at baseline, participants responded "yes" to the religious affiliation question. There were `r n_participants_att` who met these criteria.


### Identification Assumptions

To consistently estimate causal effects, we rely on three fundamental assumptions:

1. **Causal Consistency:**  we assume the potential outcomes do not depend on the specific way treatment was administered, conditional on measured covariates [@vanderweele2009]. 

2. **Conditional Exchangability:**  we assume that given the observed covariates, treatment assignment is independent of potential outcomes. In simpler terms, we assume "no unmeasured confounding."

3. **Positivity:** we assume that every subject has a non-zero probability of receiving the treatments to be compared. [Appendix C](#appendix-exposures) shows a change in the treatment variable between NZAVS Time 10 and NZAVS Time 11.  
Notably, the causal estimands in Study 1 and Study 2 require projecting non-religious people into an initial state in which they are religious, an assumption that may arguably strain the positivity assumption. For this reason, in Study 2 we estimate a Conditional Average Treatment effect for the religiously affiliated sub-sample at baseline. The causal quantity that we obtain in Study 2 amounts to a contrast between the average effect among the religiously affiliated had they all lost affiliation with the average effect among the religiously affiliated had they remained affiliated. In New Zealand, the conditional average treatment effect (CATE) requires weaker assumptions that the average or marginal treatment effect in the population as a whole (ATE) because only 40% of the sample population reports religious affiliation. 

**Considerations:**

Because unmeasured confounding cannot be verified in the data, we perform sensitivity analysis by estimating the magnitude of unmeasured confounding required to explain away a result (E-values), see @vanderweele2017.


### Analysis

Model specification for NZAVS three-wave panel designs operate broadely within a targeted learning framework [@vanderlaan2011; @vanderlaan2018; @díaz2021] and are pre-specified in [@bulbulia2024PRACTICAL].


### Confounding Control Strategy

We adopt @vanderweele2020's *minimally modified disjunctive criteria* for confounding control.

1. **Initial identify confounders**: using causal diagrams, we began by enumerating all covariates that may influence either the treatment (exposure) or outcomes, spanning five domains. This includes variables directly affecting the exposure or outcome and potential effects of these variables (i.e. proxies.)

2. **Remove instrumental variables**: we removed variables identified as instrumental variables, i.e., those influencing the exposure but not the outcome. Their inclusion can reduce the efficiency of the analysis.

3. **Inclusion of proxy variables**: for unmeasured variables that affect both the exposure and outcome, include proxy variables wherever possible. These proxies act as indicators for the unmeasured common causes. [Appendix B](#appendix-demographics) lists covariates we used for confounding control. 

Again, these protocols follow the advice in [@bulbulia2024PRACTICAL] as prespecified in [https://osf.io/ce4t9/](https://osf.io/ce4t9/).

4. **Baseline exposure control**: we include baseline outcome measures to adjust for unmeasured confounding and obtain an incident exposure effect estimate [@vanderweele2020; @danaei2012; @hernan2023].

5. **Baseline outcome control**: we also include baseline measures of the outcome to adjust for unmeasured confounding [@vanderweele2020]. This strategy also minimises the possibility of reverse causation.

Thus, for any unmeasured variable to be associated with both the treatment in the exposure wave and the outcome one year later, this association would need to be independent of the measured baseline responses for the treatment and the outcome as well as for the demographic and personality covariates included in the model [@vanderweele2020].

Overall, then, our approach utilises *three-wave* of New Zealand Attitudes and Values Study (NZAVS) panel data: 

- **Wave 0**: the baseline wave (NZAVS Time 10, years 2018-2019)

- **Wave 1**: the treatment wave (NZAVS Time 11, years 2019-2020)

- **Wave 2**: and the outcome wave (NZAVS Time 12, years 2019-2020)

 Because in observational studies, we cannot be confident there is no unmeasured confounding, we perform additional sensitivity analysis using the E-value as described below [@vanderweele2017]. For further discussion of our three-wave confounding control strategy, see @vanderweele2020; @bulbulia2024PRACTICAL.

### Missing Data 

**Handling missing data**: to better ensure that the analyses are robust to potential biases introduced by missing data, we adopt the following strategies:

  - **Baseline missingness**: we impute missing data to recover baseline missing data for all variables except the treatment (`mice` package in R) [@vanbuuren2018]. Overall 1.6% of responses at baseline data were missing.
  
  - **Follow-up missingness**: we used the `lmtp` package's built-in non-parametric  inverse probability of censoring weighting feature to adjust for loss-to-follow-up/sample attrition [@williams2021]. Note this includes right-censoring including for the treatment variable in the treatment wave, enabling us to recover a *per protocol effect* that is more robust to informed censoring after baseline, including censoring of the treatment.

We obtain valid inferences for the target population - the New Zealand general population -- by applying post-stratification census weights for age, ethnicity, and gender using the census weights specified in @sibley2021.


#### Estimator 

We employ a semi-parametric estimator known as Sequentially Doubly Robust Estimation (SDR), which can estimate the causal effect of modified treatment policies on outcomes over time [@díaz2021]. This estimator belongs broadly to the "TMLE-verse" class of doubly-robust targeted learning framework developed by @vanderlaan2011; @vanderlaan2018. Estimation was performed using `lmtp` package [@williams2021; @díaz2021; @hoffman2023]. SDR is a robust method that flexibly combines machine learning techniques with traditional statistical models to estimate causal effects, while also providing valid statistical uncertainty measures for these estimates.

SDR operates through a two-step process involving outcome and treatment (exposure) models. Initially, it employs machine learning algorithms to flexibly model the relationship between treatments, covariates, and outcomes. This flexibility allows SDR to account for complex, high-dimensional covariate spaces without imposing restrictive parametric modelling assumptions. The outcome of this step is a set of initial estimates for these relationships.

The second step of SDR involves "targeting" these initial estimates by incorporating information about the observed data distribution to improve the accuracy for the causal effect-estimates of interest. This is achieved through an iterative updating process, which adjusts the initial estimates towards the true causal effect. In targeted learning, updating is guided by the efficient influence function, ensuring that the final estimate is as close as possible to the true causal effect, while remaining robust to model misspecification to either the outcome or treatment model.

A central feature of SDR is its double-robustness property, meaning that if either the model for the treatment or the outcome is correctly specified, the SDR estimator will still consistently estimate the causal effect. Additionally, SDR uses cross-validation to avoid over-fitting and ensure that the estimator performs well in finite samples. Each of these steps contributes to a robust methodology for examining the *causal* effects of interventions on outcomes. The marriage of SDR and machine learning technologies reduces the dependence on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021].


#### Estimation

We used the `lmtp` R package to estimate treatment the causal effects of pre-specified interventions on religious affiliation for the two political orientation responses that the New Zealand Attitudes and Values Study measures: political conservativism and right-wing orientation.  

As mentione, our models included an indicator for censoring in the next wave (right-censoring), which reduces bias from loss to follow-up/attrition conditional on measured covariates, subject to the covariates in the model.


Again in Study 1 and Study 2 we compare two hypothetical interventions:

- **TREATMENT: all start with religious affiliation and then lose religious affiliation**

- **CONTRAST: all start with religious affiliation and then remain religiously affiliated** 


In Study 3, we contrast a shift from affiliation to disaffiliation with an intervention in which everyone is constantly disaffiliated. This approach allows us to evaluate the religious residue hypothesis, according to which features of religious cognition and behaviour remain with disaffiliates even after they have lost their religion [@vantongeren2020]. Thus our causal contrast compares:

- **TREATMENT: all start with religious affiliation and then lose religious affiliation**

- **CONTRAST: all start with religious disaffiliation and then remain religiously disaffiliated** 


`lmtp` draws on the `SuperLearner` library, which comprises a growing set machine learning algorithms [@SuperLearner2023]. Given the size of the data, we used the `Ranger` estimator, which uses causal forests to estimate effects [@Ranger2017].  Causal forests excel in identifying complex, non-linear relationships between variables without presupposing a specific model form and thus making fewer assumptions about the underlying data distribution.

#### Cross-Validation

We implemented a 10-fold cross-validation in `lmtp` [@williams2021]. This method partitions the data into ten subsets of approximately equal size. During the cross-validation process, nine subsets are used to train the model, and the remaining subset is used for testing. This process is repeated ten times, with each of the ten subsets used exactly once as the test set. Using different subsets for training and validation minimises the risk of the model over-fit. Because each observation is used for training and validation the approach is efficient. The cross-validation process averages over the ten model estimates.


#### Sensitivity Analysis Using the E-value

To assess sensitivity to unmeasured confounding, we report VanderWeele and Ding's "E-value" in all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain away the observed exposure-outcome association  [@vanderweele2020; @linden2020EVALUE]. To evaluate strength of evidence, we use the bound of the E-value's 95% confidence interval that is closest to 1.


{{< pagebreak >}}

## Study 1: Causal Effect of Religious Disaffiliation on both Political Conservativism and Right-Wing Orientation in the New Zealand Population

```{r}
#| label: fig-ate
#| fig-cap: "Figure describing the effect of disaffiliation on political conservativism and right-wing orientation (both standardised) + 1 year after disaffiliation (ATE)."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_full <- here_read("plot_group_tab_full")
plot_group_tab_full
```


::: {#tbl-ate}
```{r}
#| label: tbl-ate
#| tbl-cap: "Table describing the effect of disaffiliation on political conservativism and right-wing orientation (both standardised) + 1 year after disaffiliation (ATE)."
#| eval: true
#| include: true
#| echo: false


tab_full <- here_read("tab_full")

tab_full |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

# group_tab_full <- here_read("group_tab_full")
# margot_interpret_table(group_tab_full, "causal_difference", "ATE")

```
:::

The Average Treatment Effect (ATE) represents the expected difference in outcomes between treatment and control groups for the population.

For the outcome 'Dis-affiliation Effect: Right Wing', the ATE is -0.08 [-0.14,-0.02]. The E-value for this effect estimate is 1.36 with a lower bound of 1.16. At this bound, an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.16 each could explain away the observed effect; weaker confounding would not. Here, we find evidence for causality.

For the outcome 'Dis-affiliation Effect: Pol.Conservative', the ATE is -0.19 [-0.25,-0.14]. The E-value for this effect estimate is 1.66 with a lower bound of 1.51. At this bound, an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.51 each could explain away the observed effect; weaker confounding would not. Here again, we find evidence for causality.


{{< pagebreak >}}

## Study 2: Causal Effect of Religious Disaffiliation on Political Conservativism and Right-Wing Orientation (both outcomes were standardised) in the Religious-Only Baseline Sample + 1 year after disaffiliation (CATE).


```{r}
#| label: fig-att
#| fig-cap: "Figure describing the effect of disaffiliation on political conservativism and right-wing orientation (both outcomes were standardised) among the religious-only sample+ 1 year after dis-affiliation (CATE)"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_att <- here_read("plot_group_tab_att")
plot_group_tab_att
```


::: {#tbl-att}
```{r}
#| label: tbl-att
#| tbl-cap: "Table describing the effect of disaffiliation on political conservativism and right-wing orientation among the religious-only sample  (both outcomes were standardised) + 1 year after dis-affiliation (CATE)"
#| eval: true
#| include: true
#| echo: false


tab_att <- here_read("tab_att")

tab_att |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

# group_tab_att <- here_read("group_tab_att")
# margot_interpret_table(group_tab_att, "causal_difference", "CATE")

```
:::

The Conditional Average Treatment Effect (CATE) represents the expected difference in outcomes between treatment and control groups in a stratum of the population -- here those who were religiously affiliated at baseline.

For the outcome 'Dis-affiliation Effect (CATE): Right Wing', the CATE is -0.07 [-0.12,-0.01]. The E-value for this effect estimate is 1.33 with a lower bound of 1.11. At this bound, an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.11 each could explain away the observed effect; weaker confounding would not. Here, we find evidence for causality.

For the outcome 'Dis-affiliation Effect (CATE): Pol.Conservative', the CATE is -0.22 [-0.27,-0.17]. The E-value for this effect estimate is 1.74, with a lower bound of 1.59. At this bound, an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.59 each could explain away the observed effect; weaker confounding would not. Here again, we find evidence for causality.


## Study 3 Causal Effect of Religious Disaffiliation on Political Conservativism and Right Wing Orientation  (both outcomes were standardised) contrasted with prevalence effect of consistent disaffiliation


```{r}
#| label: fig-contrast
#| fig-cap: "Figure describing the effect of disaffiliation on political conservativism and right-wing orientation (both outcomes were standardised) contrasted with the prevalence effect of constant dissafiliation + 1 year after treatment (ATE)"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_contrast <- here_read("plot_group_tab_contrast")
plot_group_tab_contrast
```


::: {#tbl-att}
```{r}
#| label: tbl-contrast
#| tbl-cap: "Table describing the effect of disaffiliation on political conservativism and right-wing orientation (both outcomes were standardised) contrasted with the prevalence effect of dissafiliation + 1 year after contrast (ATE)"
#| eval: true
#| include: true
#| echo: false


tab_contrast <- here_read("tab_contrast")

tab_contrast |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

# group_tab_contrast <- here_read("group_tab_contrast")
# margot_interpret_table(group_tab_contrast, "causal_difference", "ATE")

```
:::

For the outcome 'Dis-affiliation Contrast (ATE): Pol.Conservative', the ATE is 0.07 [0.02,0.12]. The E-value for this effect estimate is 1.33 with a lower bound of 1.11. At this bound, an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.11 each could explain away the observed effect; weaker confounding would not. Here, we find some evidence for causality: a hypothetical world in which everyone in the population was religiously affiliated and became disaffiliated would be a world in which the population is higher in political conservativism compared with a world in which people had not been religiously affiliated in the first place, and remained unaffiliated.  

For the outcome 'Dis-affiliation Contrast (ATE): Right Wing', given the lower bound of the E-value equals 1, we find no reliable evidence for causality. That is, we do not find any difference only right/left-wing political orientation between the hypothetical world in which everyone in the population was religiously affiliated and became disaffiliated compared with a world in which people had not been religiously affiliated in the first place, and remained unaffiliated.  


## Discussion

### Religious change affects political attitudes, although prior religious affiliation may leave a "residue"

Here, we target valid inference for three causal effects of religious disaffiliation on political attitudes by: (1) clearly specifying the causal quantities of interest; (2) stating the assumptions required to obtain these quantity from data; (3) providing a robust non-parametric statistical estimator specified with minimal modelling assumptions; (4) obtaining the relevant panel data and ordering these data such that confounders are measured before causes and causes are measured before their effects [@bulbulia2023].   

For political conservatism and right-wing orientation, disaffiliation shifts people to the liberal/left spectrum of orientation. Notably, estimates are similar for both the population as a whole (ATE) and the religious sub-population (CATE). In both cases, effects are more pronounced for political conservatism, with about a .2 SD expected to shift toward liberal orientation. Notably, effects were much weaker for (reduced) right-wing orientation, and both effects were replicated in the religious-at-baseline subsample (Study 2).

The findings in Study 1 and Study 2 apply to a hypothetical intervention in which religious people disaffiliate compared with a hypothetical intervention in which they do not.  In Study 3, we compare a hypothetical intervention in which religious people disaffiliate with one in which the population was disaffiliated at baseline and remained disaffiliated in the treatment wave.  We observe that the recently disaffiliated population would be somewhat more politically conservative than the permanently disaffiliated population. However we do not observed a reliable contrast for right-wing political orientation; notably, effects on this parameter were muted in Studies 1 and 2. Overall, our finding that political orientations are more conservative under the religious disaffiliation intervention compared with the stably disaffiliated intervention makes sense in light of @vantongeren2020, which finds that psychological features associated with religious behaviour and cognition remain after de-conversion, what @vantongeren2020 calls a "religious residue."  


### We do not evaluate the political sources of religious change 

Notably, our models identify the specific effect of religious disaffiliation on political orientations. It might be asked, what are the effects of changes in political orientations on religious affiliation? Such a question could be made precise, and if all identification assumptions are satisfied, a statistical estimand could be modelled and estimated. However, our interest here is to evaluate the effects of religious disaffiliation on political orientation, controlling for the prior effects of political orientation on disaffiliation.

### Why did the previous studies not detect a residue effect? 

The findings in Study 3 differ from those reported earlier in this manuscripts. Whereas Study 3 points to religious residue, our previous studies suggest that "dones" are more liberal than their long-time secular counterparts. We offer the following speculation:

1. **Cross-sectional data limitations**: such data do not provide a reliable standard for evidence for causality because the timing of confounders, treatments, and outcomes cannot be ensured. Correlations in cross sectional data may distort true causal relationships and are known to run in opposition to true causal associations [@westreich2013; @bulbulia2023]. 

2. **Measurement bias owing to recall**: when treatments and pre-treatment responses are based on memory, there is considerable scope for measurement bias [@hernan2009MEASUREMENT]. For instance, individuals recalling their religiosity may misremember both their religiosity and their past political conservatism or liberalism -- introducing directed measurement error.

3. **Unmeasured confounders**: valid causal inferences require satisfying the "no unmeasured confounder" assumption. It is plausible there are common causes of religious change and left-leaning or liberal ideologies that are not captured in the previous data -- such as "openness."  Failure to appropriately control for openness, every other potential confounder will yield unreliable causal effect estimates.

4. **Challenges in observing the incident-exposure effect**: accurately modelling the causal effects of religious change requires repeated measures time-series data in conversion is observed *in vivo*. Despite the large sample size of over 42,000 individuals, the New Zealand study identified approximately 1,100 instances of religious disaffiliation, as noted in @tbl-transition. That religious disaffiliation is rare poses considerable data collection challenges.

5. **One-year effect vs multi-year effect**: the New Zealand Study reports a one-year effect of religious stability or change. However, processes that begin with disaffiliation might continue for longer than one-year, such that over many years, religious dones are as liberal, or perhaps even more liberal, than long-disaffiliated counterparts.

6. **Model misspecification**: methods for obtaining causal effect estimates differ between the studies; one, several, or indeed all of the models employed may be misspecified in the sense that they do not converge to valid inferences for the populations from which data were collected  

7. **Sampling characteristics**: differences in sampling and in features of the underlying popolations from which the data were variously draw across the studies may ultimately explain the differences we observe across our studies. If there is treatment-effect heterogeneity, there is no need to appeal to reverse-causation, unmeasured confounding, measurement-error bias, or model-misspecification bias to explain these differences. 

Unfortunately, the data do not clarify whether one or any combination of these possible explanations, or perhaps entirely different explanations, underpin the observed discrepancies across the studies reported in this manuscript. Documenting and explaining variation in the effects of religious disaffiliation on political attitudes are matters for future longitudinal and comparative research. 

### Summary

This study provides robust evidence that religious disaffiliation leads to a shift towards liberal political orientations and strong reduction in political conservatism. These findings reveal a causal pathway from changes in religious affiliation to change in political attitudes. As New Zealand society grows more disenchanted with religion, we may expect that it will grow more disenchanted with political conservativism, at least in the near term. 

Again, whether the causal effects we observe in New Zealand transport to other societies remains unknown, an open matter for future investigations.

{{< pagebreak >}}

### Ethics

The NZAVS is reviewed every three years by the University of Auckland Human Participants Ethics Committee. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.

### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religion Trust (TRT0418). JB received support from the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.

### Author Statement

TBA... e.g. 


- DVT conceived of the study.
- CS led NZAVS data collection. 
- JB developed the inferential approach and did the analysis.  

{{< pagebreak >}}

## References

::: {#refs}
:::


{{< pagebreak >}}

## Appendix A: Measurues {#appendix-measures}

#### Age (waves: 1-15)

We asked participants' ages in an open-ended question ("What is your age?" or "What is your date of birth").


#### Charitable Donations

Using one item from @hoverd_religious_2010, we asked participants, "How much money have you donated to charity in the last year?". To stabilise this indicator, we took the natural log of the response + 1.

#### Education Attainment (waves: 1, 4-15)

We asked participants, "What is your highest level of qualification?". We coded participans highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing) See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf

#### Employment (waves: 1-3, 4-11)

We asked participants, "Are you currently employed? (This includes self-employed or casual work)". \* note: This question disappeared in the updated NZAVS Technical documents (Data Dictionary).

<!-- #### European (waves: 1-15) -->

<!-- Participants were asked "Which ethnic group do you belong to (NZ census question)?" or "Which ethnic group(s) do you belong to? (Open-ended)" (wave: 3). Europeans were coded as 1, whereas other ethnicities were coded as 0. -->

#### Ethnicity

Based on the New Zealand Census, we asked participants, "Which ethnic group(s) do you belong to?". The responses were: (1) New Zealand European; (2) Māori; (3) Samoan; (4) Cook Island Māori; (5) Tongan; (6) Niuean; (7) Chinese; (8) Indian; (9) Other such as DUTCH, JAPANESE, TOKELAUAN. Please state:. We coded their answers into four groups: Maori, Pacific, Asian, and Euro (except for Time 3, which used an open-ended measure).

#### Felt Belongingness

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument [@hagerty1995]: (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded.

#### Gender (waves: 1-15)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male was coded as 1, and gender diverse coded as 3 [@fraser_coding_2020]. (or 0.5 = neither female nor male)

Here, we coded all those who responded as Male as 1, and those who did not as 0.


#### Honesty-Humility-Modesty Facet (waves: 10-14)

Participants indicated the extent to which they agree with the following four statements from @campbell2004 , and @sibley2011 (1 = Strongly Disagree to 7 = Strongly Agree)

```
i.  I want people to know that I am an important person of high status, (Waves: 1, 10-14)
ii. I am an ordinary person who is no better than others.
iii. I wouldn't want people to treat me as though I were superior to them.
iv. I think that I am entitled to more respect than the average person is.
```
#### Has Siblings

"Do you have siblings?" [@stronge2019onlychild]

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours Volunteering

We measured hours of volunteering using one item from @sibley2011: "Hours spent … voluntary/charitable work."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Work

We measured hours of work using one item from @sibley2011:"Hours spent … working in paid employment."

To stabilise this indicator, we took the natural log of the response + 1.

#### Income (waves: 1-3, 4-15)

Participants were asked "Please estimate your total household income (before tax) for the year XXXX". To stabilise this indicator, we first took the natural log of the response + 1, and then centred and standardised the log-transformed indicator.

<!-- #### Job Security (waves: 1-3,4-7,9-15) -->

<!-- Participants indicated their feeling of job security by answering "How secure do you feel in your current job?" on a scale from 1 (not secure) to 7 (very secure). -->


#### Living in an Urban Area (waves: 1-15)

We coded whether they are living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

We coded whether they were living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

#### Mini-IPIP 6 (waves: 1-3,4-15)

We measured participants' personalities with the Mini International Personality Item Pool 6 (Mini-IPIP6) [@sibley2011], which consists of six dimensions and each dimension is measured with four items:

1.  agreeableness,

    i.  I sympathize with others' feelings.
    ii. I am not interested in other people's problems. (r)
    iii. I feel others' emotions.
    iv. I am not really interested in others. (r)

2.  conscientiousness,

    i.  I get chores done right away.
    ii. I like order.
    iii. I make a mess of things. (r)
    iv. I often forget to put things back in their proper place. (r)

3.  extraversion,

    i.  I am the life of the party.
    ii. I don't talk a lot. (r)
    iii. I keep in the background. (r)
    iv. I talk to a lot of different people at parties.

4.  honesty-humility,

    i.  I feel entitled to more of everything. (r)
    ii. I deserve more things in life. (r)
    iii. I would like to be seen driving around in a very expensive car. (r)
    iv. I would get a lot of pleasure from owning expensive luxury goods. (r)

5.  neuroticism, and

    i.  I have frequent mood swings.
    ii. I am relaxed most of the time. (r)
    iii. I get upset easily.
    iv. I seldom feel blue. (r)

6.  openness to experience

    i.  I have a vivid imagination.
    ii. I have difficulty understanding abstract ideas. (r)
    iii. I do not have a good imagination. (r)
    iv. I am not interested in abstract ideas. (r)

Each dimension was assessed with four items and participants rated the accuracy of each item as it applies to them from 1 (Very Inaccurate) to 7 (Very Accurate). Items marked with (r) are reverse coded.


#### NZ-Born (waves: 1-2,4-15)

We asked participants, "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15).


#### NZ Deprivation Index (waves: 1-15)

We used the NZ Deprivation Index to assign each participant a score based on where they live [@atkinson2019]. This score combines data such as income, home ownership, employment, qualifications, family structure, housing, and access to transport and communication for an area into one deprivation score.

#### Opt-in

The New Zealand Attitudes and Values Study allows opt-ins to the study. Because the opt-in population may differ from those sampled randomly from the New Zealand electoral roll; although the opt-in rate is low, we include an indicator (yes/no) for this variable.

#### NZSEI-13 (waves: 8-15)

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017]. This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation.

We asked participants, "If you are a parent, what is the birth date of your eldest child?".

#### Parent (waves: 5-15) -->

We asked participants, "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-15). Parents were coded as 1, while the others were coded as 0. -->

<!-- #### Number of Children (waves: 1-3, 4-15) -->

<!-- We measured the number of children using one item from @Bulbulia_2015. We asked participants, "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, or adopted?" or ""How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, and/or parented?" (waves: 12-15). -->

<!-- Here, we used responses to create a binary indicator for parenthood: Parent Yes = 1/ No = 0.  -->

#### Politically Conservative 

We measured participants' political conservative orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically liberal versus conservative you see yourself as being."

(1 = Extremely Liberal to 7 = Extremely Conservative)

#### Politically Right Wing 

We measured participants' political right-wing orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically left-wing versus right-wing you see yourself as being.."

(1 = Extremely left-wing to 7 = Extremely right-wing)


#### Relationship status

"What is your relationship status?" (e.g., single, married, de-facto, civil union, widowed, living together, etc.)



#### Religion Affiliation

Participants were asked to indicate their religion identification ("Do you identify with a religion and/or spiritual group?") on a binary response (1 = Yes, 0 = No). We then asked, "What religion or spiritual group?" These questions are used in the New Zealand Census


#### Support

Participants' perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated the extent to which they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item.

#### Volunteers

We asked participants,"Please estimate how many hours you spent doing each of the following things last week" and responded to an item ("voluntary/charitable work") from [@sibley2011].



## Appendix B. Baseline Demographic Statistics {#appendix-demographics}

::: {#tbl-B}

```{r}
#| label: tbl-table-demography-code
#| tbl-cap: "Baseline demographic statistics"
#| eval: false
#| echo: false
#| include: false

table_baseline <- read_here("table_baseline")
table_baseline |>  kbl(format = "markdown")
```

:::


|**Exposure + Demographic Variables** |**N = 47,202** |
|:------------------------------------|:--------------|
|__Age__                              |               |
|Mean (SD)                            |49 (14)        |
|Range                                |18, 99         |
|IQR                                  |39, 60         |
|__Agreeableness__                    |               |
|Mean (SD)                            |5.35 (0.99)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.00     |
|__Belong__                           |               |
|Mean (SD)                            |5.14 (1.08)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.33, 6.00     |
|__Belong Beliefs__                   |               |
|1                                    |794 (1.7%)     |
|2                                    |2,011 (4.3%)   |
|3                                    |4,273 (9.2%)   |
|4                                    |11,673 (25%)   |
|5                                    |13,505 (29%)   |
|6                                    |11,659 (25%)   |
|7                                    |2,769 (5.9%)   |
|__Born Nz__                          |36,834 (78%)   |
|__Charity Donate log__               |               |
|Mean (SD)                            |1.16 (1.61)    |
|Range                                |0.00, 5.13     |
|IQR                                  |0.00, 2.40     |
|__Conscientiousness__                |               |
|Mean (SD)                            |5.11 (1.06)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.50, 6.00     |
|__Education Level Coarsen__          |               |
|no_qualification                     |1,217 (2.6%)   |
|cert_1_to_4                          |16,581 (36%)   |
|cert_5_to_6                          |5,941 (13%)    |
|university                           |12,510 (27%)   |
|post_grad                            |5,099 (11%)    |
|masters                              |3,901 (8.4%)   |
|doctorate                            |1,120 (2.4%)   |
|__Employed__                         |37,497 (80%)   |
|__Eth Cat__                          |               |
|euro                                 |37,660 (81%)   |
|maori                                |5,345 (11%)    |
|pacific                              |1,108 (2.4%)   |
|asian                                |2,466 (5.3%)   |
|__Extraversion__                     |               |
|Mean (SD)                            |3.91 (1.20)    |
|Range                                |1.00, 7.00     |
|IQR                                  |3.00, 4.75     |
|__Have Siblings__                    |44,081 (95%)   |
|__Hlth Bmi__                         |               |
|Mean (SD)                            |27.2 (5.9)     |
|Range                                |12.3, 73.6     |
|IQR                                  |23.2, 30.1     |
|__Honesty Humility__                 |               |
|Mean (SD)                            |5.41 (1.18)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.25     |
|__Hours Charity log__                |               |
|Mean (SD)                            |0.43 (0.78)    |
|Range                                |0.00, 4.62     |
|IQR                                  |0.00, 0.69     |
|__Hours Children log__               |               |
|Mean (SD)                            |4.57 (2.50)    |
|Range                                |0.00, 13.12    |
|IQR                                  |3.26, 6.22     |
|__Hours Exercise log__               |               |
|Mean (SD)                            |1.54 (0.85)    |
|Range                                |0.00, 4.39     |
|IQR                                  |1.10, 2.08     |
|__Hours Housework log__              |               |
|Mean (SD)                            |2.14 (0.78)    |
|Range                                |0.00, 5.13     |
|IQR                                  |1.70, 2.71     |
|__Hours Work log__                   |               |
|Mean (SD)                            |2.66 (1.58)    |
|Range                                |0.00, 4.62     |
|IQR                                  |1.39, 3.71     |
|__Household Inc log__                |               |
|Mean (SD)                            |11.39 (0.77)   |
|Range                                |0.69, 14.92    |
|IQR                                  |11.00, 11.92   |
|__Male__                             |29,744 (63%)   |
|__Neighbourhood Community__          |               |
|1                                    |2,973 (6.3%)   |
|2                                    |5,877 (13%)    |
|3                                    |6,927 (15%)    |
|4                                    |9,734 (21%)    |
|5                                    |9,773 (21%)    |
|6                                    |7,949 (17%)    |
|7                                    |3,712 (7.9%)   |
|__Neuroticism__                      |               |
|Mean (SD)                            |3.49 (1.15)    |
|Range                                |1.00, 7.00     |
|IQR                                  |2.75, 4.25     |
|__Nz Dep2018__                       |              |
|Mean (SD)                            |4.77 (2.73)    |
|Range                                |1.00, 10.00    |
|IQR                                  |2.00, 7.00     |
|__Nzsei 13 l__                       |               |
|Mean (SD)                            |54 (17)        |
|Range                                |10, 90         |
|IQR                                  |41, 69         |
|__Openness__                         |               |
|Mean (SD)                            |4.96 (1.12)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.25, 5.75     |
|__Parent__                           |33,444 (71%)   |
|__Partner__                          |34,660 (75%)   |
|__Rural Gch 2018 l__                 |               |
|1                                    |29,050 (62%)   |
|2                                    |8,851 (19%)    |
|3                                    |5,779 (12%)    |
|4                                    |2,647 (5.6%)   |
|5                                    |569 (1.2%)     |
|__Sample Frame Opt in__              |1,379 (2.9%)   |
|__Support__                          |               |
|Mean (SD)                            |5.95 (1.12)    |
|Range                                |1.00, 7.00     |
|IQR                                  |5.33, 7.00     |
|__Urban__                            |37,901 (81%)   |

: Baseline demography statistics {#tbl-table-demography}


@tbl-table-demography presents baseline demographic statistics for couples who met inclusion criteria.




##  Appendix C: Baseline and Treatment Wave Exposure Statistics {#appendix-exposures}


```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and treatment wave"
#| eval: false
#| echo: false
#| include: false
table_exposures <- here_read("table_exposures")

#table_exposures |>  kbl(format = "markdown")
```



|**Baseline/Outcome Variables** |**2018**, N = 47,202 |**2019**, N = 47,202 |
|:------------------------------|:--------------------|:--------------------|
|__Alert Level Combined__       |                     |                     |
|no_alert                       |47,202 (100%)        |24,761 (71%)         |
|early_covid                    |0 (0%)               |3,895 (11%)          |
|alert_level_1                  |0 (0%)               |3,008 (8.7%)         |
|alert_level_2                  |0 (0%)               |864 (2.5%)           |
|alert_level_2_5_3              |0 (0%)               |569 (1.6%)           |
|alert_level_4                  |0 (0%)               |1,646 (4.7%)         |
|Unknown                        |0                    |12,459               |
|__Religious Affiliation__      |17,141 (36%)         |11,807 (34%)         |
|Unknown                        |0                    |12,459               |

: Baseline and Exposure Wave Responses {#tbl-table-exposures}


@tbl-table-exposures presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: religious affiliation (yes/no). Those who did not respond to religious affiliation were coded as "0". Because the treatment wave (NZAVS time 11) occurred to New Zealand's COVID-19 pandemic, all models adjusted for the pandemic alert-level. The pandemic is a not a "confounder" because a confounder must be related to the treatment and the outcome. At the end of the study measurement, all had been exposed to the pandemic.  However, to satisfy the causal consistency assumption, all treatments must be conditionally equivalent within levels of all covariates [@vanderweele2013]. Evidence shows that dangerous natural events may affect religious affiliation [@sibley2012a]. To better enable conditional independence within levels of the treatment variable, we conditioned on the lead value of COVID-alert level at baseline. To mitigate systematic biases arising from attrition, and missingness, `lmtp` package uses inverse probability of censoring weights, which were use when estimating the causal effects of the exposure on the outcome. 





### Transition Table for The Treatment

```{r}
#| label: transition-table
#| eval: false
#| echo: false

transition_table <- here_read("transition_table")
#transition_table
```

@tbl-transition shows a transition matrix that captures stability and movement between disinhibition responses from the baseline (NZAVS time 10) wave and exposure wave (NZAVS time 11) among all participants who responded to the religion question at baseline and in the following year (the censored sample). Entries on the diagonal (in bold) indicate the number of individuals who stayed in their initial state. In contrast, the off-diagonal shows the transitions from the initial state (bold) to another state the following wave (off-diagonal). A cell located at the intersection of row $i$ and column $j$, where $i \neq j$, shows the count of individuals moving from state $i$ to state $j$.


|     From      | Not Religious | Religious |
|:-------------:|:-------------:|:---------:|
| Not Religious |   **20959**   |   1183    |
|   Religious   |     1977      | **10624** |

: Transition matrix for change in treatment from baseline to the treatment wave. The diagonal shows stability in response. The off-diagonal shows the number of transitions from the initial state (bold) to another state in the following wave (off-diagonal). {#tbl-transition}





##  Appendix D: Baseline and End of Study Outcome Statistics {#appendix-outcomes}


```{r}
#| label: tbl-table-demographic_vars-code
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: false
#| echo: false
#| include: false
table_outcomes <- here_read("table_outcomes")
table_outcomes |>  kbl(format = "markdown")

```

|**Baseline/Outcome Variables** |**2018**, N = 47,202 |**2020**, N = 47,202 |
|:------------------------------|:--------------------|:--------------------|
|__Political Right Wing__       |                     |                     |
|1                              |1,819 (4.1%)         |1,293 (4.3%)         |
|2                              |6,736 (15%)          |5,309 (18%)          |
|3                              |8,456 (19%)          |6,226 (21%)          |
|4                              |16,288 (37%)         |10,033 (33%)         |
|5                              |6,812 (15%)          |4,740 (16%)          |
|6                              |3,186 (7.2%)         |2,042 (6.8%)         |
|7                              |768 (1.7%)           |480 (1.6%)           |
|Unknown                        |3,137                |17,079               |
|__Political Conservative__     |                     |                     |
|1                              |2,510 (5.6%)         |1,885 (6.2%)         |
|2                              |8,662 (19%)          |6,336 (21%)          |
|3                              |8,802 (20%)          |6,576 (21%)          |
|4                              |13,866 (31%)         |9,552 (31%)          |
|5                              |6,679 (15%)          |4,082 (13%)          |
|6                              |3,292 (7.4%)         |1,793 (5.9%)         |
|7                              |736 (1.7%)           |409 (1.3%)           |
|Unknown                        |2,655                |16,569               |

: Baseline and Outcome Wave Responses for the Outcome {#tbl-table-outcomes}

@tbl-table-outcomes presents baseline and wave 1 statistics for the outcome variables.  We see an overall shift in levels of conservativism and right-wing political orientation. Additionally, we observe 16,569 missing values. To mitigate systematic biases arising from attrition and missingness, `lmtp` package uses inverse probability of censoring weights.  







