---
title: "One-year Causal Effect of Religious Deconversion on Political Orientation"
abstract: |
  We leverage longitudinal (panel) data from a large national probability sample of New Zealanders to estimate the one-year causal effect of religious disaffiliation on political conservativism and political right-wing orientation (N = 47,202). We find a shift from disaffiliation to the liberal/left end of the spectrum.  Effects were considerably stronger on (reducing) political conservativism. Equivalent effects were observed in the religious-only-at-baseline sub-sample (N = 17,141). We infer that in New Zealand, religious disaffiliation affects political orientations by reducing conservative and (to a lesser extent) right-wing political orientations. Whether such effects transport to other settings remains unknown.
  
author: 
  - name: Daryl Van Tongeren
    affiliation: Hope College
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University
    orcid_id: 0000-0003-3169-6576
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
keywords:
  - Causal Inference
  - Religion
  - Political Orientation 
editor_options: 
  chunk_output_type: console
format:
  pdf:
    sanitize: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [singlecolumn]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

#tinytex::tlmgr_update()

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
source("/Users/joseph/GIT/templates/functions/libs2.R")

# WARNING:  COMMENT THIS OUT. JB DOES THIS FOR WORKING WITHOUT WIFI
source("/Users/joseph/GIT/templates/functions/funs.R")

# ALERT: UNCOMMENT THIS AND DOWNLOAD THE FUNCTIONS FROM JB's GITHUB
# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )

# source(
#   "https://raw.githubusercontent.com/go-bayes/templates/main/functions/experimental_funs.R"
# )


## WARNING SET THIS PATH TO YOUR DATA ON YOUR SECURE MACHINE. 
pull_path <-
  fs::path_expand(
    #"/Users/joseph/v-project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs_refactor/nzavs_data_23"
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/Joseph\ Bulbulia/00Bulbulia\ Pubs/DATA/nzavs-current/r-data/nzavs_data"
  )

# read data: note that you need use the arrow package in R
### WARNING: THIS PATH WILL NOT WORK FOR YOU. PLEASE SET A PATH TO YOUR OWN COMPUTER!! ###
### WARNING: FOR EACH NEW STUDY SET UP A DIFFERENT PATH OTHERWISE YOU WILL WRITE OVER YOUR MODELS
push_mods <-
  fs::path_expand(
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/nzvs_mods/24/daryl-dones"
  )



# tables
library(gtsummary)
transition_table <-here_read("transition_table")
table_baseline <- here_read("table_baseline")

# exposures
table_exposures <- here_read("table_exposures")

# outcomes
table_outcomes <- here_read("table_outcomes")


# n participants
n_participants <-  here_read("n_participants")
n_participants_att <- here_read( "n_participants_att")

# models
# full sample conservative
# tab_contrast_t2_political_conservative_z <- here_read("tab_contrast_t2_political_conservative_z")
# output_tab_contrast_t2_political_conservative_z <- here_read("output_tab_contrast_t2_political_conservative_z")
# 
# 
# # full sample pol wing
# tab_contrast_t2_pol_wing_z <- here_read("tab_contrast_t2_pol_wing_z")
# output_tab_contrast_t2_pol_wing_z <- here_read("output_tab_contrast_t2_pol_wing_z")
# 


# att conservative
# tab_contrast_t2_political_conservative_z_rels <- here_read("tab_contrast_t2_political_conservative_z_rels")
# output_tab_contrast_t2_political_conservative_z_rels <- here_read("output_tab_contrast_t2_political_conservative_z_rels")
# tab_contrast_t2_political_conservative_z_rels
# output_tab_contrast_t2_political_conservative_z_rels

# att pol wing
# tab_contrast_t2_pol_wing_z_rels <- here_read("tab_contrast_t2_pol_wing_z_rels")
# output_tab_contrast_t2_pol_wing_z_rels <- here_read("output_tab_contrast_t2_pol_wing_z_rels")
# tab_contrast_t2_pol_wing_z_rels
# output_tab_contrast_t2_pol_wing_z_rels



#### GRAPHS
# tab_full <- here_read("tab_full")
# group_tab_full <- here_read("group_tab_full")
# 
# tab_att <- here_read("tab_att")
# tab_att
# group_tab_att <- here_read("group_tab_att")
# group_tab_att
# 
# Combined
sub_title_ate = "ATE: N = 47,202"
# N = n_participants
# N

# plot_group_tab_full <- margot_plot(
#   group_tab_full,
#   type = "RD",
#   title = "+1 year Effect of Religious Disaffiliation on
#   Political Orientation ",
#   subtitle = sub_title_ate,
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 11,
#   text_size = 3.0,
#   point_size = .5,
#   title_size = 15,
#   subtitle_size = 12,
#   legend_text_size = 12,
#   legend_title_size = 12,
#   x_offset = -1,
#   x_lim_lo = -1,
#   x_lim_hi =  .5
# )
# plot_group_tab_full
# here_save(plot_group_tab_full, "plot_group_tab_full")

# sub_title_att = "ATT: N = 17,141"
# plot_group_tab_att <- margot_plot(
#   group_tab_att,
#   type = "RD",
#   title = "+1 year Effect of Religious Disaffiliation on
#   Political Orientation",
#   subtitle = sub_title_att,
#   xlab = "",
#   ylab = "",
#   estimate_scale = 1,
#   base_size = 11,
#   text_size = 3.0,
#   point_size = .5,
#   title_size = 15,
#   subtitle_size = 12,
#   legend_text_size = 12,
#   legend_title_size = 12,
#   x_offset = -1,
#   x_lim_lo = -1,
#   x_lim_hi =  .5
# )
# plot_group_tab_att
# here_save(plot_group_tab_att, "plot_group_tab_att")
# plot_group_tab_att


# tab_full <- here_read("tab_full")
# group_tab_full <- here_read("group_tab_full")
# plot_group_tab_full <- here_read("plot_group_tab_full")
# 
# 
# tab_att <- here_read("tab_att")
# group_tab_att <- here_read("group_tab_att")
# plot_group_tab_att <- here_read("plot_group_tab_att")

#####
#library(glue)

# not this table is too long in most instance 
# table_baseline<- here_read("table_baseline")
# 
# # personality
# table_baseline_personality <- here_read("table_baseline_personality")
# 
# # demographic vars
# table_demographic_vars <- here_read("table_demographic_vars")
# 
# # self reported behaviours 
# table_virtue_vars <- here_read("table_virtue_vars")
# 
# # prejudice 
# table_acceptance_vars<- here_read("table_acceptance_vars")
# 
# # help received 
# table_selected_sorted_names_help_received_vars <- here_read("table_selected_sorted_names_help_received_vars")
# 1] "The table presents a transition matrix describing stability and movement between the treatment from the baseline to the treatment wave. Entries on the diagonal (in bold) indicate the number of individuals who stayed in their initial state. In contrast, the off-diagonal shows the transitions from the initial state (bold) to another state in the following wave (off-diagonal). A cell located at the intersection of row $i$ and column $j$, where $i \neq j$, presents the count of individuals moving from state $i$ to state $j$."

```

## Introduction

Religious disaffiliation is commonplace. How does religious disaffiliation affect political orientation? Because correlation is not causation, this question cannot be straightforwardly addressed with cross-sectional data. 

Longitudinal (panel) data may help to clarify causation, however, only if certain data and modelling assumptions are satisfied.

Here, we systematically investigate the one-year causal effect of religious disaffiliation on two dimensions of political orientation by drawing on responses from a national longitudinal probability sample of New Zealanders (New Zealand Attitudes and Values Study, years 2018-2021, N = `r n_participants`), aligning our data, performing doubly-robust non-parametric estimation, and assessing sensitivity of results to unmeasured confounding. 


## Method

### Sample

Data were collected as part of The New Zealand Attitudes and Values Study (NZAVS) is an annual longitudinal national probability panel study of social attitudes, personality, ideology and health outcomes. The NZAVS began in 2009. It includes questionnaire responses from over 72,290 New Zealand residents. The study includes researchers from many New Zealand universities, including the University of Auckland, Victoria University of Wellington, the University of Canterbury, the University of Otago, and Waikato University. Because the survey asks the same people to respond each year, it can track subtle changes in attitudes and values and is an important resource for researchers in New Zealand and worldwide. The NZAVS is university-based, not-for-profit and independent of political or corporate funding.https://doi.org/10.17605/OSF.IO/75SNB

[Appendix A](#appendix-measures) provides statistical information about the measures used in this study.

[Appendix B](#appendix-demographics) provides information about demographic covariates used for confounding control (NZAVS time 10, years 2018-2019).

[Appendix C](#appendix-exposures) provides information about the treatment responses at baseline (NZAVS time 10) and in the treatment wave (NZAVS time 11, years 2019-2020).

[Appendix D](#appendix-outcomes) provides information about about the outcome variables responses at baseline (NZAVS time 10) and in the treatment wave (NZAVS time 12, years 2020-2021).

### Treatment/Exposure

A hypothetical experiment in which all start as religiously affiliated in the baseline wave and then either: (1) lose religious affiliation in the following wave or (2) retain religious affiliation in the following wave.

### Outcomes

The New Zealand Attitudes and Values Study has two measures of political orientation, both adapted from @jost_end_2006-1:

**Politically Conservative**: "Please rate how politically liberal versus conservative you see yourself as being." (1 = Extremely Liberal to 7 = Extremely Conservative)

**Politically Right Wing**: "Please rate how politically left-wing versus right-wing you see yourself as being" (1 = Extremely left-wing to 7 = Extremely right-wing)

Outcomes were assessed one-year after the exposure wave. 

### Causal Contrast of Interest

Our causal estimand assumes the form of a "shift function" or "modified treatment policy":

$$ \text{Average Treatment Effect} = E[Y(a^*)|\textcolor{red}{f(A)},L] - E[Y(a)|A,L] $$

**Intervention Functions:**

* **TREATMENT: start affiliated and then disaffiliate:**
   $$
   f(A) = \begin{cases} a^* & \text{if religious affiliation is retained, lose religious affiliation} \\ 
   \tilde{a} & \text{if religious affiliation is lost, do not shift} \end{cases}
   $$

* **CONTRAST: start affiliated and remain affiliated:**
   $$
   f(A) = \begin{cases} a^* & \text{if religious affilliation is lost, shift to religious affiliation} \\ 
   \tilde{a} & \text{if religious affiliation is retained, do not shift} \end{cases}
   $$
  

### Eligibility criteria 

The sample consisted of respondents to NZAVS times 10 (baseline, years 2018-2019), time 11 (treatment wave, years 2019-2020), and time 12 (outcome wave, years 2020-2021).

Criteria for inclusion were: enrolled in the New Zealand Attitudes and Values Study at Time 10 (baseline) and provided complete responses to the religious affiliation question ("Do you Identify with a religious or spiritual group: y/n") Participants may have been lost to follow up.  

In Study 1, there were `r n_participants` NZAVS participants who met these criteria.

Study 2 had the same eligibility criteria and the further requirement that, at baseline, participants responded "yes" to the religious affiliation question.


### Identification Assumptions

To consistently estimate causal effects, we rely on three key assumptions:

1. **Causal Consistency:** potential outcomes must align with observed outcomes under the treatments in our data. Essentially, we assume potential outcomes do not depend on the specific way treatment was administered as long as we consider measured covariates. 

2. **Conditional Exchangability:** given the observed covariates, we assume treatment assignment is independent of potential outcomes. In simpler terms, this means "no unmeasured confounding."

3. **Positivity:** for unbiased estimation, every subject must have a non-zero chance of receiving the treatment, regardless of their covariate values. [Appendix C](#appendix-exposures) shows a change in the treatment variable between NZAVS Time 10 and NZAVS Time 11.  A statistical estimand that involves projecting non-religious people into an initial state in which they are religious may strain the positivity assumption: such an intervention at the initial state might seem implausible. For this reason, we conducted an additional sensitivity analysis on the religiously affiliated sample at baseline (**Study 2**). The causal quantity that we obtain in Study 2 is the average treatment effect in the treated (ATT) -- that is, a contrast between the average effect among the religiously affiliated had they all lost affiliation in contrast with the average effect among the religiously affiliated had they remained affiliated? In New Zealand, the average treatment effect in the treated (ATT) holds weaker assumptions than the the treatment effect in the population as a whole (ATE), only 40% of which report religious affiliation. 

**Considerations:**

Because unmeasured confounding cannot be verified in the data, we perform sensitivity analysis by estimating the magnitude of unmeasured confounding required to explain away a result (E-values).


### Analysis

Criteria for for NZAVS three-wave panel designs are reported in [@bulbulia2024PRACTICAL].


### Confounding Control Strategy

We followed @vanderweele2020's *minimally modified disjunctive criteria* for confounding control.

1. **Initial identify confounders**: using causal diagrams, we began by enumerating all covariates that may influence either the treatment (exposure) or outcomes, spanning five domains. This includes variables directly affecting the exposure or outcome and potential effects of these variables (i.e. proxies.)

2. **Remove instrumental variables**: we removed variables identified as instrumental variables, i.e., those influencing the exposure but not the outcome. Their inclusion can reduce the efficiency of the analysis.

3. **Inclusion of proxy variables**: for unmeasured variables that affect both the exposure and outcome, include proxy variables wherever possible. These proxies act as indicators for the unmeasured common causes. [Appendix B](#appendix-demographics) lists covariates we used for confounding control. 

These protocols follow the advice in [@bulbulia2024PRACTICAL] as prespecified in [https://osf.io/ce4t9/](https://osf.io/ce4t9/).

4. **Baseline exposure control**: we included baseline outcome measures to adjust for unmeasured confounding and obtain an incident exposure effect estimate [@vanderweele2020; @danaei2012; @hernan2023].

5. **Baseline outcome control**: we also included baseline measures of the outcome to adjust for unmeasured confounding [@vanderweele2020]. This strategy also minimises the possibility of reverse causation.

### Missing Data 

**Handling missing data**: to better ensure that the analyses were robust to potential biases introduced by missing data, we adopted the following strategies:

  - **Baseline missingness**: we imputed missing data to recover baseline missing data for all variables except the treatment (religious affiliation) (`mice` package in R) [@vanbuuren2018]. Overall there was 1.6% of the baseline data were missing.
  
  - **Follow-up missingness**: We used the `lmtp` package's built-in censoring weighting to adjust for loss-to-follow-up/sample attrition [@williams2021], including for the treatment variable in the exposure wave.

We obtain valid inferences for the New Zealand population by applying post-stratification census weights for age, ethnicity, and gender.

Overall our approach utilises *three-wave* of panel data: 1. The baseline wave (NZAVS Time 10), the treatment wave (NZAVS Time 11), and the outcome wave (NZAVS Time 12). By controlling for baseline exposure, and baseline outcome, as well as for demographic and personality variables at baseline, we obtain powerful confounding control. Specifically, for any unmeasured variable to be associated with both the treatment in the exposure wave and the outcome one year later, this association would need to be independent of the measured baseline responses for the treatment and the outcome. However, because in observational studies, we cannot be certain there is no unmeasured confounding, we perform additional sensitivity analysis using the E-value as described below. (For further discussion of this strategy, see: @vanderweele2020; @bulbulia2023; @bulbulia2024PRACTICAL).


#### Estimator 

We employ a semi-parametric estimator known as Sequentially Doubly Robust Estimation (SDR), which is able to estimate the causal effect of modified treatment policies on outcomes over time [@vanderlaan2011; @vanderlaan2018]. Estimation was performed using `lmtp` package [@williams2021; @díaz2021; @hoffman2023]. SDR is a robust method that combines machine learning techniques with traditional statistical models to estimate causal effects while providing valid statistical uncertainty measures for these estimates.

SDR operates through a two-step process involving outcome and treatment (exposure) models. Initially, it employs machine learning algorithms to flexibly model the relationship between treatments, covariates, and outcomes. This flexibility allows SDR to account for complex, high-dimensional covariate spaces without imposing restrictive model assumptions. The outcome of this step is a set of initial estimates for these relationships.

The second step of SDR involves "targeting" these initial estimates by incorporating information about the observed data distribution to improve the accuracy of the causal effect estimate. This is achieved through an iterative updating process, which adjusts the initial estimates towards the true causal effect. This updating process is guided by the efficient influence function, ensuring that the final SDR estimate is as close as possible to the true causal effect while remaining robust to model misspecification in either the outcome or treatment model.

A central feature of SDR is its double-robustness property, meaning that if either the model for the treatment or the outcome is correctly specified, the SDR estimator will still consistently estimate the causal effect. Additionally, SDR uses cross-validation to avoid over-fitting and ensure that the estimator performs well in finite samples. Each of these steps contributes to a robust methodology for examining the *causal* effects on of interventions on outcomes. The marriage of SDR and machine learning technologies reduces the dependence on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021]


#### Estimation

We used the `lmtp` R package to estimate treatments' causal effects on the two political orientation outcomes for which the NZAVS measures: political conservativism and right-wing orientation.  


Our models included an indicator for censoring in the next wave, which adjusted for loss to follow-up/attrition conditional on measured covariates. Two custom shift functions were defined to model the hypothetical intervention of setting the treatment variable (religious affiliation) to "yes" in the baseline wave and then to "no" in the exposure wave; these estimates were contrasted against the expectation from a 'null' in which religious affiliation was set to "yes" in the baseline wave and "yes" in the exposure wave, without outcomes recorded in the wave that followed the exposure wave -- the "outcome" wave. 

Thus we have two interventions:

- **TREATEMENT: all start with religious affiliation and then lose religious affiliation**

- **CONTRAST: all start with religious affiliation and then remain religiously affiliated** 

`lmtp` draws on the the `SuperLearner` library, which comprises various machine learning algorithms [@SuperLearner2023]. Given the size of the data, we used the `Ranger` estimator, which uses causal forests to estimate effects [@Ranger2017].  Causal forests excel in identifying complex, non-linear relationships between variables without presupposing a specific model form and thus making fewer assumptions about the underlying data distribution.


#### Cross-Validation

We implemented a 10-fold cross-validation. This method partitions the data into ten subsets of approximately equal size. During the cross-validation process, nine subsets are used to train the model, and the remaining subset is used for testing. This process is repeated ten times, with each of the ten subsets used exactly once as the test set. Using different subsets for training and validation minimises the risk of the model being overly complex and fitting the noise in the training dataset, which can lead to poor performance on new data. Moreover, each observation is used for training and validation, which is particularly beneficial in scenarios where the amount of data is limited. The cross-validation process results in ten estimates of model accuracy, which can be averaged to provide a more comprehensive measure of model performance.


#### Sensitivity Analysis Using the E-value

To measure sensitivity to unmeasured confounding, we report VanderWeele and Ding's "E-value" for all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain away the observed exposure-outcome association  [@vanderweele2020; @linden2020EVALUE]. For example, if the E-value were 1.3, it would mean an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.3 each (or 30% increase in risk) could explain away the observed effect. Weaker confounding would not suffice. We also report the bound of the E-value closest to 1. If the lower bound of the CI were 1.1, to explain away the result, the strength of an unmeasured confounder in its association with both the treatment and the outcome would need to be at least 1.1 on the risk ratio scale (or a 10% increase in risk) to explain away the result [@vanderweele2017]. 


{{< pagebreak >}}

## Study 1: Causal Effect of Religious Disaffiliation on both Political Conservativism and Right-Wing Orientation in the New Zealand Population

```{r}
#| label: fig-ate
#| fig-cap: "Figure describing the effect of disaffiliation on political conservativism and right-wing orientation (both standardised) + 1 year after dis-affiliation (ATE)."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_full <- here_read("plot_group_tab_full")
plot_group_tab_full
```


::: {#tbl-ate}
```{r}
#| label: tbl-ate
#| tbl-cap: "Table describing the effect of disaffiliation on political conservativism and right-wing orientation (both standardised) + 1 year after dis-affiliation (ATE)."
#| eval: true
#| include: false
#| echo: false


tab_full <- here_read("tab_full")

tab_full |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

# group_tab_full <- here_read("group_tab_full")
# interpret_table(group_tab_full, "causal_difference", "ATE")

```
:::


The Average Treatment Effect (ATE) represents the expected difference in outcomes between treatment and control groups for the population. Results for the total population are presented in @fig-ate and @tbl-ate. Causal contrasts are given in standardised units for the outcome; the contrasts are therefore interpretable as standardised effect sizes.

For the outcome 'Dis-affiliation Effect: Right Wing', the ATE causal contrast is -0.08. The E-value for this outcome is 1.36; The confidence interval ranges from -0.14 to -0.02. Overall, we find reliable evidence for causality. Religious disaffiliation is expected to cause the disaffiliated to become less right-wing (more politically left-wing).

For the outcome 'Dis-affiliation Effect: Pol. Conservative’,  the ATE causal contrast is -0.19. The E-value for this outcome is 1.67; The confidence interval ranges from -0.25 to -0.14. We find even more substantial evidence for causality. Religious disaffiliation is expected to cause the disaffiliated to become less politically conservative (more politically liberal).


{{< pagebreak >}}

## Study 2: Causal Effect of Religious Disaffiliation on Political Conservativism and Right Wing Orientation  (both outcomes were standardised) in the Religious-Only Baseline Sample + 1 year after dis-affiliation (ATT).


```{r}
#| label: fig-att
#| fig-cap: "Figure describing the effect of disaffiliation on political conservativism and right-wing orientation (both outcomes were standardised) among the religious-only sample+ 1 year after dis-affiliation (ATT)"
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_group_tab_att <- here_read("plot_group_tab_att")
plot_group_tab_att
```


::: {#tbl-att}
```{r}
#| label: tbl-att
#| tbl-cap: "Table describing the effect of disaffiliation on political conservativism and right-wing orientation among the religious-only sample  (both outcomes were standardised) + 1 year after dis-affiliation (ATT)"
#| eval: true
#| include: false
#| echo: false


tab_att <- here_read("tab_att")

tab_att |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

group_tab_att <- here_read("group_tab_att")
interpret_table(group_tab_att, "causal_difference", "ATT")

```
:::

The Average Treatment Effect (ATT) represents the expected difference in outcomes between treatment and control groups in the treated population, in this case, those who were religiously affiliated at baseline. Results for the full population are presented in @fig-att and @tbl-att. Again, the causal contrasts we report are given in standardised units for the outcome; the contrasts are therefore interpretable as standardised effect-sizes.  

For the outcome 'Dis-affiliation Effect (ATT): Right Wing', the ATT causal contrast is -0.07. The E-value for this outcome is 1.32; The confidence interval ranges from -0.12 to -0.01. Overall, we find reliable evidence for causality. Here, in the religious sub-sample, we again find reliable evidence for causality. Religious disaffiliation is expected to cause the disaffiliated to become less right-wing (more politically left-wing).

For the outcome 'Dis-affiliation Effect: Pol.Conservative.', the ATT causal contrast is -0.22. The E-value for this outcome is 1.74; The confidence interval ranges from -0.27 to -0.17. Here, in the religious sub-sample, we again find even stronger evidence for a one-year causal effect on (lower) political conservativism (more liberal).

## Discussion

Here, we attempt to obtain valid inference for the causal effect of religious disaffiliation on political attitudes by (1) clearly specifying the causal quantity of interest, (2) stating the assumptions required to obtain this quantity from data, (3) providing a robust statistical estimator specified with minimal modelling assumptions (4) and of course, obtaining the relevant panel data and ordering such that confounders are measured before causes, and causes are measured before their effects [@bulbulia2023].   

We find that for political conservatism and political right-wing orientation, disaffiliation shifts people to the liberal/left spectrum of orientation. Notably, estimates are similar for both the population as a whole (ATE) and the religious subpopulation (ATT). In both cases, effects are more pronounced for political conservatism, with about a .2 SD expected to shift toward liberal orientation. Notably, effects were much weaker for (reduced) right-wing orientation. 

Notably, our models identify the specific effect of religious disaffiliation on political orientations. It might be asked, what are the effects of changes in political orientations on religious affiliation? Such a question could be made precise, and if all identification assumptions are satisfied, a statistical estimand could be modelled and estimated. However, our interest here is to eavlaute the effects of religious disaffiliation on political orientation, controlling for the prior effects of political orientation on disaffiliation.

In conclusion, this study provides robust evidence that religious disaffiliation leads to a shift towards liberal political orientations, and particularly strong reduction in political conservatism. These findings identify a causal pathway from change in religious identity to political attitudes. As New Zealand society grows more disenchanted with religion, we may expect that it will grow more disenchanted with political conservativism, at least in the near term, because religious disaffiliation affects political orientations by reducing conservative and (to a lesser extent) right-wing political orientations. 

Whether the effects we observe here transport to other societies remains unknown.

{{< pagebreak >}}

### Ethics

The NZAVS is reviewed every three years by the University of Auckland Human Participants Ethics Committee. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.

### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the TempletoReligion Trust (TRT0196; TRT0418). JB received support from the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.

### Author Statement

TBA... e.g. 


- DVT conceived of the study
- CS led data collection. 
- JB developed the inferential approach and did the analysis.  

{{< pagebreak >}}

## References

::: {#refs}
:::


{{< pagebreak >}}

## Appendix A: Measurues {#appendix-measures}

#### Age (waves: 1-15)

We asked participants' ages in an open-ended question ("What is your age?" or "What is your date of birth").


#### Charitable Donations

Using one item from @hoverd_religious_2010, we asked participants, "How much money have you donated to charity in the last year?". To stabilise this indicator, we took the natural log of the response + 1.

#### Education Attainment (waves: 1, 4-15)

We asked participants, "What is your highest level of qualification?". We coded participans highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing) See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf

#### Employment (waves: 1-3, 4-11)

We asked participants, "Are you currently employed? (This includes self-employed or casual work)". \* note: This question disappeared in the updated NZAVS Technical documents (Data Dictionary).

<!-- #### European (waves: 1-15) -->

<!-- Participants were asked "Which ethnic group do you belong to (NZ census question)?" or "Which ethnic group(s) do you belong to? (Open-ended)" (wave: 3). Europeans were coded as 1, whereas other ethnicities were coded as 0. -->

#### Ethnicity

Based on the New Zealand Census, we asked participants, "Which ethnic group(s) do you belong to?". The responses were: (1) New Zealand European; (2) Māori; (3) Samoan; (4) Cook Island Māori; (5) Tongan; (6) Niuean; (7) Chinese; (8) Indian; (9) Other such as DUTCH, JAPANESE, TOKELAUAN. Please state:. We coded their answers into four groups: Maori, Pacific, Asian, and Euro (except for Time 3, which used an open-ended measure).

#### Felt Belongingness

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument [@hagerty1995]: (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded.

#### Gender (waves: 1-15)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male was coded as 1, and gender diverse coded as 3 [@fraser_coding_2020]. (or 0.5 = neither female nor male)

Here, we coded all those who responded as Male as 1, and those who did not as 0.


#### Honesty-Humility-Modesty Facet (waves: 10-14)

Participants indicated the extent to which they agree with the following four statements from @campbell2004 , and @sibley2011 (1 = Strongly Disagree to 7 = Strongly Agree)

```
i.  I want people to know that I am an important person of high status, (Waves: 1, 10-14)
ii. I am an ordinary person who is no better than others.
iii. I wouldn't want people to treat me as though I were superior to them.
iv. I think that I am entitled to more respect than the average person is.
```
#### Has Siblings

"Do you have siblings?" [@stronge2019onlychild]

#### Hours of Childcare

We measured hours of exercising using one item from @sibley2011: 'Hours spent … looking after children."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Housework

We measured hours of exercising using one item from @sibley2011: "Hours spent … housework/cooking"

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours of Exercise

We measured hours of exercising using one item from @sibley2011: "Hours spent … exercising/physical activity" 

To stabilise this indicator, we took the natural log of the response + 1.

#### Hours Volunteering

We measured hours of volunteering using one item from @sibley2011: "Hours spent … voluntary/charitable work."

To stabilise this indicator, we took the natural log of the response + 1.


#### Hours of Work

We measured hours of work using one item from @sibley2011:"Hours spent … working in paid employment."

To stabilise this indicator, we took the natural log of the response + 1.

#### Income (waves: 1-3, 4-15)

Participants were asked "Please estimate your total household income (before tax) for the year XXXX". To stabilise this indicator, we first took the natural log of the response + 1, and then centred and standardised the log-transformed indicator.

<!-- #### Job Security (waves: 1-3,4-7,9-15) -->

<!-- Participants indicated their feeling of job security by answering "How secure do you feel in your current job?" on a scale from 1 (not secure) to 7 (very secure). -->


#### Living in an Urban Area (waves: 1-15)

We coded whether they are living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

We coded whether they were living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

#### Mini-IPIP 6 (waves: 1-3,4-15)

We measured participants' personalities with the Mini International Personality Item Pool 6 (Mini-IPIP6) [@sibley2011], which consists of six dimensions and each dimension is measured with four items:

1.  agreeableness,

    i.  I sympathize with others' feelings.
    ii. I am not interested in other people's problems. (r)
    iii. I feel others' emotions.
    iv. I am not really interested in others. (r)

2.  conscientiousness,

    i.  I get chores done right away.
    ii. I like order.
    iii. I make a mess of things. (r)
    iv. I often forget to put things back in their proper place. (r)

3.  extraversion,

    i.  I am the life of the party.
    ii. I don't talk a lot. (r)
    iii. I keep in the background. (r)
    iv. I talk to a lot of different people at parties.

4.  honesty-humility,

    i.  I feel entitled to more of everything. (r)
    ii. I deserve more things in life. (r)
    iii. I would like to be seen driving around in a very expensive car. (r)
    iv. I would get a lot of pleasure from owning expensive luxury goods. (r)

5.  neuroticism, and

    i.  I have frequent mood swings.
    ii. I am relaxed most of the time. (r)
    iii. I get upset easily.
    iv. I seldom feel blue. (r)

6.  openness to experience

    i.  I have a vivid imagination.
    ii. I have difficulty understanding abstract ideas. (r)
    iii. I do not have a good imagination. (r)
    iv. I am not interested in abstract ideas. (r)

Each dimension was assessed with four items and participants rated the accuracy of each item as it applies to them from 1 (Very Inaccurate) to 7 (Very Accurate). Items marked with (r) are reverse coded.


#### NZ-Born (waves: 1-2,4-15)

We asked participants, "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15).


#### NZ Deprivation Index (waves: 1-15)

We used the NZ Deprivation Index to assign each participant a score based on where they live [@atkinson2019]. This score combines data such as income, home ownership, employment, qualifications, family structure, housing, and access to transport and communication for an area into one deprivation score.

#### Opt-in

The New Zealand Attitudes and Values Study allows opt-ins to the study. Because the opt-in population may differ from those sampled randomly from the New Zealand electoral roll; although the opt-in rate is low, we include an indicator (yes/no) for this variable.

#### NZSEI-13 (waves: 8-15)

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017]. This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation.

We asked participants, "If you are a parent, what is the birth date of your eldest child?".

#### Parent (waves: 5-15) -->

We asked participants, "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-15). Parents were coded as 1, while the others were coded as 0. -->

<!-- #### Number of Children (waves: 1-3, 4-15) -->

<!-- We measured the number of children using one item from @Bulbulia_2015. We asked participants, "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, or adopted?" or ""How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, and/or parented?" (waves: 12-15). -->

<!-- Here, we used responses to create a binary indicator for parenthood: Parent Yes = 1/ No = 0.  -->

#### Politically Conservative 

We measured participants' political conservative orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically liberal versus conservative you see yourself as being."

(1 = Extremely Liberal to 7 = Extremely Conservative)

#### Politically Right Wing 

We measured participants' political right-wing orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically left-wing versus right-wing you see yourself as being.."

(1 = Extremely left-wing to 7 = Extremely right-wing)


#### Relationship status

"What is your relationship status?" (e.g., single, married, de-facto, civil union, widowed, living together, etc.)



#### Religion Affiliation

Participants were asked to indicate their religion identification ("Do you identify with a religion and/or spiritual group?") on a binary response (1 = Yes, 0 = No). We then asked, "What religion or spiritual group?" These questions are used in the New Zealand Census


#### Support

Participants' perceived social support was measured using three items from @cutrona1987 and @williams_cyberostracism_2000: (1) "There are people I can depend on to help me if I really need it"; (2) "There is no one I can turn to for guidance in times of stress"; (3) "I know there are people I can turn to when I need help." Participants indicated the extent to which they agreed with those items (1 = Strongly Disagree to 7 = Strongly Agree). The second item was negatively worded, so we reversely recorded the responses to this item.

#### Volunteers

We asked participants,"Please estimate how many hours you spent doing each of the following things last week" and responded to an item ("voluntary/charitable work") from [@sibley2011].



## Appendix B. Baseline Demographic Statistics {#appendix-demographics}


```{r}
#| label: tbl-table-demography-code
#| tbl-cap: "Baseline demographic statistics"
#| eval: false
#| echo: false
#| include: false

table_baseline <- read_here("table_baseline")
#table_baseline |>  kbl(format = "markdown")

```


|**Demographic Variables**            |**N = 47,202** |
|:------------------------------------|:--------------|
|__Age__                              |NA             |
|Mean (SD)                            |49 (14)        |
|Range                                |18, 99         |
|IQR                                  |39, 60         |
|__Agreeableness__                    |NA             |
|Mean (SD)                            |5.35 (0.99)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.00     |
|__Belong__                           |NA             |
|Mean (SD)                            |5.14 (1.08)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.33, 6.00     |
|__Belong Beliefs__                   |NA             |
|1                                    |794 (1.7%)     |
|2                                    |2,011 (4.3%)   |
|3                                    |4,273 (9.2%)   |
|4                                    |11,673 (25%)   |
|5                                    |13,505 (29%)   |
|6                                    |11,659 (25%)   |
|7                                    |2,769 (5.9%)   |
|__Born NZ__                          |36,834 (78%)   |
|__Charity Donate log__               |NA             |
|Mean (SD)                            |1.16 (1.61)    |
|Range                                |0.00, 5.13     |
|IQR                                  |0.00, 2.40     |
|__Conscientiousness__                |NA             |
|Mean (SD)                            |5.11 (1.06)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.50, 6.00     |
|__Education Level Coarsen__          |NA             |
|no_qualification                     |1,217 (2.6%)   |
|cert_1_to_4                          |16,581 (36%)   |
|cert_5_to_6                          |5,941 (13%)    |
|university                           |12,510 (27%)   |
|post_grad                            |5,099 (11%)    |
|masters                              |3,901 (8.4%)   |
|doctorate                            |1,120 (2.4%)   |
|__Employed__                         |37,497 (80%)   |
|__Eth Cat__                          |NA             |
|Euro                                 |37,660 (81%)   |
|Maori                                |5,345 (11%)    |
|Pacific                              |1,108 (2.4%)   |
|Asian                                |2,466 (5.3%)   |
|__Extraversion__                     |NA             |
|Mean (SD)                            |3.91 (1.20)    |
|Range                                |1.00, 7.00     |
|IQR                                  |3.00, 4.75     |
|__Has Siblings__                     |44,081 (95%)   |
|__Hlth Bmi__                         |NA             |
|Mean (SD)                            |27.2 (5.9)     |
|Range                                |12.3, 73.6     |
|IQR                                  |23.2, 30.1     |
|__Honesty Humility__                 |NA             |
|Mean (SD)                            |5.41 (1.18)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.75, 6.25     |
|__Hours Charity log__                |NA             |
|Mean (SD)                            |0.43 (0.78)    |
|Range                                |0.00, 4.62     |
|IQR                                  |0.00, 0.69     |
|__Hours Children log__               |NA             |
|Mean (SD)                            |4.57 (2.50)    |
|Range                                |0.00, 13.12    |
|IQR                                  |3.26, 6.22     |
|__Hours Exercise log__               |NA             |
|Mean (SD)                            |1.54 (0.85)    |
|Range                                |0.00, 4.39     |
|IQR                                  |1.10, 2.08     |
|__Hours Housework log__              |NA             |
|Mean (SD)                            |2.14 (0.78)    |
|Range                                |0.00, 5.13     |
|IQR                                  |1.70, 2.71     |
|__Hours Work log__                   |NA             |
|Mean (SD)                            |2.66 (1.58)    |
|Range                                |0.00, 4.62     |
|IQR                                  |1.39, 3.71     |
|__Household Inc log__                |NA             |
|Mean (SD)                            |11.39 (0.77)   |
|Range                                |0.69, 14.92    |
|IQR                                  |11.00, 11.92   |
|__Male__                             |29,744 (63%)   |
|__Neighbourhood Community__          |NA             |
|1                                    |2,973 (6.3%)   |
|2                                    |5,877 (13%)    |
|3                                    |6,927 (15%)    |
|4                                    |9,734 (21%)    |
|5                                    |9,773 (21%)    |
|6                                    |7,949 (17%)    |
|7                                    |3,712 (7.9%)   |
|__Neuroticism__                      |NA             |
|Mean (SD)                            |3.49 (1.15)    |
|Range                                |1.00, 7.00     |
|IQR                                  |2.75, 4.25     |
|__NZ Dep2018__                       |NA             |
|Mean (SD)                            |4.77 (2.73)    |
|Range                                |1.00, 10.00    |
|IQR                                  |2.00, 7.00     |
|__Nzsei 13 l__                       |NA             |
|Mean (SD)                            |54 (17)        |
|Range                                |10, 90         |
|IQR                                  |41, 69         |
|__Openness__                         |NA             |
|Mean (SD)                            |4.96 (1.12)    |
|Range                                |1.00, 7.00     |
|IQR                                  |4.25, 5.75     |
|__Parent__                           |33,444 (71%)   |
|__Partner__                          |34,660 (75%)   |
|__Rural Gch 2018 l__                 |NA             |
|1                                    |29,050 (62%)   |
|2                                    |8,851 (19%)    |
|3                                    |5,779 (12%)    |
|4                                    |2,647 (5.6%)   |
|5                                    |569 (1.2%)     |
|__Sample Frame Opt in__              |1,379 (2.9%)   |
|__Support__                          |NA             |
|Mean (SD)                            |5.95 (1.12)    |
|Range                                |1.00, 7.00     |
|IQR                                  |5.33, 7.00     |
|__Urban__                            |37,901 (81%)   |
: Baseline demography statistics {#tbl-table-demography}


@tbl-table-demography presents baseline demographic statistics for couples who met inclusion criteria.




##  Appendix C: Baseline and Treatment Wave Exposure Statistics {#appendix-exposures}


```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and treatment wave"
#| eval: false
#| echo: false
#| include: false
table_exposures <- here_read("table_exposures")

#table_exposures |>  kbl(format = "markdown")
```



|**Baseline/Outcome Variables** |**2018**, N = 47,202 |**2019**, N = 47,202 |
|:------------------------------|:--------------------|:--------------------|
|__Alert Level Combined__       |NA                   |NA                   |
|no_alert                       |47,202 (100%)        |24,761 (71%)         |
|early_covid                    |0 (0%)               |3,895 (11%)          |
|alert_level_1                  |0 (0%)               |3,008 (8.7%)         |
|alert_level_2                  |0 (0%)               |864 (2.5%)           |
|alert_level_2_5_3              |0 (0%)               |569 (1.6%)           |
|alert_level_4                  |0 (0%)               |1,646 (4.7%)         |
|Unknown                        |0                    |12,459               |
|__Religious Affiliation__      |17,141 (36%)         |11,807 (34%)         |
|Unknown                        |0                    |12,459               |

: Baseline and Exposure Wave Responses {#tbl-table-exposures}


@tbl-table-exposures presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: religious affiliation (yes/no). Those who did not respond to religious affiliation were coded as "0". Because the treatment wave (NZAVS time 11) occurred to New Zealand's COVID-19 pandemic, all models adjusted for the pandemic alert-level. The pandemic is a not a "confounder" because a confounder must be related to the treatment and the outcome. At the end of the study measurement, all had been exposed to the pandemic.  However, to satisfy the causal consistency assumption, all treatments must be conditionally equivalent within levels of all covariates [@vanderweele2013]. Evidence shows that dangerous natural events may affect religious affiliation [@sibley2012]. To better enable conditional independence within levels of the treatment variable, we conditioned on the lead value of COVID-alert level at baseline. To mitigate systematic biases arising from attrition, and missingness, `lmtp` package uses inverse probability of censoring weights, which were use when estimating the causal effects of the exposure on the outcome. 





### Transition Table for The Treatment

```{r}
#| label: transition-table
#| eval: false
#| echo: false

transition_table <- here_read("transition_table")
#transition_table
```

@tbl-transition shows a transition matrix that captures stability and movement between disinhibition responses from the baseline (NZAVS time 10) wave and exposure wave (NZAVS time 11) among all participants who responded to the religion question at baseline and in the following year (the censored sample). Entries on the diagonal (in bold) indicate the number of individuals who stayed in their initial state. In contrast, the off-diagonal shows the transitions from the initial state (bold) to another state the following wave (off-diagonal). A cell located at the intersection of row $i$ and column $j$, where $i \neq j$, shows the count of individuals moving from state $i$ to state $j$.


|     From      | Not Religious | Religious |
|:-------------:|:-------------:|:---------:|
| Not Religious |   **20959**   |   1183    |
|   Religious   |     1977      | **10624** |

: Transition matrix for change in treatment from baseline to the treatment wave. The diagonal shows stability in response. The off-diagonal shows the number of transitions from the initial state (bold) to another state in the following wave (off-diagonal). {#tbl-transition}





##  Appendix D: Baseline and End of Study Outcome Statistics {#appendix-outcomes}


```{r}
#| label: tbl-table-demographic_vars-code
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: false
#| echo: false
#| include: false
table_outcomes <- here_read("table_outcomes")
table_outcomes |>  kbl(format = "markdown")

```



|**Baseline/Outcome Variables** |**2018**, N = 47,202 |**2020**, N = 47,202 |
|:------------------------------|:--------------------|:--------------------|
|__Political Right Wing__       |NA                   |NA                   |
|1                              |1,819 (4.1%)         |1,293 (4.3%)         |
|2                              |6,736 (15%)          |5,309 (18%)          |
|3                              |8,456 (19%)          |6,226 (21%)          |
|4                              |16,288 (37%)         |10,033 (33%)         |
|5                              |6,812 (15%)          |4,740 (16%)          |
|6                              |3,186 (7.2%)         |2,042 (6.8%)         |
|7                              |768 (1.7%)           |480 (1.6%)           |
|Unknown                        |3,137                |17,079               |
|__Political Conservative__     |NA                   |NA                   |
|1                              |2,510 (5.6%)         |1,885 (6.2%)         |
|2                              |8,662 (19%)          |6,336 (21%)          |
|3                              |8,802 (20%)          |6,576 (21%)          |
|4                              |13,866 (31%)         |9,552 (31%)          |
|5                              |6,679 (15%)          |4,082 (13%)          |
|6                              |3,292 (7.4%)         |1,793 (5.9%)         |
|7                              |736 (1.7%)           |409 (1.3%)           |
|Unknown                        |2,655                |16,569               |


: Baseline and Outcome Wave Responses for the Outcome {#tbl-table-outcomes}

@tbl-table-outcomes presents baseline and wave 1 statistics for the outcome variables.  We see an overall shift in levels of conservativism and right-wing political orientation. Additionally, we observe 16,569 missing values. To mitigate systematic biases arising from attrition and missingness, `lmtp` package uses inverse probability of censoring weights.  






