---
title: "Causal effect of Psychopathy on Partner Well-Being: A National Longitudinal Study"
abstract: |
author: 
  - name: Aaron Hissey
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0000-0000-0000
    email: aaron.hissey@vuw.ac.nz
    corresponding: yes
  - name: Hedwig Eisenbarth
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0000-0000-0000
    email: hedwig.eisenbarth@vuw.ac.nz
    corresponding: yes
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0002-4064-8800
  - name: Matthew Hammond
    affiliation: School of Psychology, University of Auckland
    orcid_id: 0000-0000-0000-0000
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid_id: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
keywords:
  - Causal Inference
  - Relationships
  - Panel
  - Psychopathy
  - Personality
  - Well-being
  - Outcome-wide
editor_options: 
  chunk_output_type: console
format:
  pdf:
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    sanitize: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [singlecolumn]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true
#| 
# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")
library("margot")
library("glue")
library("patchwork")
library("kableExtra")



push_mods <-
  fs::path_expand(
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/lmtp/24/aaron_psychopathy"
  )

push_figs <-
  fs::path_expand(
    "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/lmtp/24/aaron_psychopathy/figs"
  )
# reproducibility
set.seed(123)

nice_exposure_name <- glue::glue("Psychopathy Facet")

nice_exposure_name_psychopathy_combined <- glue::glue("Psychopathy")
nice_exposure_name_antagonism <- glue::glue("Antagonism")
nice_exposure_name_disinhibition <- glue::glue("Disinhibition")
nice_exposure_name_emotional_stability <- glue::glue("Emotional Stability")
nice_exposure_name_narcissism<- glue::glue("Narcissism")

# load the glue package




# name_exposure_binary <- margot::here_read("name_exposure_binary")

# read exposure vars
t0_name_exposure <- here_read("t0_name_exposure")
t1_name_exposure <- here_read("t1_name_exposure")

# original df for interpretations
original_df <- margot::here_read("df_wide_encoded_ordered")


# nice labels
shift =   glue::glue('one point')
scale_exposure <- glue::glue("(1-7)")
nice_scale_exposure <- glue::glue("from 1 to 7")

exposure_lower_bound <- glue::glue("1")
exposure_upper_bound <- glue::glue("7")
# shift =   glue::glue('two points')
nice_baseline_wave <- glue::glue("NZAVS time 10, years 2018-2019")
nice_exposure_waves <- glue::glue("NZAVS time 11, years 2019-2020")
nice_outcome_wave <- glue::glue("NZAVS time 12, years 2020-2021")


# labels
# # outcome vars
# t4_outcome_vars_z <- paste0("t4_", outcome_vars, "_z")
df_lmtp <- here_read_qs("df_lmtp", push_mods)

# read health outcomes
outcome_vars <- here_read("outcome_vars")
# get partner outcomes
t2_partner_outcomes_z <- paste0("t2_partner_", outcome_vars, "_z")

# check
t2_partner_outcomes_z

# get personal outcomes
t2_outcomes_z <- paste0("t2_", outcome_vars, "_z")

# check
t2_outcomes_z <- sort(t2_outcomes_z)
# t2_outcomes_z

# label mappings for  outcomes
label_mapping_partner <- list(
  "t2_partner_conflict_in_relationship_z" = "Conflict in Relationship: Partner",
  "t2_partner_kessler_latent_anxiety_z" = "Kessler 6 Anxiety: Partner",
  "t2_partner_kessler_latent_depression_z" = "Kessler 6 Depression: Partner",
  "t2_partner_kessler6_sum_z" = "Kessler 6 Distress: Partner",
  "t2_partner_lifesat_z" = "Life Satisfaction: Partner",
  "t2_partner_pwi_z" = "Personal Well-Being Index: Partner",
  "t2_partner_sat_relationship_z" = "Satisfaction with Relationship: Partner",
  "t2_partner_self_esteem_z" = "Self Esteem: Partner"
)

label_mapping_self <- list(
  "t2_conflict_in_relationship_z" = "Conflict in Relationship: Self",
  "t2_kessler_latent_anxiety_z" = "Kessler 6 Anxiety: Self",
  "t2_kessler_latent_depression_z" = "Kessler 6 Depression: Self",
  "t2_kessler6_sum_z" = "Kessler 6 Distress: Self",
  "t2_lifesat_z" = "Life Satisfaction: Self",
  "t2_pwi_z" = "Personal Well-Being Index: Self",
  "t2_sat_relationship_z" = "Satisfaction with Relationship: Self",
  "t2_self_esteem_z" = "Self Esteem: Self"
)



treatment_contrast <- glue::glue("What would be the average responses the population of partners in New Zealand couples increased + {shift} in a {nice_exposure_name} up to a maximum of {exposure_upper_bound} ({nice_scale_exposure}), compared with a -{shift} in the same facet?")


treatment_loss_null <- glue::glue("-{shift} compared with observed levels of intervention?")

treatment_gain_null <- glue::glue("+{shift} compared with observed levels of intervention?")

# create the titles with automatic exposure insertion
title <- glue("Shift {shift} up vs. Shift {shift} point down (baseline)")


baseline_wave <- margot::here_read("baseline_wave")
exposure_waves <- margot::here_read("exposure_waves")
exposure_wave <- margot::here_read("exposure_waves")
outcome_wave <- margot::here_read("outcome_wave")

# get table/graphs
summary_tables <- margot::here_read_qs("summary_tables", push_mods)

# n_this study
n_participants <- summary_tables$n_participants

# percentage missing at baseline
# percent_missing_baseline <- here_read("percent_missing_baseline")
# percentage_missing_baseline <- glue::glue(round(percent_missing_baseline,2))


# nice labels
scale_exposure <- glue::glue('1-7')
nice_scale_exposure =  glue::glue('ordinal scale 1-7')
exposure_lower_bound =  glue::glue('1')
exposure_upper_bound =  glue::glue('7')

nice_baseline_wave = glue::glue("NZAVS time 12, years 2020-2021")
nice_exposure_wave =  glue::glue("NZAVS time 13, years 2021-2022")
nice_outcome_wave =  glue::glue("NZAVS time 14, years 2022-2023")


## correlation not causation
# load required libraries
library(ggplot2)
library(ggokabeito)

# plot options
# plot options ------------------------------------------------------------
options_base_partner <- list(
  type = "RD",
  title = title,
  order = "default",
  colors = c(
    "positive" = "#E69F00",
    "not reliable" = "grey50",
    "negative" = "#56B4E9"
  ),
  x_offset = -1, # will be set based on type
  x_lim_lo = -1, # will be set based on type
  x_lim_hi = .5, # will be set based on type
  text_size = 5,
  linewidth = 0.5,
  estimate_scale = 1,
  base_size = 18,
  point_size = 4,
  title_size = 18,
  subtitle_size = 18,
  legend_text_size = 10,
  legend_title_size = 10,
  label_mapping_partner = label_mapping_partner
)


options_base_self <- list(
  type = "RD",
  title = title,
  order = "default",
  colors = c(
    "positive" = "#E69F00",
    "not reliable" = "grey50",
    "negative" = "#56B4E9"
  ),
  x_offset = -1, # will be set based on type
  x_lim_lo = -1, # will be set based on type
  x_lim_hi = .5, # will be set based on type
  text_size = 5,
  linewidth = 0.5,
  estimate_scale = 1,
  base_size = 18,
  point_size = 4,
  title_size = 18,
  subtitle_size = 18,
  legend_text_size = 10,
  legend_title_size = 10,
  label_mapping_partner = label_mapping_self)

options_antagonism_partner<- margot_plot_create_options(
  subtitle = "Antagonism Effect on Partner",
  base_defaults = options_base_partner,
  filename_prefix = "antagonism_partner"
)


options_antagonism_self<- margot_plot_create_options(
  subtitle = "Antagonism Effect on Self",
  base_defaults = options_base_self,
  filename_prefix = "antagonism_self"
)

options_disinhibition_partner<- margot_plot_create_options(
  subtitle = "Disinhibition Effect on Partner",
  base_defaults = options_base_partner,
  filename_prefix = "disinhibition_partner"
)

options_disinhibition_self<- margot_plot_create_options(
  subtitle = "Disinhibition Effect on Self",
  base_defaults = options_base_self,
  filename_prefix = "disinhibition_self"
)


options_emotional_stability_partner<- margot_plot_create_options(
  subtitle = "Emotional Stability Effect on Partner",
  base_defaults = options_base_partner,
  filename_prefix = "emotional_stability_partner"
)

options_emotional_stability_self<- margot_plot_create_options(
  subtitle = "Emotional Stability Effect on Self",
  base_defaults = options_base_self,
  filename_prefix = "emotional_stability_self"
)


options_narcissism_partner<- margot_plot_create_options(
  subtitle = "Narcissism Effect on Partner",
  base_defaults = options_base_partner,
  filename_prefix = "narcissism_partner"
)

options_narcissism_self<- margot_plot_create_options(
  subtitle = "Narcissism Effect on Self",
  base_defaults = options_base_self,
  filename_prefix = "narcissism_self"
)

options_psychopathy_combined_partner<- margot_plot_create_options(
  subtitle = "Psychopathy Combined Effect on Partner",
  base_defaults = options_base_partner,
  filename_prefix = "psychopathy_combined_partner"
)

options_psychopathy_combined_self<- margot_plot_create_options(
  subtitle = "Psychopathy Combined Effect on Self",
  base_defaults = options_base_self,
  filename_prefix = "psychopathy_combined_self"
)


# partner options
# plot options
options_antagonism_partner_osf<- margot_plot_create_options(
  subtitle = "Antagonism Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "antagonism_partner_osf"
)


options_disinhibition_partner_osf<- margot_plot_create_options(
  subtitle = "Disinhibition Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "disinhibition_partner_osf"
)




# lmtp models -------------------------------------------------------------
psychopathy_combined_partner_lmtp_output <- here_read_qs('psychopathy_combined_partner_lmtp_output', push_mods)
psychopathy_combined_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_shift_up


plot_psychopathy_combined_partner_lmtp_output <- margot_plot(
  psychopathy_combined_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_shift_up,
  options = options_psychopathy_combined_partner,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)

push_mods
# margot_save_png(
#   plot_psychopathy_combined_partner_lmtp_output$plot,
#   save_path = here::here(push_figs),
#   width = 12,
#   height = 12,
#   base_filename = "plot_psychopathy_combined_partner"
# )

# antagonism partner
antagonism_partner_lmtp_output <- here_read_qs('antagonism_partner_lmtp_output', push_mods)


plot_antagonism_partner_lmtp_output <- margot_plot(
  antagonism_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_antagonism_partner,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)
cat(plot_antagonism_partner_lmtp_output$interpretation)
# save file
# margot::margot_save_png(
#   plot_antagonism_partner_lmtp_output$plot,
#   width = 12,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "plot_antagonism_partner"
# )



# disinhibition partner
disinhibition_partner_lmtp_output <- here_read_qs('disinhibition_partner_lmtp_output', push_mods)


plot_disinhibition_partner_lmtp_output <- margot_plot(
  disinhibition_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_disinhibition_partner,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
  
)
plot_disinhibition_partner_lmtp_output$interpretation

# margot::margot_save_png(
#   plot_disinhibition_partner_lmtp_output$plot,
#   width = 12,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "plot_disinhibition_partner"
# )

# emotional stability partner
emotional_stability_partner_lmtp_output <- here_read_qs('emotional_stability_partner_lmtp_output', push_mods)

plot_emotional_stability_partner_lmtp_output <- margot_plot(
  emotional_stability_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_emotional_stability_partner,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
  )
plot_emotional_stability_partner_lmtp_output$plot

# margot::margot_save_png(
#   plot_emotional_stability_partner_lmtp_output$plot,
#   width = 12,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "plot_emotional_stability_partner"
# )

# narcissism
narcissism_partner_lmtp_output <- here_read_qs('narcissism_partner_lmtp_output', push_mods)

plot_narcissism_partner_lmtp_output <- margot_plot(
  narcissism_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_narcissism_partner,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)


# margot::margot_save_png(
#   plot_narcissism_partner_lmtp_output$plot,
#   width = 12,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = " plot_narcissism_partner"
# )


# self --------------------------------------------------------------------

psychopathy_combined_self_lmtp_output <- here_read_qs('psychopathy_combined_self_lmtp_output', push_mods)

plot_psychopathy_combined_self_lmtp_output <- margot_plot(
  psychopathy_combined_self_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_shift_up,
  options = options_psychopathy_combined_self,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_self

)

compare_partner_self_plot_psychopathy_combined <-plot_psychopathy_combined_self_lmtp_output$plot + plot_psychopathy_combined_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

compare_partner_self_plot_psychopathy_combined

# margot::margot_save_png(
#   compare_partner_self_plot_psychopathy_combined,
#   width = 24,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "compare_parter_self_plot_psychopathy_combined"
# )


# antagonism  self

antagonism_self_lmtp_output <- here_read_qs('antagonism_self_lmtp_output', push_mods)
antagonism_self_lmtp_output$combined_tables
  
  
plot_antagonism_self_lmtp_output <- margot_plot(
  antagonism_self_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_antagonism_self,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_self
)

compare_partner_self_plot_antagonism <-
plot_antagonism_self_lmtp_output$plot + plot_antagonism_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

compare_partner_self_plot_antagonism
# margot::margot_save_png(
#   compare_partner_self_plot_antagonism,
#   width = 24,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "compare_partner_self_plot_antagonism"
# )

# disinhibition  self


disinhibition_self_lmtp_output <- here_read_qs('disinhibition_self_lmtp_output', push_mods)

plot_disinhibition_self_lmtp_output <- margot_plot(
  disinhibition_self_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_disinhibition_self,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_self
)


compare_partner_self_plot_disinhibition <- plot_disinhibition_self_lmtp_output$plot+ plot_disinhibition_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

# margot::margot_save_png(
#   compare_partner_self_plot_disinhibition,
#   width = 24,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "compare_partner_self_plot_disinhibition"
# )
# 


# emotional stability self

emotional_stability_self_lmtp_output <- here_read_qs('emotional_stability_self_lmtp_output', push_mods)

plot_emotional_stability_self_lmtp_output <- margot_plot(
  emotional_stability_self_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_emotional_stability_self,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_self
)



compare_partner_self_plot_emotional_stability <- plot_emotional_stability_self_lmtp_output$plot + plot_emotional_stability_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

# margot::margot_save_png(
#   compare_partner_self_plot_emotional_stability,
#   width = 24,
#   height = 12,
#   save_path = here::here(push_figs),
#   base_filename = "compare_partner_self_plot_emotional_stability"
# )




# narcissism self
narcissism_self_lmtp_output <- here_read_qs('narcissism_self_lmtp_output', push_mods)

plot_narcissism_self_lmtp_output <- margot_plot(
  narcissism_self_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_narcissism_self,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_self
)

plot_narcissism_self_lmtp_output
compare_partner_self_plot_narcissism <- plot_narcissism_self_lmtp_output$plot  + plot_narcissism_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

compare_partner_self_plot_narcissism
margot::margot_save_png(
  compare_partner_self_plot_narcissism,
  width = 24,
  height = 12,
  save_path = here::here(push_figs),
  base_filename = "compare_partner_self_plot_narcissism"
)






# models OSF original SPEC
# plot options
options_antagonism_partner_osf<- margot_plot_create_options(
  subtitle = "Antagonism Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "antagonism_partner_osf"
)




options_disinhibition_partner_osf<- margot_plot_create_options(
  subtitle = "Disinhibition Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "disinhibition_partner_osf"
)



options_emotional_stability_partner_osf <- margot_plot_create_options(
  subtitle = "Emotional Stability Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "emotional_stability_partner_osf"
)




options_narcissism_partner_osf <- margot_plot_create_options(
  subtitle = "Narcissism Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "narcissism_partner_osf"
)



options_psychopathy_combined_partner_osf<- margot_plot_create_options(
  subtitle = "Psychopathy Combined Effect on Partner (OSF SPEC)",
  base_defaults = options_base_partner,
  filename_prefix = "psychopathy_combined_partner_osf"
)


psychopathy_combined_partner_osf_lmtp_output <- here_read_qs('psychopathy_combined_partner_osf_lmtp_output', push_mods)


plot_psychopathy_combined_partner_osf_lmtp_output <- margot_plot(
  psychopathy_combined_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_shift_up,
  options = options_psychopathy_combined_partner_osf,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)
plot_psychopathy_combined_partner_osf_lmtp_output$plot/plot_psychopathy_combined_partner_lmtp_output$plot



antagonism_partner_osf_lmtp_output <- here_read_qs('antagonism_partner_osf_lmtp_output', push_mods)

plot_antagonism_partner_osf_lmtp_output <- margot_plot(
  antagonism_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_antagonism_partner_osf,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)
plot_antagonism_partner_osf_lmtp_output$plot + plot_antagonism_partner_lmtp_output$plot



disinhibition_partner_osf_lmtp_output <- here_read_qs('disinhibition_partner_osf_lmtp_output', push_mods)


plot_disinhibition_partner_osf_lmtp_output<- margot_plot(
  disinhibition_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_disinhibition_partner_osf,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
  
)
plot_disinhibition_partner_osf_lmtp_output$plot  + plot_disinhibition_partner_lmtp_output$plot


emotional_stability_partner_osf_lmtp_output <- here_read_qs('emotional_stability_partner_osf_lmtp_output', push_mods)

plot_emotional_stability_partner_osf_lmtp_output<- margot_plot(
  emotional_stability_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_emotional_stability_partner_osf,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)
plot_emotional_stability_partner_osf_lmtp_output$plot + plot_emotional_stability_partner_lmtp_output$plot


narcissism_partner_osf_lmtp_output <- here_read_qs('narcissism_partner_osf_lmtp_output', push_mods)



plot_narcissism_partner_osf_lmtp_output<- margot_plot(
  narcissism_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_shift_down,
  options = options_narcissism_partner_osf,
  save_output = TRUE,
  save_path = here::here(push_mods),
  original_df = original_df,
  label_mapping = label_mapping_partner
)
plot_narcissism_partner_osf_lmtp_output$plot+ plot_narcissism_partner_lmtp_output$plot

n_participants
n_couples = 1012 / 2

```

## Introduction

A central question in psychopathy research is how it affects relationships. 

To address this question, we develop an outcome-wide longitudinal study using a national longitudinal probability sample in New Zealand. 


## Method

### Sample

Data were collected as part of The New Zealand Attitudes and Values Study (NZAVS) is an annual longitudinal national probability panel study of social attitudes, personality, ideology and health outcomes. The NZAVS began in 2009. It includes questionnaire responses from over 72,000 New Zealand residents. The study includes researchers from many New Zealand universities, including the University of Auckland, Victoria University of Wellington, the University of Canterbury, the University of Otago, and Waikato University. Because the survey asks the same people to respond each year, it can track subtle changes in attitudes and values over time and is an important resource for researchers in New Zealand and worldwide. The NZAVS is university-based, not-for-profit and independent of political or corporate funding.https://doi.org/10.17605/OSF.IO/75SNB


[Appendix A](#appendix-demographics) provides information about demographic covariates used for confounding control (NZAVS time 10, years 2018-2019).


### Eligibility criteria 

The sample consisted of respondents to NZAVS times 10 (baseline, years 2018-2019), time 11 (treatment wave, years 2019-2020), and time 12 (outcome wave) (years 2020-2021).


- Participants who were identified as being in couples.   
- Participants who provided full information to the psychopathy measures in the `r baseline_wave` and `r exposure_wave`. Participants may have been lost to follow up in the outcome wave `r outcome_wave`. 
- Missing data for all variables at baseline were allowed. Missing data at baseline were imputed through the `mice` package [@vanbuuren2018], and an extra column indicator was included to denote whether a variable was missing, following protocols described by [@williams2021].
- Inverse probability of censoring weights were calculated as part of estimation in `lmtp` to adjust for missing outcomes at NZAVS Time 12 (years 2020-2021, the outcome wave [@williams2021])

There were `r n_participants` NZAVS participants who met these criteria in `r n_couples` couples.


### Causal Contrast

Here, we leverage panel data to estimate potential outcomes in a hypothetical world in which partners in couples were shifted up or down on a one-point scale for psychopathic personality.

Our causal estimand takes a "shift function" or "modified treatment policy":

$$ \text{Average Treatment Effect} = E\{(E[Y(\text{Shift up})|A,L) - (E[Y(\text{Shift down})|A,L)\} $$

**Intervention Functionals:**

* **Shift up by +1 point:**


$$
\text{Shift up} \equiv f(A) = a^* = 
\begin{cases}
  A + 1 & \text{if the observed value of the exposure + 1 < 7 point, else:} \\
  7 &
\end{cases}
$$

* **Shift down by -1 point:**

$$
\text{Shift down} \equiv f(A) = \tilde{a} =  
\begin{cases}
  A - 1 & \text{if the the observed value of the exposure - 1 > 1, else:} \\
  1 &
\end{cases}
$$


* **Null**

$$
\text{Null} \equiv f(A) = a =  
\begin{cases}
  A & \text{the observed value of the exposure} \\
\end{cases}
$$

Our primary interest is in the contrast of these intervention functionals.  {Appendix XX](#appendix-XX) reports contrast for each intervention against the NULL of no intervention.



### Identification Assumptions

To consistently estimate causal effects, we rely on three key assumptions:

1. **Causal Consistency:** potential outcomes must align with observed outcomes under the treatments in our data. Essentially, we assume potential outcomes do not depend on the specific way treatment was administered, as long as we consider measured covariates. 

2. **Conditional Exchangeability:** given the observed covariates, we assume treatment assignment is independent of potential outcomes. In simpler terms, this means "no unmeasured confounding" – any factors influencing both treatment assignment and outcomes must be included in our measured covariates.

3. **Positivity:** for unbiased estimation, every subject must have a non-zero chance of receiving the treatment, regardless of their covariate values. We evaluate this assumption in each study by examining changes in psychopathy "treatments" from baseline (NZAVS time 10) to the treatment wave (NZAVS time 11). It is the initiation of a psychopathic trait and their combinations, whose causal effect we seek to quantify



### Analysis

Criteria for analysis were pre-specified and pre-registered in advance of the analysis at:[https://osf.io/ce4t9/](https://osf.io/ce4t9/), and followed standard NZAVS protocols for three-wave causal inference designs [@bulbulia2024PRACTICAL].


### Confounding Control Strategy

We followed @vanderweele2020's *minimally modified disjunctive criteria* for confounding control.

1. **Initial identify confounders**: using causal diagrams, we began by enumerating all covariates that may influence either the treatment (exposure) or outcomes, spanning five domains. This includes variables directly affecting the exposure or outcome, as well as potential consequences of these variables (i.e. proxies.)

2. **Remove instrumental variables**: we removed variables that are identified as instrumental variables, i.e., those influencing the exposure but not the outcome. Their inclusion can reduce the efficiency of the analysis.

3. **Inclusion of proxy variables**: for unmeasured variables that affect both the exposure and outcome, include proxy variables wherever possible. These proxies act as indicators for the unmeasured common causes. [Appendix B](#appendix-demographics) lists covariates we used for confounding control. These protocols follow advice in [@bulbulia2024PRACTICAL] as prespecified in [https://osf.io/ce4t9/](https://osf.io/ce4t9/).

4. **Baseline exposure control**: we included baseline measures of the outcome to adjust for unmeasured confounding [@vanderweele2020]. Accounting for prior exposure also allowed us to recover an incident exposure effect as opposed to a prevalent exposure effect: that is, we were able to focus on the one-year effect on partner well-being from initiation to psychopathy or a psychopathic trait [@danaei2012; @hernan2023].

5. **Baseline outcome control**: we also included baseline measures of the outcome to adjust for unmeasured confounding [@vanderweele2020]. This strategy also minimises the possibility of reverse causation.


Additionally, we adopted the following confounding-control strategies:


6. **Partner's covariates**: because the effect of changes in one individual's variables on their partner's outcomes is assessed, we include baseline measures for the partner's covariates, as we as of the focal partner. Additionally, we clustered individuals with their dyadic partnerships, which helps adjust for unmeasured features that might affect results.


<!-- Valid inferences for the New Zealand population were secured by applying post-stratification census weights for age, European ethnicity, and gender. -->



#### Estimator 

We employ a semi-parametric estimator known as Targeted Minimum Loss-based Estimation (TMLE), which is able to estimate the causal effect of modified treatment policies on outcomes over time [@vanderlaan2011; @vanderlaan2018]. Estimation was performed using `lmtp` package [@williams2021; @díaz2021; @hoffman2023]. TMLE is a robust method that combines machine learning techniques with traditional statistical models to estimate causal effects while providing valid statistical uncertainty measures for these estimates.

TMLE operates through a two-step process involving outcome and treatment (exposure) models. Initially, it employes machine learning algorithms to flexibly model the relationship between treatments, covariates, and outcomes. This flexibility allows TMLE to account for complex, high-dimensional covariate spaces without imposing restrictive model assumptions. The outcome of this step is a set of initial estimates for these relationships.

The second step of TMLE involves "targeting" these initial estimates by incorporating information about the observed data distribution to improve the accuracy of the causal effect estimate. This is achieved through an iterative updating process, which adjusts the initial estimates towards the true causal effect. This updating process is guided by the efficient influence function, ensuring that the final TMLE estimate is as close as possible to the true causal effect while remaining robust to model misspecification in either the outcome or treatment model.

A central feature of TMLE is its double-robustness property, meaning that if either the model for the treatment or the outcome is correctly specified, the TMLE estimator will still consistently estimate the causal effect. Additionally, TMLE uses cross-validation to avoid overfitting and ensure that the estimator performs well in finite samples. Each of these steps contributes to a robust methodology for examining the *causal* effects on of interventions on outcomes. The marriage of TMLE and machine learning technologies reduces the dependence on restrictive modelling assumptions and introduces an additional layer of robustness. For further details see [@hoffman2022; @hoffman2023; @díaz2021]


#### Estimation

We used the `lmtp` R package to estimate treatments' causal effects on well-being outcomes for each of the four trait dimensions and the composite measure. Measurements of the treatments and outcomes were included with baseline covariates. Our models included an indicator for censorship in the next wave (C) which adjusted for loss to follow-up/attrition conditional on measured covariates. As mentioned, models estimated the effects of shifting the treatment variable by one unit (both increase and decrease) using custom shift functions (f and f_1), which adjust the treatment variable within the observed range. Two custom shift functions were defined to model the hypothetical intervention of increasing or decreasing the treatment variable by one unit, and these were contrasted against the expectation from a 'null' model.

- **increase +1 function**: increases the treatment variable by one unit, capped at the maximum observed score.
- **decrease -1 function**: decreases the treatment variable by one unit, floored at the minimum observed score.
- **'null' model:** the null model was specified identically to the treatment effect models in terms of the included covariates, outcome variable, and model structure but without applying any shift to the treatment variable. This ensured that any differences observed between the treatment models and the null model can be attributed to the treatment effect rather than differences in model specification. By estimating a model under the assumption of no change in the treatment status, we obtained clear baselines against which the effects of treatment shifts could be assessed.

`lmtp` draws on the the `SuperLearner` library, which comprises various machine learning algorithms [@SuperLearner2023]. Given the relatively sample (N=1070) we used the `Ranger` and `randomForest` estimators - both causal forest estimators --  a subset of non-parametric estimators distinguished by their resilience against overfitting [Ranger2017; randomForest2002].  Causal forests excel in identifying complex, non-linear relationships between variables without presupposing a specific model form, and thus making fewer assumptions about the underlying data distribution.


#### Cross-Validation

We implemented a 10-fold cross-validation. This method partitions the data into ten subsets of approximately equal size. During the cross-validation process, nine subsets are used to train the model, and the remaining subset is used for testing. This process is repeated ten times, with each of the ten subsets used exactly once as the test set. Using different subsets for training and validation, minimises the risk of the model being overly complex and fitting the noise in the training dataset, which can lead to poor performance on new data. Moreover, each observation is used for training and validation, which is particularly beneficial in scenarios where the amount of data is limited. The cross-validation process results in ten estimates of model accuracy, which can be averaged to provide a more comprehensive measure of model performance.


#### Sensitiviy Analysis Using the E-value

To measure senstivity to unmeasured confounding, we report VanderWeele and Ding's "E-value" for all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain-away the observed exposure-outcome association  [@vanderweele2020; @linden2020EVALUE]. For example, if the E-value were 1.3, it would mean an unmeasured confounder associated with both the treatment and outcome by a risk ratio of 1.3 each (or 30% increase in risk) could explain away the observed effect. Weaker confounding would not suffice. We also report the bound of the E-value closest to 1. If the lower bound of the CI were 1.1, to explain away the result, the strength of an unmeasured confounder in its association with both the treatment and the outcome would need to be at least 1.1 on the risk ratio scale (or a 10% increase in risk) to explain away the result [@vanderweele2017]. 


### Deviations from pre-registered research

Our current research deviates from our preregistered plan in three areas, all aimed at enhancing the robustness and validity of our findings.

First, we adjusted our primary analysis to compare the average treatment effects of increasing psychopathic personality traits by one point for each individual (without exceeding the scale’s maximum) versus decreasing these traits by one point. This adjustment allows for greater separation between exposure conditions, closely emulating an experimental design and providing clearer insights into the effects of personality trait variations. Although it is possible to estimate numerous causal contrasts, comparing gains or losses with the null hypothesis of no change would yield contrasts that are not directly comparable. For completeness, all contrasts, including those from our original plan, are reported in Appendix XXX.

Second, we incorporated an additional baseline confounder: reported health disability. The decision is based on the assumption that health disability may be a common cause influencing both personality traits and well-being outcomes.

Third, we introduced indicators to denote when baseline variables had been imputed. This methodological refinement aligns our approach with the recommendations of @hoffman2023, ensuring transparency and robustness in handling missing data. By accounting for imputed data, we mitigate potential biases and strengthen the reliability of our results.

The adjustments made in the second and third deviations resulted in a slight attenuation of the causal effect estimates, which was anticipated. Importantly, these changes did not alter our substantive conclusions, yet support credibility for our findings. Comprehensive reporting for the original preregistered study is provided in Appendix XXX.

## Results Study 1: Antagonism  


### Results Study 1A: Antagonism: Partner

```{r}
#| label: fig-results-antagonism-partner
#| fig-cap: "Results for antagonism effect on partner multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16


plot_antagonism_partner_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-antagonism-partner
#| tbl-cap: "Table for antagonism effect on partner multi-dimensional well-being"
#| eval: true
#| include: false
#| echo: false

plot_antagonism_partner_lmtp_output$transformed_table |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_antagonism_partner_lmtp_output$interpretation)
```



### Results Study 1B: Antagonism: Self vs Partner 

```{r}
#| label: fig-results-antagonism-self
#| fig-cap: "Results for antagonism effect for self vs partner on multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_antagonism_self_lmtp_output$plot + plot_antagonism_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-antagonism-self
#| tbl-cap: "Table for antagonism effect on for self on multi-dimensional well-being"
#| eval: true
#| include: false
#| echo: false

plot_antagonism_self_lmtp_output$transformed_table |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_antagonism_self_lmtp_output$interpretation)
```


{{< pagebreak >}}
## Results Study 2 Disinhibition 

### Results Study 2A Disinhibition: Partner

```{r}
#| label: fig-results-disinhibition-partner
#| fig-cap: "Results for disinhibition effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12


plot_disinhibition_partner_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-disinhibition-partner
#| tbl-cap: "Table for disinhibition effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_disinhibition_partner_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_disinhibition_partner_lmtp_output$interpretation)

```


{{< pagebreak >}}
### Results Study 2B Disinhibition: Self vs Partner

```{r}
#| label: fig-results-disinhibition-self
#| fig-cap: "Results for disinhibition effect for self vs partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24


plot_disinhibition_self_lmtp_output$plot + plot_disinhibition_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-disinhibition-self
#| tbl-cap: "Table for disinhibition effect for self on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_disinhibition_self_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_disinhibition_self_lmtp_output$interpretation)

```


{{< pagebreak >}}
## Results Study 3 Emotional Stability 


### Results Study 3A  Emotional Stability: Partner

```{r}
#| label: fig-results-emotional-partner
#| fig-cap: "Results for emotional stability effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12

plot_emotional_stability_partner_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-emotional-partner
#| tbl-cap: "Table for emotional stability effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_emotional_stability_partner_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_emotional_stability_partner_lmtp_output$interpretation)

```

{{< pagebreak >}}

### Results Study 3B  Emotional Stability: Self vs Partner

```{r}
#| label: fig-results-emotional-self
#| fig-cap: "Results for emotional stability effect for self vs partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24


plot_emotional_stability_self_lmtp_output$plot + plot_emotional_stability_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-emotional-self
#| tbl-cap: "Table for emotional stability effect for self on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_emotional_stability_self_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_emotional_stability_self_lmtp_output$interpretation)

```

{{< pagebreak >}}




## Results Study 4 Narcissism 



### Results Study 4A Narcissism: Partner

```{r}
#| label: fig-results-narcissism-partner
#| fig-cap: "Results for Narcissism stability gain for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12
plot_narcissism_partner_lmtp_output$plot
```

{{< pagebreak >}}

```{r}
#| label: tbl-results-narcissism-partner
#| tbl-cap: "Table for Narcissism gain for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_narcissism_partner_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_narcissism_partner_lmtp_output$interpretation)


```
{{< pagebreak >}}


### Results Study 4B Narcissism: Self vs Partner

```{r}
#| label: fig-results-narcissism-self
#| fig-cap: "Results for Narcissism stability effect for self vs partner on multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_narcissism_self_lmtp_output$plot + plot_narcissism_partner_lmtp_output$plot   + plot_annotation(tag_levels = "A")
```

{{< pagebreak >}}

```{r}
#| label: tbl-results-narcissism-self
#| tbl-cap: "Table for Narcissism effect for self on multi-dimensional well-being"
#| eval: true
#| include: true
#| echo: false

plot_narcissism_self_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_narcissism_self_lmtp_output$interpretation)
```
{{< pagebreak >}}

## Results Study 5: Psychopathy Combined Score


### Results Study 5A: Psychopathy Combined Score: Partner
```{r}
#| label: fig-results-psychopathy-partner
#| fig-cap: "Results forPsychopathy Combined Score effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12

plot_psychopathy_combined_partner_lmtp_output$plot


```

{{< pagebreak >}}

```{r}
#| label: tbl-results-psychopathy-partner
#| tbl-cap: "Table for Psychopathy Combined Score effect for partner on multi-dimensional well-being."
#| include: true
#| echo: false

plot_psychopathy_combined_partner_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_psychopathy_combined_partner_lmtp_output$interpretation)
```


### Results Study 5B  Psychopathy Combined Score: Self vs Partner

```{r}
#| label: fig-results-psychopathy-self
#| fig-cap: "Results for  Psychopathy Combined Score effect for self vs partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_psychopathy_combined_self_lmtp_output$plot + plot_psychopathy_combined_partner_lmtp_output$plot   + plot_annotation(tag_levels = "A")
```

{{< pagebreak >}}

```{r}
#| label: tbl-results-psychopathy-self
#| tbl-cap: "Table for Psychopathy Combined Score effect on self multi-dimensional well-being"
#| eval: true
#| include: true
#| echo: false

plot_narcissism_self_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_psychopathy_combined_self_lmtp_output$interpretation)
```

## Discussion 

### Antagonism 

Antagonism affects partners and the self in different ways. For the **partner**, the only reliable result is a reduction in life satisfaction ($-0.10$, 95% CI: $[-0.17, -0.04]$).

For the **self**, antagonism reliably increases life satisfaction ($0.10$, 95% CI: $[0.04, 0.15]$), personal well-being ($0.13$, 95% CI: $[0.07, 0.19]$), satisfaction with the relationship ($0.11$, 95% CI: $[0.05, 0.17]$), and self-esteem ($0.08$, 95% CI: $[0.02, 0.13]$).

### Disinhibition

For the **partner**, none of the results are reliable as all confidence intervals cross zero.

For the **self**, disinhibition reliably reduces anxiety ($-0.10$, 95% CI: $[-0.15, -0.05]$) and distress ($-0.05$, 95% CI: $[-0.10, -0.01]$). There is also a marginally reliable improvement in personal well-being ($0.05$, 95% CI: $[0.00, 0.11]$) and self-esteem ($0.05$, 95% CI: $[0.00, 0.10]$).


### Emotional Stability

For the **partner**, none of the results are reliable as all confidence intervals cross zero.

For the **self**, emotional stability reliably reduces anxiety ($-0.10$, 95% CI: $[-0.15, -0.05]$) and distress ($-0.06$, 95% CI: $[-0.10, -0.01]$). There is also a marginally reliable improvement in personal well-being ($0.05$, 95% CI: $[0.00, 0.11]$) and self-esteem ($0.05$, 95% CI: $[0.01, 0.10]$).


### Narcicissm: 

For **partners**, narcissism reliably increases personal well-being ($0.07$, 95% CI: $[0.02, 0.12]$) and satisfaction with the relationship ($0.08$, 95% CI: $[0.01, 0.14]$). No other reliable effects are observed.

For the **self**, narcissism reliably reduces anxiety ($-0.08$, 95% CI: $[-0.14, -0.03]$), depression ($-0.10$, 95% CI: $[-0.16, -0.04]$), and distress ($-0.10$, 95% CI: $[-0.15, -0.04]$). It also reliably improves life satisfaction ($0.12$, 95% CI: $[0.06, 0.18]$) and self-esteem ($0.11$, 95% CI: $[0.06, 0.17]$).



### Psychopathy Combined 

For the **partner**, psychopathy reliably increases anxiety ($0.07$, 95% CI: $[0.03, 0.11]$) and distress ($0.05$, 95% CI: $[0.01, 0.09]$). It also reliably decreases personal well-being ($-0.11$, 95% CI: $[-0.14, -0.07]$), satisfaction with the relationship ($-0.06$, 95% CI: $[-0.10, -0.03]$), and self-esteem ($-0.09$, 95% CI: $[-0.13, -0.06]$).

For the **self**, psychopathy reliably reduces anxiety ($-0.08$, 95% CI: $[-0.14, -0.03]$), depression ($-0.10$, 95% CI: $[-0.16, -0.04]$), and distress ($-0.10$, 95% CI: $[-0.15, -0.04]$). It also reliably improves life satisfaction ($0.12$, 95% CI: $[0.06, 0.18]$) and self-esteem ($0.11$, 95% CI: $[0.06, 0.17]$).


### Importance of Research

These findings reveal that focussing on specific traits of psychopathic personality affects people differently than does psychopathy as a whole.  When we target individual traits such as antagonism or narcissism, people often feel better about themselves -- they report higher life satisfaction, well-being, and self-esteem, and experience reduced depression and anxiety. However, these personal improvements can come at a cost to their partners, who may suffer from decreased life satisfaction and increased distress. This indicates a complex interplay between self-improvement and relational harm.

On the other hand, when psychopathy is treated as a single, combined dimension, partners tend to suffer more. They experience more anxiety and distress, along with lower personal well-being and satisfaction in the relationship. Although, individuals for whom psychopathy increases might feel less depressed, they do not experience an overall improvement in well-being. As for partners, self-esteem, relationship satisfaction, and personal well being declines.  Moreover, life satisifaction declines and anxiety increases. There is a darker effects of psychopathy in couples are not restricted to the partners, but also extend to those in whom psychopathy personality is manifest. 

{{< pagebreak >}}

### Ethics

The NZAVS is reviewed every three years by the University of Auckland Human Participants Ethics Committee. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.

### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the TempletoReligion Trust (TRT0196; TRT0418). JB received support from the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or the decision to publish.

### Author Statement

TBA... e.g. 


- AH conceived of the study, led the validation study, and developed the theory 
- HE ...  MH ...contributed to the theory.
- CS led data collection. 
- JB developed the inferential approach and did the analysis.  
- All authors contributed to writing the final version of the manuscript, which was substantially AH’s work.

{{< pagebreak >}}

## References

::: {#refs}
:::


{{< pagebreak >}}

## Appendix A: Measurues {#appendix-measures}

#### Age (waves: 1-15)

We asked participants' age in an open-ended question ("What is your age?" or "What is your date of birth").

<!-- #### Disability (waves: 5-15) -->

<!-- We assessed disability with a one item indicator adapted from @verbrugge1997, that asks "Do you have a health condition or disability that limits you, and that has lasted for 6+ months?" (1 = Yes, 0 = No). -->

#### Disability


We assessed disability with a one-item indicator adapted from @verbrugge1997. It asks, "Do you have a health condition or disability that limits you and that has lasted for 6+ months?" (1 = Yes, 0 = No).


#### Education Attainment (waves: 1, 4-15)

Participants were asked "What is your highest level of qualification?". We coded participans highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school quals coded as Level 3, and all other ancillary categories coded as missing) See:https://www.nzqa.govt.nz/assets/Studying-in-NZ/New-Zealand-Qualification-Framework/requirements-nzqf.pdf

<!-- #### Employment (waves: 1-3, 4-11) -->

<!-- We asked participants "Are you currently employed? (This includes self-employed or casual work)". \* note: This question disappeared in the updated NZAVS Technical documents (Data Dictionary). -->

<!-- #### European (waves: 1-15) -->

<!-- Participants were asked "Which ethnic group do you belong to (NZ census question)?" or "Which ethnic group(s) do you belong to? (Open-ended)" (wave: 3). Europeans were coded as 1, whereas other ethnicities were coded as 0. -->

#### Ethnicity (waves: 3)

Based on the New Zealand Census, we asked participants "Which ethnic group(s) do you belong to?". The responses were: (1) New Zealand European; (2) Māori; (3) Samoan; (4) Cook Island Māori; (5) Tongan; (6) Niuean; (7) Chinese; (8) Indian; (9) Other such as DUTCH, JAPANESE, TOKELAUAN. Please state:. We coded their answers into four groups: Maori, Pacific, Asian, and Euro (except for Time 3, which used an open-ended measure).

#### Gender (waves: 1-15)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male was coded as 1, and gender diverse coded as 3 [@fraser_coding_2020]. (or 0.5 = neither female nor male)

Here, we coded all those who responded as Male as 1, and those who did not as 0.

#### Income (waves: 1-3, 4-15)

Participants were asked "Please estimate your total household income (before tax) for the year XXXX". To stablise this indicator, we first took the natural log of the response + 1, and then centred and standardised the log-transformed indicator.

<!-- #### Job Security (waves: 1-3,4-7,9-15) -->

<!-- Participants indicated their feeling of job security by answering "How secure do you feel in your current job?" on a scale from 1 (not secure) to 7 (very secure). -->

#### Parent (waves: 5-15) -->

Participants were asked "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-15). Parents were coded as 1, while the others were coded as 0. -->

<!-- #### Number of Children (waves: 1-3, 4-15) -->

<!-- We measured the number of children using one item from @Bulbulia_2015. We asked participants "How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, or adopted?" or ""How many children have you given birth to, fathered, or adopted. How many children have you given birth to, fathered, and/or parented?" (waves: 12-15). -->

<!-- Here, we used responses to create a binary indicator for parenthood: Parent Yes = 1/ No = 0.  -->

#### Political Orientation

We measured participants' political orientation using a single item adapted from @jost_end_2006-1.

"Please rate how politically liberal versus conservative you see yourself as being."

(1 = Extremely Liberal to 7 = Extremely Conservative)

#### NZSEI-13 (waves: 8-15)

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) [@fahy2017]. This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation.

Participants were asked "If you are a parent, what is the birth date of your eldest child?".

#### Living with Partner

Participants were asekd "Do you live with your partner?" (1 = Yes, 0 = No).

#### Living in an Urban Area (waves: 1-15)

We coded whether they are living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

We coded whether they are living in an urban or rural area (1 = Urban, 0 = Rural) based on the addresses provided.

#### NZ Deprivation Index (waves: 1-15)

We used the NZ Deprivation Index to assign each participant a score based on where they live [@atkinson2019]. This score combines data such as income, home ownership, employment, qualifications, family structure, housing, and access to transport and communication for an area into one deprivation score.

<!-- #### NZ-Born (waves: 1-2,4-15) -->

<!-- We asked participants "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15). -->

<!-- #### Mini-IPIP 6 (waves: 1-3,4-15) -->

<!-- We measured participants' personality with the Mini International Personality Item Pool 6 (Mini-IPIP6) [@sibley2011] which consists of six dimensions and each dimension is measured with four items: -->

<!-- 1.  agreeableness, -->

<!--     i.  I sympathize with others' feelings. -->
<!--     ii. I am not interested in other people's problems. (r) -->
<!--     iii. I feel others' emotions. -->
<!--     iv. I am not really interested in others. (r) -->

<!-- 2.  conscientiousness, -->

<!--     i.  I get chores done right away. -->
<!--     ii. I like order. -->
<!--     iii. I make a mess of things. (r) -->
<!--     iv. I ften forget to put things back in their proper place. (r) -->

<!-- 3.  extraversion, -->

<!--     i.  I am the life of the party. -->
<!--     ii. I don't talk a lot. (r) -->
<!--     iii. I keep in the background. (r) -->
<!--     iv. I talk to a lot of different people at parties. -->

<!-- 4.  honesty-humility, -->

<!--     i.  I feel entitled to more of everything. (r) -->
<!--     ii. I deserve more things in life. (r) -->
<!--     iii. I would like to be seen driving around in a very expensive car. (r) -->
<!--     iv. I would get a lot of pleasure from owning expensive luxury goods. (r) -->

<!-- 5.  neuroticism, and -->

<!--     i.  I have frequent mood swings. -->
<!--     ii. I am relaxed most of the time. (r) -->
<!--     iii. I get upset easily. -->
<!--     iv. I seldom feel blue. (r) -->

<!-- 6.  openness to experience -->

<!--     i.  I have a vivid imagination. -->
<!--     ii. I have difficulty understanding abstract ideas. (r) -->
<!--     iii. I do not have a good imagination. (r) -->
<!--     iv. I am not interested in abstract ideas. (r) -->

<!-- Each dimension was assessed with four items and participants rated the accuracy of each item as it applies to them from 1 (Very Inaccurate) to 7 (Very Accurate). Items marked with (r) are reverse coded. -->

<!-- #### Honesty-Humility-Modesty Facet (waves: 10-14) -->

<!-- Participants indicated the extent to which they agree with the following four statements from @campbell2004 , and @sibley2011 (1 = Strongly Disagree to 7 = Strongly Agree) -->

<!-- ```          -->
<!-- i.  I want people to know that I am an important person of high status, (Waves: 1, 10-14) -->
<!-- ii. I am an ordinary person who is no better than others. -->
<!-- iii. I wouldn't want people to treat me as though I were superior to them. -->
<!-- iv. I think that I am entitled to more respect than the average person is. -->
<!-- ``` -->


## Appendix B. Baseline Demographic Statistics {#appendix-demographics}


```{r}
#| label: tbl-table-demography-code
#| tbl-cap: "Baseline demographic statistics"
#| eval: false
#| echo: false
#| include: false

summary_tables$baseline_table |> kbl(format = "markdown")


```






##  Appendix C: Baseline and Treatment Wave Exposure Statistics {#appendix-exposures}


```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and treatment wave"
#| eval: false
#| echo: false
#| include: false

summary_tables$exposure_tables |> kbl(format = "markdown")
```




##  Appendix E: Baseline and End of Study Outcome Statistics {#appendix-outcomes}


```{r}
#| label: tbl-table-demographic_vars-code
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: false
#| echo: false
#| include: false
summary_tables$outcome_table |> kbl(format = "markdown")

```



##  Appendix XX: Baseline and End of Study Outcome Statistics {#appendix-xx}

### Results for OSF planned study 

## Results Study 1 Antagonism OSF (Partner)

```{r}
#| label: fig-results-antagonism-partner_osf
#| fig-cap: "Results for antagonism effect on osf partner multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16


plot_antagonism_partner_osf_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-antagonism-partner-osf
#| tbl-cap: "Table for antagonism effect on osf partner multi-dimensional well-being"
#| eval: true
#| include: false
#| echo: false

plot_antagonism_partner_osf_lmtp_output$transformed_table |>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_antagonism_partner_osf_lmtp_output$interpretation)
```



### Results Study 1: Compare Antagonism OSF vs Reported (Partner)

```{r}
#| label: fig-results-antagonism-osf-compare
#| fig-cap: "Results for antagonism effect for osf vs reported partner on multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_antagonism_partner_osf_lmtp_output$plot + plot_antagonism_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}
## Results Study 2 Disinhibition OSF (Partner)

```{r}
#| label: fig-results-disinhibition-partner-osf
#| fig-cap: "Results for disinhibition effect for osf partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12


plot_disinhibition_partner_osf_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-disinhibition-partner-osf
#| tbl-cap: "Table for disinhibition effect for osf partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_disinhibition_partner_osf_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_disinhibition_partner_osf_lmtp_output$interpretation)

```


{{< pagebreak >}}
### Results Study Compare Disinhibition OSF vs Reported (Partner)

```{r}
#| label: fig-results-disinhibition-osf-compare
#| fig-cap: "Results for disinhibition effect for osf vs reported partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24


plot_disinhibition_partner_osf_lmtp_output$plot + plot_disinhibition_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}

## Results Study 3  Emotional Stability: OSF (Partner) 

```{r}
#| label: fig-results-emotional-partner-osf
#| fig-cap: "Results for emotional stability effect for osf partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12

plot_emotional_stability_partner_osf_lmtp_output$plot

```

{{< pagebreak >}}

```{r}
#| label: tbl-results-emotional-partner-osf
#| tbl-cap: "Table for emotional stability effect for osf partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_emotional_stability_partner_osf_lmtp_output$transformed_table|> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_emotional_stability_partner_osf_lmtp_output$interpretation)

```

{{< pagebreak >}}

### Results Study 3  Emotional Stability OSF vs Reported (Partner)

```{r}
#| label: fig-results-emotional-osf-compare
#| fig-cap: "Results for emotional stability effect for osf vs reported partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24


plot_emotional_stability_self_lmtp_output$plot + plot_emotional_stability_partner_lmtp_output$plot + plot_annotation(tag_levels = "A")

```

{{< pagebreak >}}




## Results Study 4 Narcissism 



### Results Study 4A Narcissism: Partner

```{r}
#| label: fig-results-narcissism-partner-osf
#| fig-cap: "Results for Narcissism stability gain for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12
plot_narcissism_partner_osf_lmtp_output$plot
```

{{< pagebreak >}}

```{r}
#| label: tbl-results-narcissism-partner-osf
#| tbl-cap: "Table for Narcissism gain for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false

plot_narcissism_partner_osf_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_narcissism_partner_osf_lmtp_output$interpretation)


```
{{< pagebreak >}}


### Results Study B Narcissism OSF vs Reported (Partner)

```{r}
#| label: fig-results-narcissism-self-osf
#| fig-cap: "Results for Narcissism stability effect for self vs partner on multi-dimensional well-being: z-transformed"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_narcissism_partner_osf_lmtp_output$plot + plot_narcissism_partner_lmtp_output$plot   + plot_annotation(tag_levels = "A")
```

## Results Study 5: Psychopathy Combined Score: OSF (Partner)

```{r}
#| label: fig-results-psychopathy-partner-osf
#| fig-cap: "Results forPsychopathy Combined Score effect for partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 12

plot_psychopathy_combined_partner_osf_lmtp_output$plot


```

{{< pagebreak >}}

```{r}
#| label: tbl-results-psychopathy-partner-osf
#| tbl-cap: "Table for Psychopathy Combined Score effect for partner on multi-dimensional well-being."
#| include: true
#| echo: false

plot_psychopathy_combined_partner_osf_lmtp_output$transformed_table |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")

cat(plot_psychopathy_combined_partner_osf_lmtp_output$interpretation)
```


### Psychopathy Combined Score OSF vs Reported (Partner)

```{r}
#| label: fig-results-psychopathy-osf-compare
#| fig-cap: "Results for  Psychopathy Combined Score effect for self vs partner on multi-dimensional well-being."
#| eval: true
#| include: true
#| echo: false
#| fig-height: 12
#| fig-width: 24

plot_psychopathy_combined_partner_osf_lmtp_output$plot + plot_psychopathy_combined_partner_lmtp_output$plot   + plot_annotation(tag_levels = "A")
```



## Appendix XXX Null Contrasts for the reported model

###  Antagonism
#### Antagonism (Partner): Gain vs Null

```{r}
#| label: tbl-results-antagonism-partner-null-gain
#| tbl-cap: "Table for Antagonism on partner multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

antagonism_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Antagonism (Partner): Loss vs Null

```{r}
#| label: tbl-results-antagonism-null-loss
#| tbl-cap: "Table for Antagonism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

antagonism_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```

###  Disinhibition
#### Disinhibition (Partner): Gain vs Null

```{r}
#| label: tbl-results-disinhibition-partner-null-gain
#| tbl-cap: "Table for Disinhibition effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

disinhibition_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Disinhibition (Partner): Loss vs Null

```{r}
#| label: tbl-results-disinhibition-null-loss
#| tbl-cap: "Table for Disinhibition on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

disinhibition_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```



###  Emotional Stability

#### Emotional Stability (Partner): Gain vs Null

```{r}
#| label: tbl-results-emotional-stability-partner-null-gain
#| tbl-cap: "Table for Disinhibition effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

emotional_stability_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Emotional Stability (Partner): Loss vs Null

```{r}
#| label: tbl-results-emotional-stability-null-loss
#| tbl-cap: "Table for Disinhibition on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

emotional_stability_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```



### Narcissism
####Narcissism (Partner): Gain vs Null

```{r}
#| label: tbl-results-narcissism-partner-null-gain
#| tbl-cap: "Table for Narcissism effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

narcissism_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Narcissism (Partner): Loss vs Null

```{r}
#| label: tbl-results-psychopathy_combined_null-loss
#| tbl-cap: "Table for Narcissism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


### Psychopathy Combined Score
#### Psychopathy Combined Score (Partner): Gain vs Null

```{r}
#| label: tbl-results-psychopathy_combined_partner-null-gain
#| tbl-cap: "Table for Narcissism effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Psychopathy Combined Score (Partner): Loss vs Null

```{r}
#| label: tbl-results-narcissism-null-loss
#| tbl-cap: "Table for Narcissism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```



## Appendix XXX Null Contrasts for the OSF model

###  Antagonism OSF
#### Antagonism OSF (Partner): Gain vs Null

```{r}
#| label: tbl-results-antagonism-partner-null-gain-osf
#| tbl-cap: "Table for Antagonism on partner multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

antagonism_partner_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Antagonism OSF (Partner): Loss vs Null

```{r}
#| label: tbl-results-antagonism-null-loss-osf
#| tbl-cap: "Table for Antagonism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

antagonism_partner_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```

###  Disinhibition OSF
#### Disinhibition OSF (Partner): Gain vs Null

```{r}
#| label: tbl-results-disinhibition-partner-null-gain-osf
#| tbl-cap: "Table for Disinhibition effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

disinhibition_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Disinhibition OSF (Partner): Loss vs Null

```{r}
#| label: tbl-results-disinhibition-null-loss-osf
#| tbl-cap: "Table for Disinhibition on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

disinhibition_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```



###  Emotional Stability OSF

#### Emotional Stability OSF (Partner): Gain vs Null

```{r}
#| label: tbl-results-emotional-stability-partner-null-gain-osf
#| tbl-cap: "Table for Disinhibition effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

emotional_stability_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Emotional Stability OSF (Partner): Loss vs Null

```{r}
#| label: tbl-results-emotional-stability-null-loss-osf
#| tbl-cap: "Table for Disinhibition on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

emotional_stability_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```



### Narcissism OSF

####Narcissism OSF (Partner): Gain vs Null

```{r}
#| label: tbl-results-narcissism-partner-null-gain-osf
#| tbl-cap: "Table for Narcissism effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

narcissism_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Narcissism OSF (Partner): Loss vs Null

```{r}
#| label: tbl-results-psychopathy_combined_null-loss-osf
#| tbl-cap: "Table for Narcissism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


### Psychopathy Combined Score OSF 

#### Psychopathy Combined Score OSF (Partner): Gain vs Null

```{r}
#| label: tbl-results-psychopathy_combined_partner-null-gain-osf
#| tbl-cap: "Table for Narcissism effect for partner on multi-dimensional well-being: gain vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_up_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


#### Psychopathy Combined Score OSF (Partner): Loss vs Null

```{r}
#| label: tbl-results-narcissism-null-loss-osf
#| tbl-cap: "Table for Narcissism on partner multi-dimensional well-being: loss vs null."
#| include: true
#| echo: false

psychopathy_combined_partner_osf_lmtp_output$combined_tables$combined_outcomes_shift_down_vs_null |> 
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```