---
title: "Causal Effect of Religious Service Attendance On Fertility in an American Sample"
abstract: |
 According to 
   **KEYWORDS**: *Causal Inference*; *Longitudinal*; *Religion*; *Fertility*.
author: 
  - name: Radim Chvaja
    affiliation: 	European Research University, Masaryk University, Faculty of Arts, University of Otago
    orcid: 000-0002-1560-1197
    corresponding: yes
  - name: John H Shaver
    affiliation: Baylor University
    orcid: 0000-0002-9522-4765
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
editor_options: 
  chunk_output_type: console
format:
  docx:
  pdf:
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
date: last-modified
pdf-engine: lualatex
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/apa-7.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# link to data
pull_mods <- fs::path_expand(
  "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-radim-usa/24-05-data_for_joe.csv"
)

# set to your directory
push_mods <- fs::path_expand(
  "/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-radim-usa"
)

# use p_load to load / install the packages
pacman::p_load(
  skimr,
  naniar,
  WeightIt,
  clarify,
  MatchThem,
  cobalt,
  MatchIt,
  kableExtra,
  janitor,
  SuperLearner,
  ranger,
  xgboost,
  glmnet,
  doParallel,
  ggplot2,
  here,
  naniar,
  gtsummary,
  grf,
  progressr,
  tidyverse,
  ggplot2,
  parameters,
  kableExtra,
  tidyr,
  stringr,
  patchwork,
  margot,
  glue,
  gtsummary
)
# update.packages()

# import data
# dat_raw <- read.csv(pull_mods)

# for latex graphs
# for making graphs
library("tinytex")
library("extrafont")

loadfonts(device = "all")


## Read in data from analysis
# imports
n_participants <- readRDS(here::here(push_mods, "n_participants"))
n_total <- margot::here_read("n_total")

# missing graph baseline
missing_total <- here_read("missing_total")

# propensity score graph
love_plot_one <- margot::here_read("love_plot_one")

# transition table
transition_table <- margot::here_read("transition_table")


# baseline table
table_baseline <- margot::here_read("table_baseline")

# exposure table
table_exposures <- margot::here_read("table_exposures")

# outcome table
table_outcomes <- margot::here_read("table_outcomes")


#  results
tab_contrast_t2_two_or_more_kids <- margot::here_read("tab_contrast_t2_two_or_more_kids")

# contrasts for subgroups
tab_contrast_male_t2_two_or_more_kids <- margot::here_read("tab_contrast_male_t2_two_or_more_kids")
tab_contrast_female_t2_two_or_more_kids <- margot::here_read("tab_contrast_female_t2_two_or_more_kids")


sub_group_compare_binary <- margot::here_read("sub_group_compare_binary")

# display the result


# Risk Ratio
tab_contrast_t2_two_or_more_kids <- here_read("tab_contrast_t2_two_or_more_kids")
tab_contrast_t2_two_or_more_kids


tab_binary_ate <- margot::margot_interpret_table(tab_contrast_t2_two_or_more_kids , causal_scale = "risk_ratio", estimand = "ATE", 
order = "alphabetical")

tab_contrast_female_t2_two_or_more_kids <- here_read("tab_contrast_female_t2_two_or_more_kids")
tab_contrast_male_t2_two_or_more_kids <- here_read("tab_contrast_male_t2_two_or_more_kids")

tab_binary_female <- margot::margot_interpret_table(tab_contrast_female_t2_two_or_more_kids , causal_scale = "risk_ratio", estimand = "ATE", 
order = "alphabetical")


tab_binary_male <- margot::margot_interpret_table(tab_contrast_male_t2_two_or_more_kids , causal_scale = "risk_ratio", estimand = "ATE", 
order = "alphabetical")


# bind result tables
df_bind_results <- rbind(tab_contrast_female_t2_two_or_more_kids, tab_contrast_male_t2_two_or_more_kids, tab_contrast_t2_two_or_more_kids)

# table
tab_binary <- margot::margot_interpret_table(df_bind_results , causal_scale = "risk_ratio", estimand = "ATE", 
order = "alphabetical")

tab_binary
# check N
N <- n_participants
N
Title <- "Eight-year Causal Effect of Regular Religious Attendance on Achieving Replacement-level Fertility (Two or More Children), N = 3,365"
sub_title = 'N = 3,365'

# plot

plot_binary <- margot::margot_plot(df_bind_results, 
  title = Title, 
  subtitle= sub_title, 
  type = "RR",
  order = "alphabetical",
  x_lim_hi = 7,
  x_lim_lo = -7, 
  x_offset = -1.75,
  estimate_scale  = 4)


#plot_binary

```

## Introduction

Human societies are characterised by intricate divisions of labour that extend beyond family ties. Humans are obligate cooperators, unable to live apart from society [@sterelny2007social]
These involve both direct and indirect reciprocal relationships across multiple domains, such as survival skills, communal protection, building, learning, food gathering, hunting, commerce across long distances, and tool-craft [@sterelny2011hominins]. It has been noted that children rely on this cooperative framework for nearly two decades post-birth [@gurven2006energetic], and parents require the broader social group's support to provide for their offspring [@shaver2019alloparenting], evolutionary models of religion posit that religion evolved deep in the human lineage to foster cooperation [@sosis2003cooperation], with gene-culture co-evolution elaborating dispositions to religion, and religous cultural systems, whose ultimate functions promote fertility [@rowthorn2011religion]. The cooperative functions of these systems may include mate selection [@blume2009reproductive] and collective child-rearing [@shaver2020church]. Indeed, the function of mate-selection and collective child rearing might not be easily disentangled; offspring unite parents to a common genetic fate.  Sex in humans requires parenting through a long period of adolesence, with demands selecting for extensions of the human life-span through grandparenting [@kaplan2000theory].  In this sense, it has been claimed that 'reproduction poses biology’s most fundamental cooperation problem' 

Notably, sexual signalling theory posits that differences in reproductive potentials between males and females lead to different valuations of traits in prospective mates. For example, it has been proposed that males may be are more concerned about female infidelity given that males face risks of cuckoldry, whereas and may be concerned about resource diversion resulting from male infidelity or diminished social prestige. This gradient evolved within humans to extend cooperative networks beyond immediate reciprocity [@alexander2017biology]. It has been proposed that males engage in high-risk cooperation, such as hunting and defence to obtain and signal prestige to mates [@gurven2009men]. Indeed preliminary cross-sectional evidence suggests that highly religious males attend religious service more frequently than highly religious females and that church attendance is associated with greater offspring [@Bulbulia_2015]. Although such associational evidence is consistent with the predictions of both an alloparenting model of cooperation and sexual-specific models of cooperation, presently, we are unaware of any systematic attempt to leverage time series data to assess whether religious behaviours cause greater offspring.  

Here, we address this gap in understanding by quantitatively assessing the causal effects of regular religious service attendance on offspring in a population of [**Radim to clarify ...US citizens...**] (2) Male-specific enhancements to fertility within this same population. 


## Method

### Sample

Data were collected by [**Radim to clarify**]

We adopted a three-wave model for causal identification in which baseline measures were obtained in 2003, the exposure or treatment was measured in 2005, and the outcome, fertility was measured in 2013.  Notably we included measures of religious service in the baseline wave (year 2003), allowing us to estimate a specific 'incident exposure effect' for religious service, rather than merely estimating the prevalence effect (described below.)


## Baseline covariates 

[**Radim to clarify**]

- Age: measured by  [**Radim to clarify wording of measure**] (continuous)
- Faith scale: measured by [**Radim to clarify the wording of measure**] (ordinal)
- Female Gender:  [**Radim to clarify the wording of measure**] (binary)
- Parent Education:  [**Radim to clarify the wording of measure**] (ordered factor)
- Parent Income:  [**Radim to clarify the wording of measure**]  (continuous)
- Race:  [**Radim to clarify the wording of measure**] (factor)
- Religious Category:  [**Radim to clarify the wording of measure**] (factor)
- Ritual Scale:  [**Radim to clarify the wording of measure**] (ordinal)


### Exposure or Treatment Variable

Our study aimed to estimate causal effects by comparing two potential interventions:

1. The marginal effect of religious service attendance across the entire sampled population.
2. The stratum-specific effects for females and males separately.

We addressed four main causal questions:

1. Population-wide effect: What would be the magnitude of the effect on offspring if everyone in the population attended religious services weekly, compared to if no one attended at all?

2. Female-specific effect: What is the magnitude of this effect (comparing regular religious service attendance to no attendance) among females, as proposed by cooperative-breeding models of religion?

3. Male-specific effect: What is the magnitude of this effect (comparing regular religious service attendance to no attendance) among males, as proposed by cooperative breeding models of religion?

4. Sex differences: Do the expected effects of religious service attendance differ between females and males, as suggested by sexual signalling models?

Important note: We used the ritual scale variable at baseline for our analysis, as it provides more detailed information than the binary 'religious service attendance at least weekly' variable.

### Outcome Measures 

Offspring measures were collected in the final study wave in 2013, eight years after exposure and ten years after baseline. Since participants were between 13 and 17 years old at baseline, their fertility was assessed from 23 to 27 years old. [**Radim to check**]

We effects on the probability of having two or more children in 2013 -- or ‘replacement fertility’.... [JS SAY MORE].


### Inclusion Criteria

- No missing data in the baseline wave for the religious service attendance.
- Missing covariate data at baseline were permitted.

A total of `r n_participants` individuals met these criteria and were included in the study.




{{< pagebreak >}}

```{r}
#| label: tbl-table-demography
#| tbl-cap: "Baseline demographic statistics"
#| eval: true
#| include: true
#| echo: false

table_baseline <- margot::here_read("table_baseline")
table_baseline |>  kbl(format = "markdown")
```


@tbl-table-demography reports baseline sample characteristics. 

{{< pagebreak >}}

```{r}
#| label: tbl-table-exposures-code
#| tbl-cap: "Exposures at baseline and baseline + 1 (treatment) wave"
#| eval: true
#| include: true
#| echo: false

table_exposures <-margot::here_read("table_exposures")
table_exposures |>  kbl(format = "markdown")
```


tbl-table-exposures-code presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: forgiveness (range 1-7). All models adjusted for the pandemic alert level because the treatment wave (NZAVS time 11) occurred during New Zealand's COVID-19 pandemic. The pandemic is not a "confounder" because a confounder must be related both to the treatment and the outcome. At the end of the study, however, all participants had been exposed to the global pandemic.  However, to satisfy the causal consistency assumption, all treatments must be conditionally equivalent within levels of all covariates [@vanderweele2013]. Because COVID may have changed the quality and accessibility of forgiveness, we included each participants lockdown condition as a covariate [@sibley2021]. We employed inverse probability of censoring weights, which were computed non-parametrically within our statistical models using the `lmtp` package to mitigate systematic biases arising from attrition and missingness.


{{< pagebreak >}}

```{r}
#| label: tbl-table-outcomes
#| tbl-cap: "Outcomes at baseline and end-of-study"
#| eval: true
#| include: true
#| echo: false
table_outcomes <-margot::here_read("table_outcomes")
table_outcomes |>  kbl(format = "markdown")
```



@tbl-table-outcomes presents baseline and end-of-study descriptive statistics for the outcome variables.  


{{< pagebreak >}}


### Handling Missing Data

Participants may have been lost to follow-up at the exposure wave or outcome wave.  To address bias from missing responses and attrition, we implement the following strategies:

**Baseline missingness**: we employed the `ppm` algorithm from the `mice` package in R [@vanbuuren2018] to impute missing baseline data. This method allowed us to reconstruct incomplete datasets by estimating a plausible value for missing observation. Because we could only pass one data set to the `lmtp`, we employed single imputation. 1.3% of covariate values were missing at baseline. Eligibility for the study required fully observed baseline treatment measures as well as treatment wave treatment measures. Again, we only used baseline data to impute baseline missingness (refer to @zhang2023shouldMultipleImputation).


**Exposure missingness**: to address bias for the target population arising from missing values in the exposures, we used inverse probability of censoring weights [@hernan2024WHATIF].  


**Outcome missingness**: to address confounding and selection bias arising from missing responses and panel attrition, we applied censoring weights obtained using nonparametric machine learning ensembles afforded by the `lmtp` package (and its dependencies) in R [@williams2021]. 

We implemented inverse probability of censoring weights to address potential bias from loss to follow-up between the baseline and exposure waves. The process involved the following steps:

1. First, we used a set of baseline variables as predictors for loss to follow-up, excluding the outcome variable itself.

2. Second, we used one-hot encoding to render categorical variables binary. These were: 'religious group affiliation', 'parent education', and race'.

3. Next, we employed the SuperLearner algorithm to predict the probability of being lost to follow-up. Our SuperLearner ensemble included various machine learning methods: generalised linear models with regularisation (glmnet), extreme gradient boosting (xgboost), random forests (ranger), multivariate adaptive regression splines (earth), and polynomial spline regression (polymars).

4. The SuperLearner model was trained using 10-fold cross-validation to optimise prediction accuracy.

5. Based on the predicted probabilities, we calculated inverse probability weights. For participants lost to follow-up, the weight was one divided by their predicted probability of being lost. For those not lost, the weight was one divided by their predicted probability of not being lost.

6. To mitigate extreme weights, we computed stabilised weights by multiplying the inverse probability weights by the marginal probability of not being lost to follow-up.

7. These stabilized censoring weights were then combined with the original sample weights to create composite weights for subsequent analyses.

8. Participants lost to follow-up between baseline and the exposure wave were removed from the analysis dataset.


Our method assumes that loss to follow-up/missing responses are ‘missing at random’ (MAR) conditional on the observed variables. This means that all variables influencing dropout are captured by the variables used in the prediction model. If there are unobserved variables influencing dropout, some bias may remain. This approach is also subject to model misspecification bias.  However, by using SuperLearner - which combines multiple machine learning methods -- our models can flexibly (and non-parametrically) capture complex relationships in the data and reduce the risk of model misspecification compared to using a single model.

### Causal Assumptions

To estimate causal effects reliably, our study must credibly meet three fundamental assumptions:

1. **Causal Consistency**: we assume that the potential outcomes for a given individual under a particular treatment correspond to the observed outcomes when that treatment is administered. This assumes that potential outcomes are solely a function of the treatment and measured covariates, without interference by the mode of treatment administration (@vanderweele2009; @vanderweele2013).

2. **Exchangeability**: we assume that, conditional on observed covariates, assignment to the treatment group is independent of the potential outcomes; that is, we assume no unmeasured confounding (@hernan2024WHATIF; @chatton2020).

3. **Positivity**: we assume that within all covariate-defined strata necessary for achieving exchangeability, there is a non-zero probability of receiving each possible treatment level. The positivity assumption ensures that treatment effects are estimable across the spectrum of observed covariate combinations (@westreich2010).

The target population for this study comprises [**Radim to check**]


###  Causal Identification 

::: {#tbl-02}
```{=latex}
\threevanderweeele
```
This table presents a causal diagram using @vanderweele2020's approach for confounding control in a three-wave panel design. By including baseline measures of all outcomes in every model, as well as including the baseline treatment, and a rich array of covariates, we assume may back door paths between the treatment and outcomes will be blocked. However, because confounding cannot be ensured, we also perform sensitivity analyses.
:::


To address confounding in our analysis, we implement @vanderweele2019's *modified disjunctive cause criterion* by following these steps:

  1. **Identified all common causes** of both the treatment and outcomes to ensure a comprehensive approach to confounding control.
  2. **Excluded instrumental variables** that affect the exposure but not the outcome. Instrumental variables do not contribute to controlling confounding and can reduce the efficiency of the estimates.
  3. **Included proxies for unmeasured confounders** affecting both exposure and outcome. According to the principles of d-separation, using proxies allows us to control for their associated unmeasured confounders indirectly.
  4. **Controlled for baseline exposure** and **baseline outcome**. Both are used as proxies for unmeasured common causes, enhancing the robustness of our causal estimates.

@tbl-02 presents a causal diagram describing the general confounding control of @vanderweele2020. 


These methods adhere to the guidelines provided in [@bulbulia2024PRACTICAL].


### Statistical Estimator

We employed Targeted Maximum Likelihood Estimation (TMLE), a double-robust method for causal inference. TMLE operates through a two-step process that involves modeling both the outcome and treatment (exposure).

In the first step, we used machine learning algorithms with ten-fold cross-validation to flexibly model the relationships between treatments, covariates, and outcomes. This approach allows us to efficiently account for complex, high-dimensional covariate spaces without imposing restrictive model assumptions. We used the `SuperLearner` package [@polley2023], and implemented an ensemble learning method combining the following algorithms:

1. `SL.glmnet`: for regularised linear models, managing high dimensionality and collinearity [@glmnet_2023; @glmnet_use_2010].
2. `SL.xgboost`: employing gradient boosting to capture complex interactions [@xgboost2023].
3. `SL.ranger`:  random forests to handle non-linear relationships [@ranger_2017].
4. `SL.earth`: using multivariate adaptive regression splines for flexible modelling [@earth_2024].
5. `SL.polymars`: applying polynomial spline regression for additional flexibility [@polspline_2024].

The ensemble approach of `SuperLearner` optimally combines predictions from these models, leveraging their strengths while mitigating individual weaknesses.

In the second step, TMLE 'targets' the initial estimates by incorporating information about the observed data distribution, improving the accuracy of the causal effect estimate. This targeting step is crucial for reducing bias in the final estimate.

We implemented TMLE using the `lmtp` package, which is designed for longitudinal causal inference. The package allows for the estimation of causal effects under complex, time-varying treatments and confounding [@williams2021].

To handle potential bias from loss to follow-up, we implemented inverse probability of censoring weights. These weights were calculated using the same `SuperLearner` ensemble to predict the probability of being lost to follow-up based on baseline characteristics [@williams2021; @hoffman2023].

For binary outcomes (having two or more children), we estimated risk ratios. For continuous outcomes (total number of children), we estimated additive effects. We conducted separate analyses for the overall sample and stratified by sex to explore examine heterogeneity in treatment effects.

This approach combines the strengths of machine learning for flexible modelling with the statistical properties of TMLE, providing robust and efficient estimates of causal effects while accounting for complex confounding and loss to follow-up.

Tables and Graphs were produced using the `margot` package [@margot2024]


### Sensitivity Analysis Using the E-value

To assess the sensitivity of results to unmeasured confounding, we report VanderWeele and Ding's E-value in all analyses [@vanderweele2017]. The E-value quantifies the minimum strength of association (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and the outcome (after considering the measured covariates) to explain away the observed exposure-outcome association [@vanderweele2020; @linden2020EVALUE]. To evaluate the strength of evidence, we use the bound of the E-value 95% confidence interval closest to 1. Although the E-value provides an approximate sensitivity analysis, its interpretation is straightforward. 


### Evidence for Change in the Treatment Variable

@tbl-transition clarifies the change in the treatment variable from the baseline wave to the baseline + 1 wave across the sample.  Assessing change in a variable is essential for evaluating the positivity assumption and recovering evidence for the incident-exposure effect of the treatment variable [@vanderweele2020; @danaei2012; @hernan2024WHATIF].  

```{r}
#| label: tbl-transition
#| tbl-cap: "This transition matrix captures shifts in states across the treatment intervals. Each cell in the matrix represents the count of individuals transitioning from one state to another. The rows correspond to the treatment at baseline (From), and the columns correspond to the state at the following wave (To). **Diagonal entries** (in **bold**) correspond to the number of individuals who remained in their initial state across both waves. **Off-diagonal entries** correspond to the transitions of individuals from their baseline state to a different state in the treatment wave.\nA higher number on the diagonal relative to the off-diagonal entries in the same row indicates greater stability in a state. Conversely, higher off-diagonal numbers suggest more frequent shifts from the baseline state to other states.."
#| eval: true
#| echo: false

transition_table <- margot::here_read("transition_table")
transition_table$table
```


{{< pagebreak >}}


## Results

#### Population-wide effect: What would be the magnitude of effect on the probability of two or more offspring if everyone in the population attended religious services weekly, compared to if no one attended at all?

```{r}
#| eval: true
#| include: false
#| echo: false

tab_binary_ate
```


The effect estimate for regular religious service on having two or more offspring eight years later on the risk_ratio scale is 1.622 [1.14, 2.306]. The E-value for this estimate is 2.626, with a lower bound of 1.539. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 1.539 to negate the observed effect. Weaker confounding would not overturn it. Here, **there is evidence for causality**.


#### Female-specific effect: What is the magnitude of effect on the probability of two or more offspring (comparing regular religious service attendance to no attendance) among females, as proposed by cooperative-breeding models of religion?


```{r}
#| eval: true
#| include: false
#| echo: false

tab_binary_female
```

For females, the effect estimate for regular religious service on having two or more offspring eight years later on the risk_ratio scale is 1.303 [0.804, 2.112]. The E-value for this estimate is 1.931, with a lower bound of 1. Here, **the evidence for causality is not reliable**.


#### Male-specific effect: What is the magnitude of effect on the probability of two or more offspringcomparing regular religious service attendance to no attendance) among males, as proposed by cooperative-breeding models of religion?


```{r}
#| eval: true
#| include: false
#| echo: false

tab_binary_male
```


For males, the effect estimate for regular religious service  on the risk_ratio scale is 3.142 [1.435, 6.878]. The E-value for this estimate is 5.736, with a lower bound of 2.225. At this lower bound, unmeasured confounders would need a minimum association strength with both the intervention sequence and outcome of 2.225 to negate the observed effect. Weaker confounding would not overturn it. Here, **the evidence for causality is strong**.


#### Sex differences: Do the expected effects of religious service attendance on the probability of two or more offspring differ between females and males, as suggested by sexual signalling models?


```{r}
#| eval: true
#| include: false
#| echo: false

sub_group_compare_binary
```


The difference in the relative risk ratio between the focal group (males) and the reference group (females) is 2.41 with a standard error of 0.228 and a 95% ci of [1.54, 3.77].  We infer that males who regularly attend religious service have a 2.41 fold-increase in the probability of having at least two offspring in an eight year follow up windown as compared with females. 


@fig-1_1 *A* and @tbl-1_1: describe both the marginal and sex-specific results for the eight-year causal effect of religious service on the probability of having two or more offspring. 


```{r}
#| label: tbl-1_1
#| tbl-cap: "This table presents the causa leffect estimates for the population, males, and females on the probability of replacement-level fertility eight years after 'treatment'."
#| eval: true
#| include: true
#| echo: false

df_bind_results|>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  arrange(desc(`E[Y(1)]/E[Y(0)]`)) |>
  kable(format = "markdown") 

# margot_interpret_table(df_bind_results, estimand = "ATE",causal_scale = "risk_ratio" )
#margot_interpret_table(out_tab_contrast_t2_smoker_binary, estimand = "ATE", causal_scale = "risk_ratio" )
```

{{< pagebreak >}}



```{r}
#| label: fig-1_1
#| fig-cap: "This figure presents the causa effect estimates for the population, males, and females on the probability of replacement-level fertility eight years after 'treatement'.  "
#| eval: true
#| include: true
#| echo: false
#| fig-width: 10
#| fig-height: 12

plot_binary
```

{{< pagebreak >}}


Our findings may be summaries as follows:

1. Overall Effect (ATE):
   Regular religious service attendance is associated with a 62.2% increase in the probability of having two or more children (RR = 1.622, 95% CI: 1.140 to 2.306). This effect is statistically reliable as the confidence interval does not include 1. The E-vale of 1.549 suggests moderate robustness to unmeasured confounding. 

2. Sex-Specific Effects:
   - Females: There's a 30.3% increase in probability for females (RR = 1.303, 95% CI: 0.804 to 2.112), but this effect is not statistically reliable as the confidence interval includes 1.  We cannot rule out the prospect that religious service attendance has no effect in the eight years following the regular religious service 'treatment.'
   - Males: There's a much larger 214.2% increase in probability for males (RR = 3.142, 95% CI: 1.435 to 6.878). This effect is statistically reliable and considerably more robust than for females.

We infer that the causal effect of religious service attendance in the target population (defined to be the sample population at baseline) is considerably stronger in males than in females. Recall that the E-values indicate the strength of unmeasured confounding needed to explain away these effects. The male-specific result (E-value = 5.736) appears more robust to potential unmeasured confounding than the female-specific or overall results. This finding suggests potential sex-specific differences in how religious service attendance relates to fertility. 

In summary, regular religious service attendance causally associateed with increased probability of having two or more children, with a powerful effect observed in males. However, the wide confidence intervals, especially for the sex-specific estimates, suggest some uncertainty in the precise magnitude of these effects.


## Discussion 

### Limitation

- Population
- Sample (young)
- Measures -- can be unmeasured confounding 


### Strengths
- Evidence for causal effect of religious service on fertility
- Clear difference by gender: males replacing earlier. 


### Evidence for Gender-Specific Effects? 

Not necessarily 


Future research... etc. 



{{< pagebreak >}}

### Ethics

*include*

### Data Availability


### Acknowledgements

*Authors to fill out*
JS...
RC...
JB recewived support from Templeton Religious Trust (TRT0196; TRT0418) and the Max Planck Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement

*Authors to amend*

RC conceived of the study and wrote the first draft of the manuscript, with John Shaver advising. 
JB and RC did the analysis.
All authors contributed to the study.



{{< pagebreak >}}


##  Appendix A:  Imbalance of Confounding Covariates Treatments {#appendix-exposures}

@fig-match_1 shows imbalance of covariates on the treatment at the treatment wave. The variable on which there is strongest imbalance is the baseline measure of religious service attendance. It is important to adjust for this measure both for confounding control and to better estimate an incident exposure effect for the religious service at the treatment wave (in contrast to merely estimating a prevalence effect). See @vanderweele2020. 

```{r}
#| label: fig-match_1
#| fig-cap: "This figure shows the imbalance in covariates on the treatment" 
#| eval: true 
#| include: true 
#| echo: false 
#| fig-width: 10 
#| fig-height: 10 

love_plot_one <- margot::here_read("love_plot_one")
love_plot_one
```



tbl-table-exposures-code presents baseline (NZAVS time 10) and exposure wave (NZAVS time 11) statistics for the exposure variable: weekly religious service. Graph was produced using the `WeightIt` and `Cobalt` packages in R [@greifer2023a; @greifer2023b].



## References {.appendix}















