---
title: "Evidence for Declining Trust in Science From A Large National Panel Study in New Zealand (years 2019-2024)"
abstract: |
   The public's perception of science has wide-ranging effects, from the adoption public health behaviours to climate action. This study uses nationally representative panel data from New Zealand to assess changes in trust in science and trust in scientists between late 2019 and late 2024. We draw on a large cohort (N = 42,681, New Zealand Attitudes and Values Study). Study 1 uses multiple imputation to address biases from systematic attrition, ensuring more accurate population estimates of trust. Study 2 models average trust responses over time, showing that trust in science and scientists rose following New Zealand’s Covid-19 response, but later declined. We also find evidence for variation above and below the mean. Study 3 assesses proportional shifts across low, medium, and high levels of trust, revealing rising mistrust at both the lower and higher ends. Considering these shifts at a population level suggests that nearly 60,000 New Zealanders who once held moderate trust may now exhibit low trust in science.
   **KEYWORDS**: *Conservativism*; *Institutional Trust*; *Longitudinal*; *Panel*; *Political*; *Science*.
author: 
  - name: Authors 
    affiliation: New Zealand
    orcid: 0000-0003-3169-6576 
editor_options: 
    chunk_output_type: console
format:
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % Adjust this value as needed
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
library(margot)
packageVersion(pkg = "margot")
library(boilerplate)

# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# combo_marginal_graph_ord
# combo_pols_graph_ord
# combo_eth_graph_ord
# combo_male_graph_ord


# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("tinytable")
# set seed
options(tinytable_tt_digits = 3)

# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)


# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-john-kerr-trust-science-growth-simple")

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)

measures_path_rstudio <- here::here("manuscripts", "24-john-kerr-growth-trust-science")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))


baseline_vars <- margot::here_read("baseline_vars")
exposure_vars <- margot::here_read("exposure_vars")
outcome_vars <- margot::here_read("outcome_vars")
n_participants <-here_read("n_participants")
# read waves
all_waves <- c(2019:2023)

# get table/graphs
summary_vars <- margot::here_read_qs("summary_vars", push_mods)

# hist outcomes
histogram_outcomes <- here_read_qs("histogram_trust_science_trust_scientists_by_wave", push_mods)

# get range of pol conservative
max_pol <- here_read('max_pol')
min_pol<- here_read('min_pol')

max_pol <- round(max_pol,2)
min_pol <- round(min_pol,2)
#outcome table means

# # read cat
# tbl_combined_cat <- here_read_qs("tbl_combined_cat", push_mods)
# 
# # first model
# tbl_combined_cat
# # first imputed dataset
# tbl_combined_imp <- here_read_qs("tbl_combined_imp", push_mods)
# tbl_combined_imp
# 
# # mean
# summary_table_mean <- here_read_qs("summary_table_mean", push_mods)
# summary_table_mean_imp <- here_read_qs("summary_table_mean_imp", push_mods)
# 
# summary_table_mean
# summary_table_mean_imp

tt_summary_mean_response_sample <- here_read("tt_summary_mean_response_sample")
tt_summary_mean_response_sample
tt_summary_mean_response_sample_imp <- here_read("tt_summary_mean_response_sample_imp")
tt_summary_mean_response_sample_imp


# alluvial plots
# tba
sankey_plot_science_sample<-  here_read_qs("sankey_plot_science_sample", push_mods)
sankey_plot_st_sample <-  here_read_qs("sankey_plot_st_sample", push_mods)
sankey_plot_science_imp<-  here_read_qs("sankey_plot_science_imp", push_mods)
sankey_plot_st_imp<-  here_read_qs("sankey_plot_st_imp", push_mods)
tt_summary_table_cat<- here_read("tt_summary_table_cat")
tt_summary_table_cat_imp <- here_read("tt_summary_table_cat_imp")
tt_summary_table_cat
# tt_summary_table_cat_imp |> print("markdown")

# n_total
n_total <- here_read("n_total")

scale_outcome <- glue::glue("(1-7)")
outcome_lower_bound <- glue::glue("1")
outcome_upper_bound <- glue::glue("7")
# shift =   glue::glue('two points')
nice_baseline_wave <- glue::glue("New Zealand Attitudes and Values Study Time 11, years 2019-2024")
nice_all_waves <- glue::glue("New Zealand Attitudes and Values Study Time Time 11-14, years 2019-2024")
nice_end_of_study_wave <- glue::glue("Zealand Attitudes and Values Study Time 14, years 2022-2024")

# not used
prop_missing_baseline <- here_read("prop_missing_baseline")
# percentage_missing_baseline <- glue::glue(round(percent_missing_baseline,2))
# percentage_missing_baseline
# labels
#

# read models

# read models -------------------------------------------------------------
# nnet. 
# trust_science_marginal <- here_read("trust_science_marginal")
# trust_scientists_marginal <- here_read("trust_scientists_marginal")
# trust_science_pols <- here_read("trust_science_pols")
# trust_scientists_pols <- here_read("trust_scientists_pols")
# trust_science_eth <- here_read("trust_science_eth")
# trust_scientists_eth <- here_read("trust_scientists_eth")
# trust_science_male <- here_read("trust_science_male")
# trust_scientists_male <- here_read("trust_scientists_male")
# trust_science_marginal

# gee
# trust_science_marginal_gee <- here_read("trust_science_marginal_gee")
# trust_scientists_marginal_gee <- here_read("trust_scientists_marginal_gee")
# trust_science_pols_gee <- here_read("trust_science_pols_gee")
# trust_scientists_pols_gee <- here_read("trust_scientists_pols_gee")
# trust_science_eth_gee <- here_read("trust_science_eth_gee")
# trust_scientists_eth_gee <- here_read("trust_scientists_eth_gee")
# trust_science_male_gee <- here_read("trust_science_male_gee")
# trust_scientists_male_gee <- here_read("trust_scientists_male_gee")

predictions_gee_science_observed <- here_read("predictions_gee_science_observed")
predictions_gee_scientists_observed <- here_read("predictions_gee_scientists_observed")


predictions_gee_science_imputed <- here_read("predictions_gee_science_imputed")
predictions_gee_scientists_imputed <- here_read("predictions_gee_scientists_imputed")

predictions_gee_science_imputed
predictions_gee_scientists_observed

# plot ordinal marginal ---------------------------------------------------

plot_trust_science_observed_continuous <- plot (
  predictions_gee_science_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Trust in Science (1-7)", title = "Average Trust in Science: Retained Sample") + theme_bw() 

plot_trust_scientists_observed_continuous <- plot (
  predictions_gee_scientists_observed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Average Trust in Scientists: Retained Sample") + theme_bw() 


plot_trust_science_imputed_continuous <- plot (
  predictions_gee_science_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x = "Years: 2019-2024", y = "Trust in Science (1-7)", title = "Trust in Science: Imputed Missing Responses") + theme_bw() 



plot_trust_scientists_imputed_continuous <- plot (
  predictions_gee_scientists_imputed,
  show_data = TRUE,
  jitter = .4,
  dot_alpha  = .025,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(.5, 7.5),
  dot_size = 2)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (1-7)", title = "Trust in Scientists: Imputed Missing Responses") + theme_bw()



# patchwork graph
combo_science_continuous<-   plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study): ", tag_levels = "A", title = "Trust in Science: Observed vs Imputed Missing Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  


combo_scientists_continuous <- plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous   + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Science: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom') 


combo_all_continuous <- (plot_trust_science_observed_continuous + plot_trust_science_imputed_continuous)/ (plot_trust_scientists_observed_continuous + plot_trust_scientists_imputed_continuous) + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2023), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Science and Scientists over Time: Observed vs Imputed Continuous Responses") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  

# margot::margot_save_png(combo_all,base_filename = "trust_science_observed_vs_imputed", save_path = here::here(push_mods), height=20,width=20, dpi = 500)


# Categorical-------------------------------------------------------------------
# 
observed_trust_science_factor <- here_read("observed_trust_science_factor")
observed_trust_scientists_factor <- here_read("observed_trust_scientists_factor")

predictions_ord_science_imp <- here_read("predictions_ord_science_imp")
predictions_ord_scientists_imp <- here_read("predictions_ord_scientists_imp")

observed_trust_science_factor |> print_md()

# plot
observed_trust_science_factor_plot =  plot(
  observed_trust_science_factor,
  jitter = .4,
  dot_alpha  = .025,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Science: Retained Sample", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()

observed_trust_scientists_factor_plot =  plot(
  observed_trust_scientists_factor,
  alpha = 3,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "quadro",
  limits = c(0, 1),
  dot_size = 2
)+ geom_point(aes(x = x, y = predicted), color = "red", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Scientists (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Retained Sample",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()




imputed_trust_science_factor_plot <- plot(
  predictions_ord_science_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
  )+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Science: Imputed Missing Responses", 
       subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()


imputed_trust_science_factor_plot

imputed_trust_scientists_factor_plot <- plot(
  predictions_ord_scientists_imp,
  show_ci = TRUE,
  ci_style = "dash",
  colors = "us",
  limits = c(0, 1)
)+ geom_point(aes(x = x, y = predicted), color = "dodgerblue", size = .2, alpha =1)  +
  labs(x =  "Years: 2019-2024", y = "Trust in Science (low = 1-3, med = 4-5, high = 6-7)", title = "Average Trust in Scientists: Imputed Missing Responses",
            subtitle = "Predicted Probability of Response: Low = 1-3, Medium = 4-5, High = 6-7") + theme_bw()  +   theme(
    plot.title = element_text(face = "bold", size = 14))


combo_science_factor <- observed_trust_science_factor_plot  + imputed_trust_science_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Science: Observed vs Imputed Missing Responses: Multinomial Response Model\n'Our Society Places Too Much Emphasis on Science'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

combo_scientists_factor <- observed_trust_scientists_factor_plot  + imputed_trust_scientists_factor_plot + plot_annotation(subtitle = "NZAVS waves 11-14 (years 2019-2024), N = 42,681 (New Zealand Attitudes and Values Study)", tag_levels = "A", title = "Trust in Scientists: Observed vs Imputed Missing Responses: Multinomial Response Model\n'I have a high degree of confidence in the scientific community'") +
  plot_layout(guides = "collect") &  theme(legend.position = 'bottom')  +   theme(
    plot.title = element_text(face = "bold", size = 14))

```



## Introduction



<!-- ::: {#tbl-categorical-science-observed} -->
<!-- ```{r, results='asis'} -->
<!-- #| eval: true -->
<!-- #| include: true -->
<!-- #| echo: false -->
<!-- observed_trust_science_factor_x <- format(observed_trust_science_factor,  collapse_tables = FALSE, collapse_ci = TRUE, -->
<!--  digits = 2)  -->

<!-- observed_trust_science_factor_x|>  -->
<!--      as.data.frame() %>%  -->
<!--     dplyr::rename(Years = years) %>%     -->
<!--     dplyr::rename(Response = groups) %>%     -->
<!--   kbl( -->
<!--     format = "latex", -->
<!--     booktabs = TRUE, -->
<!--     digits = 3, -->
<!--     row.names = FALSE                      # remove row names to eliminate any unintended columns -->
<!--   ) |> -->
<!--   kable_styling( -->
<!--     font_size = 12, -->
<!--     latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h] -->
<!--   ) -->
<!-- ``` -->
<!-- Predicted probability of classification for trust in science without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7). -->
<!-- ::: -->



### Study 2: Categorical Mistrust in Science and Scientists over Time


```{r}
#| label: fig-science-categorical
#| fig-cap: "(A) Predicted probability of classification for trust in science without adjusting for missingness (B) Predicted probability of classification for trust in science adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

combo_science_factor
```

::: {#tbl-categorical-science-observed}
```{r, results='asis'}
#| eval: true
#| include: true
#| echo: false
observed_trust_science_factor_x <- format(observed_trust_science_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

observed_trust_science_factor_x|> 
     as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 12,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in science without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::



::: {#tbl-categorical-science-imputed}
```{r}
#| eval: true
#| include: true
#| echo: false
predictions_ord_science_imp_x <- format(predictions_ord_science_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

predictions_ord_science_imp_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 12,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )


```
Predicted probability of classification for trust in science, now adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


@fig-science-categorical displays predicted probabilities for being classified as low (1–3), medium (4–5), or high (6–7) in trust in science over time. @tbl-categorical-science-observed presents these probabilities without adjusting for missing responses. Under these conditions, the proportion of respondents in the high category appears to increase from the baseline and remain relatively elevated. However, after adjusting for missingness (see @tbl-categorical-science-imputed), the predicted probabilities show a less stable pattern, with gains in the high category tempered and the medium category becoming more prominent. These differences underscore that ignoring missing responses can create an artificial appearance of stronger sustained trust.





```{r}
#| label: fig-scientists-categorical
#| fig-cap: "(A) Predicted probability of classification for trust in scientists without adjusting for missingness (B) Predicted probability of classification for trust in scientists adjusting for missingness. Responses categories are low (1-3), medium (4-5), high (6-7) on the 1-7 ordinal scale."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12

combo_scientists_factor
```

::: {#tbl-categorical-scientists-observed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
observed_trust_scientists_factor_x <- format(observed_trust_scientists_factor,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

observed_trust_scientists_factor_x|> 
     as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 12,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists without adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


::: {#tbl-categorical-scientists-imputed}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false
predictions_ord_scientists_imp_x <- format(predictions_ord_scientists_imp,  collapse_tables = FALSE, collapse_ci = TRUE,
 digits = 2) 

predictions_ord_scientists_imp_x |> 
 as.data.frame() %>% 
    dplyr::rename(Years = years) %>%    
    dplyr::rename(Response = groups) %>%    
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 12,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Predicted probability of classification for trust in scientists, now adjusting for missing responses. Responses categories are low (1-3), medium (4-5), high (6-7).
:::


A similar pattern emerges for trust in scientists (see @fig-scientists-categorical). Without adjusting for missingness, the results (shown in @tbl-categorical-scientists-observed) suggest that the proportion classified as high in trust remains consistently large. Once we incorporate multiple imputation to address missing responses, as presented in @tbl-categorical-scientists-imputed, the predicted probabilities shift. The high category becomes less dominant over time, and more respondents fall into the medium category than would have been inferred without proper adjustment. In other words, accounting for nonresponse reveals a more nuanced pattern of trust that does not simply trend upward, and this correction prevents inflated estimates of stable or increasing trust in scientists.

## Discussion 

These findings illustrate that failing to adjust for missing responses can systematically distort estimates of public trust in science and scientists. Compared to models that ignore missingness, our imputation-adjusted analyses indicate that the unadjusted models overestimate the proportion of the population with high trust by several hundred thousand New Zealand adults and underestimate the proportion with low trust by tens to hundreds of thousands. For example, when trust in science is not adjusted for missingness, we appear to be underestimating the share of the population with low trust by about 82,000 people, while simultaneously overstating the size of the high-trust group by roughly 330,000. Similarly, for trust in scientists, the discrepancy in the low-trust population is about 165,000 individuals, and the high-trust category may be inflated by about 248,000. These differences represent substantial distortions in population-level estimations.

It is important to note that these are minimum bounds of potential misestimation, as we cannot measure trust among individuals who declined participation at the study’s outset. Moreover, we have not established any causal relationships, and our objective is not to identify determinants of changing trust, but rather to describe and characterise differences in trust distributions over time. Still, given that trust in science can influence critical actions related to climate change, vaccine uptake, and following health advice, accurately documenting trust and its temporal shifts is a pressing concern. By accounting for missing data, we gain a clearer and more credible picture of population-level trust in science and scientists, enabling better-informed public health messaging and policy decisions.

