---
title: "Religious Service Attendance Increases Meaning, Gratitude, and Sexual Satisfaction (But Not Much Else) in Secular New Zealand"
abstract: |
  We use nationally representative longitudinal data from 46,377 New Zealanders (2018–2023) to estimate the causal effects of religious service attendance on multiple dimensions of well-being. We compare two hypothetical four-year scenarios: (a) everyone attends services at least four times per month, and (b) no one attends. We adjust for baseline covariates measured before these interventions, including baseline service attendance and baseline measurements of multi-dimensional well-being. We also adjust for time-varying counfounders such as employment, disability, relationships, and parenting. We assess outcomes in the year following these sequential treatment regimes. Using cross-validation and doubly robust machine-learning estimators, we find that regular attendance increases a sense of meaning and purpose, enhances gratitude, and, somewhat unexpectedly, improves sexual satisfaction. Effects on mental health are modest, and effects on biological health and social well-being are negligible or small. In a mostly secular population, attending religious services modestly boosts meaning, gratitude, and sexual satisfaction, and more strongly increases a sense of purpose. Beyond these gains, religious attendance yields few other measurable benefits to health and well-being.
  
  **KEYWORDS**: *Causal Inference*; *Church*; *Cross-validation*; *Distress*; *Health*; *Longitudinal*; *Machine Learning*; *Religion*; *Semi-parametric*; *Targeted Learning*.
author: 
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
  - name: Don E Davis
    affiliation: Georgia State University, Matheny Center for the Study of Stress, Trauma, and Resilience
    orcid: 0000-0003-3169-6576 
  - name: Cyrstal Park
    affiliation: University of Connecticut, Department of Psychological Sciences
    orcid: 0000-0001-6572-7321
    email: crystal.park@uconn.edu
  - name: Kenneth G. Rice
    affiliation: Georgia State University, Matheny Center for the Study of Stress, Trauma, and Resilience
    orcid: 0000-0002-0558-2818
  - name: Geoffrey Troughton
    affiliation: School of Social and Cultural Studies, Victoria University of Wellington
    orcid: 0000-0001-7423-0640
  - name: Daryl R. Van Tongeren
    affiliation: Hope College
    orcid: 0000-0002-1810-9448
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland
    orcid: 0000-0002-4064-8800
keywords: [Use, use]
editor_options: 
    chunk_output_type: console
format:
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
      - /Users/joseph/GIT/templates/measures/data
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % Adjust this value as needed
  docx:
    default: false
date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# detach("package:margot", unload = TRUE)
#devtools::install_github("go-bayes/boilerplate")
# for making graphs
#   docx:
#    default: false
# reproducibility
set.seed(123)

# for latex graphs
# for making graphs
library("tinytex")
library("extrafont")
library("tidyverse")
library("kableExtra")
library("gtsummary")


# fonts
loadfonts(device = "all")

# rendering document
library("margot")
library("boilerplate")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("table1")
library("printr")
packageVersion(pkg = 'margot')

# measures path
measures_path <- fs::path_expand("/Users/joseph/GIT/templates/measures/data/merged_db.rds")

# read measures data
measures_data <- read_rds(measures_path)
measures_data
# set data paths
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-jb-church-fl-long")


# missing at baseline
percent_missing_baseline <- margot::here_read("percent_missing_baseline")



# read variable names
baseline_vars <- margot::here_read("baseline_vars")
exposure_var <- margot::here_read("exposure_var")
outcome_vars <- margot::here_read("outcome_vars")


# read by domain
raw_outcomes_health <- here_read("raw_outcomes_health")
raw_outcomes_psych <- here_read("raw_outcomes_psych")
raw_outcomes_present <- here_read("raw_outcomes_present")
raw_outcomes_life <- here_read("raw_outcomes_life")
raw_outcomes_social <- here_read("raw_outcomes_social")

raw_outcomes_life
# exposure
raw_exposure_vars = "religion_church"

# baseline vars
baseline_vars

# exposure
raw_exposure_vars

#outcomes
raw_outcomes_health

# labels
var_labels_baseline <- c(
  "sdo" = "Social Dominance Orientation",
  "born_nz_binary" = "Born in NZ",
  "rural_gch_2018_l" = "Rural Gch 2018 Levels",
  "education_level_coarsen" = "Education Level",
  "employed_binary" = "Employed (binary)",
  "eth_cat" = "Ethnicity",
  "household_inc" = "Household Income",
  "log_household_inc" = "Log Household Income",
  "male_binary" = "Male (binary)",
  "nz_dep2018" = "NZ Deprevation Index 2018",
  "nzsei_13_l" = "NZSEI (Occupational Prestige Index)",
  "parent_binary" = "Parent (binary)",
  "partner_binary" = "Partner (binary)",
  "rwa" = "Right Wing Authoritarianism",
  "sample_frame_opt_in_binary" = "Sample Frame Opt-In (binary)",
  "sdo" = "Social Dominance Orientation",
  "smoker_binary" = "Smoker (binary)",
  "support" = "Social Support (perceived)"
)


# labels
label_mapping_raw_exposure = c("religion_church" = "Monthly Religious Service")


# labels
label_mapping_raw_exposure = c("religion_church" = "Monthly Religious Service")

# outcomes
label_mapping_raw_health <- list(
  "hlth_bmi" = "BMI", 
  "hlth_sleep_hours" = "Sleep", 
  "log_hours_exercise" = "Hours of Exercise (log)",
  "short_form_health" = "Short Form Health" 
)

label_mapping_raw_psych<- list(
  "hlth_fatigue" = "Fatigue", 
  "kessler_latent_anxiety" = "Anxiety", 
  "kessler_latent_depression" = "Depression",  
  "rumination" = "Rumination"
)

label_mapping_raw_present<- c(
  "bodysat" = "Body Satisfaction",
  "foregiveness" = "Forgiveness",
  "perfectionism" = "Perfectionism",  
  "pwb_standard_living" = "PWB Standard Living", 
  "pwb_your_security" = "PWB Your Future Security",
  "pwb_your_health" = "PWB Your Health", 
  "pwb_your_relationships" = "PWB Your Relationships",
  "self_control_have_lots" = "Self Control Have Lots", 
  "self_control_wish_more_reversed" = "Self Control Wish More Reversed",  
  "self_esteem" = "Self-Esteem", 
  "sexual_satisfaction" = "Sexual Satisfaction"  )

label_mapping_raw_life <- list(
  "gratitude" = "Gratitude", 
  "lifesat" = "Life Satisfaction", 
  "meaning_purpose" = "Meaning Purpose", 
  "meaning_sense" = "Meaning Sense"
)

label_mapping_raw_social <- list(
  "belong" = "Belonging",
  "neighbourhood_community" = "Community Belonging", 
  "support" = "Support" 
)


label_mappings_all <- list(
  baseline = var_labels_baseline,
  exposure = label_mapping_raw_exposure,
  health = label_mapping_raw_health,
  psychological = label_mapping_raw_psych,
  present_focussed = label_mapping_raw_present,
  life_focussed = label_mapping_raw_life,
  social = label_mapping_raw_social
)


all_outcomes <- list(
  health          = raw_outcomes_health,
  psychological   = raw_outcomes_psych,
  present_focussed         = raw_outcomes_present,  # changed to match label_mappings_all
  life_focussed            = raw_outcomes_life,     # changed to match label_mappings_all
  social          = raw_outcomes_social
)

appendix_text_version_1 <- boilerplate_measures(
  baseline_vars = baseline_vars,
  exposure_var = raw_exposure_vars,
  outcome_vars = all_outcomes,
  measure_data = measures_data,
  label_mappings = label_mappings_all
  )

cat(appendix_text_version_1)

# reproducibility
set.seed(123)

# import untransformed data for graphs
original_df <- margot::here_read_qs('df_wide_impute', push_mods)

# summary tables
summary_tables <- margot::here_read_qs("summary_tables", push_mods)

# read nzavs timeline
timeline_histogram <- margot::here_read_qs("timeline_histogram", push_mods)


# read latex tables
latex_table_baseline <- margot::here_read("latex_table_baseline")
latex_table_exposures <- margot::here_read("latex_table_exposures")
latex_table_outcomes_health <- margot::here_read("latex_table_outcomes_health")
latex_table_outcomes_psych <- margot::here_read("latex_table_outcomes_psych")
latex_table_outcomes_present <- margot::here_read("latex_table_outcomes_present")
latex_table_outcomes_life <- margot::here_read("latex_table_outcomes_life")
latex_table_outcomes_social <- margot::here_read("latex_table_outcomes_social")




# read exposure vars names
t0_name_exposure <- here_read('t0_name_exposure')
t1_name_exposure <- here_read('t1_name_exposure')
t2_name_exposure <- here_read('t2_name_exposure')
t3_name_exposure <- here_read('t3_name_exposure')
t4_name_exposure <- here_read('t4_name_exposure')

# names of waves
baseline_wave <- margot::here_read("baseline_wave")
exposure_waves <- margot::here_read("exposure_waves")
outcome_wave <- margot::here_read("outcome_wave")


## Read in data from analysis 
# imports
n_total <-  margot::here_read("n_total")

# total
n_total

# participants
n_participants <- summary_tables$n_participants

# make pretty number
n_participants<- prettyNum(n_participants,big.mark=",")
n_total<- prettyNum(n_total,big.mark=",")

# participants
n_participants

# library
library(margot)

# verify positivity 
#church
#transition_table
# binary
transition_matrix <- margot::here_read("transition_matrix")


# label mapping
# label mappings for health outcomes
label_mapping_health <- list(
  "t5_hlth_bmi_z" = "BMI", 
  "t5_hlth_sleep_hours_z" = "Sleep", 
  "t5_log_hours_exercise_z" = "Hours of Exercise (log)",
  "t5_short_form_health_z" = "Short Form Health" 
)

# label mappings for psychological well-being outcomes
label_mapping_psych <- list(
  "t5_hlth_fatigue_z" = "Fatigue", 
  "t5_kessler_latent_anxiety_z" = "Anxiety", 
  "t5_kessler_latent_depression_z" = "Depression",  
  "t5_rumination_z" = "Rumination"
)

# label mappings for present reflective outcomes
label_mapping_present <- list(
  "t5_bodysat_z" = "Body Satisfaction",
  "t5_foregiveness_z" = "Forgiveness",  
  "t5_perfectionism_z" = "Perfectionism",  
  "t5_pwb_standard_living_z" = "PWB Standard Living", 
  "t5_pwb_your_future_security_z" = "PWB Your Future Security", 
  "t5_pwb_your_health_z" = "PWB Your Health", 
  "t5_pwb_your_relationships_z" = "PWB Your Relationships",
  "t5_self_control_have_lots_z" = "Self Control Have Lots", 
  "t5_self_control_wish_more_reversed_z" = "Self Control Wish More Reversed",  
  "t5_self_esteem_z" = "Self Esteem", 
  "t5_sexual_satisfaction_z" = "Sexual Satisfaction"
)

# label mappings for life reflective outcomes
label_mapping_life <- list(
  "t5_gratitude_z" = "Gratitude", 
  "t5_lifesat_z" = "Life Satisfaction", 
  "t5_meaning_purpose_z" = "Meaning: Purpose", 
  "t5_meaning_sense_z" = "Meaning: Sense"
)

# label mappings for social outcomes
label_mapping_social <- list(
  "t5_belong_z" = "Belonging",
  "t5_neighbourhood_community_z" = "Community", 
  "t5_support_z" = "Support" 
)




# plot options ------------------------------------------------------------
nice_exposure_name <- glue::glue("Monthly Religious Service Rates")
title <- glue::glue("Causal Effects of {nice_exposure_name} on Multi-Dimensional Well-Being")


options_base <- list(
  type = "RD",
  title = title,
  order  = "alphabetical",
  colors = c(
    "positive" = "#E69F00",
    "not reliable" = "grey50",
    "negative" = "#56B4E9"
  ),
  x_offset = -1.05, # will be set based on type
  x_lim_lo = -1, # will be set based on type
  x_lim_hi = .6, # will be set based on typee
  text_size = 6,
  linewidth = 0.5,
  estimate_scale = 1,
  base_size = 20,
  point_size = 5,
  title_size = 18,
  subtitle_size = 18,
  legend_text_size = 10,
  legend_title_size = 10
)


# plote options
options_health_religious_vs_secular <- margot_plot_create_options(
  subtitle = "Health: Religious vs Secular (baseline)",
  base_defaults = options_base,
  title = title,
  filename_prefix = "health_religious_vs_secular"
)

options_psych_religious_vs_secular <- margot_plot_create_options(
  subtitle = "Psychological Well-Being: Religious vs Secular (baseline)",
  base_defaults = options_base,
  title = title,
  filename_prefix = "psych_religious_vs_secular"
)


options_present_religious_vs_secular <- margot_plot_create_options(
  subtitle = "Present-Focussed Well-Being: Religious vs Secular (baseline)",
  base_defaults = options_base,
  title = title,
  filename_prefix = "present_religious_vs_secular"
)

options_life_religious_vs_secular <- margot_plot_create_options(
  subtitle = "Life Focussed Well-Being: Religious vs Secular (baseline)",
  base_defaults = options_base,
  title = title,
  filename_prefix = "life_religious_vs_secular"
)


options_social_religious_vs_secular <- margot_plot_create_options(
  subtitle = "Social Well-Being: Religious vs Secular (baseline)",
  base_defaults = options_base,
  title = title,
  filename_prefix = "social_religious_vs_secular"
)


# read models
health_lmtp_output <- margot::here_read_qs('health_lmtp_output', push_mods)
psych_lmtp_output <- margot::here_read_qs('psych_lmtp_output', push_mods)
present_lmtp_output <- margot::here_read_qs('present_lmtp_output', push_mods)
life_lmtp_output<- margot::here_read_qs('life_lmtp_output', push_mods)
social_lmtp_output <- margot::here_read_qs('social_lmtp_output', push_mods)

# graphs

# religion vs secular: health
health_religious_vs_secular <- margot_plot(
   health_lmtp_output$combined_tables$combined_outcomes_religious_vs_secular,
   options = options_health_religious_vs_secular,
   save_output = F,
   label_mapping = label_mapping_health,
   save_path = here::here(push_mods),
   original_df = original_df
 )



# religion vs secular: psych
psych_religious_vs_secular <- margot_plot(
  psych_lmtp_output$combined_tables$combined_outcomes_religious_vs_secular,
  options = options_psych_religious_vs_secular,
  save_output = F,
  label_mapping = label_mapping_psych,
  save_path = here::here(push_mods),
  original_df = original_df
)



# religion vs secular: present
present_religious_vs_secular <- margot_plot(
  present_lmtp_output$combined_tables$combined_outcomes_religious_vs_secular,
  options = options_present_religious_vs_secular,
  save_output = F,
  label_mapping = label_mapping_present,
  save_path = here::here(push_mods),
  original_df = original_df
)

# religion vs secular: life
life_religious_vs_secular <- margot_plot(
  life_lmtp_output$combined_tables$combined_outcomes_religious_vs_secular,
  options = options_life_religious_vs_secular,
  save_output = F,
  label_mapping = label_mapping_life,
  save_path = here::here(push_mods),
  original_df = original_df
)


# religion vs secular: life
social_religious_vs_secular <- margot_plot(
  social_lmtp_output$combined_tables$combined_outcomes_religious_vs_secular,
  options =  options_social_religious_vs_secular,
  save_output = F,
  label_mapping = label_mapping_social,
  save_path = here::here(push_mods),
  original_df = original_df
)



```


## Introduction

Collective religious activity is a cornerstone of many cultures, and questions about whether attending religious services affects well-being are long-standing [@koenig2012handbook]. Within religious traditions, the belief that religious practices benefit adherents permeates theology, moral teaching, and community life, from Buddhism's emphasis on meditation as a path to enlightenment and reduced suffering to Christianity's view of prayer and worship as a means of spiritual growth and divine connection, to Indigenous spiritual practices that are seen as maintaining harmony between people, nature, and the sacred. More recently, empirical science has linked religious service attendance with multiple dimensions of well-being, including social integration, life satisfaction, mental health, and meaning-making [@vanderweele2022invited_church; @chen2020religious; @dunbar2021religiosity; @highland2022national]. These studies have demonstrated that, on average, individuals who attend services report more favourable social and psychological outcomes. Is this association causal?  For whom?  

Although recent studies have applied causal methods [@pawlikowski2019religious; @vanderweele2021effectsReligiousServiceMetanalysis; @vanderweele2020vchenrespond; @vanderweelechurchmortality], debates over their implications persist [@highland2019attitudes]. Some investigations do not fully adjust for baseline health or baseline religious service attendance, nor other critical confounders affecting both religious participation and well-being [@brown2023_church; @jokela2024_church]. For instance, in their analysis of four U.S. cohort studies (N=6,592), Jokela and Laakasuo found that poorer health trajectories preceded religious disengagement, suggesting that declining practice might reflect underlying health problems rather than cause them. Similarly, Iyer and Rosso's global data ($N=65,564$) show that frequent attenders report about five per cent higher life satisfaction than infrequent attenders, though unmeasured confounding or reverse causation could explain this association.

Existing work often relies on non-representative samples or predominantly religious settings, rarely incorporates time-varying confounders, and seldom examines populations outside the north atlantic [@price2024science], or in societies with relatively low religious engagement. @brown2023_church propose that community-building itself, rather than specifically religious community-building, may explain well-being benefits. Supporting evidence from New Zealand shows that following the 2010–2011 Christchurch earthquakes, people who lost faith exhibited declining subjective health, whereas those who maintained faith or lacked faith altogether did not [@sibley2012]. These findings hint that broader communal factors -- not necessarily religious practice -- could drive well-being, especially in secular contexts where religious service might not confer distinctive advantages. Moreover, Chen and colleagues observe that well-being encompasses emotional, physical, social, and psychological domains and that some—such as meaning, purpose, and social connection—may be more sensitive to religious engagement than others [@chen2022_church_longitudinal]. However, most studies on religious attendance and well-being focus on just one or a few of these domains.

In sum, there is a need for studies that move beyond cross-sectional correlations to address causation, address causation robustly, and systematically evaluate multiple dimensions of well-being, particularly in secular societies. Without such evidence, policy, clinical, and theoretical questions about the causal role of religious service attendance in supporting human flourishing remain unanswered.

Here, we address these gaps by estimating the causal effects of regular religious service attendance on multidimensional well-being using data from a large, nationally diverse sample of New Zealanders (baseline $N=46,377$) enrolled in the New Zealand Attitudes and Values Study (NZAVS). New Zealand is characterised by high religious diversity and secularisation. Over half the population identifies as having no religion, while multiple religious traditions -- Christianity, Hinduism, Islam, Buddhism, and others -- coexist [@shaver2016religion]. We draw on an informationally rich national panel study and recent innovations in causal inference [@hernan2024WHATIF; @williams2021; @bulbulia2023] and systematically examine whether a settings in which the adult population attended services regularly differs in its effects across multi-dimensional well-being from a setting in which no-one attends. We estimate these effects using a workflow in which non-parametric machine learning reduces statistical modelling assumptions. We contribute to understanding by estimating causal effects in a distinctly secular, and culturally diverse context and obtain generalisable inferences about whether religious service attendance causally affects a well-being phenotype across dimensions of biological health, psychological health, person-focussed well-being, life-focussed well-being and social well-being. 

## Method

### Sample
Data were collected as part of the New Zealand Attitudes and Values Study (NZAVS), an annual longitudinal national probability panel assessing New Zealand residents’ social attitudes, personality, ideology, and health outcomes. The panel began in 2009 and has since expanded to include over fifty researchers, with responses from `r n_total` participants to date. The study operates independently of political or corporate funding and is based at a university. It employs prize draws to incentivise participation. The NZAVS tends to slightly under-sample males and individuals of Asian descent and to over-sample females and Māori (the Indigenous people of New Zealand). To enhance the representativeness of our sample population estimates for the target population of New Zealand, we apply census-based survey weights that adjust for age, gender, and ethnicity (New Zealand European, Asian, Māori, Pacific) [@sibley2021]. For more information about the NZAVS, visit: [OSF.IO/75SNB](https://doi.org/10.17605/OSF.IO/75SNB). Refer to [Appendix A](#appendix-timeline) for a histogram of daily responses for this cohort.


### Target Population

The target population for this study comprises the cohort of New Zealand residents in New Zealand Attitudes and Values Study wave 10 (years 2018-2019) [@sibley2021].

### Treatment Indicator


The New Zealand Attitudes and Values Study assesses religious service attendance using the following question:


- *Do you identify with a religion and/or spiritual group? If yes, how many times did you attend a church or place of worship during the last month?*


We rounded response to the nearest whole number. Because few participants reported attending more than eight times, we capped responses above eight at eight (refer to [Appendix B](#appendix-baseline), variable `religion_church_round`).

### Baseline Covariates

We adjusted for a rich set of demographic, personality, and behavioural indicators measured at the baseline wave, NZAVS time 10 (Wave 2018, years 2018-2019) (see [Appendix B](#appendix-baseline) for full measures). These variables included age, gender, ethnicity, education level, personality traits (Agreeableness, Conscientiousness, Extraversion, Honesty-Humility, Neuroticism, and Openness), household income, employment status, parenting status, relationship status, religious belonging, and health-related behaviours (e.g. smoking, alcohol use, hours spent exercising). We selected only those outcome variables measured in the baseline wave and controlled for these variables. Moreover we controlled for religious service attendance at baseline (refer to [Appendix B](#appendix-baseline) and [Appendix C](#appendix-confounding)) This strategy of confounding control is powerful because for any confounder to affect subsequent treatments and the outcome, it would need to do so independently of the baseline outcome variables, the baseline exposure, and the rich set of demographic indicators measured at baseline [@vanderweele2020].


### Outcomes

We investigate well-being using an VanderWeele and colleague's "outcome-wide" approach. We categorised outcomes into five domains—health, psychological well-being, present-reflective outcomes, life-reflective outcomes, and social outcomes—based on validated scales and measures. Outcomes were based on those modelled in an earlier outcome-wide paper @pedro_2024effects. @tbl-outcomes summarises each domain and its associated measures. For instance, health outcomes included BMI and hours of sleep, whereas psychological well-being included anxiety and depression. Outcomes were converted to z-scores (standardised), and the causal effect estimates may be therefore be interpreted as effect sizes.


| Domain          | Example Measures                           |
|-----------------|--------------------------------------------|
| Health          | BMI, Hours of Sleep, Hours of Exercise, SF-Health   |
| Psychological Well-Being | Anxiety, Depression, Fatigue, Rumination         |
| Present-Reflective        | Body Satisfaction, Forgiveness, Self-Esteem, Sexual Satisfaction |
| Life-Reflective           | Gratitude, Life Satisfaction, Meaning (Sense & Purpose)           |
| Social                    | Belonging, Support, Community          |

: Outcome domains and example dimensions. Data summaries for all measures used in this study are provided in [Appendix B](#appendix-baseline). {#tbl-outcomes}





### Causal Interventions

When psychologists analyse time-series data, they often use growth models to describe how variables evolve over time. However, many questions are causal: we want to know what would happen if we could intervene on certain variables (such as religious service attendance). To investigate these questions with observational time-series data, we must clearly define our causal question and design our analysis to emulate a hypothetical randomised controlled trial—often called a **target trial** [@hernan2016]. A target trial asks, setting aside practicalities and ethics, *what experiment are we attempting to emulate with our data?* Without explicitly stating this hypothetical experiment, it can be unclear which causal effect we are actually estimating.

Here, we ask:

> "What if, at each wave, we intervened to set religious service attendance to a certain level, and then measured everyone's well-being at the final wave?"

To answer this, we compare two hypothetical interventions. Each intervention shifts religious service attendance across four exposure waves, with outcomes measured after the year following the fourth exposure wave. Again, we control for a rich set of baseline covariates in the wave before the first exposure wave as well as for time-varying confounders at each exposure wave (refer to @@tbl-plan). We using the six most recent NZAVS waves (Times 10–15) because Wave 10 containes the largest cohort, enabling us to maximise power.

Following the modified treatment policies approach, we define **shift functions** describing each intervention:

1. **Regular Religious Service Attendance**  
   At each wave, if attendance is below four times per month, we shift it to four; otherwise, we leave it unchanged.

2. **No Religious Service Attendance**  
   At each wave, if attendance is above zero, we shift it to zero; otherwise, we leave it unchanged.

::: {#tbl-plan}
```{=latex}
\vizfive
```
We contrast well-being outcomes from two treatment regimes: (1) regular attendance for four years and (2) no service attendance for four years. $a^{+}$ denotes regular religious service attendance; $a^{-}$ denotes no attendance. Our statistical models control for baseline-wave confounders, and subsequent time-varying confounders for all exposure waves. We include baseline measurments of religious service attendance and baseline measurements of all outcomes as confounders. We assume that conditional on these confounders, treatment assignment is "as good as random." Outcomes, here denoted $Y_\tau$, are measured in the wave following the final treatment.
:::


We then organise our data to resemble a randomised sequential experiment that assigns each person to one of two longitudinal treatment strategies "regular religious service for four waves" and "no religious service for four waves." We define a "confounder" is a variable that, once included in the model, along with other included variables, removes any non-causal association between the treatment and outcome. Here, as mentioned, we adjust for a rich set of demographic and personality variables, as well as baseline religious service attendance and baseline measures of all outcomes. We also adjust for time-varying confounders at each wave (physical disability, employment status, partner status, and parenting status). We assume these time-varying confounders can influence religious service attendance and later well-being, potentially biasing our estimates. We ensure there is no reverse causation by measuring the outcomes at the end of the study, one year after the final treatment wave. 


### Causal Contrasts

We compared the average expected outcome under the "regular attendance" and "no attendance" regime, estimating contrasts on the difference scale (i.e., subtracting the expected outcome under "no attendance" from that under "regular attendance"). We obtained confidence intervals using the cross-fitted influence-function approach in the `lmtp` package [@williams2021]. This approach employs sequentially doubly robust (SDR) estimator, as developed by @diaz2021_non_parametric_lmtp, which remains valid if either the outcome model or the propensity model is correctly specified, thereby requiring weaker assumptions than standard approaches. By setting up our data as if it came from a hypothetical experiment, we gain clarity about which causal effects we are estimating and as well as confidence about our causal effect estimates (refer to @hernan2024WHATIF, @bulbulia2022; @bulbulia2023).

We estimate this average treatment effect on the difference scale for each outcome across five well-being domains under the assumptions of no-unmeasured confounding, causal consistency, and positivity (refer to [Appendix D](#appendix-assumptions) and [Appendix E](#appendix-transition)). Again, the target population is all adults in New Zealand from the years 2018-2024.


### Comment on Methods for Causal Inference Compared with Growth Curve Analysis

Although both growth models and causal inference methods pertain to questions about changes over time, each method focusses on different questions. Growth models quantify how individuals vary in their natural trajectories. In contrast, causal models explicitly ask: "What happens if we intervene?" -- which of course requires that we define the interventions we seek to compare.  By clearly defining the series of interventions to be compared,  adjusting for confounders, and ensuring that outcomes are measured after the interventions, investigators can rigorously address "what if?" questions, bridging the gap between observational data and actionable insights. Causal inference opens new avenues for psychological scientists to design and evaluate interventions that might guide policy and practice with evidence. Such advice cannot be obtained from statistical models of change over time, no matter how sophisticated [@hernan2024WHATIF].

### Statistical Estimator

We estimate causal effects of time-varying treatment policies using a Sequential Doubly-Robust (SDR) estimator in the `lmtp` package [@williams2021]. SDR proceeds in two steps. First, we use machine learning to flexibly model relationships among treatments, covariates, and outcomes. This approach captures complex, high-dimensional structures without strict assumptions [@díaz2021]. Second, SDR "targets" these initial estimates by incorporating information from the observed data distribution. This step iteratively refines the accuracy of our causal estimates.

The SDR estimator is multiply robust when treatments repeat over multiple waves [@diaz2023lmtp; @hoffman2023]. This design maintains consistency if either the outcome model or treatment model is correctly specified. The `lmtp` package relies on the `SuperLearner` library in R [@SuperLearner2023]. We used `SL.ranger`, `SL.glmnet`, and `SL.xgboost` [@polley2023; @xgboost2023; @Ranger2017] as our base learners. **`SL.ranger`**: implements a random forest algorithm, capturing non-linear relationships and complex interactions. **`SL.glmnet`**: provides regularised linear models for high-dimensional data. **`SL.xgboost`**: uses gradient boosting to capture intricate patterns without over-fitting.

`SuperLearner` combines these learners adaptively to optimise predictive performance. We created graphs, tables, and output reports with the `margot` package [@margot2024]. For more details on targeted learning with `lmtp`, see [@hoffman2022; @hoffman2023; @díaz2021].

### Handling of Missing Data

#### Baseline Missingness

We used predictive mean matching from the `mice` package [@vanbuuren2018] to impute missing baseline values (comprising `r percent_missing_baseline` of the baseline data). Following [@zhang2023shouldMultipleImputation], we performed single imputation using only baseline data. For each column with missing values, we created a binary indicator of missingness so that the machine learning algorithms we employed could condition on missingness information during estimation (see `lmtp` documentation [@williams2021]).

#### Missingness in Time-Varying Variables

When a time-varying value was missing in any wave but a future value was observed, we carried forward the previous response and included a missingness indicator. Again, this approach let the patterns of missingness inform nonparametric machine learnering. If no future value was observed, we considered the participant censored and used inverse probability of treatment weights to address attrition.

#### Outcome Missingness

Finally, to handle confounding and selection bias arising from missing outcomes and panel attrition, we applied inverse probability of censoring weights, estimated via nonparametric machine learning ensembles in the `lmtp` package [@williams2021].


### Sensitivity Analysis

We perform sensitivity analyses using the E-value metric [@vanderweele2017; @linden2020EVALUE]. The E-value represents the minimum association strength (on the risk ratio scale) that an unmeasured confounder would need to have with both the exposure and outcome—after adjusting for measured covariates—to explain away the observed exposure-outcome association [@vanderweele2020; @linden2020EVALUE].


## Results


### Health

```{r}
#| label: fig-health
#| fig-cap: "Health effects"
#| eval: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

health_religious_vs_secular$plot
```

```{r}
#| label: tbl-health
#| tbl-cap: "Health effects"
#| eval: true
#| echo: false

health_religious_vs_secular$transformed_table|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


```{r, results = 'asis'}
cat(health_religious_vs_secular$interpretation)
```

{{< pagebreak >}}

### Psychological Well-Being

```{r}
#| label: fig-psych
#| fig-cap: "Effects on Psychological Well-Being"
#| eval: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

psych_religious_vs_secular$plot
```

```{r}
#| label: tbl-psych
#| tbl-cap: "Effects on Psychological Well-Being"
#| eval: true
#| echo: false

psych_religious_vs_secular$transformed_table|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


```{r, results = 'asis'}
cat(psych_religious_vs_secular$interpretation)
```

{{< pagebreak >}}
### Present-Focussed Well-Being

```{r}
#| label: fig-present
#| fig-cap: "Effects on Person-Focussed Well-Being"
#| eval: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

present_religious_vs_secular$plot
```

```{r}
#| label: tbl-present
#| tbl-cap: "Effects on Person-Focussed Well-Being"
#| eval: true
#| echo: false

present_religious_vs_secular$transformed_table|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


```{r, results = 'asis'}
cat(present_religious_vs_secular$interpretation)
```


{{< pagebreak >}}
### Life-Focussed Well-Being

```{r}
#| label: fig-life
#| fig-cap: "Effects on Life-Focussed Well-Being"
#| eval: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

life_religious_vs_secular$plot
```

```{r}
#| label: tbl-life
#| tbl-cap: "Effects on Life-Focussed Well-Being"
#| eval: true
#| echo: false

life_religious_vs_secular$transformed_table|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


```{r, results = 'asis'}
cat(life_religious_vs_secular$interpretation)
```


{{< pagebreak >}}
### Social-Focussed Well-Being

```{r}
#| label: fig-social
#| fig-cap: "Effects on Life-Focussed Well-Being"
#| eval: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

social_religious_vs_secular$plot
```

```{r}
#| label: tbl-social
#| tbl-cap: "Effects on Social Well-Being"
#| eval: true
#| echo: false

social_religious_vs_secular$transformed_table|>
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  kbl(format = "markdown")
```


```{r, results = 'asis'}
cat(social_religious_vs_secular$interpretation)
```

{{< pagebreak >}}
## Discussion

Our findings indicate that regular religious service attendance modestly enhances meaning, gratitude, and sexual satisfaction while more strongly boosting a sense of purpose, even in a predominantly secular setting such as New Zealand. Although these causal effect estimates withstand our robust handling of the time-series data, we suggest the following caution when interpreting the results.

First, the absence of consistent or robust evidence for many hypothesised effects does not imply that they do not exist. Uncertainties may arise from unmeasured confounding or non-directed measurement error, the later of which can bias estimates toward the null [@hernan2009MEASUREMENT; @vanderweele2012a; @bulbulia2024wierd]. In other words, what appears to be a lack of effect might reflect limitations in our measures rather than a genuine absence of causation.

Second, our causal effect estimates average over the entire target population. We do not consider how causal effects of religious attendance vary by faith tradition, intensity of belief, or aspects of religious (or secular) upbringing. To investigate such heterogeneity requires theoretically guiding interests, and the availability of relevant observations for the populations of interest -- matters for future interdisciplinary research.

Third, we contrast (1) regular religious service attendance for four years with (2) no attendance for four years, however, numerous other causal contrasts exist. For example, investigators might consider "conversion" or "de-conversion" scenarios. That is, one might examine how well-being would differ if everyone in a target population regularly attended for two waves and then ceased for two waves or if everyone did not attend for two waves but then took up the attendance for two waves [@vantongeren2020]. Contrasts could be made between these regimes and those we considered here, or perhaps others.  One might focus on longer or shorter periods of exposure, as well. One might also limit analyses to sub-populations for whom shifting attendance is more realistic (e.g., those already identifying as religious). Because there are innumerable combinations of potential interventions and target population, it is important to clarify specific research questions rather than vaguely referring to the "causal effects of religious service."  Although we cannot estimate all possible hypothetical experiments in a single study, framing questions as target trials helps identify possible interventions, another matter for future investigations.

Finally, it is important to emphasise that, at an individual level, responses to religious activities are likely to differ. Our findings represent an average treatment effect for the adult population of New Zealand, but they should not be construed as life advice for any particular individual.

Setting aside these limitations, one of the most robust results is that religious engagement fosters a sense of meaning and purpose (and especially of purpose). This finding supports long-standing traditions of theory and empirical research on religion and well-being [@park2005religion; @park2005religion_meaning; @eliade1984quest]. However, understanding why religious service attendance enhances a sense of meaning, particularly a sense of purpose, remains an open question. Evolutionary investigators might ask why a sense of meaning and purpose, if it confers biological advantages, is not a universal "default" cognitive state [@sterelny2018religion]? Some evolutionary investigators suggest that religious activity might function partly as a hard-to-fake signal of commitment rather than as an end in itself [@sosis2003cooperation]. Another compatible explanation is that religious service confers purpose by assigning tasks and objective -- both secular and collective. Charitable work, shared worship, and other communal practices can promote reciprocity, mutual support, and a shared responsibility that elevates abstract values into concrete goals. Notably, collective purposes have been shown to enhance the cooperative effects of collective activities, which may favour religious service in processes of cultural selection [@reddish2013let]. However, why religious attendance might amplify meaning and (especially) purpose awaits future inquiries.

We also observe a positive causal association between regular attendance and gratitude. In classical Christian thought (for example, Aquinas), religion is described as a virtue of justice called "piety," acknowledging the sources of one's existence as a debt that cannot be repaid [@bulbulia2023understanding]. In contemporary language, we might call this a conception of religion as a virtue of gratitude. Across religious traditions, gratitude to the gods and ancestors, or a single God, is commonly rehearsed [@watts2015pulotu]. Some hypothesise that religions actively cultivate gratitude to foster prosocial behaviours [@mccullough2001gratitude; @bono2004gratitude]. In any case, the conjecture that religious participation expresses gratitude is consistent with our findings.

We find that religious attendance is causally associated with modest improvements in sexual satisfaction. Earlier cross-sectional studies have observed correlations between religiosity and marital sexual satisfaction [@dew2020joint]. Although these studies do reliably estimate causal magnitudes, our results suggest a genuine but modest causal effect. Again, our purposes here are descriptive, and we do not evaluate specific theories. Speculating, religious service might align values, enhance communication, and foster stable social support networks within couples, thereby strengthening intimacy and trust and reduce conflict. From a bio-functional perspective, such intimacy and trust would be especially valuable for parenting -- the "ultimate cooperation problem" that links biological fates [@Bulbulia_2015].  On the other hand, a darker side to such cooperation might be sexual inequality [@deak2021individuals]. Again, our findings should not be considered as offering life advice. However, our findings motivate future investigation within the emerging field of religion, biological sex, and parenting (refer to @shaver2020church; @shaver2024religious; @shaver2022integrating). We contribute to these horizons by observing that in contemporary New Zealand, the relationship between religious attendance and sexual satisfaction is likely causal. 

Finally, we observe small but detectable effects on certain health behaviours. Specifically, regular attendance slightly reduced hours of exercise. This finding is consistent with a lost-opportunity explanation, whereby time spent at religious services replaces other physical activities. We observed a modest negative effect on BMI, suggesting that religious engagement might offset this reduced exercise through dietary practices or social norms around body weight, although this is speculative and mechanisms await further investigation. These smaller estimates illustrate how religious involvement can reallocate time and attention in ways that shape diverse health outcomes, even if the overall patterns are more modest than those observed for meaning, gratitude, sexual satisfaction, and a sense of purpose

In summary, our findings show that in New Zeland, religious service attendance can enrich certain facets of well-being, especially meaning, gratitude, sexual satisfaction, and a sense of purpose. We find only limited evidence for broader effects on other domains. Future research might consider different populations and subgroups within New Zealand, and a wider range of populations [@price2024science; @johnson2022global]. Such research should also examine proximate and evolutionary mechanisms by which religious service affects well-being, broadly conceived to include biological well-being [@shaver2024religious; @spake2024practical] and where possible, include objective measures both of attendance and well-being [@shaver2021comparison]. As with all causal investigations, we should also remain mindful that a practice that benefits certain groups, on average, may not benefit every individual similarly. We recommend care when imparting individual advice.

### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study. Members of the NZAVS management team and research group hold full copies of the NZAVS data. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author or any member of the NZAVS advisory board for replication or checking of any published study using NZAVS data. The code for the analysis can be found at [OSF link](https://osf.io/ab7cx/). 


### Acknowledgements

The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Plank Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement


All authors had input into the manuscript. JB did the analysis and developed the approach. CGS led data collection. JB, DD, CGS, KR, at GT obtained funding.


{{< pagebreak >}}
## Appendix A: Daily Data Collection  {#appendix-timeline}

{{< pagebreak >}}

@fig-timeline presents the New Zealand Attitudes and Values Study Data Collection (2018 retained cohort) from years 2018-2014 (NZAVS time 10--time 15).

```{r}
#| label: fig-timeline
#| fig-cap: "Historgram of New Zealand Attitudes and Values Study Daily Data Collection for Time 10 cohort: years 2018-2024."
#| eval: true
#| include: true
#| echo: false
#| fig-width: 12
#| fig-height: 12
#| 
timeline_histogram
```


{{< pagebreak >}}

## Appendix B: Measures and Demographic Statistics {#appendix-baseline}


### Measures 

```{r, results='asis'}
cat(appendix_text_version_1)
```



### Sample Demographic Statistics

@tbl-baseline presents sample demographic statistics.

::: {#tbl-baseline}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

library(gtsummary)
library(kableExtra)

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_baseline |> 
  kable_styling(
    font_size = 6,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Demographic statistics for New Zealand Attitudes and Values Cohort waves 2018.
:::


### Exposure Variable: Religious Service Attendance  {#appendix-exposure} 

@tbl-sample-exposures presents sample statistics for the exposure variable, religious service attendance, during the baseline and exposure waves. This variable was not measured in part of NZAVS time 12 (years 2020-2021) and part of NZAVS time 13 (years 2021-2022). To address missingness, if a value was observed after NZAVS time 14, we carried the previous observation forward and created and NA indicator. If there was no future observation, the participant was treated as censored, and inverse probability of censoring weights were applied, following our standard method for handling missing observations (see mansucript **Method**/**Handling of Missing Data**). Here, our carry-forward imputation approach may result in conservative causal effect estimation because it introduces measurement error. However, this approach would not generally bias causal effect estimation away from the null because the measurement error is unsystematic and random and unrelated to the outcomes.  

::: {#tbl-sample-exposures}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_exposures |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Demographic statistics for New Zealand Attitudes and Values Cohort waves 2018.
:::

{{< pagebreak >}}

### Outcome Variables {#appendix-outcomes} 


### Health Outcome Variables

::: {#tbl-sample-outcomes-health}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_outcomes_health |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Health variables measured at baseline (NZAVS time 10, years 2018-2019, and time 15, years 2023-2024).
:::



### Psychological Well-Being Outcome Variables

::: {#tbl-sample-outcomes-psych}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_outcomes_psych |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Psychological well-being variables measured at baseline (NZAVS time 10, years 2018-2019, and time 15, years 2023-2024).
:::



### Present-Focussed Well-Being Indicators

::: {#tbl-sample-outcomes-present}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_outcomes_present |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Present-focussed well-being variables measured at baseline (NZAVS time 10, years 2018-2019, and time 15, years 2023-2024).
:::



### Life-Focussed Well-Being Indicators

::: {#tbl-sample-outcomes-life}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_outcomes_life |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Life-reflective well-being variables measured at baseline (NZAVS time 10, years 2018-2019, and time 15, years 2023-2024).
:::


### Social Well-Being Indicators

::: {#tbl-sample-outcomes-social}
```{r, results = 'asis'}
#| eval: true
#| include: true
#| echo: false

# summary_tables$variables |> 
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     digits = 2,
#     row.names = FALSE                      # remove row names to eliminate any unintended columns
#   ) 


latex_table_outcomes_social |> 
  kable_styling(
    font_size = 12,
    latex_options = c("hold_position", "repeat_header", "striped", "longtable")
   ) 

```
Life-reflective well-being variables measured at baseline (NZAVS time 10, years 2018-2019, and time 15, years 2023-2024).
:::


{{< pagebreak >}}

## Appendix C: Confouding Control {#appendix-confounding}

::: {#tbl-C}
```{=latex}
\tvtable
```
@tbl-C presents single-world intervention graphs showing time-fixed and time-varying sources of bias in our six waves (baseline, four exposure waves, followed by the outcome wave.) Time-fixed confounders are included in the baseline wave. Time-varying confounders are included in each of the four treatment waves (abbreviated here by '$\dots$' to declutter the graph). When there is more than one exposure wave, identifying causal effects requires adjustment for time-varying confounders [@robins2008estimation; @bulbulia2024swigstime; @richardson2013].
:::

For confounding control, we employ a modified disjunctive cause criterion [@vanderweele2019], which involves:


1.	Identifying all common causes of both the treatment and outcomes.
2.	Excluding instrumental variables that affect the exposure but not the outcome.
3.	Including proxies for unmeasured confounders affecting both exposure and outcome.
4.	Controlling for baseline exposure and baseline outcome, serving as proxies for unmeasured common causes [@vanderweele2020].

Additionally, we control for time-varying confounders at each exposure wave [@robins2008estimation; @bulbulia2024swigstime; @richardson2013].


The covariates included for confounding control are described in @pedro_2024effects.

Where there are multiple exposures, causal inference may be threatened by time-varying confounding [@bulbulia2024swigstime]. 

{{< pagebreak >}}

## Appendix D: Causal Contrasts and Causal Assumptions {#appendix-assumptions}

### Notation

  - $A_k$: Observed religious service attendance at Wave $k$, for $k = 1, \dots, 4$.  
  - $Y_\tau$: Outcome measured at the end of the study (Wave 5).  
  - $W_0$: Confounders measured at baseline (Wave 0).  
  - $L_k$: Time-varying confounders measured at Wave $k$ (for $k = 1, \dots, 4$).  

### Shift Functions

Let $\boldsymbol{\text{d}}(a_k)^+$ represent the **regular attendance** treatment sequence and $\boldsymbol{\text{d}}(a_k)^-$ the **no attendance** treatment sequence, where the interventions occure at each wave $k = 1\dots 4; k\in \{0\dots 5\}$. Formally:

#### Regular Attendance Regime $\bigl(\boldsymbol{\text{d}}(a_k^+)\bigr)$

$$
\boldsymbol{\text{d}} (a_k^+) 
\;=\; 
\begin{cases}
4, & \text{if } A_k < 4,\\[6pt]
A_k, & \text{otherwise.}
\end{cases}
$$

#### No Attendance Regime $\bigl(\boldsymbol{\text{d}}(a_k^-)\bigr)$

$$
\boldsymbol{\text{d}}(a_k^-) 
\;=\; 
\begin{cases}
0, & \text{if } A_k > 0,\\[6pt]
A_k, & \text{otherwise.}
\end{cases}
$$

Here, $A_k$ is the observed attendance at Wave $k$. The shift function $\boldsymbol{\text{d}}$ "nudges" $A_k$ to a target level (four times per month or zero) only if the current value is below (for regular attendance) or above (for no attendance) that target. Across the four waves, these shifts form a sequence $\boldsymbol{\bar{\boldsymbol{\text{d}}}}$, which defines a complete intervention regime. 



### Causal Contrast

We compare the average well-being under the **regular attendance** regime, $\boldsymbol{\bar{\boldsymbol{\text{d}}}}(a^+)$, to the average well-being under the **no attendance** regime, $\boldsymbol{\bar{\boldsymbol{\text{d}}}}(a^-)$. Specifically, the average treatment effect (ATE) is given:

$$
\text{ATE}^{\text{wellbeing}} 
\;=\; 
\mathbb{E}
\Bigl[
  Y_\tau\!\bigl(\boldsymbol{\text{d}}(a^+)\bigr) 
  \;-\; 
  Y_\tau\!\bigl(\boldsymbol{\text{d}}(a^-)\bigr)
\Bigr].
$$


### Assumptions

To estimate this effect from observational data, we assume:

1. **Conditional Exchangeability:** Once we condition on $W_0$ and each $L_k$, the interventions $\boldsymbol{\bar{\boldsymbol{\text{d}}}}(a^+)$ or $\boldsymbol{\bar{\boldsymbol{\text{d}}}}(a^-)$ are effectively random with respect to potential outcomes.
2. **Consistency:** The potential outcome under a given regime matches the observed outcome when that regime is followed.
3. **Positivity:** Everyone has a non-zero probability of receiving each level of attendance (i.e., a chance to be "shifted" up or down) given their covariates. The positivity assumption is the only causal assumption that can be evaluated with data. We evaluate this assumption in [Appendix E](#appendix-transition)).

Mathematically, for conditional exchangeability, we write:

$$
\Bigl\{
  Y\bigl(\boldsymbol{\text{d}}(a^+)\bigr), 
  \; 
  Y\bigl(\boldsymbol{\text{d}}(a^-)\bigr)
\Bigr\}
\coprod
A_k |
W_0,
L_k
$$
That is, we assume the potential outcomes under each treatment regime are independent of each treatment at every time point, conditional on baseline confounders and time-varying confounders.

Under these assumptions, our statistical models permit us to estimate $\text{ATE}^{\text{wellbeing}}$ from observational data. That contrast is (1) *regularly attend religious service for four wave* versus (2) *never attend religious service at least four waves*.  We define the target population as the New Zealand Population from 2019-2024, the years in which measurements were taken. 


{{< pagebreak >}}

## Appendix E: Transition Matrix to Check The Positivity Assumption {#appendix-transition}


```{r, results = 'asis'}
#| label: tbl-transition
#| tbl-cap: "Transition Matrix Showing Change"
#| eval: true
#| include: true
#| echo: false

transition_matrix$table
```

```{r, results = 'asis'}
cat(transition_matrix$explanation)
```


{{< pagebreak >}}

## References {.appendix-refs}


