---
title: "The New Zealand Muslim Diversity Study: Baseline Sample Information 2023-2024"
abstract: |
  In this study, we present demographic statistics from the first wave of the New Zealand Muslim Diversity Study (MSD 2023-2024). We describe patterns of religious identity and behaviour within the MDS cohort and examine self-reported multi-dimensional well-being. By comparing Muslim demographic, religious, and psychological diversity with that of the broader New Zealand population, we discover patterns of variation within and between religious groups that are not evident from group affiliations alone. Our findings uncover new commonalities between Muslims and other highly religious groups in New Zealand, and clarify that group-level averages do not capture the full spectrum of individual religious differences.
   **KEYWORDS**: Diversity; Longitudinal; Muslim, Panel; Religion
author: 
  - name: M. Usman Afzali
    affiliation: University of Otago New Zealand
    orcid: 0000-0003-3169-6576 
    corresponding: yes
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-4064-8800
  - name: Aarif Rasheed
    affiliation:  University of Otago New Zealand
    orcid: 0000-0000-0000-0000
  - name: Kumar Yogeswaaran
    affiliation: University of Canterbury New Zealand
    orcid: 0000-0002-1978-5077
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
editor_options: 
    chunk_output_type: console
format:
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % Adjust this value as needed

date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# reproducibility
set.seed(123)

# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("table1")
library("printr")
packageVersion(pkg = 'margot')

# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-muslims")


# exposure_var = "muslim_binary"
# name_exposure = "muslim_binary"

# table for muslim on religion
table_demographics <- here_read("table_demographics")
table_health <- here_read("table_health")
table_psych <- here_read("table_psych")
table_present<- here_read("table_present")

table_social<- here_read("table_social")
table_life <- here_read("table_life")
table_religion <- here_read("table_religion")
table_discrim <- here_read("table_discrim")

prop_miss_case <- here_read("prop_miss_case")
results_religion <- here_read_qs("results_religion", push_mods)
results_life <- margot::here_read_qs("results_life", push_mods)
results_psych <- margot::here_read_qs("results_psych", push_mods)

n_total <- here_read("n_total")
n_15 <- here_read("n_15")
n_mus_pre<- here_read("n_mus_pre")

n_participants<- here_read("n_participants")
# varaibles
exposure_var = "religion_bigger_denominations"
name_exposure = "religion_bigger_denominations"

baseline_vars <- here_read("baseline_vars")
outcome_vars <- here_read("outcome_vars")

n_mus_post <- here_read("n_mus_post")

n_mus_post

# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)
prop_miss_case<- here_read("prop_miss_case")


# read models
results_religion<- here_read_qs("results_religion", push_mods)
results_discrim <- here_read_qs("results_discrim", push_mods)
results_life <- here_read_qs("results_life", push_mods)
results_psych<- here_read_qs("results_psych", push_mods)
results_social<- here_read_qs("results_social", push_mods)
results_present<- here_read_qs("results_present", push_mods)
results_health<- here_read_qs("results_health", push_mods)



# plots
psy_fatigue = results_psych$plots$Fatigue + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))



psy_anxiety = results_psych$plots$Anxiety + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))


psy_depression = results_psych$plots$Depression + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))



psy_rumination = results_psych$plots$Rumination + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))

#predictions
present_forgive <- results_present$plots$Forgiveness+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

present_perfect <- results_present$plots$Perfectionism+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

present_esteem <- results_present$plots$`Self Esteem`+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 
present_pwb_a <- results_present$plots$`PWB Your Future Security` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_b <- results_present$plots$`PWB Standard Living` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_c <- results_present$plots$`PWB Your Health` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_d  <- results_present$plots$`PWB Your Relationships`+ scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))



# access individual plots Religion
religion_relid <- results_religion$plots$`Religious Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  
religion_spirid <- results_religion$plots$`Spiritual Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  

religion_church <- results_religion$plots$`Monthly Religious Service Attendence` + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-1,30))
   
religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_prayer <- results_religion$plots$`Weekly Prayer`+ scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_god <- results_religion$plots$`Belief in God`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 

religion_spirit <- results_religion$plots$`Belief in Spirit`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


# access individual plots discrimination


discrim_ethnic <- results_discrim$plots$`Perceived Ethnic Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_religious <- results_discrim$plots$`Perceived Religious Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_gender<- results_discrim$plots$`Perceived Gender Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_age <- results_discrim$plots$`Perceived Age Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
    

# access individual plots health

health_sleep <- results_health$plots$Sleep+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(.5,15))
health_sleep

health_exercise <- results_health$plots$`Hours of Exercise (log)`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(0,6))
health_exercise

exp(1)-1
exp(2)-1
exp(3)-1
exp(4)-1

health_short_form_health <- results_health$plots$`Short Form Health`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))  +  scale_y_continuous(limits = c(.5,7))
    
health_short_form_health
# access individual plots psychological wellbeing
life_grd <- results_life$plots$Gratitude+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_sat<- results_life$plots$`Life Satisfaction`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_mp <- results_life$plots$`Meaning: Purpose`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_ms <- results_life$plots$`Meaning: Sense`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

# access individual plots life-reflective wellbeing
results_psych$plots$Anxiety+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Depression + scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Fatigue+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Rumination+ scale_y_continuous(limits = c(-.5,5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

social_belong_a <- results_social$plots$`Belonging - Acceptance` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
    
    
social_belong_b <- results_social$plots$`Belonging - Shared Beliefs` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
  
social_belong_c  <-
  results_social$plots$`Belonging - Outsider (Reversed)` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_neighbourhood <- results_social$plots$`Neighbourhood Community`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_a <- results_social$plots$`Support - Help Availability`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_b <- results_social$plots$`Support - No Guidance (Reversed)`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_c <- results_social$plots$`Support - Someone to Turn To`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


# combo plots

plot_belief <- religion_god + religion_spirit + plot_annotation(title = "Religious Beliefs",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_practice <- religion_church / (religion_prayer + religion_scripture)+ plot_annotation(title = "Religious Practice",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_id <- religion_relid + religion_spirid + plot_annotation(title = "Religious and Spiritual Identification",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_life <- life_grd + life_sat + life_mp + life_ms + plot_annotation(title = "Life-Reflective Well-Being",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_descrim <- discrim_age  + discrim_gender + discrim_ethnic + discrim_religious +
  plot_annotation(title = "Perceived Dicrimination",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_social_belong <- social_belong_a  + social_belong_b + social_belong_c  +social_neighbourhood +
  plot_annotation(title = "Social Belonging and Neighbourhood Community",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')



plot_social_support <- social_support_a + social_support_b + social_support_c   +
  plot_annotation(title = "Social Support",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_present_a <- present_esteem + present_forgive + present_perfect +
  plot_annotation(title = "Personal Well-Being Indicators",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_present_b <- present_pwb_a + present_pwb_b + present_pwb_c + present_pwb_d +
  plot_annotation(title = "Personal Well-Being Index",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_health <- health_short_form_health  + (health_exercise /  health_sleep) +
  plot_annotation(title = "Biological Health",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_psych <- psy_anxiety + psy_depression + psy_fatigue + psy_rumination + plot_annotation(title = "Psychological Well-Being",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')



n_mus_post

# checks
# plot_belief
# plot_id
# plot_practice
# plot_descrim
# plot_present_a
# plot_present_b
# plot_life
# plot_social_belong
# plot_social_support
# plot_psych
# plot_health

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)


measures_path_rstudio <- here::here("manuscripts", "24-usman-mus-diversity")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))

outcome_measures <- boilerplate::boilerplate_report_measures(
  baseline_vars = baseline_vars,
  outcome_vars = outcome_vars,
  measure_data = measure_data
)

```

The New Zealand Attitudes and Values Study (NZAVS) is a nationally representative panel study of New Zealanders, initiated in 2009. It has measured responses from `r n_total` participants annually and is currently tracking `r n_15` participants, almost 1% of New Zealand’s adult population.

The Muslim Diversity Study (MSD) is a longitudinal Muslim booster housed within the New Zealand Attitudes and Values Study. The Muslim Diversity Study was established in 2024, in partnership with the Muslim community, as part of the New Zealand Attitudes and Values Study, to enhance Muslim representation within the NZAVS (MSD n = `r n_mus_pre`, year 1 = 2024, NZAVS wave 15). Despite strong coverage in the New Zealand Attitudes and Values Study of the country’s Christian religious diversity, prior to the Muslim Diversity Study, less than 0.01% of New Zealand’s Muslims were participating in the NZAVS. To put this into perspective, Muslims were under-represented by nearly an order of magnitude compared to major Christian denominations.

 The Muslim Diversity Study boosted the Muslim sample by almost 600 participants, achieving coverage of about 1% of the New Zealand Census estimate of the Muslim population (n = `r n_mus_post`).

The purpose of the Muslim Diversity Study is to extend national longitudinal research to explore Muslim diversity and flourishing over the lifespan. It seeks to document the scope of this diversity, identify challenges to flourishing, and examine how Muslims overcome adversity. Although previous studies present a clear picture of Muslim adversity (@shaver2016religion; @shaver2017news; @highland2019attitudes; @sibley2020; @shanaah2023hate; @greaves2020comparative;  @bulbulia2023a), how such adversity shapes and affects the experiences of Muslims across New Zealand, over time, remains unclear. The Muslim Diversity Study addresses this gap.

More generally, the Muslim Diversity Study aims to contribute to the consultable historical record that the New Zealand Attitudes and Values Study is gradually accumulating for the people of New Zealand during. By amplifying the voices of Muslim diversity in New Zealand, we hope to enable a richer understanding of the diversity of resident beliefs, attitudes, and behaviours during our time so that people in the future will know who we were, what we cared about, and why we acted as we did. 

This article presents a comprehensive examination of religion, experiences of prejudice, and multi-dimensional well-being among Muslims in New Zealand, situated within a comparative framework that includes other religious and secular groups across New Zealand. Our aim is to clarify the religious diversity of the Muslim sample, experiences of prejudice, and initial levels of psycho-social well-being. Although we speculate in the Discussion about future horizons of research, our approach here is descriptive. 

We begin by with a report of sample statistics across multiple demographic indicators, values, employment, and behaviours.  Subsequently, Studies 1 through 7 investigator the relationships between religious beliefs and practices, perceived discrimination across multiple domains—including religious, ethnic, gender, and age discrimination -- and multiple dimensions of psycho-social well-being. These facets of interest encompass religion, experiences of prejudice, biological health, psychological well-being, self-related well-being, life-related well-being, and social well-being -- in accordance with the Muslim Diversity Study's interests on Muslim faith and human flourishing, and alinging with New Zealand Attitudes and Values Study broader interests in religion and human flourishing (@Bulbulia2024PracticalGuide). To more accurately isolate the relationship of religious or secular affiliation to these domains, our statistical models control for demographic indicators unlikely to be influenced by religious identification: age, gender, New Zealand birth status, and ethnicity. This approach allows us to clarify distinct patterns of well-being associated with religious and secular group membership, independent of demographic differences that comprise features of our sample that, although correlated with each participant's religious or secular affiliation, are not be caused by this affiliation. Although our models hope to better identify differences and similarities as the parameters of interest are associated with religious and secular affiliation, we emphasise, again, that our work here is descriptive. We aim to document patterns of association without inferring causality to provide a baseline from which we may build understanding of Muslim diversity, and flourishing, in New Zealand.

<!-- The Muslim Diversity Study is important first, because the lifelong dynamics of Muslim flourishing are poorly understood; this project addresses an urgent scientific need. -->

<!-- Second, our work with the Muslim community revealed a strong community interest in Muslim flourishing science; this project is an attempt to address this interest by involving Muslims in this science.  -->

<!-- Third, this project serves as a model for developing panel studies of under-represented groups in Aoteoroa New Zealand and in other countries.  -->


## Method

Measures are presented in [Appendix A](#appendix-a)

### Sample

The New Zealand Attitudes and Values Study is a national probability study designed to accurately reflect the broader New Zealand population. It uses prize draws to incentivise participation. Although the New Zealand Attitudes and Values Study has good demographic representation of the country as a whole, it tends to under-sample males and individuals of Asian descent and over-sample females and Māori (the indigenous peoples of New Zealand). To address enhance the accuracy of our findings for the target population, we apply 2018 New Zealand Census survey weights to the sample data. These weights adjust for variations in age, gender, and ethnicity to better approximate the national demographic composition [@sibley2021]. 

**OFFICIAL NZAVS DOCUMENTATION HERE**


### Eligibility Criteria

Participants needed to respond to the 2023 wave of the New Zealand Attitudes Study, and to have complete information about religious identification (yes or no). A total of `r n_participants` individuals met these criteria and were included in the study. 


### Missing Data

To mitigate bias from missing data, we imputed missing responses using the `ppm` algorithm from the `mice` package in R [@vanbuuren2018].  Overall missingness accounts for less than `r prop_miss_case`% if the data. However, we note that missingness for certain indicators among Muslims is considerably higher than among other groups, including indicators for religious identification (11.1% missingness). 


### Demographic Statistics and Sample Responses

@tbl-demography presents demographic data by religious affiliation, including secular affiliation.


::: {#tbl-demography}
@tbl-religion presents sample demographic statistics. 

\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_demographics_x <- t1kable(
  table_demographics,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_demographics_x

table_demographics_x|> 
  kable_styling(
    font_size = 6,
    latex_options = c("hold_position","striped", "repeat_header", "longtable") )
```
\end{landscape}
Demographic statistics by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::

{{< pagebreak >}}
@tbl-religion presents sample response-statistics for Religious Belief, Identification, and Practice. 

::: {#tbl-religion}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_religion_x <- t1kable(
  table_religion,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_religion_x |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
  )

```
\end{landscape}
Sample statistics by for religious beliefs and behaviorus by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::
{{< pagebreak >}}

@tbl-discrim presents sample response-statistics for Perceived Discrimination.

::: {#tbl-discrim}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_discrim_x <- t1kable(
  table_discrim,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_discrim_x |>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
  )


```
\end{landscape}
Perceived discrimination by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


{{< pagebreak >}}
@tbl-health presents sample response-statistics for biological health.

::: {#tbl-health}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_health_x <- t1kable(
  table_health,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_health_x |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample biological health indicators, by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::

{{< pagebreak >}}

@tbl-psych presents sample response-statistics for psychological health.

::: {#tbl-psych}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_psych_x <- t1kable(
  table_psych,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_psych_x |>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample psychological health indicators, by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::
{{< pagebreak >}}
@tbl-present presents sample response-statistics for person-focussed well-being

::: {#tbl-present}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_present_x <- t1kable(
  table_present,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_present_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample person-focussed well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::
{{< pagebreak >}}

@tbl-life presents sample response-statistics for life-focussed well-being (NZAVS wave 15, which encompasses MSD wave 1).
:::

::: {#tbl-life}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_life_x <- t1kable(
  table_life,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_life_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample life-focussed well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::

{{< pagebreak >}}

@tbl-social presents sample response-statistics for social well-being

::: {#tbl-social}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_social_x <- t1kable(
  table_social,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_social_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample social well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).

:::
{{< pagebreak >}}

## Study 1: Comparative Models of Religious Beliefs and Behaviours

@fig-religion-id presents models for religious and spiritual identities adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-id and @tbl-spiritual-id


```{r}
#| label: fig-religion-id
#| fig-cap: "Religious and Spiritual Identity"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_id

```


{{< pagebreak >}}


::: {#tbl-religion-id}
```{r, results= 'asis'}
library(dplyr)
library(kableExtra)

#  table
results_religion$predictions$religion_identification_level |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for religious identification by religious denomination.
:::

{{< pagebreak >}}


::: {#tbl-spiritual-id}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_spiritual_identification |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # Removed "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for spiritual identification by religious denomination.
:::


{{< pagebreak >}}

@fig-religion-belief presents models for religious and spiritual beliefs adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-belief and @tbl-spiritual-belief.



```{r}
#| label: fig-religion-belief
#| fig-cap: "Religious and Spiritual Beliefs"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_belief
```

{{< pagebreak >}}

::: {#tbl-religion-belief}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_believe_god_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for belief in God by religious denomination.
:::


::: {#tbl-spiritual-belief}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_believe_spirit_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # Removed "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for sbelief in a Spirit or Life Force by religious denomination.
:::



{{< pagebreak >}}

@fig-religion-practice presents models for religious and spiritual practices, again adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-church,  @tbl-religion-prayer, and @tbl-religion-scripture.



```{r}
#| label: fig-religion-practice
#| fig-cap: "Religious and Spiritual Practices"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_practice
```

{{< pagebreak >}}


::: {#tbl-religion-church}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_church |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for monthly religious service attendance by religious denomination.
:::

{{< pagebreak >}}

::: {#tbl-religion-prayer}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_prayer |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::

{{< pagebreak >}}

::: {#tbl-religion-scripture}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_scripture |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::

{{< pagebreak >}}

## Study 2: Comparative Models of Perceived Prejudice


@fig-discrim presents models for perceived discrimination, adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-discrim-age, @tbl-discrim-gender,  @tbl-discrim-ethnicity, and @tbl-discrim-religion, 

```{r}
#| label: fig-discrim
#| fig-cap: "Religious and Spiritual Practices"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_descrim
```


{{< pagebreak >}}

::: {#tbl-discrim-age}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_age_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived age discrimination.
:::

::: {#tbl-discrim-gender}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_gend_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived gender discrimination.
:::

{{< pagebreak >}}

::: {#tbl-discrim-ethnicity}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived ethnic discrimination.
:::


::: {#tbl-discrim-religion}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_religious_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived religious discrimination.
:::

{{< pagebreak >}}
## Study 3: Comparative Models of Biological Health

@fig-health presents models for indicators related to biological health, adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-health-ex, @tbl-health-sleep, and @tbl-health-sfh.   

```{r}
#| label: fig-health
#| fig-cap: "Biological Health"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_health
```

{{< pagebreak >}}

::: {#tbl-health-ex}
```{r, results= 'asis'}

#  table
results_health$predictions$log_hours_exercise |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for log hours of weekly exercise.
:::

{{< pagebreak >}}

::: {#tbl-health-sleep}
```{r, results= 'asis'}

#  table
results_health$predictions$hlth_sleep_hours |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for weekly hours of sleep.
:::

{{< pagebreak >}}

::: {#tbl-health-sfh}
```{r, results= 'asis'}

#  table
results_health$predictions$short_form_health |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for subjective health.
:::

{{< pagebreak >}}

## Study 4: Comparative Models of Psychological Well-being

@fig-psych presents models for indicators related to psychological well-being, adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-psych-anxiety, @tbl-psych-depression, @tbl-psych-fatigue, and @tbl-psych-rumination. 

```{r}
#| label: fig-psych
#| fig-cap: "Psychological Health"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_psych
```

{{< pagebreak >}}

::: {#tbl-psych-anxiety}
```{r, results= 'asis'}

#  table
results_psych$predictions$kessler_latent_anxiety |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate 
  )
```
Adjusted means and 95% confidence intervals for psychological anxiety.
:::

::: {#tbl-psych-depression}
```{r, results= 'asis'}

#  table
results_psych$predictions$kessler_latent_depression |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for psychological depression.
:::

{{< pagebreak >}}

::: {#tbl-psych-fatigue}
```{r, results= 'asis'}

#  table
results_psych$predictions$hlth_fatigue |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for fatigue.
:::

::: {#tbl-psych-rumination}
```{r, results= 'asis'}

#  table
results_psych$predictions$rumination|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for rumination.
:::

{{< pagebreak >}}


## Study 5: Comparative Models of Person-Focussed Well-Being


@fig-self-a presents models for personal-focussed well-being in dimensions of self-esteem, forgiveness, and perfectionism, again adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-selfesteem, @tbl-forgive, and @tbl-perfect.

```{r}
#| label: fig-self-a
#| fig-cap: "Self-Esteem, Forgiveness, Perfectionism"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_present_a
```

{{< pagebreak >}}

::: {#tbl-selfesteem}
```{r, results= 'asis'}
#  table
results_present$predictions$self_esteem|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for self esteem.
:::

{{< pagebreak >}}

::: {#tbl-forgive}
```{r, results= 'asis'}
#  table
results_present$predictions$forgiveness|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for forgiveness.
:::

{{< pagebreak >}}

::: {#tbl-perfect}
```{r, results= 'asis'}
#  table
results_present$predictions$perfectionism|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perfectionism.
:::

{{< pagebreak >}}

@fig-self-b presents models for item-level responses to the personal well-being index measures. The expected means and confidence intervals for these models are presented in @tbl-pwb-fs,  @tbl-pwb-sl, @tbl-pwb-yh, @tbl-pwb-yr.

```{r}
#| label: fig-self-b
#| fig-cap: "Personal Well-Being Index"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_present_b
```


{{< pagebreak >}}

::: {#tbl-pwb-fs}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_future_security|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for personal well-being: your future security.
:::

::: {#tbl-pwb-sl}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_standard_living|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for personal well-being: your standard of living.
:::

{{< pagebreak >}}

::: {#tbl-pwb-yh}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_health|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for personal well-being: your health.
:::

::: {#tbl-pwb-yr}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_relationships|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for personal well-being: your relationships.
:::

{{< pagebreak >}}

## Study 6: Comparative Models of Life-Focussed Well-Being

@fig-life presents models for life-focussed well-being, again adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-gratitude, @tbl-ls, @tbl-purpose,  @tbl-sense.
```{r}
#| label: fig-life
#| fig-cap: "Life Related Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_life

```

{{< pagebreak >}}

::: {#tbl-gratitude}
```{r, results= 'asis'}
#  table
results_life$predictions$gratitude|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for gratitude.
:::


::: {#tbl-ls}
```{r, results= 'asis'}
#  table
results_life$predictions$lifesat|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for life satisfaction.
:::


{{< pagebreak >}}

::: {#tbl-purpose}
```{r, results= 'asis'}
#  table
results_life$predictions$meaning_purpose|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for sense of purpose in life.
:::

::: {#tbl-sense}
```{r, results= 'asis'}
#  table
results_life$predictions$meaning_sense|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for an understanding what makes life meaningful.
:::
{{< pagebreak >}}

## Study 7: Comparative Models of Social Well-Being


@fig-social-support presents models for social support, adjusting for age, New Zealand born, ethnicity, and gender.  The expected means and confidence intervals for these models are presented in @tbl-support-help, @tbl-support-guide, @tbl-support-turn.


```{r}
#| label: fig-social-support
#| fig-cap: "Social Support"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_social_support
```
{{< pagebreak >}}

::: {#tbl-support-help}
```{r, results= 'asis'}
#  table
results_social$predictions$support_help|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for social support: help.
:::

{{< pagebreak >}}

::: {#tbl-support-guide}
```{r, results= 'asis'}
#  table
results_social$predictions$support_noguidance_reversed|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for social support: guidance.
:::


{{< pagebreak >}}

::: {#tbl-support-turn}
```{r, results= 'asis'}
#  table
results_social$predictions$support_turnto|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for social support: no one to turn to (reversed).
:::
{{< pagebreak >}}

@fig-social-belong presents models for social belonging, adjusting for age, New Zealand born, ethnicity, and gender.  The expected means and confidence intervals for these models are presented in @tbl-accept, @tbl-sharedbeliefs, @tbl-outsider,  @tbl-neighbourhood, 

```{r}
#| label: fig-social-belong
#| fig-cap: "Social Belonging"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_social_belong

```
{{< pagebreak >}}

::: {#tbl-accept}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_accept|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for belonging: feels accepted.
:::

::: {#tbl-sharedbeliefs}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_accept|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for belonging: others share my beliefs.
:::

{{< pagebreak >}}

::: {#tbl-outsider}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_routside_reversed|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for belonging: feel like outsider (reversed).
:::

::: {#tbl-neighbourhood}
```{r, results= 'asis'}
#  table
results_social$predictions$neighbourhood_community|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for sense of neighbourhood community.
:::

{{< pagebreak >}}

{{< pagebreak >}}

## Conclusions 

### Discussion of Demographic and Sample Characteristics of the Muslim Diversity Study Booster.

Several similarities and differences emerge between the Muslim Diversity Study booster sample and the broader New Zealand Attitudes and Values sample at time 15 (years 2023-2024). Most immediately, Age stands out as a significant differentiator: Muslims in the study have a mean age of 33.8 years, substantially younger than the Not Religious (secular) group (51.6 years) and other religious affiliations such as Anglicans (63.3 years) and Catholics (56.4 years). This younger demographic is further reflected in the lower parental status among Muslims (45.9%) compared to Anglicans (84.6%) and Catholics (76.5%). Additionally, only 23% of Muslims were born in New Zealand, a stark contrast to the Not Religious group (79.4%) and Christian denominations, indicating a higher proportion of first-generation immigrants within the Muslim community. Ethnically, 60.6% of Muslims identify as Asian, whereas the majority of the Not Religious group (83%) and other religious groups are of European descent, indicating cultural diversity within the Muslim population.

In terms of lifestyle and attitudes, Muslims report a markedly lower frequency of alcohol consumption (mean 0.112) compared to the Not Religious group (2.05) and other religious groups, in accordance with Islamic prohibitions on alcohol. Interestingly, Muslims exhibit higher scores in Right-Wing Authoritarianism (mean 4.53) than the Not Religious group (2.88) and even other religious groups such as Anglicans (3.84) and Catholics (3.80), suggesting distinct socio-political orientations. The New Zealand Deprivation Index reveals that Muslims have higher Deprivation scores (mean 5.65) compared to the Not Religious group (4.68). Regarding personality traits, Muslims score a little lower on Honesty-Humility (mean 4.87) than other groups, but similar in Agreeableness and Conscientiousness. Patterns of missing data are notable; for example, Muslims have a higher rate of missing responses in Non-Heterosexual Orientation (18.8% missing) and Household Income (34% missing), which may reflect cultural sensitivities, a reluctance to disclose personal information on these topics, or simply a lack of familiarity with this question. The patterns of missing data in sensitive areas merits future research and addition, in collaboration with the Muslim community.

Next we discuss the findings of our statistical models. Our reported comparisons pertain to the broader sample population in the NZAVS, as weighted by 2018 New Zealand Census weights. However, whether the findings we report generalise to the population of New Zealand relies on the assumption that these weights are accurate. The question of 'representativeness' is challenging, and a matter of ongoing New Zealand Attitude and Values Study research. Every comparison we make here should be considered an approximation for the New Zealand population of unknown accuracy. 



### Discussion of Study 1: Religion 

An analysis of adjusted models for religious and spiritual identification, beliefs, and practices reveals distinct patterns among Muslims compared to other religious and secular groups in New Zealand. By adjusting for age, gender, New Zealand birth status, and ethnicity, the findings imply that differences in religious engagement are not explained by these demographic influences.

#### Religious Identification and Spirituality

Muslims exhibit the highest level of religious identification, with a predicted score of 6.21 (although note that over 11% of Muslims in the sample did not respond to this question). This strong religious identity is paralleled by a high level of spiritual identification (predicted score of 5.07), closely aligning with Buddhists and participants from Other Religions, who also report elevated spiritual identification. However, spiritual identification among Muslims is somewhat lower than among Christian No Faith Declared and Christian "Others" (smaller denominations).

##### Beliefs in God and Spirits

Belief in God is nearly universal among Muslims, with a predicted probability of 0.89, comparable to Christian groups like Christian (NFD) (0.94) and Christian (Others) (0.87). This contrasts with the Not Religious group, which has a low probability of 0.14. Regarding belief in spirits or life forces, Muslims have a predicted probability of 0.61, which is substantial but lower than that of Hindus (0.90) and participants from Other Religions (0.87). This hints at theological diversity among Muslims in spirit/life force beliefs.

##### Religious Practices

The frequency of religious practices among Muslims is markedly higher than in other groups. Muslims attend religious services with a predicted frequency of 5.43 times per month, considerably surpassing Christian denominations such as Christian (Others) (3.23) and Catholics (2.02). The Not Religious group shows negligible attendance (-0.04). Moreover, Muslims engage in prayer with an exceptionally high predicted frequency of 20.01, reflecting the centrality of daily prayers in Islam. This frequency is substantially higher than that of Christian (NFD) (8.73) and Christian (Others) (10.09), indicating a distinctiv, and particularly strong pattern of Muslim religious observance.

Scripture reading is also more prevalent among Muslims, with a predicted frequency of 4.59 times per week. This is comparable to devout Christian groups like Christian (Others) (4.19) but higher than mainstream denominations such as Anglicans (0.86) and Catholics (1.28). These findings suggest a strong engagement with religious texts within the Muslim community.


### Discussion of Study 2: Perceived Discrimination

An analysis of discrimination across domains (age, gender, ethnic, and religion-specific) reveals that Muslims in New Zealand report do not report high levels of perceived discrimination on age and gender. Muslims report higher ethnic discrimination at levels comparable to Hindus. However, perceived religious discrimination is markedly higher among Muslims than among other religious denominations.

#### Age Discrimination

Muslims report an average perceived age discrimination score of 3.28, which is only marginally higher than the Not Religious group (3.17) and Catholics (3.27). Other groups, such as Anglicans (3.04) and Christian (Others) (3.12), exhibit slightly lower perceptions of age discrimination. The overall differences across groups are small.

#### Gender Discrimination

Perceived gender discrimination among Muslims is 2.48, similar to the Not Religious group (2.50) and Catholics (2.42). Interestingly, Buddhists report a higher score (2.75), whereas Presbyterians report the lowest (2.24). This suggests that Muslims’ experiences of gender discrimination are comparable to other groups.

#### Ethnic Discrimination

For ethnic discrimination, Muslims report a score of 2.22, higher than the Not Religious group (1.70) and most Christian groups, but similar to Hindus (2.16).

#### Religious Discrimination

Religious discrimination reveals the most pronounced disparities. Muslims report the highest perceived score (3.57), significantly higher than the Not Religious group (1.66) and Anglicans (2.14). Christian (NFD) (3.08) and Christian (Others) (2.99) also report elevated religious discrimination, reflecting similar experiences with religion-centric bias.

The findings indicate that although Muslims perceive similar levels of age and gender discrimination to other groups, they report higher experiences of ethnic and markedly higher levels of religious discrimination. This finding underscores the religion-specific discrimination that Muslims face in New Zealand (@shaver2016religion; @shaver2017news), and is consistent with previous research revealing taht Muslims are the least accepted minority group in New Zealand for which the NZAVS has measures (@sibley2020).

### Discussion of Study 3: Biological Health 

Biological Health indicators, specifically sleep duration and overall health status, reveals minimal differences between Muslims and other religious and secular groups in New Zealand. Again adjusted models account for age, gender, New Zealand birth status, and ethnicity, ensuring that the comparisons are independent of these demographic factors.

#### Sleep Duration

The predicted average sleep duration for Muslims is 6.85 hours, which is slightly less than the Not Religious group at 7.01 hours. Other religious groups show similar patterns: Anglicans at 7.05 hours, Buddhists at 7.01 hours, and Catholics at 6.89 hours. The variations among all groups are minimal, suggesting that sleep habits are consistent across religious affiliations.

#### Short Form Reported Health Status

As measured by the Short Form Scale, Muslims have a predicted health score of 4.93, marginally higher than the Not Religious group at 4.77. Anglicans and Christian (NFD) participants present average scores of 4.89 and 4.86, respectively, while Catholics score 4.77 on average. Buddhists exhibit a slightly lower predicted average scores of 4.68. These differences are modest, indicating that overall health status is relatively uniform across religious and secular groups.

The findings suggest that Muslims are largely similar to other religious and secular groups in terms of reported biological health. 


### Discussion of Study 4: Psychological Health 

The data suggest that Muslims in New Zealand exhibit better-adjusted psychological well-being on certain dimensions compared to other groups, whereas depression levels are comparable. 

#### Anxiety
Adjusted scores for anxiety among Muslims are lower (1.30) compared to Not Religious individuals (1.39) and most Christian groups, with the exception of Anglicans (1.34) and Christian NFD (1.34). This finding indicates a relative buffering against anxiety in the Muslim population.

#### Fatigue

Muslims report lower fatigue (1.71) compared to Not Religious individuals (1.83) and most other groups. This pattern suggests resilience in managing psychological and physical stressors.

#### Rumination

Rumination scores for Muslims (0.90) are slightly lower than the Not Religious group (0.99) and align closely with Anglicans (0.90). Rumination, often a precursor to mental distress, appears to be less pronounced among Muslims.

#### Depression

Depression scores for Muslims (0.73) are comparable to those of other groups, such as Catholics (0.65) and Anglicans (0.64). This parity indicates that while Muslims fare better in anxiety and fatigue, depression remains similar to other populations.

Overall, these findings imply that Muslim psychological adapation is robust, raising an intriguing question: do religion-specific practices, beliefs, or the sense of community inherent to Muslim identity act as buffers against the damaging psychological effects of anti-Muslim prejudice? -- a question for future longitudinal inquiry.

### Discussion of Study 5: Self-Related Well-Being 

Muslims in New Zealand exhibit distinctive patterns on personal well-being indicators, particularly in self-esteem, personal relationships, and forgiveness. These results raise interesting questions about whether religion-specific practices, beliefs, or community factors enhance personal well-being.

#### Self-Esteem

Muslims report significantly higher self-esteem (5.46) compared to Not Religious individuals (5.00) and most other religious groups, including Anglicans (5.03) and Catholics (5.11). This elevated self-esteem may reflect the role of religious identity, belief in God, and community support in fostering personal confidence and self-worth, a matter for future investigations.

#### Forgiveness

While forgiveness scores for Muslims (4.89) are slightly lower than for Not Religious individuals (4.99) and Anglicans (5.03), they remain comparable across religious groups. Notably, it might be more difficult to forgive among those prone to greater levels of injustice, a matter for future research.

#### Perfectionism

Muslims report slightly higher perfectionism (3.32) than Not Religious individuals (3.26) and align closely with groups such as Buddhists (3.33) and Catholics (3.28). These similarities in group means again mask variation above and below the means, among individuals in all groups.

#### Future Security

Muslims have a strong sense of future security (6.07), comparable to Anglicans (6.12) and Catholics (5.97), and higher than Not Religious individuals (5.79). This might stem from community and religious frameworks providing stability and hope for the future.

#### Standard of Living and Health

Muslims report slightly lower satisfaction with their standard of living (7.17) compared to Anglicans (7.58) and Catholics (7.36) but maintain relatively high ratings of health (6.69), comparable to Anglicans (6.72) and higher than Not Religious individuals (6.49). (Recall that Muslims exhibit higher deprivation.)

#### Personal Relationships

Muslims exhibit the highest scores for satisfaction with personal relationships (7.90), surpassing Not Religious individuals (7.46) and other groups, including Anglicans (7.78) and Catholics (7.72). This finding highlights the centrality of strong interpersonal bonds in Muslim communities.


Overall the data raise the question of whether religious identity, community, and practices act both as buffers for prejudice and enhancers of personal well-being. The high self-esteem and relationship satisfaction among Muslims are particularly notable and warrant future longitudinal investigations.

### Discussion of Study 6: Life-Focussed Well-Being

Religious groups exhibit consistently higher ratings across four dimensions of life-related well-being, namely, gratitude, life satisfaction, meaning and purpose, and a sense of meaning when compared with secular people. Notably, Muslims exhibit an exceptionally strong sense of meaning and purpose.

#### Gratitude

Muslims report higher levels of gratitude (6.01) than Not Religious individuals (5.80) and Anglicans (5.89), aligning closely with Catholics (5.99) and Christian NFD (6.09). Gratitude is a cornerstone of religious faith, which may explain the elevated sense of thankfulness highly Muslims, however, it is important to acknowledge that secular people in New Zealand typically exhibit high levels of gratitude, and there is variation in gratitude across all religious groups.

#### 	Life Satisfaction

Muslims demonstrate higher average life satisfaction (5.32) compared to Not Religious individuals (5.00) and most religious groups, such as Anglicans (5.17) and Catholics (5.08).

#### Meaning and Purpose

Muslims report the highest levels of meaning and purpose (5.83) across all groups, significantly surpassing Not Religious individuals (4.76) and aligning closely with Christian Others (5.45) and Christian NFD (5.41). 

####	Sense of Meaning

Muslims also rank highly in their sense of meaning (6.03), exceeding Not Religious individuals (5.47) and most religious groups, such as Catholics (5.66) and Anglicans (5.58). This sense of coherence and understanding in life may be linked to Islamic teachings and a strong sense of community.

Overall, Muslims consistently score higher than the population average across these Life-focussed well-being. Whether a sense of meaning and purpose buffers Muslims from the slings and arrows of anti-Muslim prejudice suggests yet another pathway for future investigations.


### Discussion of Study 7: Social Well-Being 


Muslims resemble other religious groups in perceived social support and social belonging, and express higher levels of perceived neighbourhood community.

#### Social Support

**Help from Others**:  Muslims report moderate support for receiving help (5.87), slightly lower than Not Religious (5.99) and Christians (NFD: 6.12, Others: 6.12), and comparable to Buddhists (5.88).

**No Guidance (reversed)**: Muslims score lower (5.72) in perceiving guidance availability compared to Anglicans (6.01) and Catholics (5.99), but align closely with Hindus (6.00) and Buddhists (5.74).

**Turning to Others** Muslims score (5.73), slightly below Anglicans (5.97) and Catholics (5.96), but comparable to Hindus (5.94) and higher than Buddhists (5.61).

#### Belonging

**Acceptance**: Muslims exhibit moderate levels of feeling accepted (5.60), aligning with Anglicans (5.58) and Buddhists (5.55), though slightly lower than Catholics (5.71) and Christian NFD (5.67).

**Feel Like an Outsider (reversed)**: Muslims score moderately (4.96), higher than Not Religious (4.88) and similar to Anglicans (5.11), reflecting a positive perception of belonging.

**Neighbourhood Community**: Muslims score highest among groups (4.80) in neighbourhood community, surpassing Anglicans (4.49), Catholics (4.38), and Not Religious individuals (4.19), and similar (but higher than) Hindus (4.73). Do experiences of prejudice build stronger neighbhourhoods?  Do Muslims tend to migrate to neighbourhoods with higher densities of other Muslims? Relatively higher scores in neighbourhood integration suggest additional avenues for future research. 

### Final Comment

Overall, the data indicate that Muslims maintain a strong level of religious commitment, both in identity and practice. This high degree of religious engagement underscores the importance of religion as a core component of Muslim identity in New Zealand.

Nevertheless, as shown in @fig-religion-id, @fig-religion-belief, and @fig-religion-practice, there is considerable variation among Muslims in these patterns of religious commitment. A minority of Muslims do not identify strongly with their faith, do not profess beliefs in God or a Spirit, and do not partake in religious practices.

@fig-religion-id, @fig-religion-belief, and @fig-religion-practice also reveal variation above the mean, with some Muslims expressing exceptionally strong religious commitments -- "super-Muslims". Such patterns of variation indeed hold across all denominations, including among secular people who express spiritual commitments, spiritual beliefs, and occasionally even religious practices such as prayer and religious service attendance.

In the Varieties of Religious Experience, William James makes the following observation about classification (@james1902varieties).

> The first thing the intellect does with an object is to class it along with something else. But any object that is infinitely important to us and awakens our devotion feels to us also as if it must be *sui generis* and unique. Probably a crab would be filled with a sense of personal outrage if it could hear us class it without ado or apology as a crustacean, and thus dispose of it. "I am no such thing," it would say; "I am MYSELF, MYSELF alone."

In reflecting on patterns of variation within and across religious denominations in New Zealand, including Muslims, we are immediately struck by the uniqueness of each individual.  Muslims are not cast from a cookie cutter, nor are those who identify with other religious and secular groups. 

In general, allowing our intellects to focus on the broader patterns, it is notable that Muslims are more similar to the Christian No Faith Declared and Christian Others groups than are these evangelical Christian groups to other denominations such as Anglicans and (to a lesser extent) Catholics. Thus, not only do we find heterogeneity between Christian groups in the expression of religious faith, but these patterns of similarity and difference reveal certain affinities between evangelical Christians and Muslims in the central role that religion plays in their lives (for broader discussion of religious similarities and difference in New Zealand refer to @wilson2014differences and @bulbulia2020national). Whether common experiences of anti-religious prejudice and common experiences of religious faith present opporutinities for community making across theological differences is a question of previous speculation (@shaver2016religion). This study underscores the interest of this open question for future applied research.

Perhaps our most striking finding are that, despite significant evidence of anti-religious prejudice, Muslim self-esteem remains high. Moreover, Muslims possess a strong sense of what makes their lives meaningful and maintain a clear sense of purpose. Although future research is needed to examine the causes and consequences of these orientations toward self and life, it is worth pausing to reflect on these exceptional qualities that embody Muslim dignity in New Zealand.

### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study. Members of the NZAVS management team and research group hold full copies of the NZAVS data. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author or any member of the NZAVS advisory board for replication or checking of any published study using NZAVS data. The code for the analysis can be found at: OSF-lINK TBA. 


### Acknowledgements

**TBA** 
The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Plank Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement

**TBA**:  UA led the MSD and community outreach. CGS led New Zealand Attitudes and Values Study data collection. JB prepared the models, tables and graphs.  All authors contributed to the writing of this manuscript.


{{< pagebreak >}}




## Appendix A: Measures {#appendix-a}




```{r, results='asis'}
#| eval: true
#| include: true

cat(outcome_measures)

```

{{< pagebreak >}}





## References {.appendix}











