---
title: "The New Zealand Muslim Diversity Study: Baseline Sample Information 2023-2024"
abstract: |
  In this study, we present demographic statistics from the first wave of the New Zealand Muslim Diversity Study (MSD 2023-2024). We describe patterns of religious identity and behaviour within the MDS cohort and examine self-reported multi-dimensional well-being. By comparing Muslim demographic, religious, and psychological diversity with that of the broader New Zealand population, we discover patterns of variation within and between religious groups that are not evident from group affiliations alone. Our findings uncover new commonalities between Muslims and other highly religious groups in New Zealand, and clarify that group-level averages do not capture the full spectrum of individual religious differences.
   **KEYWORDS**: Diversity; Longitudinal; Muslim, Panel; Religion
author: 
  - name: M. Usman Afzali
    affiliation: University of Otago New Zealand
    orcid: 0000-0003-3169-6576 
  - name: Aarif Rasheed
    affiliation:  University of Otago New Zealand
    orcid: 0000-0000-0000-0000
  - name: Chris G. Sibley
    affiliation: School of Psychology, University of Auckland, New Zealand
    orcid: 0000-0002-4064-8800
  - name: Kumar Yogeswaaran
    affiliation: University of Canterbury New Zealand
    orcid: 0000-0002-1978-5077
  - name: Joseph A. Bulbulia
    affiliation: Victoria University of Wellington, New Zealand
    orcid: 0000-0002-5861-2056
    email: joseph.bulbulia@vuw.ac.nz
    corresponding: yes
editor_options: 
    chunk_output_type: console
format:
  pdf:  
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    pdf-engine: lualatex
    sanitise: true
    keep-tex: true
    link-citations: true
    colorlinks: true
    documentclass: article
    classoption: [single column]
    lof: false
    lot: false
    geometry:
      - top=30mm
      - left=25mm
      - heightrounded
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
    template-partials: 
      - /Users/joseph/GIT/templates/quarto/title.tex
    header-includes:
      - \input{/Users/joseph/GIT/latex/latex-for-quarto.tex}
      - \let\oldtabular\tabular
      - \renewcommand{\tabular}{\small\oldtabular}
      - \setlength{\tabcolsep}{4pt}  % Adjust this value as needed

date: last-modified
execute:
  echo: false
  warning: false
  include: true
  eval: true
fontfamily: libertinus
bibliography: /Users/joseph/GIT/templates/bib/references.bib
csl: /Users/joseph/GIT/templates/csl/camb-a.csl
---

```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# reproducibility
set.seed(123)

# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("table1")
library("printr")
packageVersion(pkg = 'margot')

# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-muslims")


# exposure_var = "muslim_binary"
# name_exposure = "muslim_binary"

# table for muslim on religion
table_demographics <- here_read("table_demographics")
table_health <- here_read("table_health")
table_psych <- here_read("table_psych")
table_present<- here_read("table_present")

table_social<- here_read("table_social")
table_life <- here_read("table_life")
table_religion <- here_read("table_religion")
table_discrim <- here_read("table_discrim")

prop_miss_case <- here_read("prop_miss_case")
results_religion <- here_read_qs("results_religion", push_mods)
results_life <- here_read_qs("results_life", push_mods)
results_psych <- here_read_qs("results_psych", push_mods)

n_total <- here_read("n_total")
n_15 <- here_read("n_15")
n_mus_pre<- here_read("n_mus_pre")

n_participants<- here_read("n_participants")
# varaibles
exposure_var = "religion_bigger_denominations"
name_exposure = "religion_bigger_denominations"

baseline_vars <- here_read("baseline_vars")
outcome_vars <- here_read("outcome_vars")
baseline_vars
outcome_vars
# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)
prop_miss_case<- here_read("prop_miss_case")



# read models
results_religion<- here_read_qs("results_religion", push_mods)
results_discrim <- here_read_qs("results_discrim", push_mods)
results_life <- here_read_qs("results_life", push_mods)
results_psych<- here_read_qs("results_psych", push_mods)
results_social<- here_read_qs("results_social", push_mods)
results_present<- here_read_qs("results_present", push_mods)
results_health<- here_read_qs("results_health", push_mods)


#predictions
present_forgive <- results_present$plots$Forgiveness+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

present_perfect <- results_present$plots$Perfectionism+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


present_esteem <- results_present$plots$`Self Esteem`+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


present_pwb_a <- results_present$plots$`PWB Standard Living` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_b <- results_present$plots$`PWB Standard Living` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_c <- results_present$plots$`PWB Your Health` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_d  <- results_present$plots$`PWB Your Relationships`+ scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


# access individual plots Religion
religion_relid <- results_religion$plots$`Religious Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  
religion_spirid <- results_religion$plots$`Spiritual Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  

religion_church <- results_religion$plots$`Monthly Religious Service Attendence` + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-1,30))
   
religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_prayer <- results_religion$plots$`Weekly Prayer`+ scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_god <- results_religion$plots$`Belief in God`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 

religion_spirit <- results_religion$plots$`Belief in Spirit`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


# access individual plots discrimination


discrim_ethnic <- results_discrim$plots$`Perceived Ethnic Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_religious <- results_discrim$plots$`Perceived Religious Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_gender<- results_discrim$plots$`Perceived Gender Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_age <- results_discrim$plots$`Perceived Age Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
    

# access individual plots health

health_sleep <- results_health$plots$Sleep+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(.5,15))
health_sleep

health_exercise <- results_health$plots$`Hours of Exercise (log)`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(0,6))
health_exercise

exp(1)-1
exp(2)-1
exp(3)-1
exp(4)-1

health_short_form_health <- results_health$plots$`Short Form Health`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))  +  scale_y_continuous(limits = c(.5,7))
    
health_short_form_health
# access individual plots psychological wellbeing
life_grd <- results_life$plots$Gratitude+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_sat<- results_life$plots$`Life Satisfaction`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_mp <- results_life$plots$`Meaning: Purpose`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_ms <- results_life$plots$`Meaning: Sense`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

# access individual plots life-reflective wellbeing
results_psych$plots$Anxiety+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Depression + scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Fatigue+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Rumination+ scale_y_continuous(limits = c(-.5,5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

social_belong_a <- results_social$plots$`Belonging - Acceptance` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
    
    
social_belong_b <- results_social$plots$`Belonging - Shared Beliefs` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
  
social_belong_c  <-
  results_social$plots$`Belonging - Outsider (Reversed)` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_neighbourhood <- results_social$plots$`Neighbourhood Community`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_a <- results_social$plots$`Support - Help Availability`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_b <- results_social$plots$`Support - No Guidance (Reversed)`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_c <- results_social$plots$`Support - Someone to Turn To`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


# combo plots

plot_belief <- religion_god + religion_spirit + plot_annotation(title = "Religious Beliefs",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_practice <- religion_church / (religion_prayer + religion_scripture)+ plot_annotation(title = "Religious Practice",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_id <- religion_relid + religion_spirid + plot_annotation(title = "Religious and Spiritual Identification",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_life <- life_grd + life_sat + life_mp + life_ms + plot_annotation(title = "Life-Reflective Well-Being",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_descrim <- discrim_age  + discrim_gender + discrim_ethnic + discrim_religious +
  plot_annotation(title = "Perceived Dicrimination",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_social_belong <- social_belong_a  + social_belong_b + social_belong_c  +social_neighbourhood +
  plot_annotation(title = "Social Belonging and Neighbourhood Community",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_social_support <- social_belong_a  + social_belong_b + social_belong_c   +
  plot_annotation(title = "Social Belonging",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_social_support <- social_belong_a  + social_belong_b + social_belong_c   +
  plot_annotation(title = "Social Belonging",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_present_a <- present_esteem + present_forgive + present_perfect +
  plot_annotation(title = "Personal Well-Being Indicators",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_present_b <- present_pwb_a + present_pwb_b + present_pwb_c + present_pwb_d +
  plot_annotation(title = "Personal Well-Being Index",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_health <- health_short_form_health  + (health_exercise /  health_sleep) +
  plot_annotation(title = "Biological Health",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_health

# checks
plot_belief
plot_id
plot_practice
plot_descrim
plot_present_a
plot_present_b
plot_life
plot_social_belong
plot_social_support

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)


measures_path_rstudio <- here::here("manuscripts", "24-usman-mus-diversity")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))

outcome_measures <- boilerplate::boilerplate_report_measures(
  baseline_vars = baseline_vars,
  outcome_vars = outcome_vars,
  measure_data = measure_data
)

```


The New Zealand Attitudes and Values Study (NZAVS) is a nationally representative panel study of New Zealanders started in 2009. It has measured `r n_total` participants annually, and is presently tracking `r n_15` participants, about 1% of New Zealand's adult population. 

Despite strong NZAVS coverage of New Zealand's denominations, in the largest NZAVS cohort, less than .01% of New Zealand's Muslims were participating (n = `r n_mus_pre`). That is, the New Zealand Attitudes and Values Study was under-representing Muslims by an order of magnitude below the New Zealand census estimates. 

The Muslim Diversity Study, led by Usman Afzali (UA), has been built through partnership with the Muslim community partnerships to boost the New Zealand Attitudes and Values Muslim sample by ~600 participants to obtain coverage of about 1% of the New Zealand Census estimate of New Zealand Muslim Community.

The premise of this study has been to enable extend national longitudinal science of the study of Muslim diversity and flourishing over the lifespan, to understand the scope of such diversity and the challenges to such flourishing, and to clarify the affordances through which Muslims overcome adversity that has been previously documented in other research (refs.)

The Muslim Diversity Study is important first, because the lifelong dynamics of Muslim flourishing are poorly understood; this project addresses an urgent scientific need.

Second, our work with the Muslim community revealed a strong community interest in Muslim flourishing science; this project is an attempt to address this interest by involving Muslims in this science. 

Third, this project serves as a model for developing panel studies of under-represented groups in Aoteoroa New Zealand and in other countries. 


## Method

Measures are presented in [Appendix A](#appendix-a)

### Sample

The New Zealand Attitudes and Values Study is a national probability study designed to accurately reflect the broader New Zealand population. It uses prize draws to incentivise participation. Although the New Zealand Attitudes and Values Study has good demographic representation of the country as a whole, it tends to under-sample males and individuals of Asian descent and over-sample females and MÄori (the indigenous peoples of New Zealand). To address enhance the accuracy of our findings for the target population, we apply 2018 New Zealand Census survey weights to the sample data. These weights adjust for variations in age, gender, and ethnicity to better approximate the national demographic composition [@sibley2021]. 

**OFFICIAL NZAVS DOCUMENTATION HERE**


### Design 

We compare and contrast the Muslim sample with other religious groups, including those secular people who do not identify as religious. This research is descriptive an exploratory. 

Study 1 reports religious beliefs, identity, and practices of Muslims and other religious groups.

Study 2 reports perceived religious, ethnic, gender, and age-based prejudice.

Study 3 reports indicators related to biological health

Study 4 reports self-related well-being

Study 5 reports life-related well-being

Study 6 reports social well-being. 


In studies 2-7, our models adjust for the following demographic indicators that are unlikely to be affected by religious (or secular affiliation): age, gender, whether someone is New Zealand Born, and ethnicity.  Such adjustments allow us to focus on those features of multi-dimensional well-being that associated with religious (and secular) affiliation. 


 
### Eligibility Criteria

Participants needed to respond to the 2023 wave of the New Zealand Attitudes Study, and to have complete information about religious identification (yes or no). A total of `r n_participants` individuals met these criteria and were included in the study. 



### Missing Data

To mitigate bias from missing data, we imputed missing responses using the `ppm` algorithm from the `mice` package in R [@vanbuuren2018].  Overall missingness accounts for less than `r prop_miss_case`% if the data. However, we note that missingness for certain indicators among Muslims is considerably higher than among other groups ... 


### Demographic Statistics and Sample Responses

@tbl-demography presents demographic data by religious affiliation, including secular affiliation.


::: {#tbl-demography}
@tbl-religion presents sample demographic statistics. 

\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_demographics_x <- t1kable(
  table_demographics,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_demographics_x|> 
  kable_styling(
    font_size = 6,
    latex_options = c("hold_position","striped", "repeat_header", "longtable") )
```
\end{landscape}
Demographic statistics by Religious Affiliation (including Muslims)
:::

{{< pagebreak >}}
@tbl-religion presents sample response-statistics for Religious Belief, Identification, and Practice. 

::: {#tbl-religion}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_religion_x <- t1kable(
  table_religion,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_religion_x |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
  )

```
\end{landscape}
Sample statistics by Religious Denomination.
:::
{{< pagebreak >}}

@tbl-discrim presents sample response-statistics for Perceived Discrimination.

::: {#tbl-discrim}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_discrim_x <- t1kable(
  table_discrim,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_discrim_x |>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
  )


```
\end{landscape}
Sample Perceived Discrimination Statistics.
:::


{{< pagebreak >}}
@tbl-health presents sample response-statistics for biological health.

::: {#tbl-health}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_health_x <- t1kable(
  table_health,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_health_x |> 
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample Biological Health Statistics.
:::

{{< pagebreak >}}

@tbl-psych presents sample response-statistics for psychological health.

::: {#tbl-psych}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_psych_x <- t1kable(
  table_psych,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 

table_psych_x |>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample Psychological Health Statistics.
:::
{{< pagebreak >}}
@tbl-present presents sample response-statistics for self-related well-being

::: {#tbl-present}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_present_x <- t1kable(
  table_present,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_present_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample Self-Related Well-Being Statistics.
:::
{{< pagebreak >}}

@tbl-life presents sample response-statistics for life-related well-being

::: {#tbl-life}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_life_x <- t1kable(
  table_life,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_life_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample Life-Related Well-Being Statistics.
:::

{{< pagebreak >}}

@tbl-social presents sample response-statistics for social well-being

::: {#tbl-social}
\begin{landscape}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "latex",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_social_x <- t1kable(
  table_social,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
) 
table_social_x|>
  kable_styling(
    font_size = 7,
    latex_options = c("hold_position", "repeat_header", "longtable")
   ) 
```
\end{landscape}
Sample Life-Related Well-Being Statistics.
:::
{{< pagebreak >}}

## Study 1: Comparative Models of Religious Beliefs and Behaviours

@fig-religion-id presents models for religious and spiritual identities adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-id and @tbl-spiritual-id


```{r}
#| label: fig-religion-id
#| fig-cap: "Religious and Spiritual Identity"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_id

```


::: {#tbl-religion-id}
```{r, results= 'asis'}
library(dplyr)
library(kableExtra)

#  table
results_religion$predictions$religion_identification_level |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for religious identification by religious denomination.
:::


::: {#tbl-spiritual-id}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_spiritual_identification |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # Removed "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for spiritual identification by religious denomination.
:::


{{< pagebreak >}}

@fig-religion-belief presents models for religious and spiritual beliefs adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-belief and @tbl-spiritual-belief



```{r}
#| label: fig-religion-belief
#| fig-cap: "Religious and Spiritual Beliefs"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_belief
```


::: {#tbl-religion-belief}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_believe_god_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for belief in God by religious denomination.
:::


::: {#tbl-spiritual-belief}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_believe_spirit_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # Removed "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for sbelief in a Spirit or Life Force by religious denomination.
:::



{{< pagebreak >}}

@fig-religion-practice presents models for religious and spiritual practices, again adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-religion-church,  @tbl-religion-prayer, and @tbl-religion-scripture.



```{r}
#| label: fig-religion-practice
#| fig-cap: "Religious and Spiritual Practices"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_practice
```


::: {#tbl-religion-church}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_church |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for monthly religious service attendance by religious denomination.
:::


::: {#tbl-religion-prayer}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_prayer |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::


::: {#tbl-religion-scripture}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_scripture |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::


{{< pagebreak >}}
## Study 2: Comparative Models of Perceived Prejudice


@fig-discrim presents models for perceived discrimination, adjusting for age, New Zealand born, ethnicity, and gender. The expected means and confidence intervals for these models are presented in @tbl-discrim-age, @tbl-discrim-gender,   @tbl-discrim-ethnicity, and @tbl-discrim-religion, 

```{r}
#| label: fig-discrim
#| fig-cap: "Religious and Spiritual Practices"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_descrim
```

::: {#tbl-discrim-age}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_age_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived age discrimination.
:::


::: {#tbl-discrim-gender}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_gend_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived gender discrimination.
:::


::: {#tbl-discrim-ethnicity}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived ethnic discrimination.
:::


::: {#tbl-discrim-religion}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_religious_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |>
  kable_styling(
    font_size = 11,
    latex_options = c("repeat_header", "striped")  # remove "hold_position" to eliminate [!h]
  )
```
Adjusted means and 95% confidence intervals for perceived religious discrimination.
:::

{{< pagebreak >}}
## Study 3: Comparative Models of Biological Health

@fig-health presents models for indicators related to biological health, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-health
#| fig-cap: "Biological Health"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_health
```
{{< pagebreak >}}


## Study 4: Comparative Models of Self-Related Well-Being


@fig-self-a presents models for personal well-being in dimensions of self-esteem, forgiveness, and perfectionism, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-self-a
#| fig-cap: "Self-Esteem, Forgiveness, Perfectionism"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_present_a

```
{{< pagebreak >}}
@fig-self-b presents models for personal well-being, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-self-b
#| fig-cap: "Personal Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_present_b

```
{{< pagebreak >}}

## Study 5: Comparative Models of Life-Related Well-Being

@fig-life presents models for personal well-being, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-life
#| fig-cap: "Personal Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_life

```
{{< pagebreak >}}

## Study 6: Comparative Models of Social Well-Being


@fig-social-support presents models for social belonging, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-social-support
#| fig-cap: "Personal Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_social_support
```
{{< pagebreak >}}
@fig-social-belong presents models for social belonging, adjusting for age, New Zealand born, ethnicity, and gender.

```{r}
#| label: fig-social-belong
#| fig-cap: "Personal Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 16
#| fig-width: 16

plot_social_belong

```
{{< pagebreak >}}


### Conclusions 



### Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.


### Data Availability

The data described in the paper are part of the New Zealand Attitudes and Values Study. Members of the NZAVS management team and research group hold full copies of the NZAVS data. A de-identified dataset containing only the variables analysed in this manuscript is available upon request from the corresponding author or any member of the NZAVS advisory board for replication or checking of any published study using NZAVS data. The code for the analysis can be found at: OSF-lINK TBA. 


### Acknowledgements

**TBA** 
The New Zealand Attitudes and Values Study is supported by a grant from the Templeton Religious Trust (TRT0196; TRT0418). JB received support from the Max Plank Institute for the Science of Human History. The funders had no role in preparing the manuscript or deciding to publish it.


### Author Statement

**TBA**:  UA led the MSD and led the interpretation of the analysis. CGS led data collection. JB prepared the tables and graphs.  All authors contributed to the writing of this manuscript.


{{< pagebreak >}}




## Appendix A: Measures {#appendix-a}




```{r, results='asis'}
#| eval: true
#| include: true

cat(outcome_measures)

```

{{< pagebreak >}}





## References {.appendix}











