---
title: "The New Zealand Muslim Diversity Study: Baseline Sample Information 2023-2024"
format:
  revealjs:
    theme: default
    embed-resources: true
    slide-number: true
    smaller: true
    scrollable: true
    incremental: false
    chalkboard: false
    buttons: false
    preview-links: auto
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
    auto-stretch: true
    html-math-method: katex
    progress: true
    highlight-style: kate
    monofont: "Fira Code"
    mainfont: "Fira Code"
    titlefont: "Fira Code"
    font-family: "Fira Code"
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
    css: styles.css
author: 
  - name: M. Usman Afzali
    affiliation: University of Otago New Zealand
    orcid: 0000-0003-3169-6576 
extra_dependencies:
  - tikz-cd
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: load-libraries
#| echo: false
#| include: false
#| eval: true

# detach("package:margot", unload = TRUE)
# devtools::install_github("go-bayes/margot")
# for making graphs
library("tinytex")
library("extrafont")
loadfonts(device = "all")

# reproducibility
set.seed(123)

# rendering document
library("margot")
library("glue")
library("patchwork")
library("kableExtra")
library("stringr")
library("ggplot2")
library("ggeffects")
library("parameters")
library("table1")
library("printr")
packageVersion(pkg = 'margot')

# path to data folder (set to path on your computer) - for pushing and pulling model results
push_mods <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/24-muslims")
push_talks <- here::here("/Users/joseph/Library/CloudStorage/Dropbox-v-project/data/00-talks")


# exposure_var = "muslim_binary"
# name_exposure = "muslim_binary"

# table for muslim on religion
table_demographics <- here_read("table_demographics")
table_health <- here_read("table_health")
table_psych <- here_read("table_psych")
table_present<- here_read("table_present")

table_social<- here_read("table_social")
table_life <- here_read("table_life")
table_religion <- here_read("table_religion")
table_discrim <- here_read("table_discrim")

prop_miss_case <- here_read("prop_miss_case")
results_religion <- here_read_qs("results_religion", push_mods)
results_life <- margot::here_read_qs("results_life", push_mods)
results_psych <- margot::here_read_qs("results_psych", push_mods)

n_total <- here_read("n_total")
n_15 <- here_read("n_15")
n_mus_pre<- here_read("n_mus_pre")

n_participants<- here_read("n_participants")
# varaibles
exposure_var = "religion_bigger_denominations"
name_exposure = "religion_bigger_denominations"

baseline_vars <- here_read("baseline_vars")
outcome_vars <- here_read("outcome_vars")

n_mus_post <- here_read("n_mus_post")

n_mus_post

# joseph.bulbulia@gmail.com (joseph.bulbulia@vuw.ac.nz)
set.seed(123)
prop_miss_case<- here_read("prop_miss_case")


# read models
results_religion<- here_read_qs("results_religion", push_mods)
results_discrim <- here_read_qs("results_discrim", push_mods)
results_life <- here_read_qs("results_life", push_mods)
results_psych<- here_read_qs("results_psych", push_mods)
results_social<- here_read_qs("results_social", push_mods)
results_present<- here_read_qs("results_present", push_mods)
results_health<- here_read_qs("results_health", push_mods)



# plots
psy_fatigue = results_psych$plots$Fatigue + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))



psy_anxiety = results_psych$plots$Anxiety + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))


psy_depression = results_psych$plots$Depression + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))



psy_rumination = results_psych$plots$Rumination + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-0.5, 4.5))

#predictions
present_forgive <- results_present$plots$Forgiveness+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

present_perfect <- results_present$plots$Perfectionism+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

present_esteem <- results_present$plots$`Self Esteem`+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 
present_pwb_a <- results_present$plots$`PWB Your Future Security` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_b <- results_present$plots$`PWB Standard Living` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_c <- results_present$plots$`PWB Your Health` + scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
present_pwb_d  <- results_present$plots$`PWB Your Relationships`+ scale_y_continuous(limits = c(-0.5,10.5))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))



# access individual plots Religion
religion_relid <- results_religion$plots$`Religious Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  
religion_spirid <- results_religion$plots$`Spiritual Identification Level`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
  

religion_church <- results_religion$plots$`Monthly Religious Service Attendence` + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) +  scale_y_continuous(limits = c(-1,30))
   
religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_scripture <- results_religion$plots$`Weekly Scripture`  + scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_prayer <- results_religion$plots$`Weekly Prayer`+ scale_y_continuous(limits = c(-1,100))+ theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


religion_god <- results_religion$plots$`Belief in God`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 

religion_spirit <- results_religion$plots$`Belief in Spirit`+ scale_y_continuous(limits = c(-0,1.1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
)) 


# access individual plots discrimination


discrim_ethnic <- results_discrim$plots$`Perceived Ethnic Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_religious <- results_discrim$plots$`Perceived Religious Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_gender<- results_discrim$plots$`Perceived Gender Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
discrim_age <- results_discrim$plots$`Perceived Age Discrimination`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))
    

# access individual plots health

health_sleep <- results_health$plots$Sleep+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(.5,15))
health_sleep

health_exercise <- results_health$plots$`Hours of Exercise (log)`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +  scale_y_continuous(limits = c(0,6))
health_exercise

exp(1)-1
exp(2)-1
exp(3)-1
exp(4)-1

health_short_form_health <- results_health$plots$`Short Form Health`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1))  +  scale_y_continuous(limits = c(.5,7))
    
health_short_form_health
# access individual plots psychological wellbeing
life_grd <- results_life$plots$Gratitude+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_sat<- results_life$plots$`Life Satisfaction`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_mp <- results_life$plots$`Meaning: Purpose`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
life_ms <- results_life$plots$`Meaning: Sense`+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

# access individual plots life-reflective wellbeing
results_psych$plots$Anxiety+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Depression + scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Fatigue+ scale_y_continuous(limits = c(-.5,5))+ theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )
results_psych$plots$Rumination+ scale_y_continuous(limits = c(-.5,5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)
  )

social_belong_a <- results_social$plots$`Belonging - Acceptance` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
    
    
social_belong_b <- results_social$plots$`Belonging - Shared Beliefs` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))
  
social_belong_c  <-
  results_social$plots$`Belonging - Outsider (Reversed)` + scale_y_continuous(limits = c(.5,7.5)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_neighbourhood <- results_social$plots$`Neighbourhood Community`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_a <- results_social$plots$`Support - Help Availability`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_b <- results_social$plots$`Support - No Guidance (Reversed)`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))

social_support_c <- results_social$plots$`Support - Someone to Turn To`+ scale_y_continuous(limits = c(.5,7.5)) + theme(
    axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) + theme(axis.text.x = element_text(
  angle = 20,
  hjust = 1,
  vjust = 1
))


# combo plots

plot_belief <- religion_god + religion_spirit + plot_annotation(title = "Religious Beliefs",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_practice <- religion_church / (religion_prayer + religion_scripture)+ plot_annotation(title = "Religious Practice",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_id <- religion_relid + religion_spirid + plot_annotation(title = "Religious and Spiritual Identification",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_life <- life_grd + life_sat + life_mp + life_ms + plot_annotation(title = "Life-Reflective Well-Being",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_descrim <- discrim_age  + discrim_gender + discrim_ethnic + discrim_religious +
  plot_annotation(title = "Perceived Dicrimination",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_social_belong <- social_belong_a  + social_belong_b + social_belong_c  +social_neighbourhood +
  plot_annotation(title = "Social Belonging and Neighbourhood Community",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')



plot_social_support <- social_support_a + social_support_b + social_support_c   +
  plot_annotation(title = "Social Support",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_present_a <- present_esteem + present_forgive + present_perfect +
  plot_annotation(title = "Person-Focussed Well-Being Indicators",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')

plot_present_b <- present_pwb_a + present_pwb_b + present_pwb_c + present_pwb_d +
  plot_annotation(title = "Personal Well-Being Index",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_health <- health_short_form_health  + (health_exercise /  health_sleep) +
  plot_annotation(title = "Biological Health",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


plot_psych <- psy_anxiety + psy_depression + psy_fatigue + psy_rumination + plot_annotation(title = "Psychological Well-Being",subtitle = "Adjusted for Age, Born in NZ,Ethnicity, and Gender", tag_levels = "A") + plot_layout(guides = 'collect')


# Save plots 
# margot::margot_save_png(
#   plot_id,
#   base_filename = "plot_religious_id",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# margot::margot_save_png(
#   plot_belief,
#   base_filename = "plot_belief",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# margot::margot_save_png(
#   plot_practice,
#   base_filename = "plot_religious_practice",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# 
# 
# margot::margot_save_png(
#   plot_descrim,
#   base_filename = "plot_descrim",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )  
# 
# margot::margot_save_png(
#   plot_health,
#   base_filename = "plot_health",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# 
# margot::margot_save_png(
#   plot_psych,
#   base_filename = "plot_psych",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# 
# margot::margot_save_png(
#   plot_present_a,
#   base_filename = "plot_person_focussed_a",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# 
# margot::margot_save_png(
#   plot_present_b,
#   base_filename = "plot_person_focussed_b",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# margot::margot_save_png(
#   plot_life,
#   base_filename = "plot_life",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# 
# margot::margot_save_png(
#   plot_social_support,
#   base_filename = "plot_social_support",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )
# 
# margot::margot_save_png(
#   plot_social_belong,
#   base_filename = "plot_social_belong",
#   save_path = here::here(push_mods),
#   width = 16,
#   height = 9
# )






# checks
# plot_belief
# plot_id
# plot_practice
# plot_descrim
# plot_present_a
# plot_present_b
# plot_life
# plot_social_belong
# plot_social_support
# plot_psych
# plot_health

# read data (use 01 script to data wrangle)
# dat_long_imputed_bounds_ordinal <- margot::here_read_qs("dat_long_imputed_bounds_ordinal", push_mods)


measures_path_rstudio <- here::here("manuscripts", "24-usman-mus-diversity")
measure_data <- readRDS(here::here(measures_path_rstudio, "merged_db.rds"))

outcome_measures <- boilerplate::boilerplate_report_measures(
  baseline_vars = baseline_vars,
  outcome_vars = outcome_vars,
  measure_data = measure_data
)

timeline_histogram_png <- margot::read_multiple_images(push_talks, ".*timeline_histogram\\.png$")

```



## Thanks


::: {.panel-tabset}


### Research Supported by

- Templeton Religion Trust
- University of Auckland
- University of Canterbury
- University of Otago
- Victoria University
- `r n_total` individuals who have participated in the New Zealand Attitudes and Values Study since 2009.
- `r n_mus_post` Muslims participating in the Muslim Diversity Study.


### Research with

- Prof. Chris G. Sibley
- Aarif Rasheed
- Associate Prof Kumar Yogeswaaran
- Prof Joseph Bulbulia
- Others on the MDS Team

:::

# Background 



## Overview {.incremental}

1. **New Zealand Attitudes and Values Study (NZAVS)**:
   - A nationally representative panel study, initiated in 2009.
   - Sample frame is drawn randomly from the NZ Electoral Roll.
   - Postal questionnaire (coverage; retention \~ 70-80%).
   - Large multidisciplinary research team (\>60).
   - Focus on personality, social attitudes, values, religion, employment, prejudice $\dots$

   - Tracks nearly 1% of New Zealand’s adult population annually, with `r n_total` participants to date.

2. **Muslim Diversity Study (MDS)**:
   - A longitudinal booster study within the NZAVS, launched in 2024.
   - Established in partnership with the Muslim community to improve representation of Muslims in the NZAVS.
   - Increased Muslim sample from less than 0.01% to nearly 1% of New Zealand’s Muslim population (~600 participants).

3. **Purpose of the MDS**:
   - Document Muslim diversity and flourishing over the lifespan.
   - Explore challenges to flourishing and examine how Muslims overcome adversity.
   - Address gaps in research on how adversity shapes the experiences of Muslims in New Zealand.

4. **Broader Contribution**:
   - Amplifies Muslim voices within NZAVS’s historical record.
   - Aims to enrich understanding of beliefs, attitudes, and behaviours of New Zealanders for future generations.

## What will we learn today? 

###  Scope:
   - Provide a descriptive examination of religion, prejudice, and multi-dimensional well-being among Muslims, compared to other groups.
   - Report on demographics, values, employment, and behaviours, followed by Studies 1–7 on various facets of well-being and discrimination.
   - Report models that control for demographics unlikely to be influenced by religious identification (age, gender, birth status, ethnicity).
   - Focus on documenting associations between religious affiliation and well-being, without inferring causality.

###  Goal 
   - Establish a baseline for understanding Muslim diversity and flourishing in New Zealand
   

###  My assumptions about you
  - You are interested in religion and science
  - You do not necessarily have a background in either. 

   
   
## NZAVS DATA Collection 


```{r}
#| fig-width: 16
#| fig-height: 8

timeline_histogram_png$timeline_histogram
```

   


## Eligibility 

- Responded to the T15 (2023-2024) wave of the New Zealand Attitudes Study
- Has complete information about religious identification (yes or no). 
- A total of `r n_participants` individuals met these criteria.



# Descriptive Data

## Sample Demography

::: {#tbl-demography}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_demographics
```
Demographic statistics by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Religion

::: {#tbl-religion}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_religion

```

Sample statistics by for religious beliefs and behaviours by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Perceived Discrimination 

::: {#tbl-discrim}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_discrim
```
Perceived discrimination by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Health

::: {#tbl-health}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_health

```
Sample biological health indicators, by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::



## Sample Psychological Well-Being

::: {#tbl-psych}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )
table_psych
```

Sample psychological health indicators, by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Person-Focussed Well-Being

::: {#tbl-present}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_present

```

Sample person-focussed well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Life-Focussed Well-Being

::: {#tbl-life}

```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )


table_life

```

Sample life-focussed well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).
:::


## Sample Social Well-Being


::: {#tbl-social}
```{r, results = 'asis'}
library(kableExtra)
# as.data.frame(table_demographics) |>
#   kbl(format="latex", booktabs = TRUE) |> 
#   kable_styling(font_size = 8)
# as.data.frame(table_demographics) |>
#   kbl(
#     format = "markdown",
#     booktabs = TRUE,
#     longtable = TRUE  # Enable longtable format
#   ) |>
#   kable_styling(
#     font_size = 8,
#     latex_options = c("striped", "hold_position", "repeat_header", "longtable")
#   )

table_social
```

Sample social well-being by religious denomination (NZAVS wave 15, which encompasses MSD wave 1).

:::



## Discussion of Demographic and Sample Characteristics of the Muslim Diversity Study Booster {.incremental}

1. **Demographics**: Muslims are younger (mean age 33.8 years) than Not Religious (51.6), Anglicans (63.3), and Catholics (56.4). Only 23% were born in New Zealand compared to 79.4% of the Not Religious group, with 60.6% identifying as Asian.

2. **Parental Status**: Lower among Muslims (45.9%) compared to Anglicans (84.6%) and Catholics (76.5%).

3. **Lifestyle and Attitudes**: Muslims report minimal alcohol consumption (mean 0.112) and score higher on Right-Wing Authoritarianism (mean 4.53) compared to Not Religious (2.88) and other religious groups.

4. **Deprivation and Personality**: Higher deprivation scores (mean 5.65) among Muslims, with similar Agreeableness and Conscientiousness but slightly lower Honesty-Humility (mean 4.87).

5. **Missing Data**: Higher missing responses for Non-Heterosexual Orientation (18.8%) and Household Income (34%) among Muslims, possibly due to cultural sensitivities or unfamiliarity.




## Study 1: Comparative Models of Religious Beliefs and Behaviours



## Religious and Spiritual Identification

```{r}
#| label: fig-religion-id
#| fig-cap: "Religious and Spiritual Identity"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_id

```



## Predicted Averages: Religious Identfication

::: {#tbl-religion-id}
```{r, results= 'asis'}
library(dplyr)
library(kableExtra)

#  table
results_religion$predictions$religion_identification_level |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for religious identification by religious denomination.
:::



## Predicted Averages: Spiritual Identification

::: {#tbl-spiritual-id}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_spiritual_identification |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for spiritual identification by religious denomination.
:::




## Religious and Spiritual Beliefs



```{r}
#| label: fig-religion-belief
#| fig-cap: "Religious and Spiritual Beliefs"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_belief
```


## Predicted Averages: Religious Beliefs


::: {#tbl-religion-belief}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_believe_god_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow") 
```
Adjusted means and 95% confidence intervals for prdicted probability for belief in God by religious denomination.
:::


## Predicted Averages: Spirit or Life-Force Beliefs

::: {#tbl-spiritual-belief}
```{r, results= 'asis'}
# table
results_religion$predictions$religion_believe_spirit_binary |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # Remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for sbelief in a Spirit or Life Force by religious denomination.
:::



## Religious Practices

```{r}
#| label: fig-religion-practice
#| fig-cap: "Religious and Spiritual Practices"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_practice
```



## Predicted Averages: Monthly Religious Service Attendance


::: {#tbl-religion-church}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_church |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for monthly religious service attendance by religious denomination.
:::



## Predicted Averages: Weekly Prayer


::: {#tbl-religion-prayer}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_prayer |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::

## Predicted Averages: Weekly Religoius Scripture Reading


::: {#tbl-religion-scripture}
```{r, results= 'asis'}

#  table
results_religion$predictions$religion_scripture |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for weekly prayer by religious denomination.
:::




## Discussion of Study 1: Religion and Spirituality {.incremental}

1. **Religious Identification**: Muslims exhibit the highest predicted religious identification score (6.21), with over 11% non-response, and high spiritual identification (5.07), comparable to Buddhists and Other Religions.

2. **Belief in God**: Nearly universal among Muslims (predicted probability 0.89), similar to Christian (NFD) (0.94) and Christian (Others) (0.87), contrasting with the Not Religious group (0.14).

3. **Belief in Spirits**: Substantial among Muslims (predicted probability 0.61), though lower than Hindus (0.90) and Other Religions (0.87), reflecting theological diversity.

4. **Religious Service Attendance**: Muslims attend services frequently (predicted 5.43 times/month), surpassing Christian (Others) (3.23) and Catholics (2.02), with the Not Religious group at negligible levels (-0.04).

5. **Prayer Frequency**: Exceptionally high among Muslims (predicted 20.01 times/month), far exceeding Christian (NFD) (8.73) and Christian (Others) (10.09).

6. **Scripture Reading**: Muslims show strong engagement (predicted 4.59 times/week), comparable to devout Christian groups like Christian (Others) (4.19) and much higher than Anglicans (0.86) and Catholics (1.28).




## Study 2: Comparative Models of Perceived Prejudice


```{r}
#| label: fig-discrim
#| fig-cap: "Perceived Discrimination"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_descrim
```



## Predicted Averages: Perceived Age Discrimination

::: {#tbl-discrim-age}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_age_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for perceived age discrimination.
:::


## Predicted Averages: Perceived Gender Discrimination


::: {#tbl-discrim-gender}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_gend_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for perceived gender discrimination.
:::

## Predicted Averages: Perceived Ethnic Discrimination


::: {#tbl-discrim-ethnicity}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for perceived ethnic discrimination.
:::


## Predicted Averages: Perceived Religious Discrimination


::: {#tbl-discrim-religion}
```{r, results= 'asis'}

#  table
results_discrim$predictions$perc_religious_discrim|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for perceived religious discrimination.
:::



## Discussion of Study 2: Perceived Discrimination {.incremental}

### Age Discrimination

1. **Age and Gender Discrimination**: Muslims report low levels of perceived discrimination:
   - **Age**: Muslims (3.28) are similar to Not Religious (3.17) and Catholics (3.27).
   - **Gender**: Muslims (2.48) align with Not Religious (2.50) and Catholics (2.42), lower than Buddhists (2.75).

2. **Ethnic Discrimination**: Muslims report elevated ethnic discrimination (2.22), comparable to Hindus (2.16) and higher than Not Religious (1.70).

3. **Religious Discrimination**: Muslims report the highest perceived religious discrimination (3.57), far exceeding Not Religious (1.66), Anglicans (2.14), and other groups, highlighting religion-specific bias.

4. **Summary**: Muslims face distinctive challenges, with religious discrimination as the most pronounced disparity, consistent with prior research on low societal acceptance of Muslims in New Zealand.




## Study 3: Comparative Models of Biological Health

```{r}
#| label: fig-health
#| fig-cap: "Biological Health"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_health
```


## Predicted Averages: Log Hours Weekly Excercise

::: {#tbl-health-ex}
```{r, results= 'asis'}

#  table
results_health$predictions$log_hours_exercise |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for log hours of weekly exercise.
:::


## Predicted Averages: Hours of Sleep


::: {#tbl-health-sleep}
```{r, results= 'asis'}

#  table
results_health$predictions$hlth_sleep_hours |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for nightly hours of sleep.
:::


## Predicted Averages: Short Form Health

::: {#tbl-health-sfh}
```{r, results= 'asis'}

#  table
results_health$predictions$short_form_health |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for subjective health.
:::


## Discussion of Study 3: Biological Health {.incremental}

1. **Sleep Duration**: Minimal variation across groups:
   - Muslims: 6.85 hours.
   - Not Religious: 7.01 hours.
   - Anglicans: 7.05 hours.
   - Catholics: 6.89 hours.

2. **Overall Health Status**: Predicted health scores are comparable:
   - Muslims: 4.93.
   - Not Religious: 4.77.
   - Anglicans: 4.89.
   - Catholics: 4.77.
   - Buddhists: Slightly lower at 4.68.



## Study 4: Comparative Models of Psychological Well-being

```{r}
#| label: fig-psych
#| fig-cap: "Psychological Health"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_psych
```

## Predicted Averages: Kessler 6 Anxiety


::: {#tbl-psych-anxiety}
```{r, results= 'asis'}

#  table
results_psych$predictions$kessler_latent_anxiety |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for psychological anxiety.
:::


## Predicted Averages: Kessler 6 Depression


::: {#tbl-psych-depression}
```{r, results= 'asis'}

#  table
results_psych$predictions$kessler_latent_depression |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for psychological depression.
:::

## Predicted Averages: Fatigue


::: {#tbl-psych-fatigue}
```{r, results= 'asis'}

#  table
results_psych$predictions$hlth_fatigue |>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for fatigue.
:::


## Predicted Averages: Rumination


::: {#tbl-psych-rumination}
```{r, results= 'asis'}

#  table
results_psych$predictions$rumination|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for rumination.
:::



## Discussion of Study 4: Psychological Well-Being {.incremental}

1. **Anxiety**: Muslims report lower anxiety (1.30) compared to Not Religious (1.39) and most Christian groups, indicating potential psychological resilience.

2. **Fatigue**: Muslims report reduced fatigue (1.71) compared to Not Religious (1.83), suggesting resilience to stressors.

3. **Rumination**: Muslims score lower on rumination (0.90) than Not Religious (0.99), aligning with Anglicans (0.90), indicating less mental distress.

4. **Depression**: Depression scores for Muslims (0.73) are comparable to other groups (e.g., Catholics 0.65, Anglicans 0.64), showing no distinct advantage in this domain.

5. **Overall**: Findings suggest that Muslims exhibit robust psychological well-being, particularly in anxiety, fatigue, and rumination, possibly linked to religion-specific factors such as practices, beliefs, or community support.



## Study 5: Comparative Models of Person-Focussed Well-Being

```{r}
#| label: fig-self-a
#| fig-cap: "Self-Esteem, Forgiveness, Perfectionism"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_present_a
```


## Predicted Averages: Self Esteem


::: {#tbl-selfesteem}
```{r, results= 'asis'}
#  table
results_present$predictions$self_esteem|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for self esteem.
:::


## Predicted Averages: Forgiveness

::: {#tbl-forgive}
```{r, results= 'asis'}
#  table
results_present$predictions$forgiveness|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for forgiveness.
:::


## Predicted Averages: Perfectionism


::: {#tbl-perfect}
```{r, results= 'asis'}
#  table
results_present$predictions$perfectionism|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for perfectionism.
:::


## Personal Well-Being Index



```{r}
#| label: fig-self-b
#| fig-cap: "Personal Well-Being Index"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_present_b
```



## Predicted Averages PWB: Your Future Security

::: {#tbl-pwb-fs}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_future_security|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for personal well-being: your future security.
:::


## Predicted Averages PWB: Your Standard of Living


::: {#tbl-pwb-sl}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_standard_living|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for personal well-being: your standard of living.
:::


## Predicted Averages PWB: Your Health


::: {#tbl-pwb-yh}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_health|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for personal well-being: your health.
:::


## Predicted Averages PWB: Your Relationships


::: {#tbl-pwb-yr}
```{r, results= 'asis'}
#  table
results_present$predictions$pwb_your_relationships|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for personal well-being: your relationships.
:::


## Discussion of Study 5: Self-Related Well-Being {.incremental}

1. **Self-Esteem**: Muslims report higher self-esteem (5.46) than Not Religious individuals (5.00) and other religious groups, potentially reflecting the influence of religious identity and community support.

2. **Forgiveness**: Scores for Muslims (4.89) are slightly lower than for Not Religious individuals (4.99) and Anglicans (5.03), raising questions about the influence of perceived injustice.

3. **Future Security**: Muslims exhibit strong future security (6.07), similar to Anglicans (6.12), higher than Not Religious individuals (5.79), possibly reflecting religious frameworks of stability and hope.

4. **Personal Relationships**: Muslims report the highest satisfaction with personal relationships (7.90), surpassing Not Religious individuals (7.46) and other religious groups.

5. **Standard of Living and Health**: Muslims rate health highly (6.69, similar to Anglicans at 6.72) but report slightly lower satisfaction with their standard of living (7.17) compared to Anglicans (7.58).

6. **Overall**: Findings suggest that religious identity, community, and practices may enhance self-esteem and relationship satisfaction while buffering against prejudice. **Future research should investigate these questions longitudinally.**




## Study 6: Comparative Models of Life-Focussed Well-Being

```{r}
#| label: fig-life
#| fig-cap: "Life Related Well-Being"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_life

```


## Predicted Averages: Gratitude


::: {#tbl-gratitude}
```{r, results= 'asis'}
#  table
results_life$predictions$gratitude|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for gratitude.
:::

## Predicted Averages: Life Satisifaction


::: {#tbl-ls}
```{r, results= 'asis'}
#  table
results_life$predictions$lifesat|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for life satisfaction.
:::


## Predicted Averages: Meaning Purpose

::: {#tbl-purpose}
```{r, results= 'asis'}
#  table
results_life$predictions$meaning_purpose|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for sense of purpose in life.
:::


## Predicted Averages: Meaning Sense


::: {#tbl-sense}
```{r, results= 'asis'}
#  table
results_life$predictions$meaning_sense|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for an understanding what makes life meaningful.
:::



## Discussion of Study 6: Life-Focussed Well-Being {.incremental}


1. **Gratitude**: Muslims report higher gratitude (6.01) than Not Religious individuals (5.80) and Anglicans (5.89), aligning with Catholics (5.99) and Christian NFD (6.09).

2. **Life Satisfaction**: Muslims have higher life satisfaction (5.32) than Not Religious individuals (5.00) and most religious groups, including Anglicans (5.17) and Catholics (5.08).

3. **Meaning and Purpose**: Muslims report the highest levels of meaning and purpose (5.83), significantly exceeding Not Religious individuals (4.76) and aligning with Christian NFD (5.41).

4. **Sense of Meaning**: Muslims rank highly in their sense of meaning (6.03), surpassing Not Religious individuals (5.47) and religious groups like Catholics (5.66) and Anglicans (5.58).

5. **Overall**: Findings underscore consistently elevated life-focused well-being among Muslims, with gratitude, life satisfaction, and meaning playing key roles. 

6. **Question**: Does meaning and purpose arise from religious community making, and does this buffer Muslim well-being?





## Study 7: Comparative Models of Social Well-Being



```{r}
#| label: fig-social-support
#| fig-cap: "Social Support"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_social_support
```


## Predicted Averages: Support Help


::: {#tbl-support-help}
```{r, results= 'asis'}
#  table
results_social$predictions$support_help|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for social support: help.
:::

## Predicted Averages: Support Guidance


::: {#tbl-support-guide}
```{r, results= 'asis'}
#  table
results_social$predictions$support_noguidance_reversed|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for social support: guidance.
:::


## Predicted Averages: Support Turn To


::: {#tbl-support-turn}
```{r, results= 'asis'}
#  table
results_social$predictions$support_turnto|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow") 
```
Adjusted means and 95% confidence intervals for social support: no one to turn to (reversed).
:::


## Social Belonging

```{r}
#| label: fig-social-belong
#| fig-cap: "Social Belonging"
#| eval: true
#| include: true
#| echo: false
#| fig-height: 9
#| fig-width: 16

plot_social_belong

```

## Predicted Averages: Belonging Acceptance 


::: {#tbl-accept}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_accept|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for belonging: feels accepted.
:::

## Predicted Averages: Belonging Shared Beliefs 


::: {#tbl-sharedbeliefs}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_beliefs|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  )|> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for belonging: others share my beliefs.
:::


## Predicted Averages: Belonging Not Outsider 


::: {#tbl-outsider}
```{r, results= 'asis'}
#  table
results_social$predictions$belong_routside_reversed|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "markdown",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for belonging: feel like outsider (reversed).
:::





## Predicted Averages: Neighbourhood Community


::: {#tbl-neighbourhood}
```{r, results= 'asis'}
#  table
results_social$predictions$neighbourhood_community|>
  as.data.frame() %>%                      # ensure it's a data frame
  dplyr::select(-group) %>%                # remove the 'group' column
  dplyr::rename(Religion = x) %>%          # rename 'x' to 'Religion'
  kbl(
    format = "latex",
    booktabs = TRUE,
    digits = 2,
    row.names = FALSE                      # remove row names to eliminate any unintended columns
  ) |> 
    kable_styling("striped", full_width = FALSE) %>% # Optional: Add additional styling
  row_spec(2, background = "yellow")
```
Adjusted means and 95% confidence intervals for sense of neighbourhood community.
:::



## Discussion of Study 7: Social Well-Being {.incremental}


- Muslims resemble other religious groups in perceived social support and social belonging, and express higher levels of perceived neighbourhood community.
- Questions:
  - Do Muslims tend to migrate to neighbourhoods with higher densities of other Muslims? 
  - Does community buffer Muslims form the slings and arrows of anti-religious prejudice? 



## Key Takeaways From This Talk {.incremental}

1. **Religious Commitment**: Muslims exhibit strong religious commitment in identity, beliefs, and practices, with considerable variation within the group:
   - Some Muslims express weaker religious engagement - faith is not a cookie-cutter.
   - Others demonstrate exceptionally strong commitments ("super-Muslims"), similar to patterns in other denominations and secular groups.

2. **Heterogeneity Across Groups**: Religious diversity is evident across all denominations:
   - Muslims show greater similarities with evangelical Christian groups (e.g., Christian NFD and Christian Others) than with Anglicans or Catholics, reflecting shared emphasis on religion in daily life.
   - Patterns of similarity and difference raise questions about shared experiences of anti-religious prejudice fostering community connections, matters for future research.

3. **Individual Uniqueness**: Each individual’s religious experience is unique:  again, faith is not like a cookie cutter. 

4. **Resilience and Dignity**: Despite experiencing anti-religious prejudice:
   - Muslim self-esteem remains high.
   - Muslims possess a strong sense of meaning and purpose, reflecting resilience and dignity.

5. **Future Research**: Open questions include:
   - The role of Muslim faith in buffering prejudice
   - The mechanisms that underpin Muslims' high self-esteem and sense of purpose and meaning.
   - The role of anti-religious prejudice in fostering interfaith communities.




# Thank you!


## Ethics

The University of Auckland Human Participants Ethics Committee reviews the NZAVS every three years. Our most recent ethics approval statement is as follows: The New Zealand Attitudes and Values Study was approved by the University of Auckland Human Participants Ethics Committee on 26/05/2021 for six years until 26/05/2027, Reference Number UAHPEC22576.








## Appendix A: Measures {#appendix-a}



```{r, results='asis'}
#| eval: false
#| include: false

cat(outcome_measures)

```

#### Age

*What is your date of birth?*

We asked participants' ages in an open-ended question ("What is your age?" or "What is your date of birth"). Developed for the NZAVS.

#### Agreeableness

*I sympathize with others' feelings.*
*I am not interested in other people's problems.*
*I feel others' emotions.*
*I am not really interested in others (reversed).*

Mini-IPIP6 Agreeableness dimension: (i) I sympathize with others' feelings. (ii) I am not interested in other people's problems. (r) (iii) I feel others' emotions. (iv) I am not really interested in others. (r) [@sibley2011].

#### Alcohol Frequency

*"How often do you have a drink containing alcohol?"*

Participants could chose between the following responses: '(1 = Never - I don't drink, 2 = Monthly or less, 3 = Up to 4 times a month, 4 = Up to 3 times a week, 5 = 4 or more times a week, 6 = Don't know)' [@Ministry_of_Health_2013].

#### Alcohol Intensity

*"How many drinks containing alcohol do you have on a typical day when drinking alcohol? (number of drinks on a typical day when drinking)"*

Participants responded using an open-ended box [@Ministry_of_Health_2013].

#### Belong

*Know that people in my life accept and value me.*
*Feel like an outsider (reversed).*
*Know that people around me share my attitudes and beliefs.*

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument (Hagerty & Patusky, 1995): (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded [@hagerty1995].

#### Born Nz (Binary)

We asked participants, "Which country were you born in?" or "Where were you born? (please be specific, e.g., which town/city?)" (waves: 6-15).. Developed for the NZAVS.

#### Charity Donate

*How much money have you donated to charity in the last year?*

Numerical: open-ended response [@hoverd_religious_2010].

#### Children Num

We measured the number of children using one item from Bulbulia et al. (2015). We asked participants, "How many children have you given birth to, fathered, or adopted?" or "How many children have you given birth to, fathered, and/or parented?" (waves: 12-15) [@Bulbulia_2015].

#### Conscientiousness

*I get chores done right away.*
*I like order.*
*I make a mess of things.*
*I often forget to put things back in their proper place.*

Mini-IPIP6 Conscientiousness dimension: (i) I get chores done right away. (ii) I like order. (iii) I make a mess of things. (r) (iv) I often forget to put things back in their proper place. (r) [@sibley2011].

#### Education Level Coarsen

*What is your highest level of qualification?*

We asked participants, "What is your highest level of qualification?". We coded participans highest finished degree according to the New Zealand Qualifications Authority. Ordinal-Rank 0-10 NZREG codes (with overseas school qualifications coded as Level 3, and all other ancillary categories coded as missing). Developed for the NZAVS.

#### Employed (Binary)

*Are you currently employed (This includes self-employed of casual work)?*

Binary response: (0 = No, 1 = Yes). Stats NZ Census Question.

#### Ethnic Category (Categorical)

*Which ethnic group(s) do you belong to?*

Coded string: (1 = New Zealand European; 2 = Māori; 3 = Pacific; 4 = Asian). NZ Census coding.

#### Extraversion

*I am the life of the party.*
*I don't talk a lot (reversed).*
*I keep in the background (reversed).*
*I talk to a lot of different people at parties.*

Mini-IPIP6 Extraversion dimension: (i) I am the life of the party. (ii) I don't talk a lot. (r) (iii) I keep in the background. (r) (iv) I talk to a lot of different people at parties [@sibley2011].

#### Hlth Disability (Binary)

We assessed disability with a one-item indicator adapted from Verbrugge (1997). It asks, "Do you have a health condition or disability that limits you and that has lasted for 6+ months?" (1 = Yes, 0 = No) [@verbrugge1997].

#### Honesty Humility

*I feel entitled to more of everything (reversed).*
*I deserve more things in life (reversed).*
*I deserve more things in life (reversed).*
*I would get a lot of pleasure from owning expensive luxury goods (reversed).*

Mini-IPIP6 Honesty-Humility dimension: (i) I feel entitled to more of everything. (r) (ii) I deserve more things in life. (r) (iii) I would like to be seen driving around in a very expensive car. (r) (iv) I would get a lot of pleasure from owning expensive luxury goods. (r) [@sibley2011].

#### Hours Children

No information available for this variable.

#### Hours Charity

*Hours spent … voluntary/charitable work.*

Numerical: open-ended response [@sibley2011].

#### Hours Commute

*Hours spent...travelling/commuting.*

Numerical: open-ended response [@sibley2011].

#### Hours Exercise

...

#### Hours Housework

...

#### Hours Work

...

#### Household Income

*Please estimate your total household income (before tax) for the year XXXX.*

Not provided. Developed for the NZAVS.


#### Male (Binary)

We asked participants' gender in an open-ended question: "what is your gender?" or "Are you male or female?" (waves: 1-5). Female was coded as 0, Male was coded as 1, and gender diverse coded as 3 (Fraser et al., 2020). (or 0.5 = neither female nor male). Here, we coded all those who responded as Male as 1, and those who did not as 0 [@fraser_coding_2020].

#### Neuroticism

*I have frequent mood swings.*
*I am relaxed most of the time (reversed).*
*I get upset easily.*
*I seldom feel blue (reversed).*

Mini-IPIP6 Neuroticism dimension: (i) I have frequent mood swings. (ii) I am relaxed most of the time. (r) (iii) I get upset easily. (iv) I seldom feel blue. (r) [@sibley2011].

#### Not Heterosexual (Binary)

*How would you describe your sexual orientation? (e.g., heterosexual, homosexual, straight, gay, lesbian, bisexual, etc.)*

Open-ended question, coded as binary (not heterosexual = 1) [@greaves2017diversity].

#### NZ Deprevation Index 2018

*New Zealand Deprivation - Decile Index - Using 2018 Census Data*

Numerical: (1-10) [@atkinson2019].

#### NZSEI

We assessed occupational prestige and status using the New Zealand Socio-economic Index 13 (NZSEI-13) (Fahy et al., 2017). This index uses the income, age, and education of a reference group, in this case, the 2013 New Zealand census, to calculate a score for each occupational group. Scores range from 10 (Lowest) to 90 (Highest). This list of index scores for occupational groups was used to assign each participant a NZSEI-13 score based on their occupation [@fahy2017].

#### Openness

*I have a vivid imagination.*
*I have difficulty understanding abstract ideas (reversed).*
*I do not have a good imagination (reversed).*
*I am not interested in abstract ideas (reversed).*

Mini-IPIP6 Openness to Experience dimension: (i) I have a vivid imagination. (ii) I have difficulty understanding abstract ideas. (r) (iii) I do not have a good imagination. (r) (iv) I am not interested in abstract ideas. (r) [@sibley2011].

#### Parent (Binary)

We asked participants, "If you are a parent, what is the birth date of your eldest child?" or "If you are a parent, in which year was your eldest child born?" (waves: 10-current). Parents were coded as 1, while the others were coded as 0.. Developed for the NZAVS.

#### Partner (Binary)

*What is your relationship status? (e.g., single, married, de-facto, civil union, widowed, living together, etc.)*

Coded as binary (has partner = 1).. Developed for the NZAVS.

#### Political Conservative

*Please rate how politically liberal versus conservative you see yourself as being.*

Ordinal response: (1 = Extremely Liberal, 7 = Extremely Conservative) [@jost_end_2006-1].

#### Rural Classification

*High Urban Accessibility = 1, Medium Urban Accessibility = 2, Low Urban Accessibility  = 3, Remote = 4, Very Remote = 5.*

"Participants residence locations were coded according to a five-level ordinal categorisation ranging from Urban to Rural." [@whitehead2023unmasking].

#### Right Wing Authoritarianism

*It is always better to trust the judgment of the proper authorities in government and religion than to listen to the noisy rabble-rousers in our society who are trying to create doubt in people's minds.*
*It would be best for everyone if the proper authorities censored magazines so that people could not get their hands on trashy and disgusting material.*
*Our country will be destroyed some day if we do not smash the perversions eating away at our moral fibre and traditional beliefs.*
*People should pay less attention to The Bible and other old traditional forms of religious guidance, and instead develop their own personal standards of what is moral and immoral.*
*Atheists and others who have rebelled against established religions are no doubt every bit as good and virtuous as those who attend church regularly.*
*Some of the best people in our country are those who are challenging our government, criticizing religion, and ignoring the "normal way" things are supposed to be done (reversed).*

[@altemeyer1996authoritarian].

#### Social Dominance Orientation

*It is OK if some groups have more of a chance in life than others.*
*Inferior groups should stay in their place.*
*To get ahead in life, it is sometimes okay to step on other groups.*
*We should have increased social equality (reversed).*
*It would be good if groups could be equal (reversed).*
*We should do what we can to equalise conditions for different groups (reversed).*

Not provided [@@sidanius1999social].

#### Belonging: Acceptance

*Know that people in my life accept and value me.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@hagerty1995].

#### Belonging: Shared Beliefs

*Know that people around me share my attitudes and beliefs.*

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument (Hagerty & Patusky, 1995): (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded [@hagerty1995].

#### Belonging: Outsider (Reversed)

*Feel like an outsider (reversed).*

We assessed felt belongingness with three items adapted from the Sense of Belonging Instrument (Hagerty & Patusky, 1995): (1) "Know that people in my life accept and value me"; (2) "Feel like an outsider"; (3) "Know that people around me share my attitudes and beliefs". Participants responded on a scale from 1 (Very Inaccurate) to 7 (Very Accurate). The second item was reversely coded [@hagerty1995].

#### Forgiveness

*Sometimes I can't sleep because of thinking about past wrongs I have suffered.*
*I can usually forgive and forget when someone does me wrong.*
*I find myself regularly thinking about past times that I have been wronged.*

We assessed participants' forgiveness using reversed scores of a the NZAVS "vengeful rumination scale." This scale contains three items, adapted from @caprara_indicators_1986 and @berry_forgivingness_2005, and developed for NZAVS, ordinal response scale 1-7 (1 = Strongly Disagree to 7 = Strongly Agree).. Developed for the NZAVS.

#### Gratitude

*I have much in my life to be thankful for.*
*When I look at the world, I don't see much to be grateful for (reversed).*
*I am grateful to a wide variety of people.*

Ordinal response scale 1 = Strongly Disagree to 7 = Strongly Agree [@mccullough_grateful_2002].

#### Fatigue

*During the last 30 days, how often did ... you feel exhausted?*

Ordinal response: (0 = None Of The Time; 1 = A Little Of The Time; 2= Some Of The Time; 3 = Most Of The Time; 4 = All Of The Time) [@sibley2020].

#### Sleep Hours

*During the past month, on average, how many hours of actual sleep did you get per night?*

Open ended response [@buysse1989pittsburgh].

#### Kessler Latent Anxiety

*During the past 30 days, how often did...you feel restless or fidgety?*
*During the past 30 days, how often did...you feel that everything was an effort?*
*During the past 30 days, how often did...you feel nervous?*

Ordinal response: (0 = None Of The Time; 1 = A Little Of The Time; 2= Some Of The Time; 3 = Most Of The Time; 4 = All Of The Time) [@kessler2002].

#### Kessler Latent Depression

*During the past 30 days, how often did...you feel hopeless?*
*During the past 30 days, how often did...you feel so depressed that nothing could cheer you up?*
*During the past 30 days, how often did...you feel you feel restless or fidgety?*

Ordinal response: (0 = None Of The Time; 1 = A Little Of The Time; 2= Some Of The Time; 3 = Most Of The Time; 4 = All Of The Time) [@kessler2002].

#### Life Satisfaction

*I am satisfied with my life.*
*In most ways my life is close to ideal.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@diener1985a].

#### Log Hours Exercise

*Hours spent...exercising/physical activity.*

We took the natural log of the response + 1 [@sibley2011].

#### Meaning Purpose

*My life has a clear sense of purpose*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@steger_meaning_2006].

#### Meaning Sense

*I have a good sense of what makes my life meaningful.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@steger_meaning_2006].

#### Neighbourhood Community

*I feel a sense of community with others in my local neighbourhood.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@sengupta2013].

#### Percieved Age Discrimation

*I feel that I am often discriminated against because of my age.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree)

#### Perc Discrim

*I feel that I am often discriminated against because of my ethnicity.*


#### Perc Gender Discrimination

*I feel that I am often discriminated against because of my gender.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree)

#### Perceived Religious Discrimination

*I feel that I am often discriminated against because of my religious/spiritual beliefs.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) 
#### Perfectionism

*Doing my best never seems to be enough.*
*My performance rarely measures up to my standards.*
*I am hardly ever satisfied with my performance.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@rice_short_2014].

#### PWB Standard Living

*Please rate your level of satisfaction with the following aspects of your life...Your standard of living.*

Ordinal response (0 = completely dissatisfied to 10 = completely satisfied) [@cummins_developing_2003].

#### PWB your Future Security

*Please rate your level of satisfaction with the following aspects of your life...Your future security.*

Ordinal response (0 = completely dissatisfied to 10 = completely satisfied) [@cummins_developing_2003].

#### PWB your Health

*Please rate your level of satisfaction with the following aspects of your life...Your health.*

Ordinal response (0 = completely dissatisfied to 10 = completely satisfied) [@cummins_developing_2003].

#### PWB your Relationships

*Please rate your level of satisfaction with the following aspects of your life...Your personal relationships.*

Ordinal response (0 = completely dissatisfied to 10 = completely satisfied) [@cummins_developing_2003].

#### Religion Believe God (Binary)

*Do you believe in a God?*

Binary response: (0 = No, 1 = Yes) [@eurobarometer2005b].

#### Religion Believe Spirit (Binary)

*Do you believe in some form of spirit or life force?*

Binary response: (0 = No, 1 = Yes) [@eurobarometer2005b].

#### Religion Religious Service Attendance

*Do you identify with a religion and/or spiritual group? -> How many times did you attend a church or place of worship in the last month?*

Numerical: open-ended response [@sibley2012].

#### Religion Identification Level

*How important is your religion to how you see yourself?*

Ordinal response: (1 = Not Important, 7 = Very Important). Developed for the NZAVS.

#### Religion Prayer

*Do you identify with a religion and/or spiritual group? -> How many times did you pray in the last week?*

Numerical: open-ended response [@Bulbulia_2015].

#### Religion Scripture

*Do you identify with a religion and/or spiritual group? -> How many times did you read religious scripture in the last week?*

Numerical: open-ended response [@bulbulia2016].

#### Religion Spiritual Identification

*I identify as a spiritual person*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@postmes2013single].

#### Rumination

*During the last 30 days, how often did...you have negative thoughts that repeated over and over?*

Ordinal responses: 0 = None of The Time, 1 = A little of The Time, 2 = Some of The Time, 3 = Most of The Time, 4 = All of The Time [@nolen-hoeksema_effects_1993].

#### Self Control Have Lots

*In general, I have a lot of self-control.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@tangney_high_2004].

#### Self Control Wish more Reversed

*I wish I had more self-discipline.*

Ordinal response (1 = Strongly Disagree to 7 = Strongly Agree) [@tangney_high_2004].

#### Self Esteem

*On the whole am satisfied with myself.*
*Take a positive attitude toward myself.*
*Am inclined to feel that I am a failure (reversed).*

Ordinal response (1 = Very inaccurate to 7 = Very accurate) [@Rosenberg1965].

#### Short Form Health

*In general, would you say your health is...*

Ordinal response: (1 = Poor, 7 = Excellent) [@instrument1992mos].

#### Support Help

*There are people I can depend on to help me if I really need it.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@cutrona1987].

#### Support Noguidance Reversed

*There is no one I can turn to for guidance in times of stress (reversed).*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@cutrona1987].

#### Support Turn To

*I know there are people I can turn to when I need help.*

Ordinal response: (1 = Strongly Disagree, 7 = Strongly Agree) [@cutrona1987].



## References {.appendix}
